### 2024-02-01 products/manuscript/Manuscript_Cassia.qmd
1 Feb 2024
First attempt at updating file

---
title: "Low Infant Birth Weight in Brazil"
subtitle: "A Historical Data Analysis Approach"
author: Cassia Roth
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../assets/Mada_bib.bib
csl: ../assets/vancouver.csl
---

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(renv)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
I will analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro. The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*. In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*. In 1922, it became the *Revista de Gynecologia e D’Obstetricia*.

The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras (50). I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). From the available issues, I recorded the following information, when available, for all patients: patient number, gravidity and parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history. Gravidity refers to the total number of pregnancies a woman has, regardless of duration. Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children (51).

Clinical notes only sporadically included infant demographic information for spontaneous abortions or stillbirths. For spontaneous abortions, stillbirths, and live births, I recorded, when available, infant sex, weight, length. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy. However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (52,53), coded as operatory or natural by physicians, as interventionist.
In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician.

The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca de Biomedicina-A, Universidade Estadual do Rio de Janeiro (BBA-UERJ), all in Rio de Janeiro, Brazil. Between January 2012 and July 2013, I manually digitized the journal by photographing each volume. Then, between January and August 2017, I manually input the data into Excel from the digital reproductions.

## Questions/Hypotheses to be addressed
I will quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in low birth weight (LBW) in Rio de Janeiro, Brazil’s first public maternity hospital, Maternidade Laranjeiras, in the 1920s.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
##| label: fig-schematic
##| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
##| echo: FALSE
#knitr::include_graphics(here("products","assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
##| label: tbl-summarytable
##| tbl-cap: "Data summary table."
##| echo: FALSE
#resulttable=readRDS("../../results/summarytable.rds")
#knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
##| label: fig-result
##| fig-cap: "Height and weight stratified by gender."
##| echo: FALSE
#knitr::include_graphics(here("results","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
##| label: tbl-resulttable2
##| tbl-cap: "Linear model fit table."
##| echo: FALSE
#resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
#knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2024-02-01 products/manuscript/Manuscript_Cassia.qmd
After meeting w/Hayley
1 Feb 2024

@@
     highlight-style: github
-bibliography: ../assets/Mada_bib.bib
-csl: ../assets/vancouver.csl
+    bibliography: assets/Mada_Bib.bib
+    csl: assets/vancouver.csl
 ---
 
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
 ```{r, echo=FALSE, message=FALSE}
-# load a few R packages
+#Loading R packages
 library(here)
@@
 library(renv)
+library(tidyverse)
 ```
@@
 # Summary/Abstract
-_Write a summary of your project._
+This project uses a linear regression model to understand disparities in infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
+I will use a linear model to estimate the relationship of maternal variables on infant birth weight.
+I hypothesize that infants born to women of color will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality.
+Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts.
+This is a significant public health concern because low birth weight is associated with higher infant mortality, and with long-term health problems such as diabetes, hypertension, and heart disease.
+
+As a historian of medicine and public health in Brazil, I have collected infant birth weight data from the country's first public maternity hospital, opened in 1904 in the then-capital city of Rio de Janeiro.
+Brazil was the last country in the Western Hemisphere to abolish slavery (1888), and it had, and still has, the largest number of African-descended peoples in the world outside of Nigeria.
+Around 51% of country's population today is of African-descent.
+
+In the post-abolition period of the 1890s to 1930s, the Brazilian government expanded public health initiatives and the provision of clinical care.
+Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil?
+To date, historians have only provided descriptive examples of health inequities.
 
@@
 The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings.
-Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras (50).
+Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras.
 I was unable to locate vol.
@@
 Gravidity refers to the total number of pregnancies a woman has, regardless of duration.
-Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children (51).
+Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posnerOxornFooteHumanLabor2013].
 
@@
 I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy.
-However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (52,53), coded as operatory or natural by physicians, as interventionist.
+However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraudMauriceauManeuverDeleterious1997), coded as operatory or natural by physicians, as interventionist.
+
 In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures.
@@
 Then, between January and August 2017, I manually input the data into Excel from the digital reproductions.
+This will be converted into a .csv file for upload into R.
 
 ## Questions/Hypotheses to be addressed
-I will quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in low birth weight (LBW) in Rio de Janeiro, Brazil’s first public maternity hospital, Maternidade Laranjeiras, in the 1920s.
+I will quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, Maternidade Laranjeiras, in the 1920s.
+I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+The outcome I will measure is infant birth weight.
+The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganizationInternationalClassificationDiseases2022].
+I will use the WHO’s classification to categorize birth weight.
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+I will use a linear model to estimate the relationship of maternal variables on infant birth weight.
+Originally, I tried using a logistic regression model with very low birthweight (VLBW), low birth weight (LBW), and normal birth weight as the outcome, but after I started taking a class with Dr.
+Swartzendruber, I realized that it is important to understand birthweight as a continuous variable.
+
+**This section was written by GitHub copilot**
+I will use the `lm` function in R to estimate the relationship of maternal variables on infant birth weight.
+I will use the `tidy` function from the `broom` package to extract the coefficients and standard errors from the model.
+I will use the `glance` function from the `broom` package to extract the R-squared value from the model.
+I will use the `augment` function from the `broom` package to extract the residuals from the model.
+I will use the `ggplot2` package to visualize the results.
 
 ## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-##| label: fig-schematic
-##| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-##| echo: FALSE
-#knitr::include_graphics(here("products","assets","antigen-recognition.png"))
-```
 
@@
 ## Data import and cleaning
+```{r}
+#Setting working directory using here package
+here("data/raw-data/MaternidadeLaranjeiras.csv")
+
+#Reading in file
+ML <- read.csv(here("data/raw-data/MaternidadeLaranjeiras.csv"))
+
+#Looking at structure of data
+str(ML)
+```
+/Users/cassiaroth/Documents/GitHub/MADARoth/Roth-MADA-project/MaternidadeLaranjeiras.csv
+
+```{r}
+```
 _Write code that reads in the file and cleans it so it's ready for analysis.
@@
 
-@tbl-summarytable shows a summary of the data.
-
 Note the loading of the data providing a **relative** path using the `../../` notation.
@@
 See examples of that below.
-
-```{r}
-##| label: tbl-summarytable
-##| tbl-cap: "Data summary table."
-##| echo: FALSE
-#resulttable=readRDS("../../results/summarytable.rds")
-#knitr::kable(resulttable)
-```
 
@@
 
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-##| label: fig-result
-##| fig-cap: "Height and weight stratified by gender."
-##| echo: FALSE
-#knitr::include_graphics(here("results","height-weight-stratified.png"))
-```
-
 ## Full analysis
@@
 Then you load them here._
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-##| label: tbl-resulttable2
-##| tbl-cap: "Linear model fit table."
-##| echo: FALSE
-#resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
-#knitr::kable(resulttable2)
-```
 
@@
 
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
 Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.

### 2024-02-23 code/eda-code/eda-v1.qmd
Part 2
Finished cleaning and EDA for Part 2 of project.

---
title: "Exploratory Analysis Script"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(ggplot2) #for plotting
library(hrbrthemes) #for nice themes
hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
#rm(list=ls()) #clear the environment
```

Load the data.

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary_processed.rds")
#load data
ML_summary <- readRDS(data_location)
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
# Create summary
summary = skimr::skim(ML_summary)
print(summary)

# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary, file = summarytable_file)
```

We are saving the results to the `results` folder. All final table/figures will have `_final` in the name. This is to make sure we don't overwrite anything we might want to keep.

# Data exploration through figures

Histogram plots for the continuous outcomes.

I will explore maternal age first.

```{r}
# Create histogram plot
age_plot <- ML_summary %>% filter(!is.na(Age)) %>% # Filter out NAs
  ggplot(aes(x = Age)) +
  geom_histogram(binwidth = 0.5, width= 0.8, fill="#69b3a2") +
  geom_vline(xintercept = c(15, 49), linetype = "dashed", color = "red") +  # Add vertical lines that mark reproductive age
  ggtitle("Maternal Age") +
  theme_ipsum() +
  theme(plot.title = element_text(size=15))
plot(age_plot)

# Save plot to file
figure_file = here("results","figures", "age_distribution.png")
ggsave(filename = figure_file, plot=age_plot, bg="white") #background is white
```

Now infant birth weight.

```{r}
# Create histogram plot
weight_plot <- ML_summary %>% filter(!is.na(Weightgrams)) %>% #filter out NAs
  ggplot(aes(x = Weightgrams)) +
  geom_histogram(binwidth = 25, fill="#69b3a2") +
  geom_vline(xintercept = c(1500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
  ggtitle("Infant birthweight") +
  theme_ipsum() +
  theme(plot.title = element_text(size=15))
plot(weight_plot)

# Save plot to file
figure_file = here("results","figures", "weight_distribution.png")
ggsave(filename = figure_file, plot=weight_plot, bg="white") #background is white
```
Now infant birth length.
```{r}
# Create histogram plot
length_plot <- ML_summary %>% filter(!is.na(Lengthcentimeters)) %>% #filter out NAs
  ggplot(aes(x = Lengthcentimeters)) +
  geom_histogram(binwidth = 1, fill="#69b3a2") +
  geom_vline(xintercept = c(49), linetype = "dashed", color = "red") +  # Add vertical lines that mark average birth length for both sexes
  ggtitle("Infant birth length") +
  theme_ipsum() +
  theme(plot.title = element_text(size=15))
plot(length_plot)

# Save plot to file
figure_file = here("results","figures","length_distribution.png")
ggsave(filename = figure_file, plot=length_plot, bg="white") #background is white
```
It appears there are some outliers, so I will need to go back and look at the data.

For now, let's look at height as function of weight. We can see that length is positively correlated with weight, which is unsurprising.

```{r}
#Create plot
length_weight <- ML_summary %>%
  filter(!is.na(Lengthcentimeters) & !is.na(Weightgrams)) %>% #filter out NAs
  ggplot(aes(x=Lengthcentimeters, y=Weightgrams)) + geom_point() + geom_smooth(method='lm')
plot(length_weight)

#Save file
figure_file = here("results","figures","length_weight.png")
ggsave(filename = figure_file, plot=length_weight, bg="white") #background is white
```

Once more height as function of weight, stratified by sex.

```{r}
#Create plot
length_weight_sex <- ML_summary %>% filter(!is.na(Lengthcentimeters) & !is.na(Weightgrams) & !is.na(Sex)) %>% 
  ggplot(aes(x=Lengthcentimeters, y=Weightgrams, color = Sex)) + geom_point() + geom_smooth(method='lm')
plot(length_weight_sex)

#Save file
figure_file = here("results","figures","length_weight_sex.png")
ggsave(filename = figure_file, plot=length_weight_sex, bg="white") #background is white
```
Now let's make some tables about the data to get a better idea of the overall numbers.

```{r}
# Summary stats of continuous variables
summary_stats <- ML_summary %>%
  summarize(
    mean_age = mean(Age, na.rm = TRUE),
    sd_age = sd(Age, na.rm = TRUE),
    mean_weight = mean(Weightgrams, na.rm = TRUE),
    sd_weight = sd(Weightgrams, na.rm = TRUE),
    mean_length = mean(Lengthcentimeters, na.rm = TRUE),
    sd_length = sd(Lengthcentimeters, na.rm = TRUE)
  )

print(summary_stats)

# Make into a table
# Create dataframe
summary_stats1 <- data.frame(
  Variable = c("Maternal Age", "Infant birthweight (grams)", "Infant birth length (cm)"),
  Mean = c(25.30435, 3087.341, 48.33493),
  StandardDev = c(5.759474, 566.5855, 3.92719)
)

# Create a gt table
summary_stats_table <- summary_stats1 %>%
  gt() %>%
  tab_header(
    title = "Summary Statistics",
    subtitle = "Mean Values and Standard Deviations of Continuous Variables"
  ) %>%
  cols_label(
    Variable = "Variable",
    Mean = "Mean",
    StandardDev = "Standard Deviation"
  )

# Print the table
print(summary_stats_table)

#Save table
saveRDS(summary_stats_table, file = here::here("results", "tables", "summary_stats_table.rds"))
```

Now let's work with our categorical variables.

```{r}
# Summary stats of categorical variables

# Calculate frequencies of color category (branca, parda, preta)
color_freq <- ML_summary %>%
  filter(!is.na(Color)) %>% # Exclude rows with NA in color variable
  count(Color, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(color_freq)
```

For maternal race, we have a sample size of n = 2695. The majority of the sample is White (branca), with 42.45% of the sample. Mixed-race (parda) is 29.2% of the sample, and Black (preta) is 28.32% of the sample. Let's now compare White patients to all patients of color by combining Black and mixed-race categories. I will create a dummy variable for color, called modified_color with two categories: AfroDescent (comprising parda and preta) and EuroDescent (comprising branca).

```{r}
# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
ML_summary$ModifiedColor <- ifelse(ML_summary$Color == "Branca", "EuroDescent",
  ifelse(ML_summary$Color == "Preta", "AfroDescent",
  ifelse(ML_summary$Color == "Parda", "AfroDescent", NA)))

# Or using recode factor from dplyr (easier)
# ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")

# View the updated dataframe
print(head(ML_summary))

# Calculate frequencies of color category (branca, parda, preta)
color_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedColor)) %>% # Exclude rows with NA in color variable
  count(ModifiedColor, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(color_freq2)
```
We can see that of the clinic patients, 1551 (57.6%) were of African descent (defined as preta or parda) and 1144 (42.4%) were of European descent (branca).

Now lets get summary statistics of the other categorical variables, including `Status`, `Nationality`, `Birth`, `MaternalOutcome`, `FetalOutcome` (excluding abortion), and `Sex`. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. For the `Status` variable, we will combine "multigesta" and "trigesta" into "multigesta", since they mean the same thing (more than two pregnancies). For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern". "Brazilian" will remain the same.

```{r}
# Calculate frequencies of Status category
status_freq <- ML_summary %>%
  filter(!is.na(Status)) %>% # Exclude rows with NA in status variable
  count(Status, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(status_freq)

# Create dummy variables for status
ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Trigesta" = "Multigesta")

# Calculate frequencies of ModifiedStatus category
status_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedStatus)) %>% # Exclude rows with NA in status variable
  count(ModifiedStatus, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(status_freq2)
```

```{r}
# Calculate frequencies of Nationality category
nationality_freq <- ML_summary %>%
  filter(!is.na(Nationality)) %>% # Exclude rows with NA in nationality variable
  count(Nationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(nationality_freq)

# Create dummy variables for status
ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")

# Calculate frequencies of ModifiedStatus category
nationality_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedNationality)) %>% # Exclude rows with NA in nationality variable
  count(ModifiedNationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(nationality_freq2)
```
```{r}
# Calculate frequencies of birth category with abortion
birth_freq <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(birth_freq)
```
We can see that of a sample size of n = 2761, 88% of outcomes were natural deliveries. Now let's take abortion out of this sample size to just look at labor and delivery outcomes.

```{r}
# Calculate frequencies of birth category without abortion
birth_freq2 <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  filter(Birth != "aborto") %>% # Exclude rows with abortion
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(birth_freq2)
```
We can see that of a new sample size of n = 2672, 91% of outcomes were natural deliveries. Now let's look at maternal outcomes.

```{r}
# Calculate frequencies of MaternalOutcome category
maternal_freq <- ML_summary %>%
  filter(!is.na(MaternalOutcome)) %>% # Exclude rows with NA in MaternalOutcome variable
  count(MaternalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(maternal_freq)
```
We can see that of the 2829 women who visited the clinic, 23 died, although not necessarily in childbirth. I need to come back to maternal mortality later after I have figured out live births. The equation is MMR = MD/LB * 10,000. We can also look at stillbirth rate. The equation is SBR = SB/LB * 1000.

```{r}
# Calculate frequencies of FetalOutcome category excluding abortion
fetal_freq <- ML_summary %>%
  filter(!is.na(FetalOutcome)) %>% # Exclude rows with NA in FetalOutcome variable
  count(FetalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  filter(FetalOutcome != "aborto") %>% # Exclude rows with abortion
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(fetal_freq)

#MMR Maternal mortality ratio
maternal_deaths <-23
live_births <- 2440
MMR <- (maternal_deaths / live_births) * 10000

#Stillbirth rate SBR
still_births <- 226
total_births <- 2666
SBR <- (still_births / total_births) * 1000
```
The MMR is 94.3% in this specific clinic. The SBR is 84.8% in this specific clinic.

Finally, lets look at infant sex.

```{r}
# Calculate frequencies of Sex category
sex_freq <- ML_summary %>%
  filter(!is.na(Sex)) %>% # Exclude rows with NA in Sex variable
  count(Sex, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(sex_freq)
```

# Notes

I will need to come back and calculate sex ratio in the clinic, but I need to first figure out how to exclude stillbirths from the sample, since sex ratio is calculated male/female * 100 for live births only.

### 2024-02-23 code/processing-code/processingfile-v1.qmd
Part 2
Finished cleaning and EDA for Part 2 of project.

---
title: "Cleaning and Processing"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

Please note I used ChatGPT and CoPilot throughout to help troubleshoot code.

# Processing script

This contains the same code and comments/information as `processingcode.R`. This quarto file does not pull code from the R script but rather includes it directly.

# Setup

Load needed packages. Make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
```

# Data loading

Note that for functions that come from specific packages (instead of base R), you can specify both package and function: `package::function()`. Although not required, specifying the package makes it clearer where the function is coming from. This text is adapted from Dr. Andreas Handel.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","MaternidadeLaranjeiras.csv")
ML_raw <- read_csv(data_location)

# Structure of Data
glimpse(ML_raw)
```
From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.

# Check data

First we can look at the Codebook and practice table making.
```{r}
# Path to codebook
codebook_location <- here::here("data", "raw-data", "Codebook.csv")
codebook <- read_csv(codebook_location)

# Full column specification
spec(codebook)

# Display the structure of the data frame
str(codebook)

# Display the first few rows of the data frame
head(codebook)

# Create table
codebook_table <- kable(codebook, caption = "Codebook") %>% kable_styling()

#Print table
print(codebook_table)
```

# Explore data

There are several ways of looking at the data. This code comes from Dr. Andreas Handel.

```{r}
# Looking at the data
dplyr::glimpse(ML_raw)
summary(ML_raw)
head(ML_raw)
skimr::skim(ML_raw)
```

# Clean data

I created this dataset from published clinical notes from a specific maternity hospital in Rio de Janeiro, Brazil for multiple projects, mostly of a qualitative nature. Thus, I was interested in the notes clinicians included on difficult deliveries and their outcomes. These variables include qualitative information that can be excluded from our processed data, since we are not looking at qualitative notes on cases. We also don't need the citation information to run regressions. So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Birth_Notes`, `Maternal_Notes`, `Fetal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.

```{r}
# Remove unnecessary variables
ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Birth_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
```

Despite removing variables we won't use because of their qualitative nature, we still have many missing observations for some of the variables. Let's check for missing data.

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
```
We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations. This is a lot. Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.

```{r}
# Remove variables with missing data not relevant to hypothesis
ML_2 <- ML_1 %>% dplyr::select(-CivilStatus, -GestationalAge_Months)
```

Now we can check for missing data again.

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_2 %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(missing_data_summary)
```

Now I will work on transforming character variables into factors.

```{r}
# Transform character variables into factors
ML_3 <- ML_2 %>%
  mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex), as.factor))

# Check variables
glimpse(ML_3)

# Check levels of factor variables
map(ML_3[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')], levels)

# Define factor variables to check frequencies
factor_variables <- c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')

# Check frequencies of factor variables
frequency_table <- ML_3 %>%
  count(across(all_of(factor_variables)))

print(frequency_table)
```

# Create datasets for analysis

There are still missing observations for all the remaining variables except `Date`. For now, I will keep these observations intact because I want the maximum number of observations for summary statistics of all variables. Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study. Here, I will create two datasets: one for summary statistics and one for the linear regression model.

The latter (linear regression) will exclude the variables `MaternalOutcome` and `Birth`.

```{r} 
# Create dataset for summary statistics (includes missing values)
ML_summary <- ML_3

# Creating dataset for linear regression
ML_linear <- ML_3 %>% tidyr::drop_na()

# Check for no missing data in linear model
ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(ML_linear_summary)
```
Now, I can see that my `ML_linear` dataset has no missing data. It has 2032 observations and 11 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.

# Save data

Finally, I will save the clean data both the `ML_Linear` dataset and the `ML_summary` dataset as RDS files
.
```{r}
# Save ML_Linear as RDS
save_data_location <- here::here("data","processed-data","ML_linear_processed.rds")
saveRDS(ML_linear, file = save_data_location)

# Save ML_summary as RDS
save_data_location <- here::here("data","processed-data","ML_summary_processed.rds")
saveRDS(ML_summary, file = save_data_location)
```

# Notes

This script is a work in progress. I will be adding more to it as I continue to work on the data.

### 2024-02-23 code/processing-code/processingfile-v2.qmd
Part 2
Finished cleaning and EDA for Part 2 of project.

--- 2024-02-23 code/processing-code/processingfile-v1.qmd
@@
 ---
-title: "Cleaning and Processing"
+title: "Cleaning and Processing with code pulled in"
 author: "Cassia Roth"
@@
 
-Please note I used ChatGPT and CoPilot throughout to help troubleshoot code.
+Please note I used ChatGPT and CoPilot throughout to help troubleshoot code. I'm not sure that I did this file and the `processingcode.R` files correctly. `processingfile-v1.qmd` is correct.
 
@@
 
-This contains the same code and comments/information as `processingcode.R`. This quarto file does not pull code from the R script but rather includes it directly.
+This contains the same code and comments/information as `processingcode.R`. This quarto file pulls in from the R script using the code chunk labels.
 
@@
 
-Load needed packages. Make sure they are installed.
+This needs to run to load the R script so we can include the code chunks below
 
-```{r}
-library(readxl) #for loading Excel files
-library(tidyverse) #for data processing/cleaning
-library(skimr) #for nice visualization of data 
-library(here) #to set paths
-library(renv) #for package management
-library(knitr) #for nice tables
-library(kableExtra) #for nice tables
+```{r, include=FALSE, cache=FALSE}
+knitr::read_chunk('processingcode.R')
+```
+
+```{r, packages, message = FALSE, warning = FALSE}
 ```
@@
 
-```{r}
-# path to data
-# note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw-data","MaternidadeLaranjeiras.csv")
-ML_raw <- read_csv(data_location)
-
-# Structure of Data
-glimpse(ML_raw)
+```{r, loaddata}
 ```
@@
 First we can look at the Codebook and practice table making.
-```{r}
-# Path to codebook
-codebook_location <- here::here("data", "raw-data", "Codebook.csv")
-codebook <- read_csv(codebook_location)
-
-# Full column specification
-spec(codebook)
-
-# Display the structure of the data frame
-str(codebook)
-
-# Display the first few rows of the data frame
-head(codebook)
-
-# Create table
-codebook_table <- kable(codebook, caption = "Codebook") %>% kable_styling()
-
-#Print table
-print(codebook_table)
+```{r, checkdata}
 ```
@@
 
-```{r}
-# Looking at the data
-dplyr::glimpse(ML_raw)
-summary(ML_raw)
-head(ML_raw)
-skimr::skim(ML_raw)
+```{r, exploredata}
 ```
@@
 
-```{r}
-# Remove unnecessary variables
-ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Birth_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
+```{r, cleandata}
 ```
@@
 
-```{r}
-#Summarize missing data using dplyr
-missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
-```
 We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations.
@@
 
-```{r}
-# Remove variables with missing data not relevant to hypothesis
-ML_2 <- ML_1 %>% dplyr::select(-CivilStatus, -GestationalAge_Months)
-```
-
 Now we can check for missing data again.
 
-```{r}
-#Summarize missing data using dplyr
-missing_data_summary <- ML_2 %>% dplyr::summarise_all(funs(sum(is.na(.))))
-View(missing_data_summary)
-```
-
 Now I will work on transforming character variables into factors.
-
-```{r}
-# Transform character variables into factors
-ML_3 <- ML_2 %>%
-  mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex), as.factor))
-
-# Check variables
-glimpse(ML_3)
-
-# Check levels of factor variables
-map(ML_3[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')], levels)
-
-# Define factor variables to check frequencies
-factor_variables <- c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')
-
-# Check frequencies of factor variables
-frequency_table <- ML_3 %>%
-  count(across(all_of(factor_variables)))
-
-print(frequency_table)
-```
 
@@
 
-```{r} 
-# Create dataset for summary statistics (includes missing values)
-ML_summary <- ML_3
-
-# Creating dataset for linear regression
-ML_linear <- ML_3 %>% tidyr::drop_na()
-
-# Check for no missing data in linear model
-ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
-View(ML_linear_summary)
+```{r, createdatasets}
 ```
@@
 .
-```{r}
-# Save ML_Linear as RDS
-save_data_location <- here::here("data","processed-data","ML_linear_processed.rds")
-saveRDS(ML_linear, file = save_data_location)
-
-# Save ML_summary as RDS
-save_data_location <- here::here("data","processed-data","ML_summary_processed.rds")
-saveRDS(ML_summary, file = save_data_location)
+```{r, savedata}
 ```

### 2024-02-23 products/manuscript/Manuscript_Cassia.qmd
Part 2
Finished cleaning and EDA for Part 2 of project.

@@
     highlight-style: github
-    bibliography: assets/Mada_Bib.bib
-    csl: assets/vancouver.csl
+    bibliography: ../../assets/Mada_Bib.bib
+    csl: ../../assets/vancouver.csl
 ---
@@
 ```{r, echo=FALSE, message=FALSE}
-#Loading R packages
-library(here)
-library(knitr)
-library(renv)
-library(tidyverse)
+#Loading packages
+library(tidyverse) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+library(renv) #for package management
 ```
@@
 # Summary/Abstract
-This project uses a linear regression model to understand disparities in infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
+This cross-sectional study uses a linear regression model to understand disparities in infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
 I will use a linear model to estimate the relationship of maternal variables on infant birth weight.
@@
 ## General Background Information
-Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality.
+Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality [@travassos2004].
 Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts.
@@
 ## Description of data and data source
+
 I will analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
@@
 Gravidity refers to the total number of pregnancies a woman has, regardless of duration.
-Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posnerOxornFooteHumanLabor2013].
+Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013].
 
@@
 I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy.
-However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraudMauriceauManeuverDeleterious1997), coded as operatory or natural by physicians, as interventionist.
+However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraud1997), coded as operatory or natural by physicians, as interventionist.
 
@@
 ## Questions/Hypotheses to be addressed
+
 I will quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, Maternidade Laranjeiras, in the 1920s.
@@
 The outcome I will measure is infant birth weight.
-The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganizationInternationalClassificationDiseases2022].
+The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
 I will use the WHO’s classification to categorize birth weight.
@@
 
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+## Data import and cleaning
 
-## Data import and cleaning
-```{r}
-#Setting working directory using here package
-here("data/raw-data/MaternidadeLaranjeiras.csv")
-
-#Reading in file
-ML <- read.csv(here("data/raw-data/MaternidadeLaranjeiras.csv"))
-
-#Looking at structure of data
-str(ML)
-```
-/Users/cassiaroth/Documents/GitHub/MADARoth/Roth-MADA-project/MaternidadeLaranjeiras.csv
-
-```{r}
-```
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
-
-## Statistical analysis
-_Explain anything related to your statistical analyses._
+Please see the qmd.file `processingfile-v1.qmd1` for full detail on the data import and cleaning process.
 
@@
 
-# Results
-
 ## Exploratory/Descriptive analysis
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+The exploratory analysis was performed on the ML_summary dataset with n = 2845 observations and 14 variables. Please see the Codebook for more details on variables.
 
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
+There were three continuous variables: maternal age, infant birth weight, and infant length.
+I am aware that this table is not rendering correctly in the word document, but I am unable to figure out why.
+If I load library(gt), then my word document is corrupted and will not open at all.
+I will continue to work on this issue.
+```{r}
+#| label: tbl-summarytablecont
+#| tbl-cap: "Continous Data Summary Table"
+#| echo: FALSE
+summary_stats_table <- readRDS(here::here("results", "tables", "summary_stats_table.rds"))
+summary_stats_table
+```
+
+Birth weight is the outcome variable of interest, so let's look at some visualizations of the data.
+Here is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO defines as normal birth weight.
+
+```{r}
+#| label: fig-schematic-1
+#| fig-cap: Histogram of birth weight
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","weight_distribution.png"))
+```
+
+Birth length is less important for our analysis, but we might want to look at the relationship between weight and length.
+Below we can see a histogram of the distribution of birth lengths in the sample.
+The red line marks the average birth length for both male and female infants (49 centimeters).
+
+```{r}
+#| label: fig-schematic-2
+#| fig-cap: Histogram of birth length
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","length_distribution.png"))
+```
+
+Finally, we will look at the relationship between infant weight and length, both not-stratified and stratified by sex. Unsurprisingly, birth weight and length are positively correlated.
+
+```{r}
+#| label: fig-schematic-3
+#| fig-cap: Scatterplot of birth weight by length
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","length_weight.png"))
+```
+Finally, since male infants are usually slightly heavier than females, let's look at a scatterplot of birth weight by length, stratified by sex.
+
+```{r}
+#| label: fig-schematic-4
+#| fig-cap: Scatterplot of birth weight by length stratified by sex
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","length_weight_sex.png"))
+```
+
+For information on the categorical variables, please see eda-v1.qmd.
+Please note Dr.
+Handel, I am really struggling with making a table of these variables, and I will continue working on that moving forward.
+If you are interested in the code and data, please see the eda-v1.qmd file.
+
+{{< pagebreak >}}
 
@@
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
 ## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
 
 ## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
 

### 2024-02-23 code/eda-code/edacode.R

@@ -1,0 +1,240 @@
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(tidyverse) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+library(renv) #for package management
+library(knitr) #for nice tables
+library(kableExtra) #for nice tables
+library(gt) #for nice tables
+library(ggplot2) #for plotting
+library(hrbrthemes) #for nice themes
+hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
+rm(list=ls()) #clear the environment
+
+## ---- loaddata --------
+#Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","ML_summary_processed.rds")
+#load data
+ML_summary <- readRDS(data_location)
+
+## ---- table1 --------
+# Create summary
+summary = skimr::skim(ML_summary)
+print(summary)
+# save to file
+summarytable_file = here("results", "tables", "summarytable.rds")
+saveRDS(summary, file = summarytable_file)
+
+## ---- age --------
+# Create histogram plot
+age_plot <- ML_summary %>% filter(!is.na(Age)) %>% # Filter out NAs
+  ggplot(aes(x = Age)) +
+  geom_histogram(binwidth = 0.5, width= 0.8, fill="#69b3a2") +
+  geom_vline(xintercept = c(15, 49), linetype = "dashed", color = "red") +  # Add vertical lines that mark reproductive age
+  ggtitle("Maternal Age") +
+  theme_ipsum() +
+  theme(plot.title = element_text(size=15))
+plot(age_plot)
+# Save plot to file
+figure_file = here("results","figures", "age_distribution.png")
+ggsave(filename = figure_file, plot=age_plot, bg="white") #background is white
+
+## ---- Birthweight --------
+# Create histogram plot
+weight_plot <- ML_summary %>% filter(!is.na(Weightgrams)) %>% #filter out NAs
+  ggplot(aes(x = Weightgrams)) +
+  geom_histogram(binwidth = 25, fill="#69b3a2") +
+  geom_vline(xintercept = c(1500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
+  ggtitle("Infant birthweight") +
+  theme_ipsum() +
+  theme(plot.title = element_text(size=15))
[…]

### 2024-02-23 code/processing-code/processingcode.R

@@ -1,0 +1,93 @@
+###############################
+# processing script
+#
+#this script loads the raw data, processes and cleans it 
+#and saves it as Rds file in the processed_data folder
+#
+# Note the ## ---- name ---- notation
+# This is done so one can pull in the chunks of code into the Quarto document
+# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
+
+
+## ---- packages --------
+#Load needed packages. Make sure they are installed.
+library(readxl) #for loading Excel files
+library(tidyverse) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+library(renv) #for package management
+library(knitr) #for nice tables
+library(kableExtra) #for nice tables
+
+## ---- loaddata --------
+# path to data
+# note the use of the here() package and not absolute paths
+data_location <- here::here("data","raw-data","MaternidadeLaranjeiras.csv")
+ML_raw <- read_csv(data_location)
+# Structure of Data
+glimpse(ML_raw)
+
+## ---- checkdata --------
+# Check codebook
+# Path to codebook
+codebook_location <- here::here("data", "raw-data", "Codebook.csv")
+codebook <- read_csv(codebook_location)
+# Full column specification
+spec(codebook)
+# Display the structure of the data frame
+str(codebook)
+# Display the first few rows of the data frame
+head(codebook)
+# Create table
+codebook_table <- kable(codebook, caption = "Codebook") %>% kable_styling()
+#Print table
+print(codebook_table)
+
+## ---- exploredata --------
+# Looking at the data
+dplyr::glimpse(ML_raw)
+summary(ML_raw)
+head(ML_raw)
+skimr::skim(ML_raw)
+
+## ---- cleandata --------
+# Dataset created for multiple reasons. Need to narrow down relevant variables
+# Remove unnecessary variables
+ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Birth_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
+#Summarize missing data using dplyr
+missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
+# Remove variables with missing data not relevant to hypothesis
[…]

### 2024-02-23 products/manuscript/Manuscript_Cassia.qmd
Part 2
Fixed tables with Savannah Hammerton's help

@@
 summary_stats_table <- readRDS(here::here("results", "tables", "summary_stats_table.rds"))
-summary_stats_table
+summary_stats_table |> gt::as_raw_html()
 ```

### 2024-03-15 code/analysis-code/introanalysis-v1.qmd
Project Part 3
This is the first commit of all changes from Project Part 3.

---
title: "Statistical Analysis Script"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling
#rm(list=ls()) #clear the environment
```

Load the data

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear_processed.rds")
#load data
ML_linear <- readRDS(data_location)
```

Let's run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry.

```{r}
# Set reference level for ModifiedColor
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm1 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = ML_linear)

# Summarize model fit with tidiers
ML_lm1_summary <- summary(ML_lm1) # Summary of fit
tidy(ML_lm1, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_lm1) # Model-level statistics

# Produce predictions for RMSE and R-squared
ML_lm1_pred <- predict(ML_lm1, new_data = ML_linear %>% select(-Weightgrams))
ML_lm1_pred

# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
ML_lm1_pred <- bind_cols(ML_lm1_pred, ML_linear %>% select(Weightgrams))
ML_lm1_pred

# Plot observed versus predicted
plot1 <- ggplot(ML_lm1_pred, aes(x = Weightgrams, y = .pred)) +
  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
  geom_point(alpha = 0.5) +
  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
  coord_obs_pred()
plot1

# Create metric set including RMSE and R-squared
metrics_lm1 <- metric_set(rmse, rsq, mae) # MAE for fun
metrics_lm1(ML_lm1_pred, truth = Weightgrams, estimate = .pred)

# Create a gtsummary table directly from the model
table2 <- tbl_regression(ML_lm1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table2

# Save regression output into table
saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
```
Now let's run a second simple bivariate analysis, looking at infant birth weight as a function of maternal age. This time, when plotting weight as a function of maternal age, we can use a scatterplot.

```{r}
# Set reference level for ModifiedColor
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm2 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Age, data = ML_linear)

# Summarize model fit with tidiers
ML_lm2_summary <- summary(ML_lm2) # Summary of fit
tidy(ML_lm2, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_lm2) # Model-level statistics

# Plot data
plot2 <- ggplot(ML_linear, aes(x = Age, y = Weightgrams)) +
  geom_point(alpha = 0.5) +
  labs(x = "Maternal Age", y = "Infant Birth Weight (grams)") + #Scale and size x- and y-axis uniformly
  geom_smooth(method = "lm", se = FALSE) # Add a linear trend line
  theme_minimal() #set plot theme
plot2

# Produce predictions for RMSE and R-squared
ML_lm2_pred <- predict(ML_lm2, new_data = ML_linear %>% select(-Weightgrams))
ML_lm2_pred

# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
ML_lm2_pred <- bind_cols(ML_lm2_pred, ML_linear %>% select(Weightgrams))
ML_lm2_pred

# Create metric set including RMSE and R-squared
metrics_lm1 <- metric_set(rmse, rsq, mae) # MAE for fun
metrics_lm1(ML_lm2_pred, truth = Weightgrams, estimate = .pred)

# Create a gtsummary table directly from the model
table3 <- tbl_regression(ML_lm2, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table3

# Save regression output into table
saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
```

Now let's run a slightly more complicated model that includes maternal age, nationality, and gestational status.

```{r}
# Set reference level for ModifiedColor
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm3 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)

# Summarize model fit with tidiers
ML_lm3_summary <- summary(ML_lm3) # Summary of fit
tidy(ML_lm3, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm3) # Model-level statistics

# Create a gtsummary table directly from the model
table4 <- tbl_regression(ML_lm3, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table4

# Save regression output into table
saveRDS(table4, file = here::here("results", "tables", "table4_final.rds"))
```

In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women. In the second model, we can see that older mothers are associated with giving birth to infants with higher birth weights.

For all three models, the r-squared is very small (0.0073, 0.0061, and 0.027), indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

I am not trying to explain which general factors are trying to explain birth weight. Rather, I'm trying to understand if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil. For example, the gestational age probably explains much of the variation in the model, yet this variable was only included in very few observations in the published clinic data from which I created this dataset.

Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color, however, is also probably associated with premature birth, which is correlated to infant birth weight.

### 2024-03-15 code/eda-code/eda-v1.qmd
Project Part 3
This is the first commit of all changes from Project Part 3.

@@
 
-As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
-Figures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data.
-
-Start by exploring one variable at a time.
-Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-
-Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.
-
 # Setup
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 #load needed packages. make sure they are installed.
-library(tidyverse) #for data processing/cleaning
+library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
 library(skimr) #for nice visualization of data 
@@
 library(gt) #for nice tables
+library(gtsummary) #for summary tables
 library(ggplot2) #for plotting
@@
 
-Load the data.
+Load the data. First we will load the ML_summary set for summary statistics.
 
@@
 
-We are saving the results to the `results` folder. All final table/figures will have `_final` in the name. This is to make sure we don't overwrite anything we might want to keep.
+We are saving the results to the `results` folder.
+All final table/figures will have `_final` in the name.
+This is to make sure we don't overwrite anything we might want to keep.
+All supplemental table/figures will have a `suppl` in the name.
 
@@
 
+Before we create a summary table, lets clean up our variables for easier analysis.
+First, we need to create dummy variables for categories.
+We will start with the `Color` variable, which is maternal skin color.
+We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation.
+The original data categorized maternal racial categories as White (*Branca*), the reference group here; mixed-race (*Parda*); and Black (*Preta*).
+I maintained this categorization.
+However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category.
+For the `Status` variable, we will combine "multigesta", "secundigesta", and "trigesta" into "multigravida", since they mean the same thing (two or more pregnancies) and change the language to English (from gesta to gravida).
+We will also combine `secundipara` and `multipara` into one category `multipara`, since this means two or more viable pregnancies.
+The words in Portuguese and English here are the same.
+For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern".
+I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations.
+Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil.
+"Brazilian" will remain the same.
+
 ```{r}
+# Create dummy variables
+# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
+ML_summary$ModifiedColor <- ifelse(ML_summary$Color == "Branca", "EuroDescent",
+  ifelse(ML_summary$Color == "Preta", "AfroDescent",
+  ifelse(ML_summary$Color == "Parda", "AfroDescent", NA)))
+
+# Or using recode factor from dplyr (easier)
+# ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")
+
+# Create dummy variables for status and change to English
+ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")
+
+# Create dummy variables for nationality and change to English
+ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
+
+# View the updated dataframe
+print(head(ML_summary))
+```
+
+Let's also change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.
+
+```{r}
+# Change observations from Portuguese to English
+ML_summary <- ML_summary %>%
+  mutate(
+    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
+    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
+    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
+    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
+    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
+    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
+    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
+    )
+```
+
+I have now learned that `gtsummary` is a very easy way to create summary statistics (Table 1).
+In the note section at the end of this document, I also include the code I worked on to create the tables manually to show my learning process.
+
+```{r, output="html"}
+# Create gtsummary table 
+table1 <- 
+  ML_summary %>%
+  tbl_summary(
+    include = c(Color, ModifiedColor, ModifiedStatus, Age, Nationality, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Weightgrams, Lengthcentimeters),
+    missing = "no",
+    label = list(ModifiedColor ~ "Ancestry", ModifiedStatus ~ "Parity or Gravidity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Weightgrams ~ "Infant Birthweight (grams)", Lengthcentimeters ~ "Infant Birth Length (centimeters)")
+  ) %>%
+  add_n()
+table1
+
+#Save table
+saveRDS(table1, file = here::here("results", "tables", "table1_final.rds"))
+```
+
+For maternal race, we have a sample size of n = 2695.
+The majority of the sample is White (branca), with 42.45% of the sample.
+Mixed-race (parda) is 29.2% of the sample, and Black (preta) is 28.32% of the sample.
+Let's now compare White patients to all patients of color by combining Black and mixed-race categories.
+I will create a dummy variable for color, called modified_color with two categories: Afro-Descent (comprising parda and preta) and Euro-Descent (comprising branca).
+
+We can see that of the clinic patients, 1551 (57.6%) were of African descent (defined as preta or parda) and 1144 (42.4%) were of European descent (branca).
+
+We can see that of a sample size of n = 2761, 88% of outcomes were natural deliveries.
+
+Now let's take abortion out of this sample size to just look at labor and delivery outcomes.
+
+```{r}
+# Calculate frequencies of birth category without abortion
+birth_freq2 <- ML_summary %>%
+  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
+  filter(Birth != "abortion") %>% # Exclude rows with abortion
+  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
+  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
+  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(birth_freq2)
+```
+We can see that of a new sample size of n = 2672, 91% of outcomes were natural deliveries.
+
+Now let's look at maternal outcomes.
+
+Now let's look at some common reproductive and perinatal health statistics: maternal mortality ratio (MMR), stillbirth rate (SBR), and sex ratio at birth (SRaB).
+The equation is MMR = MD/LB * 10,000.
+We can also look at stillbirth rate.
+The equation is SBR = SB/LB * 1000.
+```{r}
+#MMR Maternal mortality ratio
+maternal_deaths <-23
+live_births <- 2440
+MMR <- (maternal_deaths / live_births) * 10000
+MMR
+
+#Stillbirth rate SBR
+still_births <- 226
+total_births <- 2666
+SBR <- (still_births / total_births) * 1000
+SBR
+
+#Sex ratio at birth (live male births/live female births)*100
+sex_ratio <- ML_summary %>%
+  filter(!Birth %in% c("Abortion") & !FetalOutcome %in% c("Stillbirth or Neonatal Death")) %>% # Exclude rows with abortion and stillbirth
+ filter(!is.na(Sex) & !is.na(Birth) & !is.na(FetalOutcome)) %>%
+  group_by(Sex) %>%
+  summarise(count = n()) %>%
+  spread(key = Sex, value = count) %>%
+  mutate(SexRatio = M / F) * 100
+sex_ratio
+```
+The MMR is 94.3% in this specific clinic.
+The SBR is 84.8% in this specific clinic.
+The sex ratio is 1.2 or there were 120 male live births per 100 female live births at this specific clinic.
+This is much higher than the current range of between 103 and 107 male births per 100 female births.
+
+Because I am working on a different subset of the data, `ML_linear` for the statistical analysis, I will also need to recategorize variables here.
+
+Load the data.
+
+```{r}
+#Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","ML_linear_processed.rds")
+#load data
+ML_linear <- readRDS(data_location)
+```
+
+Showing a bit of code to produce and save a summary table.
+
+```{r}
+# Create summary
+summary = skimr::skim(ML_linear)
+print(summary)
+
+# save to file
+lineartable_file = here("results", "tables", "lineartable.rds")
+saveRDS(summary, file = lineartable_file)
+```
+
+```{r}
+# Create dummy variables
+# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
+ML_linear$ModifiedColor <- ifelse(ML_linear$Color == "Branca", "EuroDescent",
+  ifelse(ML_linear$Color == "Preta", "AfroDescent",
+  ifelse(ML_linear$Color == "Parda", "AfroDescent", NA)))
+
+# Or using recode factor from dplyr (easier)
+# ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")
+
+# Create dummy variables for status and change to English
+ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")
+
+# Create dummy variables for nationality and change to English
+ML_linear$ModifiedNationality <- recode_factor(ML_linear$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
+
+# View the updated dataframe
+print(head(ML_linear))
+
+# Change observations from Portuguese to English
+ML_linear <- ML_linear %>%
+  mutate(
+    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
+    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
+    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
+    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
+    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
+    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
+    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
+    )
+```
+
+# Notes
+
+Here is the previous code I worked on before learning about `gtsummary`.
+
+```{r}
+# Don't use this code in future
 # Summary stats of continuous variables
@@
 #Save table
-saveRDS(summary_stats_table, file = here::here("results", "tables", "summary_stats_table.rds"))
-```
+#saveRDS(summary_stats_table, file = here::here("results", "tables", "summary_stats_table.rds"))
 
-Now let's work with our categorical variables.
-
-```{r}
 # Summary stats of categorical variables
-
 # Calculate frequencies of color category (branca, parda, preta)
@@
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-
 print(color_freq)
-```
-
-For maternal race, we have a sample size of n = 2695.
-The majority of the sample is White (branca), with 42.45% of the sample.
-Mixed-race (parda) is 29.2% of the sample, and Black (preta) is 28.32% of the sample.
-Let's now compare White patients to all patients of color by combining Black and mixed-race categories.
-I will create a dummy variable for color, called modified_color with two categories: AfroDescent (comprising parda and preta) and EuroDescent (comprising branca).
-
-```{r}
-# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
-ML_summary$ModifiedColor <- ifelse(ML_summary$Color == "Branca", "EuroDescent",
-  ifelse(ML_summary$Color == "Preta", "AfroDescent",
-  ifelse(ML_summary$Color == "Parda", "AfroDescent", NA)))
-
-# Or using recode factor from dplyr (easier)
-# ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")
-
-# View the updated dataframe
-print(head(ML_summary))
 
@@
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(color_freq2)
 
-print(color_freq2)
-```
-We can see that of the clinic patients, 1551 (57.6%) were of African descent (defined as preta or parda) and 1144 (42.4%) were of European descent (branca).
-
-Now lets get summary statistics of the other categorical variables, including `Status`, `Nationality`, `Birth`, `MaternalOutcome`, `FetalOutcome` (excluding abortion), and `Sex`.
-We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation.
-For the `Status` variable, we will combine "multigesta" and "trigesta" into "multigesta", since they mean the same thing (more than two pregnancies).
-For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern".
-"Brazilian" will remain the same.
-
-```{r}
 # Calculate frequencies of Status category
@@
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-
 print(status_freq)
-
-# Create dummy variables for status
-ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Trigesta" = "Multigesta")
 
@@
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(status_freq2)
 
-print(status_freq2)
-```
-
-```{r}
 # Calculate frequencies of Nationality category
@@
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-
 print(nationality_freq)
 
-# Create dummy variables for status
-ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
-
-# Calculate frequencies of ModifiedStatus category
+# Calculate frequencies of ModifiedNationality category
 nationality_freq2 <- ML_summary %>%
@@
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(nationality_freq2)
 
-print(nationality_freq2)
-```
-```{r}
+# Calculate frequencies of MaternalOutcome category
+maternal_freq <- ML_summary %>%
+  filter(!is.na(MaternalOutcome)) %>% # Exclude rows with NA in MaternalOutcome variable
+  count(MaternalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
+  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
+  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(maternal_freq)
+
+# Calculate frequencies of Sex category
+sex_freq <- ML_summary %>%
+  filter(!is.na(Sex)) %>% # Exclude rows with NA in Sex variable
+  count(Sex, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
+  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
+  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(sex_freq)
+
 # Calculate frequencies of birth category with abortion
@@
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(birth_freq)
 
-print(birth_freq)
-```
-We can see that of a sample size of n = 2761, 88% of outcomes were natural deliveries. Now let's take abortion out of this sample size to just look at labor and delivery outcomes.
-
-```{r}
-# Calculate frequencies of birth category without abortion
-birth_freq2 <- ML_summary %>%
-  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
-  filter(Birth != "aborto") %>% # Exclude rows with abortion
-  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-
-print(birth_freq2)
-```
-We can see that of a new sample size of n = 2672, 91% of outcomes were natural deliveries. Now let's look at maternal outcomes.
-
-```{r}
-# Calculate frequencies of MaternalOutcome category
-maternal_freq <- ML_summary %>%
-  filter(!is.na(MaternalOutcome)) %>% # Exclude rows with NA in MaternalOutcome variable
-  count(MaternalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-
-print(maternal_freq)
-```
-We can see that of the 2829 women who visited the clinic, 23 died, although not necessarily in childbirth.
-I need to come back to maternal mortality later after I have figured out live births.
-The equation is MMR = MD/LB * 10,000.
-We can also look at stillbirth rate.
-The equation is SBR = SB/LB * 1000.
-
-```{r}
 # Calculate frequencies of FetalOutcome category excluding abortion
@@
   count(FetalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  filter(FetalOutcome != "aborto") %>% # Exclude rows with abortion
+  filter(FetalOutcome != "abortion") %>% # Exclude rows with abortion
   mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
   mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-
 print(fetal_freq)
-
-#MMR Maternal mortality ratio
-maternal_deaths <-23
-live_births <- 2440
-MMR <- (maternal_deaths / live_births) * 10000
-
-#Stillbirth rate SBR
-still_births <- 226
-total_births <- 2666
-SBR <- (still_births / total_births) * 1000
 ```
-The MMR is 94.3% in this specific clinic. The SBR is 84.8% in this specific clinic.
-
-Finally, lets look at infant sex.
-
-```{r}
-# Calculate frequencies of Sex category
-sex_freq <- ML_summary %>%
-  filter(!is.na(Sex)) %>% # Exclude rows with NA in Sex variable
-  count(Sex, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-
-print(sex_freq)
-```
-
-# Notes
-
-I will need to come back and calculate sex ratio in the clinic, but I need to first figure out how to exclude stillbirths from the sample, since sex ratio is calculated male/female * 100 for live births only.

### 2024-03-15 code/processing-code/processingfile-v1.qmd
Project Part 3
This is the first commit of all changes from Project Part 3.

@@
 ```
+
 From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.
@@
 codebook <- read_csv(codebook_location)
-
-# Full column specification
-spec(codebook)
 
@@
 ```
+
 We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations.
@@
 Here, I will create two datasets: one for summary statistics and one for the linear regression model.
-
-The latter (linear regression) will exclude the variables `MaternalOutcome` and `Birth`.
+In addition, most contemporary studies exclude all birthweights less than 500 grams because they have a high chance of being stillborn.
+I do so the same here, even if they were marked as a live birth.
+I will also exclude all stillbirths or neonatal deaths as the analysis is interested in studying live births.
 
@@
 # Creating dataset for linear regression
-ML_linear <- ML_3 %>% tidyr::drop_na()
+ML_linear <- ML_3 %>% tidyr::drop_na() %>%
+  filter(Weightgrams >= 500, FetalOutcome != "morto") # Drop observations of < 500 grams and all non-live births
 
@@
 ```
+
 Now, I can see that my `ML_linear` dataset has no missing data.
-It has 2032 observations and 11 variables.
+It has 1944 observations and 11 variables.
 This might change as I begin analysis.

### 2024-03-15 products/manuscript/Manuscript_Cassia.qmd
Project Part 3
This is the first commit of all changes from Project Part 3.

--- 2024-02-23 products/manuscript/Manuscript_Cassia.qmd
@@
 # Summary/Abstract
+
 This cross-sectional study uses a linear regression model to understand disparities in infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
 I will use a linear model to estimate the relationship of maternal variables on infant birth weight.
-I hypothesize that infants born to women of color will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
+I hypothesize that infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
 
@@
 
-## General Background Information
 Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality [@travassos2004].
-Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts.
-This is a significant public health concern because low birth weight is associated with higher infant mortality, and with long-term health problems such as diabetes, hypertension, and heart disease.
+Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts [@nyarko2013, @wehby2015, @vilanova2019].
+Mothers with lower education levels and lower socioeconomic status also give birth to infants at lower birth weights [@wehby2015, @vilanova2019].
+Understanding these patterns is important as low birth weight is associated with higher rates of infant mortality in the first year of life [@cardoso2005, @vilanova2019].
 
-As a historian of medicine and public health in Brazil, I have collected infant birth weight data from the country's first public maternity hospital, opened in 1904 in the then-capital city of Rio de Janeiro.
-Brazil was the last country in the Western Hemisphere to abolish slavery (1888), and it had, and still has, the largest number of African-descended peoples in the world outside of Nigeria.
-Around 51% of country's population today is of African-descent.
+These trends hold true in the United States, as well, where Black newborns have lower birth weights than White newborns, and these trends have worsened over time [@mangold1991, @pollock2021].
+Racism is a major factor in these outcomes [@ifatunji2022], as studies demonstrate that U.S.-born Black women have infants with lower birth weights and higher rates of pre-term birth than Black immigrant women [@braveman2021, @clay2023, @mckenzie-sampson2021].
+Other racial groups, whether U.S.
+or foreign-born, also face similar trends, although more research is needed to understand confounding factors [@montoya-williams2022].
 
-In the post-abolition period of the 1890s to 1930s, the Brazilian government expanded public health initiatives and the provision of clinical care.
-Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil?
-To date, historians have only provided descriptive examples of health inequities.
+Scholars across disciplines have argued that health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@adamo1983, @otovo2016, @roth2020, @telles2022].
+In Brazil, most historical studies remain descriptive in nature and have not tested any quantitative associations between the legacies of slavery, including racism, and maternal-infant health outcomes [@otovo2016, @roth2018a, @roth2018, @roth2020, @telles2022].
+This study aims to fill that gap.
 
-## Description of data and data source
+It looks at infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Brazil’s first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola), opened in 1904 to provide free gynecological and obstetric care to the city’s poorest women.
 
-I will analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
+It relies on a unique sample of infant birth weight data from the country's first public maternity hospital, opened in 1904 in the then-capital city of Rio de Janeiro.
+Brazil was the last country in the Western Hemisphere to abolish slavery (1888), and it had, and still has, the largest number of African-descended peoples in the world outside of Nigeria [@michelle2023].
+Around 55% of country's population today is of African-descent [@belandi2023].
+
+## Background
+
+Brazil imported over 4 million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009].
+It was also the last country to abolish slavery (1888).
+Healthcare outcomes under chattel slavery were poor for enslaved individuals, evidenced by the lack of endogenous growth among the enslaved population and enslavers’ continued reliance on the transatlantic slave trade (until its abolition in 1850) for new enslaved laborers as mortality rates outpaced fertility rates [@bergad1999, @klein2009, @schwartz1985].
+
+In the immediate aftermath of the abolition of slavery, which coincided with the implementation of a republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into civil and political life occurred [@carvalho2004].
+Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020, @fischer2008, @otovo2016, @roth2020].
+
+Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level [@franken2019].
+The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe.
+Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986].
+But the major advancements in clinical medicine, that improved maternal mortality and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s.
+Worldwide, the combination of advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992, @woods2009].
+
+By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to hospitalize labor and delivery [@barreto2015, @barreto2016, @brenes1991, @martins2004, @mott2002, @roth2020].
+Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
+In hospitals, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine, including new cesarean section techniques [@martins2004, @rohden2009, @roth2021].
+Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which harnessed advances in medical knowledge and public health infrastructure to support a technocratic model towards pregnancy, delivery, and motherhood.
+This ideology relied on essentialized notions of gender, foregrounding physicians’ scientific authority [@besse1996, @freire2009, @martins2008, @otovo2016].
+
+Laranjeiras, which was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces.
+Maternidade Laranjeiras provided gynecological and obstetric care free of charge; thus, its clientele came from the city’s poor and working classes [@barreto2016, @roth2020].
+The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants.
+
+{{< pagebreak >}}
+
+## Data: Description and Sources
+
+In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
 I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO).
@@
 When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician.
+For more information on this as a source, see my other publications [@roth2020, @roth2020a].
 
@@
 Then, between January and August 2017, I manually input the data into Excel from the digital reproductions.
-This will be converted into a .csv file for upload into R.
+I then converted this into a .csv file for upload into R.
 
-## Questions/Hypotheses to be addressed
+## Questions and Hypotheses
 
-I will quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, Maternidade Laranjeiras, in the 1920s.
+I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil?
+I quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, the Maternidade Laranjeiras, in the 1920s.
 I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.
@@
 The outcome I will measure is infant birth weight.
-The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
-I will use the WHO’s classification to categorize birth weight.
+The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
+I am employing birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data.
 
@@
 
-I will use a linear model to estimate the relationship of maternal variables on infant birth weight.
-Originally, I tried using a logistic regression model with very low birthweight (VLBW), low birth weight (LBW), and normal birth weight as the outcome, but after I started taking a class with Dr.
-Swartzendruber, I realized that it is important to understand birthweight as a continuous variable.
+I use a linear model to estimate the relationship of maternal variables on infant birth weight.
 
-**This section was written by GitHub copilot**
-I will use the `lm` function in R to estimate the relationship of maternal variables on infant birth weight.
-I will use the `tidy` function from the `broom` package to extract the coefficients and standard errors from the model.
-I will use the `glance` function from the `broom` package to extract the R-squared value from the model.
-I will use the `augment` function from the `broom` package to extract the residuals from the model.
-I will use the `ggplot2` package to visualize the results.
+## Study Measures
+
+The outcome of interest, birth weight measured in grams, was measured as a continuous variable.
+Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017].
+Infant length was recorded in centimeters.
+Maternal age was recorded in years.
+
+The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*).
+I maintained this categorization.
+However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category.
+Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011, @travassos2004].
+Racial categories and skin color exist on a spectrum rather than a Black-White binary as in the United States [@telles2006].
+Analyzing both specific categories of skin color and more general categories of ancestry is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013, @wehby2015].
+
+Gestational status was divided into parity and gravidity, with categories for first, multiple, and none for each.
+Please see the exploratory data analysis code in this paper's Github repository for further information on this category's reclassifications.
+
+For maternal nationality, I created categories based on individual country and categories based on region.
+The latter broad categories, which I will use in the linear model, include: Brazilian; Latin American (Argentine, Paraguayan, Uruguayan); European (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and Middle Eastern (Syrian).
+
+Clinical outcomes were classified as spontaneous abortion, natural (normal delivery), interventionist (interventionist delivery), and operative (surgical delivery).
+Spontaneous abortion data points are included in the exploratory analysis but excluded from the statistical analysis.
+I also excluded any live births weighing \<500 grams to exclude any possible stillbirths not recorded as such [@nyarko2013, @wehby2015, @wehby2017].
 
@@
 
+If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` and `processingcode.R` in the `processing-code` folder; 2) `eda-v1.qmd` or `edacode.R` in the `eda-cde` folder; 3) `introanalyis-v1.qmd` or `introanalysis.R` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` or `fullanalysiscode.R` in the `analysis-code` folder.
+
+{{< pagebreak >}}
+
 ## Data import and cleaning
 
-Please see the qmd.file `processingfile-v1.qmd1` for full detail on the data import and cleaning process.
+The GitHub repository for this project includes all relevant materials.
+All discussions of data import are detailed in the files `processingfile-v1.qmd` and `processingcode.R` in the `processing-code` folder.
+The original raw data, the codebook, and the processed data are available in the `data` folder.
 
@@
 
-The exploratory analysis was performed on the ML_summary dataset with n = 2845 observations and 14 variables. Please see the Codebook for more details on variables.
+I performed the exploratory analysis on the ML_summary dataset with n = 2845 observations and 14 variables. Please see the Codebook for more details on variables.
 
-There were three continuous variables: maternal age, infant birth weight, and infant length.
-I am aware that this table is not rendering correctly in the word document, but I am unable to figure out why.
-If I load library(gt), then my word document is corrupted and will not open at all.
-I will continue to work on this issue.
+There are three continuous variables: maternal age, infant birth weight, and infant length.
+There are nine categorical variables: skin color; ancestry; parity or gravidity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.
+
 ```{r}
-#| label: tbl-summarytablecont
-#| tbl-cap: "Continous Data Summary Table"
+#| label: tbl-table1
+#| tbl-cap: "Table 1"
 #| echo: FALSE
-summary_stats_table <- readRDS(here::here("results", "tables", "summary_stats_table.rds"))
-summary_stats_table
+table1 <- readRDS(here::here("results", "tables", "table1_final.rds"))
+table_gt <- gtsummary::as_gt(table1)
+table_gt |> gt::as_raw_html()
 ```
 
-Birth weight is the outcome variable of interest, so let's look at some visualizations of the data.
-Here is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO defines as normal birth weight.
+For maternal race (sample size n = 2695), the majority of the sample is White (*branca*), with 42.45% of the sample.
+Mixed-race (*parda*) comprises 29.2%, and Black (*preta*) comprises 28.32%.
+If comparing White patients to all patients of color by combining Black and mixed-race categories, of the clinic patients, 1551 (57.6%) were of African descent (defined as *preta* or *parda*) and 1144 (42.4%) were of European descent (*branca*).
+
+Of all reproductive outcomes (sample size n = 2761), 88% were natural deliveries.
+Yet, this also includes abortions.
+If abortion is excluded (sample size n = 2672), 91% of outcomes were natural deliveries.
+The remaining 9% were interventionist or operative deliveries.
+
+For patients who went to the clinic to deliver their infant (thus excluding those suffering from spontaneous abortions or receiving postpartum care after an at-home delivery), 23 died.
+The Maternal Mortality Ratio (MMR), calculated as:
+
+$$MMR = \frac{MD}{LB} \times 10,000$$
+
+where maternal deaths (MD) are divided by 10,000 live births (LB), was 94.26%.
+For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018, @roth2020].But differential levels of recording must be taken into account.
+These rates could reflect a registration effect, as all births and deaths were recorded in the hospital, whereas accurate reporting at the city level was less reliable [@loudon1992].
+Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s.
+Thus, the city’s rates were probably higher.
+Nonetheless, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother.
+
+The mean stillbirth rate (SBR) (excluding spontaneous abortion but including intrapartum death defined here as stillbirth) for the hospital, calculated as:
+
+$$SBR = \frac{SB}{TB} \times 1,000$$
+where the total number of stillbirths (SB) is divided by 1000 total births (TB).
+The SBR is 84.8% in this specific clinic.
+We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% (@roth2020, @roth2018).
+Again, registration effects could explain the higher rates in the hospital (@loudon1992).
+
+The sex ratio at birth (SRaB) is calculated as:
+
+$$SRaB = \frac{M}{F} \times 100$$
+
+where the total number of live male births (M) is divided by by 100 live female births (F) in a given period.
+
+The sex ratio is 1.2: there were 120 male live births per 100 female live births at this specific clinic.
+This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
+Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring - parents were more likely to terminate a pregnancy or kill an infant if it was a female.
+Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
+However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
+The skewed sex ratio deserves further study.
+
+To better understand our outcome of interest, birth weight, my exploratory analysis includes various visual explorations.
+First, there is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO defines as normal birth weight [@worldhealthorganization2022].
 
@@
 
-Finally, we will look at the relationship between infant weight and length, both not-stratified and stratified by sex. Unsurprisingly, birth weight and length are positively correlated.
+Finally, I have visualized the relationship between infant weight and length. Unsurprisingly, birth weight and length are positively correlated.
 
@@
 ```
+
 Finally, since male infants are usually slightly heavier than females, let's look at a scatterplot of birth weight by length, stratified by sex.
@@
 
-For information on the categorical variables, please see eda-v1.qmd.
-Please note Dr.
-Handel, I am really struggling with making a table of these variables, and I will continue working on that moving forward.
-If you are interested in the code and data, please see the eda-v1.qmd file.
-
 {{< pagebreak >}}
@@
 ## Basic statistical analysis
+
+For the basic statistical analysis, I ran three linear models to understand the relationship between maternal factors and infant birth weight.
+The first model is a simple linear regression with outcome variable (birth weight in grams) and exposure variable maternal ancestry (`Euro-descent` or `Afro-descent1`).
+
+```{r}
+#| label: tbl-table2
+#| tbl-cap: "Table 2"
+#| echo: FALSE
+table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
+table_gt <- gtsummary::as_gt(table2)
+table_gt |> gt::as_raw_html()
+```
+
+The second simple bivariate linear analysis, looks at infant birth weight as a function of maternal age.
+
+```{r}
+#| label: tbl-table3
+#| tbl-cap: "Table 3"
+#| echo: FALSE
+table3 <- readRDS(here::here("results", "tables", "table3_final.rds"))
+table_gt <- gtsummary::as_gt(table3)
+table_gt |> gt::as_raw_html()
+```
+
+The third and final exploratory statistical model is a multilinear regression, looking at the relationship between maternal age, nationality, and gestational status on infant birth weight.
+
+```{r}
+#| label: tbl-table4
+#| tbl-cap: "Table 4"
+#| echo: FALSE
+table4 <- readRDS(here::here("results", "tables", "table4_final.rds"))
+table_gt <- gtsummary::as_gt(table4)
+table_gt |> gt::as_raw_html()
+```
+
+In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women.
+In the second model, older mothers are associated with giving birth to infants with higher birth weights.
+
+For all three models, the r-squared is very small (0.0073, 0.0061, and 0.027), indicating that the model does not explain much of the variance in birth weight.
+This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
+
+Moreover, this study is not trying to explain which general factors are influencing birth weight.
+Rather, it is exploring if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil.
+For example, the gestational age probably explains much of the variation in the model, yet the gestational age variable was only included in very few observations in the published clinic data from which I created this dataset (please see the Codebook and the `processingfile-v1.qmd` for information on this variable).
+
+Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color, however, is also probably associated with premature birth, which is correlated to infant birth weight.
+
+{{< pagebreak >}}
 

### 2024-03-15 code/analysis-code/introanalysis.R

@@ -1,0 +1,125 @@
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+library(renv) #for package management
+library(knitr) #for nice tables
+library(kableExtra) #for nice tables
+library(gt) #for nice tables
+library(gtsummary) #for summary tables
+library(ggplot2) #for plotting
+library(tidymodels) #for modeling
+#rm(list=ls()) #clear the environment
+
+## ---- loaddata --------
+#Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","ML_linear_processed.rds")
+#load data
+ML_linear <- readRDS(data_location)
+
+## ---- lm1 --------
+# Set reference level for ModifiedColor
+ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm1 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_lm1_summary <- summary(ML_lm1) # Summary of fit
+tidy(ML_lm1, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_lm1) # Model-level statistics
+
+# Produce predictions for RMSE and R-squared
+ML_lm1_pred <- predict(ML_lm1, new_data = ML_linear %>% select(-Weightgrams))
+ML_lm1_pred
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_lm1_pred <- bind_cols(ML_lm1_pred, ML_linear %>% select(Weightgrams))
+ML_lm1_pred
+
+# Plot observed versus predicted
+plot1 <- ggplot(ML_lm1_pred, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
[…]

### 2024-03-15 code/eda-code/edacode.R

@@ -1,16 +1,17 @@
 ## ---- packages --------
 #load needed packages. make sure they are installed.
-library(tidyverse) #for data processing/cleaning
+library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
 library(skimr) #for nice visualization of data 
 library(here) #to set paths
 library(renv) #for package management
 library(knitr) #for nice tables
 library(kableExtra) #for nice tables
 library(gt) #for nice tables
+library(gtsummary) #for summary tables
 library(ggplot2) #for plotting
 library(hrbrthemes) #for nice themes
 hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
-rm(list=ls()) #clear the environment
+#rm(list=ls()) #clear the environment
 
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
@@ -87,7 +88,133 @@ plot(length_weight_sex)
 figure_file = here("results","figures","length_weight_sex.png")
 ggsave(filename = figure_file, plot=length_weight_sex, bg="white") #background is white
 
+## ---- Dummyvariables --------
+# Create dummy variables
+# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
+ML_summary$ModifiedColor <- ifelse(ML_summary$Color == "Branca", "EuroDescent",
+                                   ifelse(ML_summary$Color == "Preta", "AfroDescent",
+                                          ifelse(ML_summary$Color == "Parda", "AfroDescent", NA)))
+
+# Or using recode factor from dplyr (easier)
+# ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")
+
+# Create dummy variables for status and change to English
+ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")
+
+# Create dummy variables for nationality and change to English
[…]

### 2024-03-15 code/processing-code/processingcode.R

@@ -28,12 +28,9 @@ ML_raw <- read_csv(data_location)
 glimpse(ML_raw)
 
 ## ---- checkdata --------
-# Check codebook
 # Path to codebook
 codebook_location <- here::here("data", "raw-data", "Codebook.csv")
 codebook <- read_csv(codebook_location)
-# Full column specification
-spec(codebook)
 # Display the structure of the data frame
 str(codebook)
 # Display the first few rows of the data frame
@@ -79,7 +76,8 @@ print(frequency_table)
 # Create dataset for summary statistics (includes missing values)
 ML_summary <- ML_3
 # Creating dataset for linear regression
-ML_linear <- ML_3 %>% tidyr::drop_na()
+ML_linear <- ML_3 %>% tidyr::drop_na() %>%
+  filter(Weightgrams >= 500, FetalOutcome != "morto") # Drop observations of < 500 grams and all non-live births
 # Check for no missing data in linear model
 ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
 View(ML_linear_summary)

### 2024-03-22 code/eda-code/eda-v1.qmd
Debugged code
After Erick's peer feedback

@@
 # Create dummy variables
-# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
-ML_summary$ModifiedColor <- ifelse(ML_summary$Color == "Branca", "EuroDescent",
-  ifelse(ML_summary$Color == "Preta", "AfroDescent",
-  ifelse(ML_summary$Color == "Parda", "AfroDescent", NA)))
 
 # Or using recode factor from dplyr (easier)
-# ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")
+ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
 
@@
 # Create dummy variables
-# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
-ML_linear$ModifiedColor <- ifelse(ML_linear$Color == "Branca", "EuroDescent",
-  ifelse(ML_linear$Color == "Preta", "AfroDescent",
-  ifelse(ML_linear$Color == "Parda", "AfroDescent", NA)))
-
-# Or using recode factor from dplyr (easier)
-# ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")
+# Create dummy variables using recode factor from dplyr
+ ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
 

### 2024-03-22 code/eda-code/eda-v1.qmd
Debugged Code Part 2
Pushing it to get some feedback

@@
 hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
-#rm(list=ls()) #clear the environment
+rm(list=ls()) #clear the environment
 ```
@@
 data_location <- here::here("data","processed-data","ML_summary_processed.rds")
+
 #load data
@@
 We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation.
-The original data categorized maternal racial categories as White (*Branca*), the reference group here; mixed-race (*Parda*); and Black (*Preta*).
+The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*).
 I maintained this categorization.
@@
 # Create dummy variables
+
+#Check pre-structure
+str(ML_linear)
+
 # Create dummy variables using recode factor from dplyr
@@
 ML_linear$ModifiedNationality <- recode_factor(ML_linear$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
-
-# View the updated dataframe
-print(head(ML_linear))
 
@@
     )
+str(ML_linear)
 ```

### 2024-03-22 code/analysis-code/introanalysis.R

@@ -15,6 +15,7 @@ library(tidymodels) #for modeling
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear_processed.rds")
+
 #load data
 ML_linear <- readRDS(data_location)
 

### 2024-03-22 code/eda-code/edacode-extra.R

@@ -1,0 +1,123 @@
+## ---- Extracode --------
+# Don't use this code in future
+# Summary stats of continuous variables
+summary_stats <- ML_summary %>%
+  summarize(
+    mean_age = mean(Age, na.rm = TRUE),
+    sd_age = sd(Age, na.rm = TRUE),
+    mean_weight = mean(Weightgrams, na.rm = TRUE),
+    sd_weight = sd(Weightgrams, na.rm = TRUE),
+    mean_length = mean(Lengthcentimeters, na.rm = TRUE),
+    sd_length = sd(Lengthcentimeters, na.rm = TRUE)
+  )
+
+print(summary_stats)
+
+# Make into a table
+# Create dataframe
+summary_stats1 <- data.frame(
+  Variable = c("Maternal Age", "Infant birthweight (grams)", "Infant birth length (cm)"),
+  Mean = c(25.30435, 3087.341, 48.33493),
+  StandardDev = c(5.759474, 566.5855, 3.92719)
+)
+
+# Create a gt table
+summary_stats_table <- summary_stats1 %>%
+  gt() %>%
+  tab_header(
+    title = "Summary Statistics",
+    subtitle = "Mean Values and Standard Deviations of Continuous Variables"
+  ) %>%
+  cols_label(
+    Variable = "Variable",
+    Mean = "Mean",
+    StandardDev = "Standard Deviation"
+  )
+
+# Print the table
+print(summary_stats_table)
+
+#Save table
+#saveRDS(summary_stats_table, file = here::here("results", "tables", "summary_stats_table.rds"))
+
+# Summary stats of categorical variables
+# Calculate frequencies of color category (branca, parda, preta)
+color_freq <- ML_summary %>%
+  filter(!is.na(Color)) %>% # Exclude rows with NA in color variable
+  count(Color, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
+  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
+  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
+print(color_freq)
+
+# Calculate frequencies of color category (branca, parda, preta)
+color_freq2 <- ML_summary %>%
+  filter(!is.na(ModifiedColor)) %>% # Exclude rows with NA in color variable
+  count(ModifiedColor, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
+  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
[…]

### 2024-03-22 code/eda-code/edacode.R

@@ -16,6 +16,7 @@ hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_summary_processed.rds")
+
 #load data
 ML_summary <- readRDS(data_location)
 
@@ -23,6 +24,7 @@ ML_summary <- readRDS(data_location)
 # Create summary
 summary = skimr::skim(ML_summary)
 print(summary)
+
 # save to file
 summarytable_file = here("results", "tables", "summarytable.rds")
 saveRDS(summary, file = summarytable_file)
@@ -65,6 +67,7 @@ length_plot <- ML_summary %>% filter(!is.na(Lengthcentimeters)) %>% #filter out 
   theme_ipsum() +
   theme(plot.title = element_text(size=15))
 plot(length_plot)
+
 # Save plot to file
 figure_file = here("results","figures","length_distribution.png")
 ggsave(filename = figure_file, plot=length_plot, bg="white") #background is white
@@ -75,6 +78,7 @@ length_weight <- ML_summary %>%
   filter(!is.na(Lengthcentimeters) & !is.na(Weightgrams)) %>% #filter out NAs
   ggplot(aes(x=Lengthcentimeters, y=Weightgrams)) + geom_point() + geom_smooth(method='lm')
 plot(length_weight)
+
 #Save file
 figure_file = here("results","figures","length_weight.png")
 ggsave(filename = figure_file, plot=length_weight, bg="white") #background is white
@@ -84,19 +88,16 @@ ggsave(filename = figure_file, plot=length_weight, bg="white") #background is wh
 length_weight_sex <- ML_summary %>% filter(!is.na(Lengthcentimeters) & !is.na(Weightgrams) & !is.na(Sex)) %>% 
   ggplot(aes(x=Lengthcentimeters, y=Weightgrams, color = Sex)) + geom_point() + geom_smooth(method='lm')
 plot(length_weight_sex)
+
 #Save file
 figure_file = here("results","figures","length_weight_sex.png")
 ggsave(filename = figure_file, plot=length_weight_sex, bg="white") #background is white
 
 ## ---- Dummyvariables --------
 # Create dummy variables
-# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
[…]

### 2024-03-22 code/analysis-code/introanalysis-v1.qmd
Debugged Analysis Code
After Zane's help

@@
 #Path to summary data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","ML_linear_processed.rds")
+data_location <- here::here("data","processed-data","ML_linear_final.rds")
 #load data
-ML_linear <- readRDS(data_location)
+ML_linear_final <- readRDS(data_location)
 ```

### 2024-03-22 code/eda-code/eda-v1.qmd
Debugged Analysis Code
After Zane's help

@@
     )
+
+# Save new summary data with recoded factor variables
+saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary_final.rds"))
 ```
@@
     )
-str(ML_linear)
+
+# Save new summary data with recoded factor variables
+saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear_final.rds"))
 ```
@@
 
-Here is the previous code I worked on before learning about `gtsummary`.
-
-```{r}
-# Don't use this code in future
-# Summary stats of continuous variables
-summary_stats <- ML_summary %>%
-  summarize(
-    mean_age = mean(Age, na.rm = TRUE),
-    sd_age = sd(Age, na.rm = TRUE),
-    mean_weight = mean(Weightgrams, na.rm = TRUE),
-    sd_weight = sd(Weightgrams, na.rm = TRUE),
-    mean_length = mean(Lengthcentimeters, na.rm = TRUE),
-    sd_length = sd(Lengthcentimeters, na.rm = TRUE)
-  )
-
-print(summary_stats)
-
-# Make into a table
-# Create dataframe
-summary_stats1 <- data.frame(
-  Variable = c("Maternal Age", "Infant birthweight (grams)", "Infant birth length (cm)"),
-  Mean = c(25.30435, 3087.341, 48.33493),
-  StandardDev = c(5.759474, 566.5855, 3.92719)
-)
-
-# Create a gt table
-summary_stats_table <- summary_stats1 %>%
-  gt() %>%
-  tab_header(
-    title = "Summary Statistics",
-    subtitle = "Mean Values and Standard Deviations of Continuous Variables"
-  ) %>%
-  cols_label(
-    Variable = "Variable",
-    Mean = "Mean",
-    StandardDev = "Standard Deviation"
-  )
-
-# Print the table
-print(summary_stats_table)
-
-#Save table
-#saveRDS(summary_stats_table, file = here::here("results", "tables", "summary_stats_table.rds"))
-
-# Summary stats of categorical variables
-# Calculate frequencies of color category (branca, parda, preta)
-color_freq <- ML_summary %>%
-  filter(!is.na(Color)) %>% # Exclude rows with NA in color variable
-  count(Color, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(color_freq)
-
-# Calculate frequencies of color category (branca, parda, preta)
-color_freq2 <- ML_summary %>%
-  filter(!is.na(ModifiedColor)) %>% # Exclude rows with NA in color variable
-  count(ModifiedColor, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(color_freq2)
-
-# Calculate frequencies of Status category
-status_freq <- ML_summary %>%
-  filter(!is.na(Status)) %>% # Exclude rows with NA in status variable
-  count(Status, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(status_freq)
-
-# Calculate frequencies of ModifiedStatus category
-status_freq2 <- ML_summary %>%
-  filter(!is.na(ModifiedStatus)) %>% # Exclude rows with NA in status variable
-  count(ModifiedStatus, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(status_freq2)
-
-# Calculate frequencies of Nationality category
-nationality_freq <- ML_summary %>%
-  filter(!is.na(Nationality)) %>% # Exclude rows with NA in nationality variable
-  count(Nationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(nationality_freq)
-
-# Calculate frequencies of ModifiedNationality category
-nationality_freq2 <- ML_summary %>%
-  filter(!is.na(ModifiedNationality)) %>% # Exclude rows with NA in nationality variable
-  count(ModifiedNationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(nationality_freq2)
-
-# Calculate frequencies of MaternalOutcome category
-maternal_freq <- ML_summary %>%
-  filter(!is.na(MaternalOutcome)) %>% # Exclude rows with NA in MaternalOutcome variable
-  count(MaternalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(maternal_freq)
-
-# Calculate frequencies of Sex category
-sex_freq <- ML_summary %>%
-  filter(!is.na(Sex)) %>% # Exclude rows with NA in Sex variable
-  count(Sex, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(sex_freq)
-
-# Calculate frequencies of birth category with abortion
-birth_freq <- ML_summary %>%
-  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
-  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(birth_freq)
-
-# Calculate frequencies of FetalOutcome category excluding abortion
-fetal_freq <- ML_summary %>%
-  filter(!is.na(FetalOutcome)) %>% # Exclude rows with NA in FetalOutcome variable
-  count(FetalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
-  filter(FetalOutcome != "abortion") %>% # Exclude rows with abortion
-  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
-  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
-print(fetal_freq)
-```
+For the previous code I worked on in relation to summary stats, see the `edacode-extra.R` file.

### 2024-03-22 code/analysis-code/introanalysis.R

@@ -14,10 +14,10 @@ library(tidymodels) #for modeling
 
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","ML_linear_processed.rds")
+data_location <- here::here("data","processed-data","ML_linear_final.rds")
 
 #load data
-ML_linear <- readRDS(data_location)
+ML_linear_final <- readRDS(data_location)
 
 ## ---- lm1 --------
 # Set reference level for ModifiedColor

### 2024-03-22 code/eda-code/edacode.R

@@ -121,6 +121,9 @@ ML_summary <- ML_summary %>%
     FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
   )
 
+# Save new summary data with recoded factor variables
+saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary_final.rds"))
+
 ## ---- Summarystats --------
 # Create gtsummary table 
 table1 <- 
@@ -134,7 +137,7 @@ table1 <-
 table1
 
 #Save table
-saveRDS(summary_stats_table, file = here::here("results", "tables", "Table1_final.rds"))
+saveRDS(table1, file = here::here("results", "tables", "table1_final.rds"))
 
 ## ---- Summarystatscont --------
 # Calculate frequencies of birth category without abortion
@@ -172,7 +175,6 @@ sex_ratio
 ##---- LinearData --------
 #Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear_processed.rds")
-
 #load data
 ML_linear <- readRDS(data_location)
 
@@ -185,8 +187,10 @@ lineartable_file = here("results", "tables", "lineartable.rds")
 saveRDS(summary, file = lineartable_file)
 
 # Create dummy variables
+#Check pre-structure
+str(ML_linear)
 
-# Or using recode factor from dplyr (easier)
+# Create dummy variables using recode factor from dplyr
 ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
 
 # Create dummy variables for status and change to English
@@ -194,9 +198,6 @@ ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Mu
 
 # Create dummy variables for nationality and change to English
[…]

### 2024-04-05 code/analysis-code/fullanalysis-v1.qmd
Project Part 4
Completed full analysis and updated all files.

---
title: "Full Analysis Script"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling

rm(list=ls()) #clear the environment
```

Load the data

```{r, message=FALSE,  warning=FALSE}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)
```

The first simple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$

whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term. Results are displayed in Table 2.

```{r}
# Model using recipes and workflows
# Define recipe for model
ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable

# Specify model
ML_spec1 <- linear_reg() %>% 
  set_engine("lm")

# Create work flow
ML_wf1 <- workflow() %>%
  add_recipe(ML_recipe1) %>%
  add_model(ML_spec1)

# Fit the work flow
ML_fit1 <- fit(ML_wf1, data = ML_linear)

# Summarize model fit with tidiers
ML_summary1 <- summary(ML_fit1) # Summary of fit
tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_fit1) # Model-level statistics

# Produce predictions for R-squared
ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
ML_fit1_pred

# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
ML_fit1_pred <- bind_cols(ML_fit1_pred, ML_linear %>% select(Weightgrams))
ML_fit1_pred

# Plot observed versus predicted
plot1 <- ggplot(ML_fit1_pred, aes(x = Weightgrams, y = .pred)) +
  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
  geom_point(alpha = 0.5) +
  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
  coord_obs_pred()
plot1

# Create metric set including RMSE and R-squared
metrics_fit1 <- metric_set(rmse, rsq)
metrics_fit1(ML_fit1_pred, truth = Weightgrams, estimate = .pred)

# Convert the output to a data frame and mutate the .estimate column to numeric
metrics_fit1 <- as.data.frame(metrics_fit1(ML_fit1_pred, truth = Weightgrams, estimate = .pred)) %>%
  mutate(.estimate = as.numeric(.estimate))
options(scipen = 999) # set opens to avoid scientific notation
metrics_fit1

# Create a gtsummary table directly from the model
table2 <-
  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
table2

#Print table
table2

# Save regression output into table
saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
```

The second multiple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$

whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), Gest is gestational status (nullipara, primipara, multipara, multigravida), Age is maternal age in years, and $\epsilon$ is the error term. Results are displayed in Table 3.

```{r}
# Using recipes and workflows
# Define recipe for model
ML_recipe2 <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>% 
  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors

# Define model specification
ML_spec2 <- linear_reg() %>% 
  set_engine("lm") %>%
  set_mode("regression")

# Create workflow
ML_wf2 <- workflow() %>%
  add_recipe(ML_recipe2) %>%
  add_model(ML_spec2)

# Fit the workflow
ML_fit2 <- fit(ML_wf2, data = ML_linear)

# Summarize model fit with tidiers
ML_lm2_summary <- summary(ML_fit2) # Summary of fit
tidy(ML_fit2, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_fit2) # Model-level statistics

# Create a gtsummary table directly from the model
table3 <- pull_workflow_fit(ML_fit2) %>%
  tbl_regression(exponentiate = FALSE, intercept = TRUE) %>%
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
  bold_p()
table3

# Save regression output into table
saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
```

The second two models are also simple and multiple linear regressions with outcome variable as birthweight in grams. However, in these two models, I maintained maternal skin color as three categories, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women. The first simple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$
whereas BW is birthweight in grams, MC is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.

```{r}
# Using recipes and workflows
# Simple linear model
# Specify recipe
ML_recipe3 <- recipe(Weightgrams ~ Color, data = ML_linear) %>%
  step_relevel(Color, ref_level = "White")

# Specify model
ML_spec3 <- linear_reg() %>%
  set_engine("lm")

# Create workflow
ML_wf3 <- workflow() %>%
  add_recipe(ML_recipe3) %>%
  add_model(ML_spec3)

# Fit the workflow
ML_fit3 <- fit(ML_wf3, data = ML_linear)

# Summarize model fit with tidiers and create a gtsummary table directly from the model
ML_summary3 <- summary(ML_fit3)
tidy(ML_fit3, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_fit3) # Model-level statistics

# Create a gtsummary table from model
table4 <- tbl_regression(ML_fit3, exponentiate = FALSE, intercept = TRUE) %>%
  bold_p() %>%
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
table4

# Save regression output into table
saveRDS(table4, file = here("results", "tables", "table4_final.rds"))
```

The second multiple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$

whereas BW is birthweight in grams, MC is maternal skin color (White, Black and mixed race), Gest is gestational status (nullipara, primipara, multipara, multigravida), Nat is maternal nationality (Brazilian, Latin American, European, Middle Eastern), Age is maternal age in years, and $\epsilon$ is the error term.

```{r}
# Using reipes and workflows
# Complex linear model
# Specify recipe
ML_recipe4 <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
  step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
  step_relevel(ModifiedStatus, ref_level = "Multipara") %>% # Set reference level for ModifiedStatus
  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors

# Specify model
ML_spec4 <- linear_reg() %>%
  set_engine("lm")

# Create workflow
ML_wf4 <- workflow() %>%
  add_recipe(ML_recipe4) %>%
  add_model(ML_spec4)

# Fit the workflow
ML_fit4 <- fit(ML_wf4, data = ML_linear)

# Summarize model fit with tidiers and create a gtsummary table directly from the model
ML_summary4 <- summary(ML_fit4)
tidy(ML_fit4, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_fit4) # Model-level statistics

# Create a gtsummary table from model
table5 <- tbl_regression(ML_fit4, exponentiate = FALSE, intercept = TRUE) %>%
  bold_p() %>%
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
table5

# Save regression output into table
saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
```

### 2024-04-05 code/analysis-code/introanalysis-v1.qmd
Project Part 4
Completed full analysis and updated all files.

--- 2024-03-15 code/analysis-code/introanalysis-v1.qmd
@@
 ---
-title: "Statistical Analysis Script"
+title: "Intro Statistical Analysis Script"
 author: "Cassia Roth"
@@
 library(tidymodels) #for modeling
+
 #rm(list=ls()) #clear the environment
@@
 #Path to summary data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","ML_linear_processed.rds")
+data_location <- here::here("data","processed-data","ML_linear.rds")
 #load data
@@
 
-Let's run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry.
+Let's run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry. I will use the recipe and workflow features of `tidymodels`.
 
 ```{r}
-# Set reference level for ModifiedColor
-ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+# Model using recipes and workflows
+# Define recipe for model
+ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
 
-# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm1 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = ML_linear)
+# Specify model
+ML_spec1 <- linear_reg() %>% 
+  set_engine("lm")
+
+# Create work flow
+ML_wf1 <- workflow() %>%
+  add_recipe(ML_recipe1) %>%
+  add_model(ML_spec1)
+
+# Fit the work flow
+ML_fit1 <- fit(ML_wf1, data = ML_linear)
 
 # Summarize model fit with tidiers
-ML_lm1_summary <- summary(ML_lm1) # Summary of fit
-tidy(ML_lm1, conf.int = TRUE) # Coefficients and confidence intervals
-glance(ML_lm1) # Model-level statistics
+ML_summary1 <- summary(ML_fit1) # Summary of fit
+tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit1) # Model-level statistics
 
-# Produce predictions for RMSE and R-squared
-ML_lm1_pred <- predict(ML_lm1, new_data = ML_linear %>% select(-Weightgrams))
-ML_lm1_pred
+# Produce predictions for R-squared
+ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
+ML_fit1_pred
 
 # Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
-ML_lm1_pred <- bind_cols(ML_lm1_pred, ML_linear %>% select(Weightgrams))
-ML_lm1_pred
+ML_fit1_pred <- bind_cols(ML_fit1_pred, ML_linear %>% select(Weightgrams))
+ML_fit1_pred
 
 # Plot observed versus predicted
-plot1 <- ggplot(ML_lm1_pred, aes(x = Weightgrams, y = .pred)) +
+plot1 <- ggplot(ML_fit1_pred, aes(x = Weightgrams, y = .pred)) +
   geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
@@
 # Create metric set including RMSE and R-squared
-metrics_lm1 <- metric_set(rmse, rsq, mae) # MAE for fun
-metrics_lm1(ML_lm1_pred, truth = Weightgrams, estimate = .pred)
+metrics_fit1 <- metric_set(rmse, rsq)
+metrics_fit1(ML_fit1_pred, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit1 <- as.data.frame(metrics_fit1(ML_fit1_pred, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit1
 
 # Create a gtsummary table directly from the model
-table2 <- tbl_regression(ML_lm1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
+table2 <-
+  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
+
+#Print table
 table2
@@
 # Save regression output into table
-saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
+saveRDS(table2, file = here::here("results", "tables", "table2.rds"))
 ```
@@
 ```{r}
-# Set reference level for ModifiedColor
-ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+# Model using recipes and workflows
+# Create recipe
+ML_recipe_age <- recipe(Weightgrams ~ Age, data = ML_linear)
 
-# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm2 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Age, data = ML_linear)
+# Specify model
+ML_spec_age <- linear_reg() %>% 
+  set_engine("lm")
+
+# Create work flow
+ML_wf_age <- workflow() %>%
+  add_recipe(ML_recipe_age) %>%
+  add_model(ML_spec_age)
+
+# Fit the work flow
+ML_fit_age <- fit(ML_wf_age, data = ML_linear)
 
 # Summarize model fit with tidiers
-ML_lm2_summary <- summary(ML_lm2) # Summary of fit
-tidy(ML_lm2, conf.int = TRUE) # Coefficients and confidence intervals
-glance(ML_lm2) # Model-level statistics
-
-# Plot data
-plot2 <- ggplot(ML_linear, aes(x = Age, y = Weightgrams)) +
-  geom_point(alpha = 0.5) +
-  labs(x = "Maternal Age", y = "Infant Birth Weight (grams)") + #Scale and size x- and y-axis uniformly
-  geom_smooth(method = "lm", se = FALSE) # Add a linear trend line
-  theme_minimal() #set plot theme
-plot2
+ML_summary_age <- summary(ML_fit_age) # Summary of fit
+tidy(ML_fit_age, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit_age) # Model-level statistics
 
 # Produce predictions for RMSE and R-squared
-ML_lm2_pred <- predict(ML_lm2, new_data = ML_linear %>% select(-Weightgrams))
-ML_lm2_pred
+ML_age_pred <- predict(ML_fit_age, new_data = ML_linear %>% select(-Weightgrams))
+ML_age_pred
 
 # Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
-ML_lm2_pred <- bind_cols(ML_lm2_pred, ML_linear %>% select(Weightgrams))
-ML_lm2_pred
+ML_age_pred <- bind_cols(ML_age_pred, ML_linear %>% select(Weightgrams))
+ML_age_pred
+
+# Plot observed versus predicted
+plot_age <- ggplot(ML_age_pred, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot_age
 
 # Create metric set including RMSE and R-squared
-metrics_lm1 <- metric_set(rmse, rsq, mae) # MAE for fun
-metrics_lm1(ML_lm2_pred, truth = Weightgrams, estimate = .pred)
+metrics_fit_age <- metric_set(rmse, rsq)
+metrics_fit_age(ML_age_pred, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit_age <- as.data.frame(metrics_fit_age(ML_age_pred, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit_age
 
 # Create a gtsummary table directly from the model
-table3 <- tbl_regression(ML_lm2, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
+table_age <-
+  tbl_regression(ML_fit_age, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
-table3
+
+#Print table
+table_age
 
 # Save regression output into table
-saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
+saveRDS(table_age, file = here::here("results", "tables", "tableage_final.rds"))
 ```
 
-Now let's run a slightly more complicated model that includes maternal age, nationality, and gestational status.
+Now let's run a slightly more complicated model that includes maternal skin color, age, nationality, and gestational status.
 
@@
 # Create a gtsummary table directly from the model
-table4 <- tbl_regression(ML_lm3, exponentiate = FALSE, intercept = TRUE) %>% 
+table3 <- tbl_regression(ML_lm3, exponentiate = FALSE, intercept = TRUE) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
-table4
+table3
 
 # Save regression output into table
-saveRDS(table4, file = here::here("results", "tables", "table4_final.rds"))
+saveRDS(table3, file = here::here("results", "tables", "table3.rds"))
 ```
@@
 Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color, however, is also probably associated with premature birth, which is correlated to infant birth weight.
+
+Now let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race).
+We'll start with a simple model of infant birth weight as a function of maternal skin color.
+Then we will run a model with all the predictors.
+
+```{r}
+# Simple linear model
+# Set reference level for Color
+ML_linear$Color <- relevel(ML_linear$Color, ref = "White")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_lm4_summary <- summary(ML_lm4) # Summary of fit
+tidy(ML_lm4, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_lm4) # Model-level statistics
+
+# Create a gtsummary table directly from the model
+table4 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table4
+
+# Save regression output into table
+saveRDS(table4, file = here::here("results", "tables", "table4.rds"))
+
+# Complex linear model
+# Set reference level for Color
+ML_linear$Color <- relevel(ML_linear$Color, ref = "White")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_lm5_summary <- summary(ML_lm5) # Summary of fit
+tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_lm5) # Model-level statistics
+
+# Create a gtsummary table directly from the model
+table5 <- tbl_regression(ML_lm5, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table5
+
+# Save regression output into table
+saveRDS(table5, file = here::here("results", "tables", "table5.rds"))
+```

### 2024-04-05 code/eda-code/eda-v1.qmd
Project Part 4
Completed full analysis and updated all files.

@@
 # Create dummy variables for status and change to English
-ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")
+ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nullipara", "Nulipara" = "Nullipara")
 
@@
 # Save new summary data with recoded factor variables
-saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary_final.rds"))
+saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
 ```
@@
 
-Showing a bit of code to produce and save a summary table.
-
-```{r}
-# Create summary
-summary = skimr::skim(ML_linear)
-print(summary)
-
-# save to file
-lineartable_file = here("results", "tables", "lineartable.rds")
-saveRDS(summary, file = lineartable_file)
-```
+Now I will create dummy variables.
 
@@
 # Create dummy variables using recode factor from dplyr
- ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
+ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Preta" = "Afro-Descent", "Parda" = "Afro-Descent", "Branca" = "Euro-Descent")
 
 # Create dummy variables for status and change to English
-ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")
+ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Nullipara", "Nulipara" = "Nullipara")
 
@@
   mutate(
-    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
     ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
@@
 
+str(ML_linear)
+
 # Save new summary data with recoded factor variables
-saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear_final.rds"))
+saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear.rds"))
 ```

### 2024-04-05 products/manuscript/Manuscript_Cassia.qmd
Project Part 4
Completed full analysis and updated all files.

@@
 library(renv) #for package management
+library(gtsummary) #for summary tables
 ```
@@
 
-This cross-sectional study uses a linear regression model to understand disparities in infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
-I will use a linear model to estimate the relationship of maternal variables on infant birth weight.
+This cross-sectional study uses linear regression models to understand disparities in infant birth weight in relation to maternal skin color or ancestry, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
 I hypothesize that infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
@@
 
-It looks at infant birth weight in relation to maternal race, age, gestational status (parity or gravidity), and nationality in Brazil’s first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola), opened in 1904 to provide free gynecological and obstetric care to the city’s poorest women.
-
+It looks at infant birth weight in relation to maternal skin color or ancestry, age, gestational status (parity or gravidity), and nationality in Brazil’s first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola), which provided free gynecological and obstetric care to the city’s poorest women.
 It relies on a unique sample of infant birth weight data from the country's first public maternity hospital, opened in 1904 in the then-capital city of Rio de Janeiro.
-Brazil was the last country in the Western Hemisphere to abolish slavery (1888), and it had, and still has, the largest number of African-descended peoples in the world outside of Nigeria [@michelle2023].
-Around 55% of country's population today is of African-descent [@belandi2023].
 
@@
 Brazil imported over 4 million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009].
-It was also the last country to abolish slavery (1888).
+It was also the last country to abolish slavery (1888) in the Western Hemisphere.
+It had, and still has, the largest number of African-descended peoples in the world outside of Nigeria [@michelle2023].
+Around 55% of country's population today is of African-descent [@belandi2023].
+
 Healthcare outcomes under chattel slavery were poor for enslaved individuals, evidenced by the lack of endogenous growth among the enslaved population and enslavers’ continued reliance on the transatlantic slave trade (until its abolition in 1850) for new enslaved laborers as mortality rates outpaced fertility rates [@bergad1999, @klein2009, @schwartz1985].
@@
 Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013].
-
-Clinical notes only sporadically included infant demographic information for spontaneous abortions or stillbirths.
-For spontaneous abortions, stillbirths, and live births, I recorded, when available, infant sex, weight, length.
-I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy.
-However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraud1997), coded as operatory or natural by physicians, as interventionist.
-
-In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures.
-Over time, clinical notes became streamlined.
-When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician.
-For more information on this as a source, see my other publications [@roth2020, @roth2020a].
-
-The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca de Biomedicina-A, Universidade Estadual do Rio de Janeiro (BBA-UERJ), all in Rio de Janeiro, Brazil.
-Between January 2012 and July 2013, I manually digitized the journal by photographing each volume.
-Then, between January and August 2017, I manually input the data into Excel from the digital reproductions.
-I then converted this into a .csv file for upload into R.
+Please see the `Supplemental_Materials` document for more information on the dataset.
 
@@
 I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil?
-I quantify how maternal race, nationality, age, and gravidity or parity explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, the Maternidade Laranjeiras, in the 1920s.
+I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, the Maternidade Laranjeiras, in the 1920s.
 I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.
@@
 The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
-I am employing birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data.
+I employ birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data.
 
@@
 
-I use a linear model to estimate the relationship of maternal variables on infant birth weight.
+I use linear models to estimate the relationship of maternal variables on infant birth weight.
 
@@
 
-The outcome of interest, birth weight measured in grams, was measured as a continuous variable.
+The outcome of interest, birth weight in grams, was measured as a continuous variable.
 Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017].
+I discuss this constraint in the limitations section below.
 Infant length was recorded in centimeters.
-Maternal age was recorded in years.
 
-The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*).
-I maintained this categorization.
-However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category.
+The covariates include maternal age (recorded as a continuous variable in years), maternal skin color, gestational status, and maternal nationality.
+
+I run two multivariate models: one in which maternal skin color is categorized into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent).
+The second categorizes maternal skin color into the dataset's original racial groups: White (the reference group), Black, and mixed race.
 Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011, @travassos2004].
@@
 
-Gestational status was divided into parity and gravidity, with categories for first, multiple, and none for each.
-Please see the exploratory data analysis code in this paper's Github repository for further information on this category's reclassifications.
-
-For maternal nationality, I created categories based on individual country and categories based on region.
-The latter broad categories, which I will use in the linear model, include: Brazilian; Latin American (Argentine, Paraguayan, Uruguayan); European (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and Middle Eastern (Syrian).
-
-Clinical outcomes were classified as spontaneous abortion, natural (normal delivery), interventionist (interventionist delivery), and operative (surgical delivery).
 Spontaneous abortion data points are included in the exploratory analysis but excluded from the statistical analysis.
@@
 
+Please see the supplemental materials (`Supplemental_Materials`) for information on the data processing and exploratory analysis.
+
 ## Schematic of workflow
 
-If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` and `processingcode.R` in the `processing-code` folder; 2) `eda-v1.qmd` or `edacode.R` in the `eda-cde` folder; 3) `introanalyis-v1.qmd` or `introanalysis.R` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` or `fullanalysiscode.R` in the `analysis-code` folder.
+If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder; 2) `eda-v1.qmd` or `edacode.R` in the `eda-code` folder; 3) `introanalyis-v1.qmd` or `introanalysis.R` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` or `fullanalysiscode.R` in the `analysis-code` folder.
+Please defer to all `.qmd` files as the master over any `.R` files.
 
@@
 The GitHub repository for this project includes all relevant materials.
-All discussions of data import are detailed in the files `processingfile-v1.qmd` and `processingcode.R` in the `processing-code` folder.
+All discussions of data import are detailed in the files `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder.
 The original raw data, the codebook, and the processed data are available in the `data` folder.
@@
 
-## Exploratory/Descriptive analysis
+## Descriptive analysis
 
-I performed the exploratory analysis on the ML_summary dataset with n = 2845 observations and 14 variables. Please see the Codebook for more details on variables.
-
-There are three continuous variables: maternal age, infant birth weight, and infant length.
-There are nine categorical variables: skin color; ancestry; parity or gravidity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.
+I performed the exploratory analysis on the ML_summary dataset with n = 2845 observations and 14 variables. Table 1 demonstrates the distribution of the data.
 
@@
 Of all reproductive outcomes (sample size n = 2761), 88% were natural deliveries.
-Yet, this also includes abortions.
-If abortion is excluded (sample size n = 2672), 91% of outcomes were natural deliveries.
+Yet, this also includes spontaneous abortions.
+If we exclude spontaneous abortion (sample size n = 2672), 91% of outcomes were natural deliveries.
 The remaining 9% were interventionist or operative deliveries.
@@
 where maternal deaths (MD) are divided by 10,000 live births (LB), was 94.26%.
-For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018, @roth2020].But differential levels of recording must be taken into account.
+For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018, @roth2020].
+But differential levels of recording must be taken into account.
 These rates could reflect a registration effect, as all births and deaths were recorded in the hospital, whereas accurate reporting at the city level was less reliable [@loudon1992].
@@
 This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
-Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring - parents were more likely to terminate a pregnancy or kill an infant if it was a female.
+Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was a female.
 Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
@@
 
-To better understand our outcome of interest, birth weight, my exploratory analysis includes various visual explorations.
-First, there is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO defines as normal birth weight [@worldhealthorganization2022].
+To better understand our outcome of interest, birth weight, the exploratory analysis includes the distribution of the data.
+Figure 1 is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO now defines as normal birth weight [@worldhealthorganization2022].
 
@@
 
-Birth length is less important for our analysis, but we might want to look at the relationship between weight and length.
-Below we can see a histogram of the distribution of birth lengths in the sample.
-The red line marks the average birth length for both male and female infants (49 centimeters).
+Because the data are missing gestational age, birth length is less important for our analysis.
+Nonetheless, Figure 2 shows a histogram of the distribution of birth lengths in the sample.
+The red line marks the average birth length for both male and female infants (49 centimeters), which is in line with current estimates on birth length globally [@jamshed2020].
 
@@
 
-Finally, I have visualized the relationship between infant weight and length. Unsurprisingly, birth weight and length are positively correlated.
+Figure 3 visualizes the relationship between infant weight and length. Unsurprisingly, birth weight and length are positively correlated.
 
@@
 
-Finally, since male infants are usually slightly heavier than females, let's look at a scatterplot of birth weight by length, stratified by sex.
+Finally, Figure 4 visualizes the relationship of birth weight and length, stratified by sex since male infants are usually slightly heavier and longer than females [@crawford1987, @marsal1996].
 
@@
 
-## Basic statistical analysis
+## Statistical analysis
 
-For the basic statistical analysis, I ran three linear models to understand the relationship between maternal factors and infant birth weight.
-The first model is a simple linear regression with outcome variable (birth weight in grams) and exposure variable maternal ancestry (`Euro-descent` or `Afro-descent1`).
+I ran four linear models to understand the relationship between maternal factors and infant birth weight.
+The first two models are a simple and multiple linear regression, respectively, with outcome variable as birthweight in grams.
+In these two models, I recoded maternal skin color as maternal ancestry (`Euro-descent` or `Afro-descent`).
+
+The first simple linear regression model is as follows:
+
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$
+whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term. Results are displayed in Table 2.
 
@@
 
-The second simple bivariate linear analysis, looks at infant birth weight as a function of maternal age.
+The second multiple linear regression model is as follows:
+
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$
+whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term.
+Results are displayed in Table 3.
 
@@
 
-The third and final exploratory statistical model is a multilinear regression, looking at the relationship between maternal age, nationality, and gestational status on infant birth weight.
+The second two models are also simple and multiple linear regressions with outcome variable as birthweight in grams.
+However, in these two models, I maintained maternal skin color as three categories, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women.
+The first simple linear regression model is as follows:
+
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$
+whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.
 
@@
 
-In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women.
-In the second model, older mothers are associated with giving birth to infants with higher birth weights.
+The second multiple linear regression model is as follows:
 
-For all three models, the r-squared is very small (0.0073, 0.0061, and 0.027), indicating that the model does not explain much of the variance in birth weight.
-This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$
+whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), Nat is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+Results are displayed in Table 5.
 
-Moreover, this study is not trying to explain which general factors are influencing birth weight.
-Rather, it is exploring if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil.
-For example, the gestational age probably explains much of the variation in the model, yet the gestational age variable was only included in very few observations in the published clinic data from which I created this dataset (please see the Codebook and the `processingfile-v1.qmd` for information on this variable).
+```{r}
+#| label: tbl-table5
+#| tbl-cap: "Table 5"
+#| echo: FALSE
+table5 <- readRDS(here::here("results", "tables", "table5_final.rds"))
+table_gt <- gtsummary::as_gt(table5)
+table_gt |> gt::as_raw_html()
+```
 
-Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color, however, is also probably associated with premature birth, which is correlated to infant birth weight.
+Given that this study is an inferential, hypothesis-supporting approach that is exploring how specific maternal predictors affect the outcome of interest, birth weight, I am excluding any predictive metrics.
+These models are not examples to be used in other contexts, which suggests I should not apply machine-learning methods such as cross-validation or test/train.
 
-{{< pagebreak >}}
+## Results
 
-## Full analysis
+All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference group) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race).
+Age and nationality were not significant.
+Women who were nulliparous (never carried a pregnancy to viability) and women who were multigravida (had multiple pregnancies but no viable deliveries) had infants with lower birth weights.
+No significance was found for women who were delivering their first viable child (primipara).
 
@@
 
-## Summary and Interpretation
-
 ## Strengths and Limitations
 
+All models based on historical data must contend with limited sample sizes and missing data.
+In this study, a major limitation was the exclusion of gestational age from the original published clinical notes upon which this dataset is based.
+A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
+Lack of data on maternal nutrition and health status, which cannot be corrected for by using race as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in past time periods [@loudon1992, @woods2009].
+
+For all models in this study, the $\(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight.
+This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors [@leimert2020]
+
+Moreover, this study is not trying to explain which general factors are influencing birth weight.
+Rather, it is exploring if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil.
+For example, the gestational age probably explains much of the variation in the model, yet the gestational age variable was only included in very few observations in the published clinic data from which I created this dataset (please see the Codebook and the `Supplemental_materials.qmd` file for information on this variable).
+
+Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color, however, is also probably associated with premature birth, which is correlated to infant birth weight.
+
 ## Conclusions
+
+Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between skin color and health in the early twentieth century.
 

### 2024-04-05 products/manuscript/supplement/Supplemental_Materials.qmd
Project Part 4
Completed full analysis and updated all files.

---
title: "Supplemental Materials"
author: Cassia Roth
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
    bibliography: ../../../assets/Mada_Bib.bib
    csl: ../../../assets/vancouver.csl
---
# Supplemental Materials

## Data Source

In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro. The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*. In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*. In 1922, it became the *Revista de Gynecologia e D’Obstetricia*.

The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). From the available issues, I recorded the following information, when available, for all patients: patient number, gravidity and parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history. Gravidity refers to the total number of pregnancies a woman has, regardless of duration. Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013].

In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician. For more information on this as a source, see my other publications [@roth2020, @roth2020a].

The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca de Biomedicina-A, Universidade Estadual do Rio de Janeiro (BBA-UERJ), all in Rio de Janeiro, Brazil. Between January 2012 and July 2013, I manually digitized the journal by photographing each volume. Then, between January and August 2017, I manually input the data into Excel from the digital reproductions. I then converted this into a .csv file for upload into R.

## Study Measures

There are three continuous variables: maternal age, infant birth weight, and infant length. There are nine categorical variables: skin color; ancestry; parity or gravidity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.

### Maternal Skin Color

The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category. Please see the `eda-v1.qmd` file for more information on the variable manipulation process.

### Gestational Status

Gestational status was divided into parity and gravidity. Originally, the data had eight categories: nullipara, primipara, secundipara, multipara; and primigravida, secundigravida, trigravida, and multigravida. I combined categories into four final categories: 1) nullipara and primigravida into `nullipara` (or any woman who has never given birth to a viable fetus); 2) primipara into `primipara` (or any woman who has given birth to one viable fetus); 3) secundipara and multipara into `multipara` (or any woman who has given birth to more than one viable fetus); and 4) secundigravida, trigravida, and multigravida into `multigravida` (or any woman who has been pregnant more than one but never given birth to a viable fetus).

### Maternal Nationality

For maternal nationality, I created categories based on individual country (`nationality`) and categories based on region (`modifiednationality`). Recategorization occurred as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian).

### Clinical Outcomes

The clinical notes upon which I based this dataset only sporadically included infant demographic information for spontaneous abortions or stillbirths. For spontaneous abortions, stillbirths, and live births, I recorded, when available, infant sex, weight, length. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy. However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraud1997), coded as operatory or natural by physicians, as interventionist. I classified clinical outcomes as spontaneous abortion, natural (normal delivery), interventionist (interventionist delivery), and operative (surgical delivery).

## Basic Statistical Analysis

For the basic statistical analysis, I ran three linear models to understand the relationship between maternal factors and infant birth weight. The first model is a simple linear regression with outcome variable (birthweight in grams) and exposure variable maternal ancestry (`Euro-descent` or `Afro-descent`).

```{r}
#| label: tbl-table2
#| tbl-cap: "Supplemental Table 1"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
table_gt <- gtsummary::as_gt(table2)
table_gt |> gt::as_raw_html()
```

The second simple bivariate linear analysis, looks at infant birth weight as a function of maternal age.

```{r}
#| label: tbl-table_age
#| tbl-cap: "Supplemental Table 2"
#| echo: FALSE
table_age <- readRDS(here::here("results", "tables", "tableage_final.rds"))
table_gt <- gtsummary::as_gt(table_age)
table_gt |> gt::as_raw_html()
```

The third and final exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, nationality, and gestational status on infant birth weight.

```{r}
#| label: tbl-table4
#| tbl-cap: "Supplemental Table 3"
#| echo: FALSE
table4 <- readRDS(here::here("results", "tables", "table4.rds"))
table_gt <- gtsummary::as_gt(table4)
table_gt |> gt::as_raw_html()
```

In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women. In the second model, older mothers are associated with giving birth to infants with higher birth weights.

For all three models, the r-squared is very small (0.007, 0.006, and 0.008), indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

### 2024-04-05 code/analysis-code/fullanalysis.R

@@ -1,0 +1,197 @@
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+library(renv) #for package management
+library(knitr) #for nice tables
+library(kableExtra) #for nice tables
+library(gt) #for nice tables
+library(gtsummary) #for summary tables
+library(ggplot2) #for plotting
+library(tidymodels) #for modeling
+
+#rm(list=ls()) #clear the environment
+
+## ---- loaddata --------
+#Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","ML_linear.rds")
+
+#load data
+ML_linear <- readRDS(data_location)
+
+## ---- Model1 --------
+# Model using recipes and workflows
+# Define recipe for model
+ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
+
+# Specify model
+ML_spec1 <- linear_reg() %>% 
+  set_engine("lm")
+
+# Create work flow
+ML_wf1 <- workflow() %>%
+  add_recipe(ML_recipe1) %>%
+  add_model(ML_spec1)
+
+# Fit the work flow
+ML_fit1 <- fit(ML_wf1, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_summary1 <- summary(ML_fit1) # Summary of fit
+tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit1) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
+ML_fit1_pred
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit1_pred <- bind_cols(ML_fit1_pred, ML_linear %>% select(Weightgrams))
+ML_fit1_pred
+
+# Plot observed versus predicted
[…]

### 2024-04-05 code/analysis-code/introanalysis.R

@@ -14,33 +14,45 @@ library(tidymodels) #for modeling
 
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","ML_linear_final.rds")
+data_location <- here::here("data","processed-data","ML_linear.rds")
 
 #load data
-ML_linear_final <- readRDS(data_location)
+ML_linear <- readRDS(data_location)
 
 ## ---- lm1 --------
-# Set reference level for ModifiedColor
-ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+# Model using recipes and workflows
+# Define recipe for model
+ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
 
-# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm1 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = ML_linear)
+# Specify model
+ML_spec1 <- linear_reg() %>% 
+  set_engine("lm")
+
+# Create work flow
+ML_wf1 <- workflow() %>%
+  add_recipe(ML_recipe1) %>%
+  add_model(ML_spec1)
+
+# Fit the work flow
+ML_fit1 <- fit(ML_wf1, data = ML_linear)
 
 # Summarize model fit with tidiers
-ML_lm1_summary <- summary(ML_lm1) # Summary of fit
-tidy(ML_lm1, conf.int = TRUE) # Coefficients and confidence intervals
-glance(ML_lm1) # Model-level statistics
+ML_summary1 <- summary(ML_fit1) # Summary of fit
+tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit1) # Model-level statistics
 
-# Produce predictions for RMSE and R-squared
-ML_lm1_pred <- predict(ML_lm1, new_data = ML_linear %>% select(-Weightgrams))
-ML_lm1_pred
+# Produce predictions for R-squared
+ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
+ML_fit1_pred
 
[…]

### 2024-04-05 code/eda-code/edacode.R

@@ -122,7 +122,7 @@ ML_summary <- ML_summary %>%
   )
 
 # Save new summary data with recoded factor variables
-saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary_final.rds"))
+saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
 
 ## ---- Summarystats --------
 # Create gtsummary table 
@@ -178,23 +178,15 @@ data_location <- here::here("data","processed-data","ML_linear_processed.rds")
 #load data
 ML_linear <- readRDS(data_location)
 
-# Create summary
-summary = skimr::skim(ML_linear)
-print(summary)
-
-# save to file
-lineartable_file = here("results", "tables", "lineartable.rds")
-saveRDS(summary, file = lineartable_file)
-
 # Create dummy variables
 #Check pre-structure
 str(ML_linear)
 
 # Create dummy variables using recode factor from dplyr
-ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
+ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Preta" = "Afro-Descent", "Parda" = "Afro-Descent", "Branca" = "Euro-Descent")
 
 # Create dummy variables for status and change to English
-ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")
+ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Nullipara", "Nulipara" = "Nullipara")
 
 # Create dummy variables for nationality and change to English
[…]

### 2024-04-18 code/analysis-code/fullanalysis-v1.qmd
Project Part 5
Finished ML analysis and redid tables in final manuscript.

@@
 
-rm(list=ls()) #clear the environment
+#rm(list=ls()) #clear the environment
 ```
@@
 
+# Preparation for Model Assessment
+
+First, I will create train and test data (75/25) to be able to evaluate the performance of our models.
+I will also set the seed to ensure reproducibility.
+For each model, I perform the following steps: 1) run the model on the train data; 2) perform cross-validation on the train data; 3) evaluate the model on the test data.
+The model outputs included in the final manuscript are on all the data.
+
+```{r, message=FALSE,  warning=FALSE}
+# Fix the random numbers by setting the seed 
+# This enables the analysis to be reproducible when random numbers are used 
+rngseed = 1234
+set.seed(rngseed)
+
+# Create training data
+# Put 3/4 of the data into the training set 
+data_split <- initial_split(ML_linear, prop = 3/4)
+
+# Create data frames for the two sets:
+train_data <- training(data_split)
+test_data  <- testing(data_split)
+```
+
+# Model 1
+
 The first simple linear regression model is as follows:
@@
 
-whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term. Results are displayed in Table 2.
+whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
+
+Full results are displayed in Table 2.
 
 ```{r}
+# Model 1: Final model for paper using all data
 # Model using recipes and workflows
@@
 ```
+## Model 1 Performance
+
+First we will evaluate the model on our training data.
+
+```{r}
+# Training model 1
+# Model using recipes and workflows
+# Define recipe for model
+ML_recipe1_train <- recipe(Weightgrams ~ ModifiedColor, data = train_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
+
+# Specify model
+ML_spec1_train <- linear_reg() %>% 
+  set_engine("lm")
+
+# Create work flow
+ML_wf1_train <- workflow() %>%
+  add_recipe(ML_recipe1_train) %>%
+  add_model(ML_spec1_train)
+
+# Fit the work flow
+ML_fit1_train <- fit(ML_wf1_train, data = train_data)
+
+# Summarize model fit with tidiers
+ML_summary1_train <- summary(ML_fit1_train) # Summary of fit
+tidy(ML_fit1_train, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit1_train) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit1_pred_train <- predict(ML_fit1_train, new_data = train_data %>% select(-Weightgrams))
+ML_fit1_pred_train
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit1_pred_train <- bind_cols(ML_fit1_pred_train, train_data %>% select(Weightgrams))
+ML_fit1_pred_train
+
+# Plot observed versus predicted
+plot1_train <- ggplot(ML_fit1_pred_train, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot1_train
+
+# Create metric set including RMSE and R-squared
+metrics_fit1_train <- metric_set(rmse, rsq)
+metrics_fit1_train(ML_fit1_pred_train, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit1_train <- as.data.frame(metrics_fit1_train(ML_fit1_pred_train, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit1_train
+
+# Create a gtsummary table directly from the model
+table2_train <-
+  tbl_regression(ML_fit1_train, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
+table2_train
+
+# Save regression output into table
+saveRDS(table2_train, file = here::here("results", "tables", "table2_train.rds"))
+```
+Our results here are not that different from the model run on all the data.
+Our adjusted R-squared is 0.007, the same as the full model.
+The RMSE is 495, which is slightly lower than the full model (502), which shows better performance.
+
+Now let's perform cross-validation on the training data.
+
+```{r, message=FALSE,  warning=FALSE}
+# CV model 1
+# Set the seed for reproducibility
+set.seed(rngseed)
+
+# Create 10-fold cross-validation
+folds <- vfold_cv(train_data, v = 10)
+folds
+
+# Define recipe for model
+ML_recipe1_cv <- recipe(Weightgrams ~ ModifiedColor, data = train_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
+
+# Set model specification
+ML_spec1_cv <- linear_reg() %>% set_engine("lm")
+
+# Create and initialize workflow that bundles together model specification and formula.
+ML_wf1_cv <- workflow() %>% 
+  add_model(ML_spec1_cv) %>% 
+  add_recipe(ML_recipe1_cv)
+  fit_resamples(ML_wf1_cv, resamples = folds)
+
+# Resample the model
+ML_wf1_cv_rs <- fit_resamples(ML_wf1_cv, resamples = folds)
+ML_wf1_cv_rs
+
+# Get model metrics
+ML_cv1_metrics <- collect_metrics(ML_wf1_cv_rs)
+
+# Filter the data frame to include only rows with "rmse" and "rsq" in the ".metric" column
+ML_cv1_metrics <- ML_cv1_metrics[ML_cv1_metrics$.metric %in% c("rmse", "rsq"), !(names(ML_cv1_metrics) %in% ".config")]
+ML_cv1_metrics
+
+# Save regression output into table
+saveRDS(ML_cv1_metrics, file = here::here("results", "tables", "ML_cv1_metrics.rds"))
+```
+Performing cross-validation results in an RMSE that is the same as our training model (495), lower than our full model.
+The r-squared (0.02) is also higher than our training model (0.008) and our full model (0.007), demonstrating better model performance.
+
+Finally, let's run the model on our test data.
+
+```{r, message=FALSE,  warning=FALSE}
+# Testing model 1
+# Model using recipes and workflows
+# Define recipe for model
+ML_recipe1_test <- recipe(Weightgrams ~ ModifiedColor, data = test_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
+
+# Specify model
+ML_spec1_test <- linear_reg() %>% 
+  set_engine("lm")
+
+# Create work flow
+ML_wf1_test <- workflow() %>%
+  add_recipe(ML_recipe1_test) %>%
+  add_model(ML_spec1_test)
+
+# Fit the work flow
+ML_fit1_test <- fit(ML_wf1_test, data = test_data)
+
+# Summarize model fit with tidiers
+ML_summary1_test <- summary(ML_fit1_test) # Summary of fit
+tidy(ML_fit1_test, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit1_test) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit1_pred_test <- predict(ML_fit1_test, new_data = test_data %>% select(-Weightgrams))
+ML_fit1_pred_test
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit1_pred_test <- bind_cols(ML_fit1_pred_test, test_data %>% select(Weightgrams))
+ML_fit1_pred_test
+
+# Plot observed versus predicted
+plot1_test <- ggplot(ML_fit1_pred_test, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot1_test
+
+# Create metric set including RMSE and R-squared
+metrics_fit1_test <- metric_set(rmse, rsq)
+metrics_fit1_test(ML_fit1_pred_test, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit1_test <- as.data.frame(metrics_fit1_test(ML_fit1_pred_test, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit1_test
+
+# Create a gtsummary table directly from the model
+table2_test <-
+  tbl_regression(ML_fit1_test, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
+table2_test
+
+# Save regression output into table
+saveRDS(table2_test, file = here::here("results", "tables", "table2_test.rds"))
+```
+
+When running the model on our testing dataset, our results are not significant, given the small sample size.
+
+# Model 2
 
@@
 
-whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), Gest is gestational status (nullipara, primipara, multipara, multigravida), Age is maternal age in years, and $\epsilon$ is the error term.
+whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), Gest is gestational status (nulliparous or multiparous), Age is maternal age in years, and $\epsilon$ is the error term.
 Results are displayed in Table 3.
@@
 ```{r}
+# Final model 2 for paper using all data
 # Using recipes and workflows
@@
 
+# Produce predictions for R-squared
+ML_fit2_pred <- predict(ML_fit2, new_data = ML_linear %>% select(-Weightgrams))
+ML_fit2_pred
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit2_pred <- bind_cols(ML_fit2_pred, ML_linear %>% select(Weightgrams))
+ML_fit2_pred
+
+# Plot observed versus predicted
+plot2 <- ggplot(ML_fit2_pred, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot2
+
+# Create metric set including RMSE and R-squared
+metrics_fit2 <- metric_set(rmse, rsq)
+metrics_fit2(ML_fit2_pred, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit2 <- as.data.frame(metrics_fit2(ML_fit2_pred, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit2
+
 # Create a gtsummary table directly from the model
@@
 ```
+Our RMSE is 498 and our r-squared is 0.02.
+
+## Model 2 Performance
+
+Now let's run the model on our training data.
+
+```{r}
+# Training model 2
+# Using recipes and workflows
+# Define recipe for model
+ML_recipe2_train <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = train_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+
+# Define model specification
+ML_spec2_train <- linear_reg() %>% 
+  set_engine("lm") %>%
+  set_mode("regression")
+
+# Create workflow
+ML_wf2_train <- workflow() %>%
+  add_recipe(ML_recipe2_train) %>%
+  add_model(ML_spec2_train)
+
+# Fit the workflow
+ML_fit2_train <- fit(ML_wf2_train, data = train_data)
+
+# Summarize model fit with tidiers
+ML_lm2_summary_train <- summary(ML_fit2_train) # Summary of fit
+tidy(ML_fit2_train, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_fit2_train) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit2_pred_train <- predict(ML_fit2_train, new_data = train_data %>% select(-Weightgrams))
+ML_fit2_pred_train
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit2_pred_train <- bind_cols(ML_fit2_pred_train, train_data %>% select(Weightgrams))
+ML_fit2_pred_train
+
+# Plot observed versus predicted
+plot2_train <- ggplot(ML_fit2_pred_train, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot2_train
+
+# Create metric set including RMSE and R-squared
+metrics_fit2_train <- metric_set(rmse, rsq)
+metrics_fit2_train(ML_fit2_pred_train, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit2_train <- as.data.frame(metrics_fit2_train(ML_fit2_pred_train, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit2_train
+
+# Create a gtsummary table directly from the model
+table3_train <- pull_workflow_fit(ML_fit2_train) %>%
+  tbl_regression(exponentiate = FALSE, intercept = TRUE) %>%
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
+  bold_p()
+table3_train
+
+# Save regression output into table
+saveRDS(table3_train, file = here::here("results", "tables", "table3_train.rds"))
+```
+Our RMSE is 490 and our r-squared is 0.03, a slightly better performance than the full model.
+
+Now let's run cross-validation.
+
+```{r, message=FALSE,  warning=FALSE}
+# CV model 2
+# Set the seed for reproducibility
+set.seed(rngseed)
+
+# Create 10-fold cross-validation
+folds <- vfold_cv(train_data, v = 10)
+folds
+
+# Define recipe for model
+ML_recipe2_cv <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = train_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+
+# Set model specification
+ML_spec2_cv <- linear_reg() %>% set_engine("lm")
+
+# Create and initialize workflow that bundles together model specification and formula.
+ML_wf2_cv <- workflow() %>% 
+  add_model(ML_spec2_cv) %>% 
+  add_recipe(ML_recipe2_cv)
+  fit_resamples(ML_wf2_cv, resamples = folds)
+
+# Resample the model
+ML_wf2_cv_rs <- fit_resamples(ML_wf2_cv, resamples = folds)
+ML_wf2_cv_rs
+
+# Get model metrics
+ML_cv2_metrics <- collect_metrics(ML_wf2_cv_rs)
+
+# Filter the data frame to include only rows with "rmse" and "rsq" in the ".metric" column
+ML_cv2_metrics <- ML_cv2_metrics[ML_cv2_metrics$.metric %in% c("rmse", "rsq"), !(names(ML_cv2_metrics) %in% ".config")]
+ML_cv2_metrics
+
+# Save regression output into table
+saveRDS(ML_cv2_metrics, file = here::here("results", "tables", "ML_cv2_metrics.rds"))
+```
+Here, our RMSE is 458, an improvement from both the full model and the training model, and our r-squared is 0.03, the same as our training model.
+
+Finally, let's run our model on the test data.
+
+```{r}
+# Testing model 2
+# Using recipes and workflows
+# Define recipe for model
+ML_recipe2_test <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = test_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+
+# Define model specification
+ML_spec2_test <- linear_reg() %>% 
+  set_engine("lm") %>%
+  set_mode("regression")
+
+# Create workflow
+ML_wf2_test <- workflow() %>%
+  add_recipe(ML_recipe2_test) %>%
+  add_model(ML_spec2_test)
+
+# Fit the workflow
+ML_fit2_test <- fit(ML_wf2_test, data = test_data)
+
+# Summarize model fit with tidiers
+ML_lm2_summary_test <- summary(ML_fit2_test) # Summary of fit
+tidy(ML_fit2_test, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_fit2_test) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit2_pred_test <- predict(ML_fit2_test, new_data = test_data %>% select(-Weightgrams))
+ML_fit2_pred_test
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit2_pred_test <- bind_cols(ML_fit2_pred_test, test_data %>% select(Weightgrams))
+ML_fit2_pred_test
+
+# Plot observed versus predicted
+plot2_test <- ggplot(ML_fit2_pred_test, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot2_test
+
+# Create metric set including RMSE and R-squared
+metrics_fit2_test <- metric_set(rmse, rsq)
+metrics_fit2_test(ML_fit2_pred_test, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit2_test <- as.data.frame(metrics_fit2_test(ML_fit2_pred_test, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit2_test
+
+# Create a gtsummary table directly from the model
+table3_test <- pull_workflow_fit(ML_fit2_test) %>%
+  tbl_regression(exponentiate = FALSE, intercept = TRUE) %>%
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
+  bold_p()
+table3_test
+
+# Save regression output into table
+saveRDS(table3_test, file = here::here("results", "tables", "table3_test.rds"))
+```
+Running the model on our test data, we see that the only significant result is for the nulliparous status. All other results are insignificant.
+
+# Model 3
 
@@
 ```{r}
+# Final model 3 for paper using all data
 # Using recipes and workflows
@@
 
+# Produce predictions for R-squared
+ML_fit3_pred <- predict(ML_fit3, new_data = ML_linear %>% select(-Weightgrams))
+ML_fit3_pred
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit3_pred <- bind_cols(ML_fit3_pred, ML_linear %>% select(Weightgrams))
+ML_fit3_pred
+
+# Plot observed versus predicted
+plot3 <- ggplot(ML_fit3_pred, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot3
+
+# Create metric set including RMSE and R-squared
+metrics_fit3 <- metric_set(rmse, rsq)
+metrics_fit3(ML_fit3_pred, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit3 <- as.data.frame(metrics_fit3(ML_fit3_pred, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit3
+
 # Create a gtsummary table from model
@@
 
+Our RMSE is 502 and our r-squared is 0.007.
+
+## Model 3 Performance
+
+Let's now evaluate the model on our training data.
+
+```{r, message=FALSE,  warning=FALSE}
+# Training model 3
+# Using recipes and workflows
+# Specify recipe
+ML_recipe3_train <- recipe(Weightgrams ~ Color, data = train_data) %>%
+  step_relevel(Color, ref_level = "White")
+
+# Specify model
+ML_spec3_train <- linear_reg() %>%
+  set_engine("lm")
+
+# Create workflow
+ML_wf3_train <- workflow() %>%
+  add_recipe(ML_recipe3_train) %>%
+  add_model(ML_spec3_train)
+
+# Fit the workflow
+ML_fit3_train <- fit(ML_wf3_train, data = train_data)
+
+# Summarize model fit with tidiers and create a gtsummary table directly from the model
+ML_summary3_train <- summary(ML_fit3_train)
+tidy(ML_fit3_train, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit3_train) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit3_pred_train <- predict(ML_fit3_train, new_data = train_data %>% select(-Weightgrams))
+ML_fit3_pred_train
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit3_pred_train <- bind_cols(ML_fit3_pred_train, train_data %>% select(Weightgrams))
+ML_fit3_pred_train
+
+# Plot observed versus predicted
+plot3_train <- ggplot(ML_fit3_pred_train, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot3_train
+
+# Create metric set including RMSE and R-squared
+metrics_fit3_train <- metric_set(rmse, rsq)
+metrics_fit3_train(ML_fit3_pred_train, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit3_train <- as.data.frame(metrics_fit3_train(ML_fit3_pred_train, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit3_train
+
+# Create a gtsummary table from model
+table4_train <- tbl_regression(ML_fit3_train, exponentiate = FALSE, intercept = TRUE) %>%
+  bold_p() %>%
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
+table4_train
+
+# Save regression output into table
+saveRDS(table4_train, file = here("results", "tables", "table4_train.rds"))
+```
+
+Our RMSE is 495 and our r-squared is 0.009, slightly better performance than the full model. Our results are not significantly different from the full model.
+
+Now let's run cross-validation.
+
+```{r, message=FALSE,  warning=FALSE}
+# CV model 3
+# Set the seed for reproducibility
+set.seed(rngseed)
+
+# Create 10-fold cross-validation
+folds <- vfold_cv(train_data, v = 10)
+folds
+
+# Define recipe for model
+ML_recipe3_cv <- recipe(Weightgrams ~ Color, data = train_data) %>%
+  step_relevel(Color, ref_level = "White")
+
+# Set model specification
+ML_spec3_cv <- linear_reg() %>% set_engine("lm")
+
+# Create and initialize workflow that bundles together model specification and formula.
+ML_wf3_cv <- workflow() %>% 
+  add_model(ML_spec3_cv) %>% 
+  add_recipe(ML_recipe3_cv)
+  fit_resamples(ML_wf3_cv, resamples = folds)
+
+# Resample the model
+ML_wf3_cv_rs <- fit_resamples(ML_wf3_cv, resamples = folds)
+ML_wf3_cv_rs
+
+# Get model metrics
+collect_metrics(ML_wf3_cv_rs)
+
+# Get model metrics
+ML_cv3_metrics <- collect_metrics(ML_wf3_cv_rs)
+
+# Filter the data frame to include only rows with "rmse" and "rsq" in the ".metric" column
+ML_cv3_metrics <- ML_cv3_metrics[ML_cv3_metrics$.metric %in% c("rmse", "rsq"), !(names(ML_cv3_metrics) %in% ".config")]
+ML_cv3_metrics
+
+# Save regression output into table
+saveRDS(ML_cv3_metrics, file = here::here("results", "tables", "ML_cv3_metrics.rds"))
+```
+Using cross-validation, our RMSE is 495 and our r-squared is 0.02, a similar performance to our training model.
+
+Finally, let's run our model on the test data.
+
+```{r, message=FALSE, warning=FALSE}
+# Testing model 3
+# Using recipes and workflows
+# Specify recipe
+ML_recipe3_test <- recipe(Weightgrams ~ Color, data = test_data) %>%
+  step_relevel(Color, ref_level = "White")
+
+# Specify model
+ML_spec3_test <- linear_reg() %>%
+  set_engine("lm")
+
+# Create workflow
+ML_wf3_test <- workflow() %>%
+  add_recipe(ML_recipe3_test) %>%
+  add_model(ML_spec3_test)
+
+# Fit the workflow
+ML_fit3_test <- fit(ML_wf3_test, data = test_data)
+
+# Summarize model fit with tidiers and create a gtsummary table directly from the model
+ML_summary3_test <- summary(ML_fit3_test)
+tidy(ML_fit3_test, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit3_test) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit3_pred_test <- predict(ML_fit3_test, new_data = test_data %>% select(-Weightgrams))
+ML_fit3_pred_test
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit3_pred_test <- bind_cols(ML_fit3_pred_test, test_data %>% select(Weightgrams))
+ML_fit3_pred_test
+
+# Plot observed versus predicted
+plot3_test <- ggplot(ML_fit3_pred_test, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot3_test
+
+# Create metric set including RMSE and R-squared
+metrics_fit3_test <- metric_set(rmse, rsq)
+metrics_fit3_test(ML_fit3_pred_test, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit3_test <- as.data.frame(metrics_fit3_test(ML_fit3_pred_test, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit3_test
+
+# Create a gtsummary table from model
+table4_test <- tbl_regression(ML_fit3_test, exponentiate = FALSE, intercept = TRUE) %>%
+  bold_p() %>%
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
+table4_test
+
+# Save regression output into table
+saveRDS(table4_test, file = here("results", "tables", "table4_test.rds"))
+```
+
+Much like in the other testing models, our results are not significant (except the intercept). Our performance is also worse (RMSE of 518 and r-squared of 0.005).
+
+# Model 4
+
 The second multiple linear regression model is as follows:
@@
 ```{r}
-# Using reipes and workflows
+# Final model 4 for paper using all data
+# Using recipes and workflows
 # Complex linear model
@@
   step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
-  step_relevel(ModifiedStatus, ref_level = "Multipara") %>% # Set reference level for ModifiedStatus
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
   step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
@@
 
+# Produce predictions for R-squared
+ML_fit4_pred <- predict(ML_fit4, new_data = ML_linear %>% select(-Weightgrams))
+ML_fit4_pred
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit4_pred <- bind_cols(ML_fit4_pred, ML_linear %>% select(Weightgrams))
+ML_fit4_pred
+
+# Plot observed versus predicted
+plot4 <- ggplot(ML_fit4_pred, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot4
+
+# Create metric set including RMSE and R-squared
+metrics_fit4 <- metric_set(rmse, rsq)
+metrics_fit4(ML_fit4_pred, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit4 <- as.data.frame(metrics_fit4(ML_fit4_pred, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit4
+
 # Create a gtsummary table from model
@@
 ```
+Our RMSE is 498 and our r-squared is 0.02.
+
+## Model 4 Performance
+
+Let's now evaluate the model on our training data.
+
+```{r, message=FALSE,  warning=FALSE}
+# Model 4 training
+# Using recipes and workflows
+# Complex linear model
+# Specify recipe
+ML_recipe4_train <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = train_data) %>%
+  step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
+  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+
+# Specify model
+ML_spec4_train <- linear_reg() %>%
+  set_engine("lm")
+
+# Create workflow
+ML_wf4_train <- workflow() %>%
+  add_recipe(ML_recipe4_train) %>%
+  add_model(ML_spec4_train)
+
+# Fit the workflow
+ML_fit4_train <- fit(ML_wf4_train, data = train_data)
+
+# Summarize model fit with tidiers and create a gtsummary table directly from the model
+ML_summary4_train <- summary(ML_fit4_train)
+tidy(ML_fit4_train, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit4_train) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit4_pred_train <- predict(ML_fit4_train, new_data = train_data %>% select(-Weightgrams))
+ML_fit4_pred_train
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit4_pred_train <- bind_cols(ML_fit4_pred_train, train_data %>% select(Weightgrams))
+ML_fit4_pred_train
+
+# Plot observed versus predicted
+plot4_train <- ggplot(ML_fit4_pred_train, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot4_train
+
+# Create metric set including RMSE and R-squared
+metrics_fit4_train <- metric_set(rmse, rsq)
+metrics_fit4_train(ML_fit4_pred_train, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit4_train <- as.data.frame(metrics_fit4_train(ML_fit4_pred_train, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit4_train
+
+# Create a gtsummary table from model
+table5_train <- tbl_regression(ML_fit4_train, exponentiate = FALSE, intercept = TRUE) %>%
+  bold_p() %>%
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
+table5_train
+
+# Save regression output into table
+saveRDS(table5_train, file = here("results", "tables", "table5_train.rds"))
+```
+Our RMSE is 490 and our r-squared is 0.03, a slightly better performance than the full model.
+
+Now let's run cross-validation.
+
+```{r, message=FALSE,  warning=FALSE}
+# CV model 4
+# Set the seed for reproducibility
+set.seed(rngseed)
+
+# Create 10-fold cross-validation
+folds <- vfold_cv(train_data, v = 10)
+folds
+
+# Define recipe for model
+ML_recipe4_cv <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
+  step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
+  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+
+# Set model specification
+ML_spec4_cv <- linear_reg() %>% set_engine("lm")
+
+# Create and initialize workflow that bundles together model specification and formula.
+ML_wf4_cv <- workflow() %>% 
+  add_model(ML_spec4_cv) %>% 
+  add_recipe(ML_recipe4_cv)
+  fit_resamples(ML_wf4_cv, resamples = folds)
+
+# Resample the model
+ML_wf4_cv_rs <- fit_resamples(ML_wf4_cv, resamples = folds)
+ML_wf4_cv_rs
+
+# Get model metrics
+collect_metrics(ML_wf4_cv_rs)
+
+# Get model metrics
+ML_cv4_metrics <- collect_metrics(ML_wf4_cv_rs)
+
+# Filter the data frame to include only rows with "rmse" and "rsq" in the ".metric" column
+ML_cv4_metrics <- ML_cv4_metrics[ML_cv4_metrics$.metric %in% c("rmse", "rsq"), !(names(ML_cv4_metrics) %in% ".config")]
+ML_cv4_metrics
+
+# Save regression output into table
+saveRDS(ML_cv4_metrics, file = here::here("results", "tables", "ML_cv4_metrics.rds"))
+```
+Here, our RMSE is 458 and our r-squared is 0.03, a better performance than both the full model and training model.
+
+Finally, let's run our model on the test data.
+
+```{r, message=FALSE, warning=FALSE}
+# Model 4 testing
+# Using recipes and workflows
+# Complex linear model
+# Specify recipe
+ML_recipe4_test <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = test_data) %>%
+  step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
+  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+
+# Specify model
+ML_spec4_test <- linear_reg() %>%
+  set_engine("lm")
+
+# Create workflow
+ML_wf4_test <- workflow() %>%
+  add_recipe(ML_recipe4_test) %>%
+  add_model(ML_spec4_test)
+
+# Fit the workflow
+ML_fit4_test <- fit(ML_wf4_test, data = test_data)
+
+# Summarize model fit with tidiers and create a gtsummary table directly from the model
+ML_summary4_test <- summary(ML_fit4_test)
+tidy(ML_fit4_test, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit4_test) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit4_pred_test <- predict(ML_fit4_test, new_data = test_data %>% select(-Weightgrams))
+ML_fit4_pred_test
+
+# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+ML_fit4_pred_test <- bind_cols(ML_fit4_pred_test, test_data %>% select(Weightgrams))
+ML_fit4_pred_test
+
+# Plot observed versus predicted
+plot4_test <- ggplot(ML_fit4_pred_test, aes(x = Weightgrams, y = .pred)) +
+  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
+  geom_point(alpha = 0.5) +
+  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
+  coord_obs_pred()
+plot4_test
+
+# Create metric set including RMSE and R-squared
+metrics_fit4_test <- metric_set(rmse, rsq)
+metrics_fit4_test(ML_fit4_pred_test, truth = Weightgrams, estimate = .pred)
+
+# Convert the output to a data frame and mutate the .estimate column to numeric
+metrics_fit4_test <- as.data.frame(metrics_fit4_test(ML_fit4_pred_test, truth = Weightgrams, estimate = .pred)) %>%
+  mutate(.estimate = as.numeric(.estimate))
+options(scipen = 999) # set opens to avoid scientific notation
+metrics_fit4_test
+
+# Create a gtsummary table from model
+table5_test <- tbl_regression(ML_fit4_test, exponentiate = FALSE, intercept = TRUE) %>%
+  bold_p() %>%
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
+table5_test
+
+# Save regression output into table
+saveRDS(table5_test, file = here("results", "tables", "table5_test.rds"))
+```
+Again, most of our results are not significant except the intercept and nulliparous variable given the small sample size.

### 2024-04-18 code/analysis-code/introanalysis-v1.qmd
Project Part 5
Finished ML analysis and redid tables in final manuscript.

@@
   tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
-  bold_p() %>% 
+  bold_p() %>% # make p-values bold
   add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%

### 2024-04-18 code/eda-code/eda-v1.qmd
Project Part 5
Finished ML analysis and redid tables in final manuscript.

@@
 However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category.
-For the `Status` variable, we will combine "multigesta", "secundigesta", and "trigesta" into "multigravida", since they mean the same thing (two or more pregnancies) and change the language to English (from gesta to gravida).
-We will also combine `secundipara` and `multipara` into one category `multipara`, since this means two or more viable pregnancies.
+
+For the `Status` variable, Table 1 will include all of the gravidity or parity categories from the original data.
+But I will also create a dummy variable for the linear analysis: one category for first pregnancy/birth and one for second or more pregnancies/births.
+With this, I am removing the distinction between parity and gravidity since it is unclear from the original data published in the clinical notes if the physicians were categorizing the women before the delivery or after.
+With this, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births.
+However, because these data are about women entering a hospital to give birth, rather than to receive prenatal care for a pregnancy, and this assumption can provide more statistical power for a linear regression, I assume this risk.
+
+For first pregnancy/birth, I combine all "primi" statuses and "nulli" statuses (gravida or parous) into `nulliparous`.
+For second or more pregnancies/births, I combine all multiple statuses (gravida or parous) into `multiparous`.
+This combines "multigesta", "secundigesta", and "trigesta" (-gesta meaning -gravida in English) into `multiparous`, since they mean the same thing (two or more pregnancies).
+I also combine "secundipara" and "multipara" into the category `multiparous`, since this means two or more births.
 The words in Portuguese and English here are the same.
+
 For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern".
@@
 
-# Or using recode factor from dplyr (easier)
+# Recode factor using dplyr (easier)
 ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
@@
 # Create dummy variables for status and change to English
-ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nullipara", "Nulipara" = "Nullipara")
+ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multiparous", "Trigesta" = "Multiparous", "Multigesta" = "Multiparous", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nulliparous", "Nulipara" = "Nulliparous", "Primipara" = "Nulliparous")
 
@@
     MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
-    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
+    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death"),
+    Status = recode_factor(Status, "Multigesta" = "Multigravida", "Trigesta" = "Trigravida", "Secundigesta" = "Secundigravida", "Secundipara" = "Secundiparous", "Multipara" = "Multiparous", "Primigesta" = "Primigravida", "Nulipara" = "Nulliparous", "Primipara" = "Primiparous")
     )
@@
   tbl_summary(
-    include = c(Color, ModifiedColor, ModifiedStatus, Age, Nationality, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Weightgrams, Lengthcentimeters),
+    include = c(Color, ModifiedColor, Status, ModifiedStatus, Age, Nationality, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Weightgrams, Lengthcentimeters),
     missing = "no",
-    label = list(ModifiedColor ~ "Ancestry", ModifiedStatus ~ "Parity or Gravidity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Weightgrams ~ "Infant Birthweight (grams)", Lengthcentimeters ~ "Infant Birth Length (centimeters)")
+    label = list(ModifiedColor ~ "Ancestry", Status ~ "Parity or Gravidity", ModifiedStatus ~ "Parity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Weightgrams ~ "Infant Birthweight (grms)", Lengthcentimeters ~ "Infant Birth Length (cms)")
   ) %>%
@@
 # Create dummy variables for status and change to English
-ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Nullipara", "Nulipara" = "Nullipara")
+ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multiparous", "Trigesta" = "Multiparous", "Multigesta" = "Multiparous", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nulliparous", "Nulipara" = "Nulliparous", "Primipara" = "Nulliparous")
 

### 2024-04-18 products/manuscript/Manuscript_Cassia.qmd
Project Part 5
Finished ML analysis and redid tables in final manuscript.

@@
 #Loading packages
+#install.packages("flextable")
+#install.packages("gdtools")
 library(tidyverse) #for data processing/cleaning
@@
 library(gtsummary) #for summary tables
+library(flextable) #for nice tables
 ```
@@
 #| label: tbl-table1
-#| tbl-cap: "Table 1"
+#| tbl-cap: "Summary Statistics"
 #| echo: FALSE
 table1 <- readRDS(here::here("results", "tables", "table1_final.rds"))
-table_gt <- gtsummary::as_gt(table1)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table1)
+flex_table
 ```
@@
 
-$$SBR = \frac{SB}{TB} \times 1,000$$
-where the total number of stillbirths (SB) is divided by 1000 total births (TB).
+$$SBR = \frac{SB}{TB} \times 1,000$$ where the total number of stillbirths (SB) is divided by 1000 total births (TB).
 The SBR is 84.8% in this specific clinic.
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$
-whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term. Results are displayed in Table 2.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
+Results are displayed in Table 2.
 
@@
 #| label: tbl-table2
-#| tbl-cap: "Table 2"
+#| tbl-cap: "Model 1 Results"
 #| echo: FALSE
 table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
-table_gt <- gtsummary::as_gt(table2)
-table_gt |> gt::as_raw_html()
+flex_table2 <- as_flex_table(table2)
+flex_table2
 ```
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$
-whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term.
 Results are displayed in Table 3.
@@
 #| label: tbl-table3
-#| tbl-cap: "Table 3"
+#| tbl-cap: "Model 2 Results"
 #| echo: FALSE
 table3 <- readRDS(here::here("results", "tables", "table3_final.rds"))
-table_gt <- gtsummary::as_gt(table3)
-table_gt |> gt::as_raw_html()
+flex_table3 <- as_flex_table(table3)
+flex_table3
 ```
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$
-whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
+Results are displayed in Table 4.
 
@@
 #| label: tbl-table4
-#| tbl-cap: "Table 4"
+#| tbl-cap: "Model 3 Results"
 #| echo: FALSE
 table4 <- readRDS(here::here("results", "tables", "table4_final.rds"))
-table_gt <- gtsummary::as_gt(table4)
-table_gt |> gt::as_raw_html()
+flex_table4 <- as_flex_table(table4)
+flex_table4
 ```
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$
-whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), Nat is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
 Results are displayed in Table 5.
@@
 #| label: tbl-table5
-#| tbl-cap: "Table 5"
+#| tbl-cap: "Model 4 Results"
 #| echo: FALSE
 table5 <- readRDS(here::here("results", "tables", "table5_final.rds"))
-table_gt <- gtsummary::as_gt(table5)
-table_gt |> gt::as_raw_html()
+flex_table5 <- as_flex_table(table5)
+flex_table5
 ```
 
-Given that this study is an inferential, hypothesis-supporting approach that is exploring how specific maternal predictors affect the outcome of interest, birth weight, I am excluding any predictive metrics.
-These models are not examples to be used in other contexts, which suggests I should not apply machine-learning methods such as cross-validation or test/train.
+Given that this study is an inferential, hypothesis-supporting approach that is exploring how specific maternal predictors affect the outcome of interest, birth weight, please see the `fullanalysis-v1.qmd` for all predictive metrics, where test/train and cross-validation were applied to test performance.
+Model performance did not significantly improve with these additional tests.
+When run on testing data, most results were insignificant.
 
@@
 Age and nationality were not significant.
-Women who were nulliparous (never carried a pregnancy to viability) and women who were multigravida (had multiple pregnancies but no viable deliveries) had infants with lower birth weights.
-No significance was found for women who were delivering their first viable child (primipara).
-
-{{< pagebreak >}}
+Women who were nulliparous (delivering their first child) had infants with lower birth weights.
 
@@
 All models based on historical data must contend with limited sample sizes and missing data.
-In this study, a major limitation was the exclusion of gestational age from the original published clinical notes upon which this dataset is based.
+In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
+This means that the birthweight variable is capturing both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation.
+This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017].
+Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights.
+
 A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].

### 2024-04-18 products/manuscript/supplement/Supplemental_Materials.qmd
Project Part 5
Finished ML analysis and redid tables in final manuscript.

@@
 #| label: tbl-table2
-#| tbl-cap: "Supplemental Table 1"
+#| tbl-cap: "Supplemental Table"
 #| echo: FALSE
@@
 #| label: tbl-table_age
-#| tbl-cap: "Supplemental Table 2"
+#| tbl-cap: "Supplemental Table"
 #| echo: FALSE
@@
 #| label: tbl-table4
-#| tbl-cap: "Supplemental Table 3"
+#| tbl-cap: "Supplemental Table"
 #| echo: FALSE
@@
 This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
+
+## Performance Measures
+
+Test/train (75-25 split) and cross-validation were performed on all models.
+Model prediction performance did not improve greatly using these tests.
+Model performance resulted in insignificant results when run on the testing data given the smaller sample size.
+All code for the train/test analysis and cross-validation analysis can be found in `fullanalysis-v1.qmd`.
+
+Here I include the cross-validation results for each model.
+
+```{r}
+#| label: tbl-table5
+#| tbl-cap: "Cross-Validation Model 1"
+#| echo: FALSE
+ML_cv1_metrics <- readRDS(here::here("results", "tables", "ML_cv1_metrics.rds"))
+ML_cv1_metrics
+```
+
+```{r}
+#| label: tbl-table6
+#| tbl-cap: "Cross-Validation Model 2"
+#| echo: FALSE
+ML_cv2_metrics <- readRDS(here::here("results", "tables", "ML_cv2_metrics.rds"))
+ML_cv2_metrics
+```
+
+```{r}
+#| label: tbl-table7
+#| tbl-cap: "Cross-Validation Model 3"
+#| echo: FALSE
+ML_cv3_metrics <- readRDS(here::here("results", "tables", "ML_cv3_metrics.rds"))
+ML_cv3_metrics
+```
+
+```{r}
+#| label: tbl-table8
+#| tbl-cap: "Cross-Validation Model 4"
+#| echo: FALSE
+ML_cv4_metrics <- readRDS(here::here("results", "tables", "ML_cv4_metrics.rds"))
+ML_cv4_metrics
+```
+
+{{< pagebreak >}}
+
+# References

### 2024-04-18 code/analysis-code/fullanalysis.R

@@ -20,7 +20,22 @@ data_location <- here::here("data","processed-data","ML_linear.rds")
 #load data
 ML_linear <- readRDS(data_location)
 
+## ---- traintest --------
+# Fix the random numbers by setting the seed 
+# This enables the analysis to be reproducible when random numbers are used 
+rngseed = 1234
+set.seed(rngseed)
+
+# Create training data
+# Put 3/4 of the data into the training set 
+data_split <- initial_split(ML_linear, prop = 3/4)
+
+# Create data frames for the two sets:
+train_data <- training(data_split)
+test_data  <- testing(data_split)
+
 ## ---- Model1 --------
+# Model 1: Final model for paper using all data
 # Model using recipes and workflows
 # Define recipe for model
 ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
@@ -85,7 +100,171 @@ table2
 # Save regression output into table
 saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
 
+## ---- Model1train --------
+# Training model 1
+# Model using recipes and workflows
+# Define recipe for model
+ML_recipe1_train <- recipe(Weightgrams ~ ModifiedColor, data = train_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
+
+# Specify model
+ML_spec1_train <- linear_reg() %>% 
+  set_engine("lm")
+
+# Create work flow
+ML_wf1_train <- workflow() %>%
+  add_recipe(ML_recipe1_train) %>%
+  add_model(ML_spec1_train)
+
+# Fit the work flow
+ML_fit1_train <- fit(ML_wf1_train, data = train_data)
+
+# Summarize model fit with tidiers
+ML_summary1_train <- summary(ML_fit1_train) # Summary of fit
+tidy(ML_fit1_train, conf.int = TRUE) # Coefficients and confidence intervals
+glance(ML_fit1_train) # Model-level statistics
+
+# Produce predictions for R-squared
+ML_fit1_pred_train <- predict(ML_fit1_train, new_data = train_data %>% select(-Weightgrams))
+ML_fit1_pred_train
+
[…]

### 2024-04-18 code/eda-code/edacode.R

@@ -11,7 +11,7 @@ library(gtsummary) #for summary tables
 library(ggplot2) #for plotting
 library(hrbrthemes) #for nice themes
 hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
-#rm(list=ls()) #clear the environment
+rm(list=ls()) #clear the environment
 
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
@@ -96,11 +96,11 @@ ggsave(filename = figure_file, plot=length_weight_sex, bg="white") #background i
 ## ---- Dummyvariables --------
 # Create dummy variables
 
-# Or using recode factor from dplyr (easier)
+# Recode factor using dplyr (easier)
 ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
 
 # Create dummy variables for status and change to English
-ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")
+ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multiparous", "Trigesta" = "Multiparous", "Multigesta" = "Multiparous", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nulliparous", "Nulipara" = "Nulliparous", "Primipara" = "Nulliparous")
 
 # Create dummy variables for nationality and change to English
 ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
@@ -118,7 +118,8 @@ ML_summary <- ML_summary %>%
[…]

### 2024-04-26 CMiller-project-review-Roth.qmd
Part 5 Review C Miller
Part 5 Review C Miller

---
title: Low Infant Birth Weight in Brazil Project Reveiw 
author: Clarke Miller
date: 23 April 2024
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Low Infant Birth Weight in Brazil, A Historical Data Analysis Approach

Name of project author(s):  Cassia Roth

Name of project reviewer:  Clarke Miller

## Background, Context and Motivation
*
*How well is the context of the project described?*

The overall project is very well described. The public health impact is immediately apparent and the background/data sources/questions and hypothesis are all well explained and explored.

*Is a comprehensive background, including summary of previous/related work given?*

A good background is given. There doesn’t seem to be a comprehensive summary of prior work done, but the appropriate citations for that work are given. My impression is that this data came from some prior work, i.e. it is a historical data set, and that was described in the data description.

*Is the project well placed into the context of existing work (including proper referencing of existing work).*

Yes. This work is well placed in the context of the existing work.

*Is it clear why the project was undertaken and what new information it hopes to provide?*

Yes, it is clear why the project is important. The importance to public health is clear.

### Feedback and Comments

Okay, the first comment that I need to make is that when I try to open the docx file, I get an error from Microsoft word that says:

“Word found unreadable content in Manuscript_Cassia. Do you want to recover the contents of this document? If you trust the source of this document, click yes.”

I get this every time I try to open the document. My Microsoft Word software is able to open the document when I click “Yes” and there is no action from my antivirus software, so I am not sure what the issue here is.  Maybe it's just an issue for me?

**Okay, now the feedback for this specific section...**

The context/motivation/background part looks great.  I wouldn't change anything at this point.  Well done!

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

I chose this one because the section looked great overall.

## Question description

*How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?*

I think that the questions/hypothesis was answered as well as the data allowed. The questions are directly related to the data.

### Feedback and Comments

The question/hypothesis looked great!

### Summary assessment

* question/hypotheses fully clear

It is clear what you want to do and why, your exploration of the data supports that.  That being said, I do have one suggestion.

Table 1 is a bit long.  I would consider a shorter, maybe a single page, summary table of some kind then put this longer table in the supplemental material.  Of course, that is a large number of variables, so it may not be possible to make the table more compact.  (Unless you format it so that each sub-characteristic went horizontal?  Beats me.)  Maybe break this into more than one table?  Make a table that highlights the data that you want to use for your analysis?  Again, I don't know what a good solution would be.  Maybe break each characteristic into a separate table?  For example, take "Parity of Gravidity" and make that one a single table?  ¯\_(ツ)_/¯

Also, why isn't the actual publication or source of the data cited in your work?

## Data description

A good background is given.  There doesn't seem to be a comprehensive summary of prior work done, but the appropriate citations for that work are given.  My impression is that this data came from some prior work, i.e. it is a historical data set, and that was described in the data description.

*Is the project well placed into the context of existing work (including proper referencing of existing work).*

Yes.  This work is well placed in the context of the existing work.

*Is it clear why the project was undertaken and what new information it hopes to provide?*

Yes, it is clear why the project is important.  The importance to public health is clear.  Again, the data is from an older study, the data probably won't provide much new insight past the work-ups that have already been done.

### Feedback and Comments

The data set looks great.  You have narrowed it down to a workable set of variables.

### Summary assessment

* source and overall structure of data well explained

The data is well explained in the data tables and in the supplementary materials.

## Data wrangling and exploratory analysis

*How well is the data cleaned/processed and explored? Are all steps reasonable and well explained?*

Very well.

*Are alternatives discussed and considered?*

Yes.

*Are meaningful exploratory results shown (e.g. in the supplementary materials)?*

Yes.  The exploration results are shown in the paper body.

### Feedback and Comments

No real feedback here.  It looks good.  The data exploration looks like something that I may have done for the same data set.  You made graphs to explore the data, I would go for a visual exploration too.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

It looks good.

## Appropriateness of Analysis

*Were the analysis methods appropriate for the data? Was the analysis done properly?*

Yes.  The analysis was appropriate for the data.  The analysis seems to be done well.

*Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?*

Okay, who am I to judge the best possible way?  That is a judgement call for any individual researcher and strictly between them and reviewer #2.  The data analysis looked appropriate to me.  The data was explored several different ways to look for correlations between material features, with African/European ancestry being an important one, and the results were evaluated.

### Feedback and Comments

I think that this section looks good.

### Summary assessment

* strong and reasonable analysis

## Presentation

*How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?*

The results were presented very well.  The only issue I had with any of the tables or figures was the length of table 1.

**A couple of issues to look at:**

In the word document, table 3 seems to be off the page on the left.  Same thing for Table 5.

The figure captions need to be included in the figure captions.  For example, for figure 3 you have "Scatterplot of birth weight by length" in the code after "fig-cap:", then you add a completely different figure caption: "Figure 3 visualizes the relationship between infant weight and length. Unsurprisingly, birth weight and length are positively correlated."  Combine these.  Also, it is traditional to have the figure description below the figure.  Just a suggestion, but it would make the paper come together a bit better.

I would also take some of the page breaks out.  They are a bit distracting.  For example, the huge gap between the abstract and the introduction leaves quite a bit of scrolling down.  Smoother pagination makes a paper flow easier.

Again, all of this is just a suggestion.  The paper looks pretty good as it is.

### Feedback and Comments

The paper looks pretty good as it is.  A few little details could be polished up a bit.  Really, it's just up to the author's taste and style.

### Summary assessment

* results are very well presented

## Discussion/Conclusions

*Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?*

Yes.  Absolutely.

### Feedback and Comments

Nothing to say, the dicussion and comments look good!

### Summary assessment

* strong, complete and clear discussion

## Further comments

None.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure

*Is the project well structured? Are files in well labeled folders?*

Yes.

*Do files have reasonable names?*

Yes.

*Are all "junk" files not needed for analysis/reproduction removed?*

Yes.

*By just looking at files and folders, can you get an idea of how things fit together?*

Yes.

### Feedback and Comments

No real feedback here.  The project is looking really good.  As a matter of fact, I learned a few things from looking at this project.

### Summary assessment

* well structured

## Documentation

*How well is the project documented?*

Yes.  I like the readme files.  I am going to use them as a model to revise my own readme.

*Are you able to understand each step of the whole analysis, each decision that was made, and each line of code?*

Yes.

*Is enough information provided as comments in code or as part of Rmd files?*

Yes.

### Feedback and Comments

No real feed back here.  Everything looks great.

### Summary assessment

* fully and well documented

## Reproducibility

*Are all results fully reproducible?*

Yes.

*Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention?*

Yes.

*Are you able to re-run the whole analysis without having to do manual interventions/edits?*

Yes.

### Feedback and Comments

No real feed back here.  Everything looks great.

### Summary assessment

* fully reproducible without issues

## Thoroughness

*How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered?*

Yes.

*Were alternatives discussed?*

Yes.

*Were the questions/hypotheses fully and thoroughly addressed?*

Yes.

### Feedback and Comments

No real feed back here.  Everything looks great.

### Summary assessment

* strong level of thorougness

## Further comments

No real feed back here.  Everything looks great.

### 2024-04-30 code/eda-code/eda-v1.qmd
Project Part 6
Updated tables per peer reviews

@@
   geom_vline(xintercept = c(15, 49), linetype = "dashed", color = "red") +  # Add vertical lines that mark reproductive age
-  ggtitle("Maternal Age") +
   theme_ipsum() +
@@
   geom_vline(xintercept = c(1500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
-  ggtitle("Infant birthweight") +
   theme_ipsum() +
@@
   geom_vline(xintercept = c(49), linetype = "dashed", color = "red") +  # Add vertical lines that mark average birth length for both sexes
-  ggtitle("Infant birth length") +
   theme_ipsum() +

### 2024-04-30 products/manuscript/Manuscript_Cassia.qmd
Project Part 6
Updated tables per peer reviews

@@
 
-This cross-sectional study uses linear regression models to understand disparities in infant birth weight in relation to maternal skin color or ancestry, age, gestational status (parity or gravidity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
+This cross-sectional study uses linear regression models to understand disparities in infant birth weight in relation to maternal skin color (ancestry), age, gestational status (parity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
 I hypothesize that infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
@@
 
-{{< pagebreak >}}
-
 ## Data: Description and Sources
@@
 Please see the `Supplemental_Materials` document for more information on the dataset.
+For full citations on the journals from which these data came, see my other publications [@roth2020, @roth2020a].
 
@@
 I employ birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data.
-
-{{< pagebreak >}}
 
@@
 
-{{< pagebreak >}}
-
 ## Data import and cleaning
@@
 The original raw data, the codebook, and the processed data are available in the `data` folder.
-
-{{< pagebreak >}}
 
@@
 
-{{< pagebreak >}}
-
 ## Statistical analysis
@@
 
+{{< pagebreak >}}
+
 ```{r}
@@
 flex_table2 <- as_flex_table(table2)
+flex_table2 <- flex_table2 %>%
+  flextable::set_table_properties(width = .5, align = "center")
 flex_table2
@@
 #| echo: FALSE
+i = 16.5
+w = i*0.3937
+font_size <- 10
 table3 <- readRDS(here::here("results", "tables", "table3_final.rds"))
 flex_table3 <- as_flex_table(table3)
+flex_table3 <- autofit(flex_table3)
+auto_widths <- dim(flex_table3)$widths/sum(dim(flex_table3)$widths) # calculate column widths
+flex_table3 <- width(flex_table3, width=w*auto_widths) # adjust column widths
+flex_table3 <- fontsize(flex_table3, size=font_size) # set font size
 flex_table3
@@
 flex_table4 <- as_flex_table(table4)
+flex_table4 <- flex_table4 %>%
+  flextable::set_table_properties(width = .5, align = "center")
 flex_table4
@@
 #| echo: FALSE
+i = 16.5
+w = i*0.3937
+font_size <- 10
 table5 <- readRDS(here::here("results", "tables", "table5_final.rds"))
 flex_table5 <- as_flex_table(table5)
+flex_table5 <- autofit(flex_table5)
+auto_widths <- dim(flex_table5)$widths/sum(dim(flex_table5)$widths) # calculate column widths
+flex_table5 <- width(flex_table5, width=w*auto_widths) # adjust column widths
+flex_table5 <- fontsize(flex_table5, size=font_size) # set font size
 flex_table5

### 2024-04-30 code/eda-code/edacode.R

@@ -35,7 +35,6 @@ age_plot <- ML_summary %>% filter(!is.na(Age)) %>% # Filter out NAs
   ggplot(aes(x = Age)) +
   geom_histogram(binwidth = 0.5, width= 0.8, fill="#69b3a2") +
   geom_vline(xintercept = c(15, 49), linetype = "dashed", color = "red") +  # Add vertical lines that mark reproductive age
-  ggtitle("Maternal Age") +
   theme_ipsum() +
   theme(plot.title = element_text(size=15))
 plot(age_plot)
@@ -49,7 +48,6 @@ weight_plot <- ML_summary %>% filter(!is.na(Weightgrams)) %>% #filter out NAs
   ggplot(aes(x = Weightgrams)) +
   geom_histogram(binwidth = 25, fill="#69b3a2") +
   geom_vline(xintercept = c(1500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
-  ggtitle("Infant birthweight") +
   theme_ipsum() +
   theme(plot.title = element_text(size=15))
 plot(weight_plot)
@@ -63,7 +61,6 @@ length_plot <- ML_summary %>% filter(!is.na(Lengthcentimeters)) %>% #filter out 
   ggplot(aes(x = Lengthcentimeters)) +
   geom_histogram(binwidth = 1, fill="#69b3a2") +
   geom_vline(xintercept = c(49), linetype = "dashed", color = "red") +  # Add vertical lines that mark average birth length for both sexes
-  ggtitle("Infant birth length") +
   theme_ipsum() +
   theme(plot.title = element_text(size=15))
 plot(length_plot)

### 2024-05-01 project-review-by CLi.qmd

---
title: Project Part 5 Review  
author: Chaohua Li
date:  04-25-2024
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project:Low Infant Birth Weight in Brazil: A Historical Data Analysis Approach

Name of project author(s):  Cassia Roth

Name of project reviewer:   Chaohua Li

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The introduction is well-structured with citations appropriately added. The selection of the research question is justified.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The question was clearly stated.

### Summary assessment

* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

Data source was described in detail. A codebook was also provided in the 'rawdata' folder.

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Data was properly cleaned and processed. Comments/Notes were provided for each procedure.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The analysis methods were appropriate given the inferential purpose of this study. Model performances were compared using cross-validation. Final model was evaluated using the test data.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The results were presented in a well-structured manner. It was easy to follow the steps without getting lost. The format for Tables 2&3&5 may need a little tweak. The first column and the last column of the tables appear to be truncated.

### Summary assessment

* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

Findings were properly discussed. Strengths and limitations were acknowledged.

### Summary assessment

* strong, complete and clear discussion

## Further comments

The content of the paper was super clear and easy to follow. The only thing that needs improvement is the format of the tables.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

All files are well label. All residual files from the template have been removed.

### Summary assessment

* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

There are Rmd files in each folder, clearly indicating the content and purpose for the files in it.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

I was able reproduce the results. Instructions were clear.

### Summary assessment

* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

The research questions have been fully addressed. This study is relatively complete and thorough.

### Summary assessment

* strong level of thorougness

## Further comments

Congratulations on completing the project so well! Enjoy your summertime.

### 2024-06-24 code/analysis-code/fullanalysis-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

@@
 
-#rm(list=ls()) #clear the environment
+rm(list=ls()) #clear the environment
 ```
@@
 Our results here are not that different from the model run on all the data.
-Our adjusted R-squared is 0.007, the same as the full model.
-The RMSE is 495, which is slightly lower than the full model (502), which shows better performance.
+Our adjusted R-squared is 0.01, the same as the full model.
+The RMSE is 488, which is slightly lower than the full model (494), which shows better performance.
 
@@
 ```
-Performing cross-validation results in an RMSE that is the same as our training model (495), lower than our full model.
-The r-squared (0.02) is also higher than our training model (0.008) and our full model (0.007), demonstrating better model performance.
+Performing cross-validation results in an RMSE that is the same as our training model (488), lower than our full model.
+The r-squared (0.015) is also higher than our training and full models (0.01), demonstrating better model performance.
 
@@
 
-When running the model on our testing dataset, our results are not significant, given the small sample size.
+When running the model on our testing dataset, our results remain significant at the 0.05 level. The RMSE is 508 and the r-squared is 0.008. This shows a slightly worse fit than our training results.
 
@@
 ```
-Our RMSE is 498 and our r-squared is 0.02.
+Our RMSE is 490 and our r-squared is 0.026.
 
@@
 ```
-Our RMSE is 490 and our r-squared is 0.03, a slightly better performance than the full model.
+Our RMSE is 483 and our r-squared is 0.03, a slightly better performance than the full model.
 
@@
 ```
-Here, our RMSE is 458, an improvement from both the full model and the training model, and our r-squared is 0.03, the same as our training model.
+Here, our RMSE is 483 and our r-squared is 0.026, the same as the training model.
 
@@
 ```
-Running the model on our test data, we see that the only significant result is for the nulliparous status. All other results are insignificant.
+Running the model on our test data, we see no significant results at any alpha level.
 
@@
 
-Our RMSE is 502 and our r-squared is 0.007.
+Our RMSE is 494 and our r-squared is 0.01.
 
@@
 
-Our RMSE is 495 and our r-squared is 0.009, slightly better performance than the full model. Our results are not significantly different from the full model.
+Our RMSE is 489 and our r-squared is 0.01, slightly better performance than the full model. Our results are not significantly different from the full model.
 
@@
 ```
-Using cross-validation, our RMSE is 495 and our r-squared is 0.02, a similar performance to our training model.
+Using cross-validation, our RMSE is 487 and our r-squared is 0.01, a similar performance to our training model.
 
@@
 
-Much like in the other testing models, our results are not significant (except the intercept). Our performance is also worse (RMSE of 518 and r-squared of 0.005).
+Much like in the other testing models, our results are not significant (except the intercept). Our performance is also worse (RMSE of 507 and r-squared of 0.008).
 
@@
 ```
-Our RMSE is 490 and our r-squared is 0.03, a slightly better performance than the full model.
+Our RMSE is 483 and our r-squared is 0.033, a slightly better performance than the full model.
 
@@
 ```
-Here, our RMSE is 458 and our r-squared is 0.03, a better performance than both the full model and training model.
+Here, our RMSE is 484 and our r-squared is 0.025, a better performance than both the full model and training model.
 
@@
 ```
-Again, most of our results are not significant except the intercept and nulliparous variable given the small sample size.
+Again, most of our results are not significant except the intercept given the small sample size.

### 2024-06-24 code/analysis-code/introanalysis-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

@@
 library(tidymodels) #for modeling
-
-#rm(list=ls()) #clear the environment
+rm(list=ls()) #clear the environment
 ```
@@
 ```
+
+## Linear Models
 
@@
 
-For all three models, the r-squared is very small (0.0073, 0.0061, and 0.027), indicating that the model does not explain much of the variance in birth weight.
+For all three models, the r-squared is very small (0.009, 0.008, and 0.026), indicating that the model does not explain much of the variance in birth weight.
 This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
@@
 ```
+
+## Logistic models
+
+Now we will still consider `Birthweight` as the outcome of interest, but we will transform it into a categorical outcome of <=2500 (low birthweight, LBW) and >2500 (normal birthweight, NBW) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
+
+First, we will split data into training and testing sets to later calculate ROC_AUC.
+
+```{r}
+set.seed(123)
+ML_log_split <- initial_split(ML_linear, prop = 0.75, strata = BirthweightCategory)
+
+# Create training and test datasets
+ML_log_train <- training(ML_log_split)
+ML_log_test <- testing(ML_log_split)
+```
+
+Now we will run the first logistic model with `ModifiedColor` as the predictor.
+
+```{r}
+# Simple logistic regression model with ModifiedColor as the predictor
+# Check factor levels
+levels(ML_linear$BirthweightCategory) # Reference is LBW
+
+# Set the reference level explicitly to NBW
+ML_linear <- ML_linear %>%
+  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))
+
+# Create recipe
+# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
+log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) %>%
+  step_dummy(all_nominal(), -all_outcomes()) %>%
+  step_normalize(all_predictors())
+
+# Define model specification
+log1_spec <- logistic_reg() %>%
+  set_engine("glm") %>%
+  set_mode("classification")
+
+# Create workflow
+log1_wf <- workflow() %>%
+  add_recipe(log1_recipe) %>%
+  add_model(log1_spec)
+
+# Fit model
+log1_fit <- log1_wf %>%
+fit(data = ML_linear)
+
+# Extract fitted model from workflow
+log1_fitted <- extract_fit_parsnip(log1_fit)$fit
+
+# Create a gtsummary table directly from the model
+table6 <- tbl_regression(log1_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table6
+
+# Save regression output into table
+saveRDS(table6, file = here::here("results", "tables", "table6.rds"))
+
+# Calculate performance metrics using pseudo-R2
+# Extract model metrics using broom
+model_metrics1 <- glance(log1_fitted)
+
+# Calculate McFadden's R-squared
+# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
+mcfadden_r2_1 <- 1 - (model_metrics1$deviance / model_metrics1$null.deviance)
+
+# Print McFadden's R-squared
+print(mcfadden_r2_1)
+
+# Calculate performance metrics using ROC_AUC
+# Fit the model on the training data
+log1_fit <- log1_wf %>% fit(data = ML_log_train)
+
+# Predict probabilities on the test set
+prob_preds1 <- predict(log1_fit, new_data = ML_log_test, type = "prob")
+
+# Combine predictions with true values
+results1 <- bind_cols(ML_log_test, prob_preds1)
+
+# Calculate ROC-AUC
+roc_auc_result1 <- roc_auc(results1, truth = BirthweightCategory, .pred_LBW)
+
+# Print ROC-AUC
+print(roc_auc_result1)
+
+# Create and plot the ROC curve
+roc_curve_data1 <- roc_curve(results1, truth = BirthweightCategory, .pred_LBW)
+
+# Plotting the ROC curve
+roc_plot1 <- ggplot(roc_curve_data1, aes(x = 1 - specificity, y = sensitivity)) +
+  geom_path() +
+  geom_abline(lty = 2) +
+  labs(title = "ROC Curve",
+       x = "1 - Specificity",
+       y = "Sensitivity")
+
+# Print the plot
+print(roc_plot1)
+```
+With an ROC_AUC of 0.463, our model is performing worse than random guessing. Our pseudo R-squared is 0.004.
+
+Now let's run another logistic model with `Color` as the predictor.
+
+```{r}
+# Simple logistic regression model with ModifiedColor as the predictor
+# Check factor levels
+levels(ML_linear$BirthweightCategory) # Reference is LBW
+
+# Set the reference level explicitly to NBW
+ML_linear <- ML_linear %>%
+  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))
+
+# Create recipe
+# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
+log2_recipe <- recipe(BirthweightCategory ~ Color, data = ML_linear) %>%
+  step_dummy(all_nominal(), -all_outcomes()) %>%
+  step_normalize(all_predictors())
+
+# Define model specification
+log2_spec <- logistic_reg() %>%
+  set_engine("glm") %>%
+  set_mode("classification")
+
+# Create workflow
+log2_wf <- workflow() %>%
+  add_recipe(log2_recipe) %>%
+  add_model(log2_spec)
+
+# Fit model
+log2_fit <- log2_wf %>%
+fit(data = ML_linear)
+
+# Extract fitted model from workflow
+log2_fitted <- extract_fit_parsnip(log2_fit)$fit
+
+# Create a gtsummary table directly from the model
+table7 <- tbl_regression(log2_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table7
+
+# Save regression output into table
+saveRDS(table7, file = here::here("results", "tables", "table7.rds"))
+
+# Calculate metrics using pseudo-R2
+# Extract model metrics using broom
+model_metrics2 <- glance(log2_fitted)
+
+# Calculate McFadden's R-squared
+# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
+mcfadden_r2_2 <- 1 - (model_metrics2$deviance / model_metrics2$null.deviance)
+
+# Print McFadden's R-squared
+print(mcfadden_r2_2)
+
+# Calculate performance metrics using ROC_AUC
+# Fit the model on the training data
+log2_fit <- log2_wf %>% fit(data = ML_log_train)
+
+# Predict probabilities on the test set
+prob_preds2 <- predict(log2_fit, new_data = ML_log_test, type = "prob")
+
+# Combine predictions with true values
+results2 <- bind_cols(ML_log_test, prob_preds2)
+
+# Calculate ROC-AUC
+roc_auc_result2 <- roc_auc(results2, truth = BirthweightCategory, .pred_LBW)
+
+# Print ROC-AUC
+print(roc_auc_result2)
+
+# Create and plot the ROC curve
+roc_curve_data2 <- roc_curve(results2, truth = BirthweightCategory, .pred_LBW)
+
+# Plotting the ROC curve
+roc_plot2 <- ggplot(roc_curve_data2, aes(x = 1 - specificity, y = sensitivity)) +
+  geom_path() +
+  geom_abline(lty = 2) +
+  labs(title = "ROC Curve",
+       x = "1 - Specificity",
+       y = "Sensitivity")
+
+# Print the plot
+print(roc_plot2)
+```
+The ROC_AUC for our second simple logistic regression model is 0.47, or still worse than random guessing. Our pseudo R-squared is 0.006.
+
+Now, we will run two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.
+
+```{r}
+# Multiple logistic regression model with ModifiedColor as the predictor
+# Check factor levels
+levels(ML_linear$BirthweightCategory) # Reference is LBW
+levels(ML_linear$ModifiedColor) # Reference is Euro-descent
+levels(ML_linear$ModifiedStatus) # Reference is multiparous
+levels(ML_linear$ModifiedNationality) # Reference is European
+
+# Set the reference level explicitly to NBW
+ML_linear <- ML_linear %>%
+  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"),
+         ModifiedNationality = relevel(ModifiedNationality, ref = "Brazilian"))
+
+# Create recipe
+# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
+log3_recipe <- recipe(BirthweightCategory ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
+  step_dummy(all_nominal(), -all_outcomes()) %>%
+  step_normalize(all_predictors())
+
+# Define model specification
+log3_spec <- logistic_reg() %>%
+  set_engine("glm") %>%
+  set_mode("classification")
+
+# Create workflow
+log3_wf <- workflow() %>%
+  add_recipe(log3_recipe) %>%
+  add_model(log3_spec)
+
+# Fit model
+log3_fit <- log3_wf %>%
+fit(data = ML_linear)
+
+# Extract fitted model from workflow
+log3_fitted <- extract_fit_parsnip(log3_fit)$fit
+
+# Create a gtsummary table directly from the model
+table8 <- tbl_regression(log3_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table8 <- bold_p(table8, t = 0.05)
+table8
+
+# Save regression output into table
+saveRDS(table8, file = here::here("results", "tables", "table8.rds"))
+
+# Calculate metrics using pseudo-R2
+# Extract model metrics using broom
+model_metrics3 <- glance(log3_fitted)
+
+# Calculate McFadden's R-squared
+# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
+mcfadden_r2_3 <- 1 - (model_metrics3$deviance / model_metrics3$null.deviance)
+
+# Print McFadden's R-squared
+print(mcfadden_r2_3)
+
+# Calculate performance metrics using ROC_AUC
+# Fit the model on the training data
+log3_fit <- log3_wf %>% fit(data = ML_log_train)
+
+# Predict probabilities on the test set
+prob_preds3 <- predict(log3_fit, new_data = ML_log_test, type = "prob")
+
+# Combine predictions with true values
+results3 <- bind_cols(ML_log_test, prob_preds3)
+
+# Calculate ROC-AUC
+roc_auc_result3 <- roc_auc(results3, truth = BirthweightCategory, .pred_LBW)
+
+# Print ROC-AUC
+print(roc_auc_result3)
+
+# Create and plot the ROC curve
+roc_curve_data3 <- roc_curve(results3, truth = BirthweightCategory, .pred_LBW)
+
+# Plotting the ROC curve
+roc_plot3 <- ggplot(roc_curve_data3, aes(x = 1 - specificity, y = sensitivity)) +
+  geom_path() +
+  geom_abline(lty = 2) +
+  labs(title = "ROC Curve",
+       x = "1 - Specificity",
+       y = "Sensitivity")
+
+# Print the plot
+print(roc_plot3)
+```
+Our ROC_AUC here is worse, at 0.46, or worse than random guessing. Our pseudo-R2 is 0.0055.
+
+```{r}
+# Multiple logistic regression model with Color as the predictor
+# Check factor levels
+levels(ML_linear$BirthweightCategory) # Reference is LBW
+levels(ML_linear$Color) # Reference is White
+levels(ML_linear$ModifiedStatus) # Reference is multiparous
+levels(ML_linear$ModifiedNationality) # Reference is European
+
+# Set the reference level explicitly to NBW
+ML_linear <- ML_linear %>%
+  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"),
+         ModifiedNationality = relevel(ModifiedNationality, ref = "Brazilian"))
+
+# Create recipe
+# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
+log4_recipe <- recipe(BirthweightCategory ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
+  step_dummy(all_nominal(), -all_outcomes()) %>%
+  step_normalize(all_predictors())
+
+# Define model specification
+log4_spec <- logistic_reg() %>%
+  set_engine("glm") %>%
+  set_mode("classification")
+
+# Create workflow
+log4_wf <- workflow() %>%
+  add_recipe(log4_recipe) %>%
+  add_model(log4_spec)
+
+# Fit model
+log4_fit <- log4_wf %>%
+fit(data = ML_linear)
+
+# Extract fitted model from workflow
+log4_fitted <- extract_fit_parsnip(log4_fit)$fit
+
+# Create a gtsummary table directly from the model
+table9 <- tbl_regression(log4_fitted, exponentiate = TRUE) # exponentiate true for ORs
+
+#Bold significant p-values
+table9 <- bold_p(table9, t = 0.05)
+table9
+
+# Save regression output into table
+saveRDS(table9, file = here::here("results", "tables", "table9.rds"))
+
+# Calculate metrics using pseudo-R2
+# Extract model metrics using broom
+model_metrics4 <- glance(log4_fitted)
+
+# Calculate McFadden's R-squared
+# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
+mcfadden_r2_4 <- 1 - (model_metrics4$deviance / model_metrics4$null.deviance)
+
+# Print McFadden's R-squared
+print(mcfadden_r2_4)
+
+# Calculate performance metrics using ROC_AUC
+# Fit the model on the training data
+log4_fit <- log4_wf %>% fit(data = ML_log_train)
+
+# Predict probabilities on the test set
+prob_preds4 <- predict(log4_fit, new_data = ML_log_test, type = "prob")
+
+# Combine predictions with true values
+results4 <- bind_cols(ML_log_test, prob_preds4)
+
+# Calculate ROC-AUC
+roc_auc_result4 <- roc_auc(results4, truth = BirthweightCategory, .pred_LBW)
+
+# Print ROC-AUC
+print(roc_auc_result4)
+
+# Create and plot the ROC curve
+roc_curve_data4 <- roc_curve(results4, truth = BirthweightCategory, .pred_LBW)
+
+# Plotting the ROC curve
+roc_plot4 <- ggplot(roc_curve_data4, aes(x = 1 - specificity, y = sensitivity)) +
+  geom_path() +
+  geom_abline(lty = 2) +
+  labs(title = "ROC Curve",
+       x = "1 - Specificity",
+       y = "Sensitivity")
+
+# Print the plot
+print(roc_plot4)
+```
+In our final model, our ROC_AUC is 0.477, still worse than random guessing. Our pseudo-R2 is 0.007.

### 2024-06-24 code/eda-code/eda-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

@@
 
+In the second part of this code chunk, I also break down infant birth weight by sex and by race. I will need to come back and rename these tables.
+
 ```{r, output="html"}
@@
 saveRDS(table1, file = here::here("results", "tables", "table1_final.rds"))
+
+#Create gtsummary table on infant birth weight stratified by sex and maternal skin color
+stratified_color <- 
+  ML_summary %>%
+  filter(!is.na(Weightgrams) & !is.na(ModifiedColor)) %>% # Exclude rows with NA in weight & MC variable
+  rename("Infant Weight (grms)" = Weightgrams) %>% # Rename Weightgrams to Infant Weight (grms)
+  tbl_summary(
+    by = ModifiedColor,  # Grouping by ModifiedColor
+    include = "Infant Weight (grms)",
+    statistic = list("Infant Weight (grms)" ~ "{mean} ({sd})")
+  )
+stratified_color
+
+stratified_sex <- 
+  ML_summary %>%
+  filter(!is.na(Weightgrams) & !is.na(Sex)) %>%
+  rename("Infant Weight (grms)" = Weightgrams) %>% # Rename Weightgrams to Infant Weight (grms)
+  tbl_summary(
+    by = Sex,  # Grouping by Sex
+    include = "Infant Weight (grms)",
+    statistic = list("Infant Weight (grms)" ~ "{mean} ({sd})")
+  )
+stratified_sex
+
+# Combine the two tables
+table1.1 <- tbl_merge(
+  tbls = list(stratified_color, stratified_sex),
+  tab_spanner = c("**Weight by Modified Color**", "**Weight by Sex**"))
+
+# Display the modified table
+table1.1
+
+#Save table
+saveRDS(table1.1, file = here::here("results", "tables", "table1.1_final.rds"))
 ```

### 2024-06-24 code/processing-code/processingfile-v1.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

@@
 library(kableExtra) #for nice tables
+rm(list=ls()) #clear the environment
 ```
@@
 We also don't need the citation information to run regressions.
-So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Birth_Notes`, `Maternal_Notes`, `Fetal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.
+So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Maternal_Notes`, `Fetal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.
+Further, we will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).
 
@@
 # Remove unnecessary variables
-ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Birth_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
+ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
 ```
@@
 #Summarize missing data using dplyr
-missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
+missing_data_summary <- ML_1 %>% dplyr::summarise_all(list(~sum(is.na(.))))
+missing_data_summary
 ```
@@
 Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.
+`Birth_Notes` also has 2330 missing values, but first I need to extract information on gestation (twin versus singleton birth) from this variable before I exclude it.
 
 ```{r}
+# Extract information on gestation from Birth_Notes and create new variable gestation with singleton and twin births
+ML_2 <- ML_1 %>%
+  mutate(Birth_Notes = str_trim(Birth_Notes)) %>% # removes any leading or trailing whitespace from Birth_Notes column
+  mutate(Gestation = if_else(is.na(Birth_Notes) | Birth_Notes == "", "singleton", # checks if Birth_Notes is NA or empty strings and assigns singleton to Gestation column
+                             if_else(str_detect(tolower(Birth_Notes), "gemelar"), "twin", "singleton"))) # checks for gemelar in case-insensitive manner and assigns twin accordingly or singleton otherwise
+ML_2
+
 # Remove variables with missing data not relevant to hypothesis
-ML_2 <- ML_1 %>% dplyr::select(-CivilStatus, -GestationalAge_Months)
+ML_3 <- ML_2 %>% dplyr::select(-CivilStatus, -GestationalAge_Months, -Birth_Notes)
 ```
@@
 #Summarize missing data using dplyr
-missing_data_summary <- ML_2 %>% dplyr::summarise_all(funs(sum(is.na(.))))
-View(missing_data_summary)
+missing_data_summary <- ML_3 %>% dplyr::summarise_all(list(~sum(is.na(.))))
+missing_data_summary
+```
+Now, I will create a new categorical variable of Normal Birthweight (NBW) for all birth weights greater than 2500 grams and Low Birthweight (LBW) for all birth weights equal to or less than 2500 grams for a future logistic regression.
+This will come from the `Weightgrams` variable.
+
+```{r}
+# Create new variable of NBW and LBW for future logistic regression
+ML_4 <- ML_3 %>%
+  mutate(BirthweightCategory = if_else(Weightgrams <= 2500, "LBW", "NBW"))
+ML_4
 ```
@@
 # Transform character variables into factors
-ML_3 <- ML_2 %>%
-  mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex), as.factor))
+ML_5 <- ML_4 %>%
+  mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex, Gestation, BirthweightCategory), as.factor))
 
 # Check variables
-glimpse(ML_3)
+glimpse(ML_5)
 
 # Check levels of factor variables
-map(ML_3[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')], levels)
+map(ML_5[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex', 'Gestation', 'BirthweightCategory')], levels)
 
 # Define factor variables to check frequencies
-factor_variables <- c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex')
+factor_variables <- c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex', 'Gestation', 'BirthweightCategory')
 
 # Check frequencies of factor variables
-frequency_table <- ML_3 %>%
+frequency_table <- ML_5 %>%
   count(across(all_of(factor_variables)))
@@
 I will also exclude all stillbirths or neonatal deaths as the analysis is interested in studying live births.
+Finally, for the linear dataset I will exclude all twin births, since most current studies restrict their regression analyses to singleton births.
 
@@
 # Create dataset for summary statistics (includes missing values)
-ML_summary <- ML_3
+ML_summary <- ML_5
 
 # Creating dataset for linear regression
-ML_linear <- ML_3 %>% tidyr::drop_na() %>%
-  filter(Weightgrams >= 500, FetalOutcome != "morto") # Drop observations of < 500 grams and all non-live births
+ML_linear <- ML_5 %>% tidyr::drop_na() %>%
+  filter(Weightgrams >= 500, FetalOutcome != "morto", Gestation != "twin") # Drop observations of < 500 grams, all non-live births, and all twin births for regression dataset
 
 # Check for no missing data in linear model
-ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
-View(ML_linear_summary)
+ML_linear_summary <- ML_linear %>% dplyr::summarise_all(list(~sum(is.na(.))))
+ML_linear_summary
 ```
@@
 Now, I can see that my `ML_linear` dataset has no missing data.
-It has 1944 observations and 11 variables.
+It has 1919 observations and 12 variables.
 This might change as I begin analysis.
@@
 
-Finally, I will save the clean data both the `ML_Linear` dataset and the `ML_summary` dataset as RDS files
-.
+Finally, I will save the clean data both the `ML_Linear` dataset and the `ML_summary` dataset as RDS files.
+
 ```{r}

### 2024-06-24 products/manuscript/Manuscript_Draft.qmd
Post-Class Changes
Log plus new tables and new manuscript draft

--- 2024-04-30 products/manuscript/Manuscript_Cassia.qmd
@@
 The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants.
+
+This study builds on an important body of work in economic history that uses hospital records to explore trends in birth weights in the past [@costa1998, @costa2004, @goldin1989, @steckel1986, @steckel1998].
 
@@
 This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017].
-Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights.
+Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights, although other historical studies have datasets that include gestational age [@costa1998, @costa2004, @goldin1989].
 
@@
 For all models in this study, the $\(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight.
-This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors [@leimert2020]
+This is likely due to the fact that birth weight is a complex trait influenced by genetic, environmental, and social factors [@leimert2020].
+Yet, as Claudia Goldin and Robert A.
+Margo have shown for nineteenth-century Philadelphia, the inclusion of gestational age greatly improved the regression fit in their studies, demonstrating that gestational age could explain much of the variance of birth weight in this study [@goldin1989].
 

### 2024-06-24 products/manuscript/Manuscript_Final.qmd
Revisions
Updated manuscript and Assets Readme and Master Readme

--- 2024-06-24 products/manuscript/Manuscript_Draft.qmd
@@
     bibliography: ../../assets/Mada_Bib.bib
-    csl: ../../assets/vancouver.csl
+    csl: ../../assets/chicago-fullnote-bibliography.csl
 ---
@@
 
-# Summary/Abstract
+# Abstract
 
 This cross-sectional study uses linear regression models to understand disparities in infant birth weight in relation to maternal skin color (ancestry), age, gestational status (parity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
-I hypothesize that infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
+Infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant.
+I hypothesize these differences were related to the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
 
@@
 # Introduction
+
+At 4 in the morning on November 17, 1922, Virginia Maria da Conceição, a twenty-six-year-old Black ("*preta*") woman gave birth to a stillborn daughter in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now called Maternidade Escola).
+This was Conceição's fifth birth; prior, she had delivered three full-term infants vaginally (described as "naturally" in the clinical notes) and one premature infant, born at seven months gestational age.
+According to the notes, Conceição had had a "good" pregnancy.
+After arriving at the hospital, she remained in the first stage of labor ("dilatação") for five hours.
+The second stage of labor ("expulsão") lasted only thirty minutes.
+The placenta was delivered ten minutes later.
+Despite the apparently easy pregancy and delivery, Conceição would not take home a new baby; rather, "hereditary syphilis," as the fetal autopsy noted, was the cause of the stillbirth.
+The unnamed infant weighed 3,180 grams at birth, a normal weight for a full-term infant.
+[PAGE 542-4, Dezembro 1922, 16 no.
+12]
+
+Virginia Maria da Conceição's experience with stillbirth is one of the 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926 that form the source base for this article.
+It is also the only one to include the patient's name.
+For historians of medicine and public health, this breach of privacy is a boon to social historical methods, which analyze these histories from the "patient's point of view" CITE.
+For contemporary medical practitioners, the inclusion of Conceição's name may come as a shock, as patient privacy is a cornerstone of medical ethics.
+I use her story here as a jumping off point to explore the relationship between social determinants of health and infant birthweight in a former slave society.
 

### 2024-06-27 products/manuscript/Manuscript_Final.qmd
Update to manuscript
Changed repository name

@@
     csl: ../../assets/chicago-fullnote-bibliography.csl
+    suppress-bibliography: true
 ---
@@
 Infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant.
-I hypothesize these differences were related to the historical legacies of slavery, only abolished in 1888, on population health in Brazil.
+I hypothesize these differences were related to the historical legacies of slavery, only abolished in 1888, including intergenerational BLANK caused by racism, on population health in Brazil.
 
@@
 The placenta was delivered ten minutes later.
-Despite the apparently easy pregancy and delivery, Conceição would not take home a new baby; rather, "hereditary syphilis," as the fetal autopsy noted, was the cause of the stillbirth.
-The unnamed infant weighed 3,180 grams at birth, a normal weight for a full-term infant.
-[PAGE 542-4, Dezembro 1922, 16 no.
-12]
+The unnamed stillborn girl weighed 3,180 grams at birth, a normal weight for a full-term infant.
+Despite the apparently easy pregnancy and delivery, Conceição would not take home a healthy baby; rather, "hereditary syphilis," as the fetal autopsy noted, was the cause of the stillbirth [@rolindo1923, pp.
+542-44].
 
-Virginia Maria da Conceição's experience with stillbirth is one of the 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926 that form the source base for this article.
-It is also the only one to include the patient's name.
-For historians of medicine and public health, this breach of privacy is a boon to social historical methods, which analyze these histories from the "patient's point of view" CITE.
-For contemporary medical practitioners, the inclusion of Conceição's name may come as a shock, as patient privacy is a cornerstone of medical ethics.
-I use her story here as a jumping off point to explore the relationship between social determinants of health and infant birthweight in a former slave society.
+Virginia Maria da Conceição's experience with stillbirth demonstrates the precarious state of maternal-infant health in early twentieth-century Brazil.
+I use Conceição's tragic but all too common story of poor perinatal and neonatal outcomes as a jumping off point to explore the historical relationship between social determinants of health and infant birth weight in a former slave society.
 
 Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality [@travassos2004].
-Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts [@nyarko2013, @wehby2015, @vilanova2019].
-Mothers with lower education levels and lower socioeconomic status also give birth to infants at lower birth weights [@wehby2015, @vilanova2019].
-Understanding these patterns is important as low birth weight is associated with higher rates of infant mortality in the first year of life [@cardoso2005, @vilanova2019].
+Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts [@nyarko2013; @wehby2015; @vilanova2019].
+Mothers with lower education levels and lower socioeconomic status also give birth to infants at lower birth weights [@wehby2015; @vilanova2019].
+Understanding these patterns is important as low birth weight is associated with higher rates of infant mortality in the first year of life [@cardoso2005; @vilanova2019].
 
-These trends hold true in the United States, as well, where Black newborns have lower birth weights than White newborns, and these trends have worsened over time [@mangold1991, @pollock2021].
-Racism is a major factor in these outcomes [@ifatunji2022], as studies demonstrate that U.S.-born Black women have infants with lower birth weights and higher rates of pre-term birth than Black immigrant women [@braveman2021, @clay2023, @mckenzie-sampson2021].
+These trends hold true in the United States, as well, where Black newborns have lower birth weights than White newborns, and these trends have worsened over time [@mangold1991; @pollock2021].
+Racism is a major factor in these outcomes [@ifatunji2022], as studies demonstrate that U.S.-born Black women have infants with lower birth weights and higher rates of pre-term birth than Black immigrant women [@braveman2021; @clay2023; @mckenzie-sampson2021].
 Other racial groups, whether U.S.
@@
 
-Scholars across disciplines have argued that health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@adamo1983, @otovo2016, @roth2020, @telles2022].
-In Brazil, most historical studies remain descriptive in nature and have not tested any quantitative associations between the legacies of slavery, including racism, and maternal-infant health outcomes [@otovo2016, @roth2018a, @roth2018, @roth2020, @telles2022].
-This study aims to fill that gap.
+Scholars across disciplines have argued that health disparities in countries with histories of race-based chattel slavery, from the United States to Brazil, are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @steckel1986].
+In Brazil, most historical studies remain descriptive in nature and have not tested any quantitative associations between the legacies of slavery, including race-based inequality and its consequences and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2018a; @roth2018; @roth2020; @telles2022].
 
-It looks at infant birth weight in relation to maternal skin color or ancestry, age, gestational status (parity or gravidity), and nationality in Brazil’s first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola), which provided free gynecological and obstetric care to the city’s poorest women.
-It relies on a unique sample of infant birth weight data from the country's first public maternity hospital, opened in 1904 in the then-capital city of Rio de Janeiro.
+Here, I am to fill this gap by examining infant birth weight in relation to maternal skin color or ancestry, age, gestational status (parity), and nationality in Rio de Janeiro, Brazil's public maternity hospital, the Maternidade Laranjeiras, which opened in 1904 and provided free gynecological and obstetric care to the then-capital city’s poor and working-class women.
+It relies on a unique sample of infant birth weight data (n=2845).
 
@@
 
-Healthcare outcomes under chattel slavery were poor for enslaved individuals, evidenced by the lack of endogenous growth among the enslaved population and enslavers’ continued reliance on the transatlantic slave trade (until its abolition in 1850) for new enslaved laborers as mortality rates outpaced fertility rates [@bergad1999, @klein2009, @schwartz1985].
+Population health was poor for all people in the nineteenth century, but particularly so for enslaved individuals, who not only experienced epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from severe nutritional deficiencies (@kiple1989).
 
 In the immediate aftermath of the abolition of slavery, which coincided with the implementation of a republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into civil and political life occurred [@carvalho2004].
-Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020, @fischer2008, @otovo2016, @roth2020].
+Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016; @roth2020].
 
-Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level [@franken2019].
+Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019].
 The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe.
-Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986].
-But the major advancements in clinical medicine, that improved maternal mortality and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s.
-Worldwide, the combination of advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992, @woods2009].
+Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986; @benchimol1999; @hochman2013].
+INFANT MORTALITY RATES IMPROVED.
+But the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s.
+Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].
 
-By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to hospitalize labor and delivery [@barreto2015, @barreto2016, @brenes1991, @martins2004, @mott2002, @roth2020].
+By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to hospitalize labor and delivery [@barreto2015; @barreto2016; @brenes1991; @martins2004; @mott2002; @roth2020].
 Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
-In hospitals, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine, including new cesarean section techniques [@martins2004, @rohden2009, @roth2021].
+In hospitals, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine, including new cesarean section techniques [@martins2004; @rohden2009; @roth2021].
 Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which harnessed advances in medical knowledge and public health infrastructure to support a technocratic model towards pregnancy, delivery, and motherhood.
-This ideology relied on essentialized notions of gender, foregrounding physicians’ scientific authority [@besse1996, @freire2009, @martins2008, @otovo2016].
+This ideology relied on essentialized notions of gender and race, foregrounding physicians’ scientific authority [@besse1996; @freire2009; @martins2008; @otovo2016].
 
 Laranjeiras, which was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces.
-Maternidade Laranjeiras provided gynecological and obstetric care free of charge; thus, its clientele came from the city’s poor and working classes [@barreto2016, @roth2020].
+Maternidade Laranjeiras provided gynecological and obstetric care free of charge; thus, its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
 The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants.
 
-This study builds on an important body of work in economic history that uses hospital records to explore trends in birth weights in the past [@costa1998, @costa2004, @goldin1989, @steckel1986, @steckel1998].
+This study builds on an important body of work in economic history that uses hospital records to explore trends in birth weights in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998].
+It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020].
 
 ## Data: Description and Sources
-
+MAYBE ADD IN HERE ON COVARIATES AND TAKE OUT SOME OTHER PARTS.
 In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
-I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO).
+[For information on the data processing, exploratory analysis, and final analysis, including all code in R and the open-access dataset, see my Github repository at BLANK.
+The `Supplemental_Materials` includes information on data processing, exploratory analysis, and model fitting.
+All materials, including the dataset, are open access and reproducible, with instructions on the schematic of the workflow.] I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO).
 The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro.
@@
 Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras.
-I was unable to locate vol.
+From the available issues, I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery ("natural" or vaginal, "interventionist" or requiring some intervention most often forceps, and "operative," or a cesarean section), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth or neonatal death, or live birth), and the mother’s reproductive history [I was unable to locate vol.
 18, nos.
@@
 4 (1926).
-From the available issues, I recorded the following information, when available, for all patients: patient number, gravidity and parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.
-Gravidity refers to the total number of pregnancies a woman has, regardless of duration.
-Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013].
-Please see the `Supplemental_Materials` document for more information on the dataset.
-For full citations on the journals from which these data came, see my other publications [@roth2020, @roth2020a].
+For full citations on the journals from which these data came, see @roth2020; @roth2020a].
 
@@
 I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, the Maternidade Laranjeiras, in the 1920s.
-I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), will have lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.
-
-The outcome I will measure is infant birth weight.
-The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
-I employ birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data.
+I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), had lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.
 
@@
 
-The outcome of interest, birth weight in grams, was measured as a continuous variable.
+The outcome I will measure is infant birth weight.
+The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
+I employ birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data.
+For my main models, I measured birthweight as a continuous variable in line with current studies [CITE].
 Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017].
@@
 
-I run two multivariate models: one in which maternal skin color is categorized into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent).
-The second categorizes maternal skin color into the dataset's original racial groups: White (the reference group), Black, and mixed race.
-Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011, @travassos2004].
+I ran two multivariate linear models: one in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent).
+In the second, I categorized maternal skin color into the dataset's original racial groups: White (the reference group), Black, and mixed race.
+Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011; @travassos2004].
 Racial categories and skin color exist on a spectrum rather than a Black-White binary as in the United States [@telles2006].
-Analyzing both specific categories of skin color and more general categories of ancestry is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013, @wehby2015].
+Analyzing both specific categories of skin color and more general categories of ancestry is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015].
 
 Spontaneous abortion data points are included in the exploratory analysis but excluded from the statistical analysis.
-I also excluded any live births weighing \<500 grams to exclude any possible stillbirths not recorded as such [@nyarko2013, @wehby2015, @wehby2017].
-
-Please see the supplemental materials (`Supplemental_Materials`) for information on the data processing and exploratory analysis.
-
-## Schematic of workflow
-
-If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder; 2) `eda-v1.qmd` or `edacode.R` in the `eda-code` folder; 3) `introanalyis-v1.qmd` or `introanalysis.R` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` or `fullanalysiscode.R` in the `analysis-code` folder.
-Please defer to all `.qmd` files as the master over any `.R` files.
-
-## Data import and cleaning
-
-The GitHub repository for this project includes all relevant materials.
-All discussions of data import are detailed in the files `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder.
-The original raw data, the codebook, and the processed data are available in the `data` folder.
+I also excluded any live births weighing \<500 grams to exclude any possible stillbirths not recorded as such [@nyarko2013; @wehby2015; @wehby2017].
 
@@
 
-For maternal race (sample size n = 2695), the majority of the sample is White (*branca*), with 42.45% of the sample.
-Mixed-race (*parda*) comprises 29.2%, and Black (*preta*) comprises 28.32%.
-If comparing White patients to all patients of color by combining Black and mixed-race categories, of the clinic patients, 1551 (57.6%) were of African descent (defined as *preta* or *parda*) and 1144 (42.4%) were of European descent (*branca*).
+For maternal race (sample size n = 2695), the majority of the sample was White (*branca*) (42.45%).
+Mixed-race (*parda*) comprised 29.2%, and Black (*preta*) comprised 28.32%.
+If comparing White patients to all patients of color by combining Black and mixed-race categories, 1551 (57.6%) were of African descent (defined as *preta* or *parda*) and 1144 (42.4%) were of European descent (*branca*).
 
 Of all reproductive outcomes (sample size n = 2761), 88% were natural deliveries.
-Yet, this also includes spontaneous abortions.
+Yet, this also included spontaneous abortions.
 If we exclude spontaneous abortion (sample size n = 2672), 91% of outcomes were natural deliveries.
 The remaining 9% were interventionist or operative deliveries.
+These included cesarean sections or hysterectomies after a difficult delivery [@roth2020; @roth2020a]
 
@@
 where maternal deaths (MD) are divided by 10,000 live births (LB), was 94.26%.
-For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018, @roth2020].
+For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018; @roth2020].
 But differential levels of recording must be taken into account.
@@
 Thus, the city’s rates were probably higher.
-Nonetheless, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother.
+Nonetheless, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures could be detrimental to maternal-infant health [@goldin1989].
 
@@
 The SBR is 84.8% in this specific clinic.
-We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% (@roth2020, @roth2018).
-Again, registration effects could explain the higher rates in the hospital (@loudon1992).
+We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% [@roth2020; @roth2018].
+Again, registration effects could explain the higher rates in the hospital [@loudon1992].
 
@@
 The skewed sex ratio deserves further study.
-
-To better understand our outcome of interest, birth weight, the exploratory analysis includes the distribution of the data.
-Figure 1 is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO now defines as normal birth weight [@worldhealthorganization2022].
-
-```{r}
-#| label: fig-schematic-1
-#| fig-cap: Histogram of birth weight
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","weight_distribution.png"))
-```
-
-Because the data are missing gestational age, birth length is less important for our analysis.
-Nonetheless, Figure 2 shows a histogram of the distribution of birth lengths in the sample.
-The red line marks the average birth length for both male and female infants (49 centimeters), which is in line with current estimates on birth length globally [@jamshed2020].
-
-```{r}
-#| label: fig-schematic-2
-#| fig-cap: Histogram of birth length
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","length_distribution.png"))
-```
-
-Figure 3 visualizes the relationship between infant weight and length. Unsurprisingly, birth weight and length are positively correlated.
-
-```{r}
-#| label: fig-schematic-3
-#| fig-cap: Scatterplot of birth weight by length
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","length_weight.png"))
-```
-
-Finally, Figure 4 visualizes the relationship of birth weight and length, stratified by sex since male infants are usually slightly heavier and longer than females [@crawford1987, @marsal1996].
-
-```{r}
-#| label: fig-schematic-4
-#| fig-cap: Scatterplot of birth weight by length stratified by sex
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","length_weight_sex.png"))
-```
 
@@
 
-Given that this study is an inferential, hypothesis-supporting approach that is exploring how specific maternal predictors affect the outcome of interest, birth weight, please see the `fullanalysis-v1.qmd` for all predictive metrics, where test/train and cross-validation were applied to test performance.
-Model performance did not significantly improve with these additional tests.
-When run on testing data, most results were insignificant.
-
 ## Results
@@
 
+paradox between taller people and racial disparities in infant birth weight 
+catch up if surviving childhood (Steckel)
+
 # Discussion
+
+Syphilis
+prenatal care had an 8.22% positive rate for prenatal syphilis in 1935 and an 8.66% positive rate in 1936 (Fontenelle 326).
+
+Between 1935 and 1938 7.92% positive tests (Fontenelle, 275 second Fontenelle).
+
+Racism
+Malnutrition during pregnancy (not linear)
 
@@
 
-All models based on historical data must contend with limited sample sizes and missing data.
+All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014].
 In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
@@
 This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017].
-Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights, although other historical studies have datasets that include gestational age [@costa1998, @costa2004, @goldin1989].
+Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989].
 
 A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
-Lack of data on maternal nutrition and health status, which cannot be corrected for by using race as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in past time periods [@loudon1992, @woods2009].
+The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009].
+But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
+Research on the Dutch famine has shown that the later health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
+Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study and in all that look at these issues in the past [@loudon1992; @woods2009].
 
-For all models in this study, the $\(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight.
+For all models in this study, the $(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight.
 This is likely due to the fact that birth weight is a complex trait influenced by genetic, environmental, and social factors [@leimert2020].
@@
 Margo have shown for nineteenth-century Philadelphia, the inclusion of gestational age greatly improved the regression fit in their studies, demonstrating that gestational age could explain much of the variance of birth weight in this study [@goldin1989].
+Nonetheless, other scholars have shown that even though differential rates of PTB explain some of the variance in birth weight according to maternal skin color in the United States, up to 91 percent of variance remained unexplained [@costa2004].
 
-Moreover, this study is not trying to explain which general factors are influencing birth weight.
+This study is not trying to explain which general factors are influencing birth weight.
 Rather, it is exploring if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil.
-For example, the gestational age probably explains much of the variation in the model, yet the gestational age variable was only included in very few observations in the published clinic data from which I created this dataset (please see the Codebook and the `Supplemental_materials.qmd` file for information on this variable).
-
 Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color, however, is also probably associated with premature birth, which is correlated to infant birth weight.
@@
 Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between skin color and health in the early twentieth century.
-
-{{< pagebreak >}}
-
-# References

### 2024-06-27 products/manuscript/supplement/Supplemental_Materials.qmd
Update to manuscript
Changed repository name

@@
     bibliography: ../../../assets/Mada_Bib.bib
-    csl: ../../../assets/vancouver.csl
+    csl: ../../../assets/chicago-fullnote-bibliography.csl
 ---
 # Supplemental Materials
+
+## Schematic of workflow
+
+If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder; 2) `eda-v1.qmd` or `edacode.R` in the `eda-code` folder; 3) `introanalyis-v1.qmd` or `introanalysis.R` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` or `fullanalysiscode.R` in the `analysis-code` folder.
+Please defer to all `.qmd` files as the master over any `.R` files.
+
+## Data import and cleaning
+
+The GitHub repository for this project includes all relevant materials.
+All discussions of data import are detailed in the files `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder.
+The original raw data, the codebook, and the processed data are available in the `data` folder.
 
@@
 When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician.
-For more information on this as a source, see my other publications [@roth2020, @roth2020a].
+For more information on this as a source, see my other publications [@roth2020; @roth2020a].
 
@@
 ## Basic Statistical Analysis
+
+To better understand our outcome of interest, birth weight, the exploratory analysis includes the distribution of the data.
+Figure 1 is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO now defines as normal birth weight [@worldhealthorganization2022].
+
+```{r}
+#| label: fig-schematic-1
+#| fig-cap: Histogram of birth weight
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","weight_distribution.png"))
+```
+
+Because the data are missing gestational age, birth length is less important for our analysis.
+Nonetheless, Figure 2 shows a histogram of the distribution of birth lengths in the sample.
+The red line marks the average birth length for both male and female infants (49 centimeters), which is in line with current estimates on birth length globally [@jamshed2020].
+
+```{r}
+#| label: fig-schematic-2
+#| fig-cap: Histogram of birth length
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","length_distribution.png"))
+```
+
+Figure 3 visualizes the relationship between infant weight and length. Unsurprisingly, birth weight and length are positively correlated.
+
+```{r}
+#| label: fig-schematic-3
+#| fig-cap: Scatterplot of birth weight by length
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","length_weight.png"))
+```
+
+Finally, Figure 4 visualizes the relationship of birth weight and length, stratified by sex since male infants are usually slightly heavier and longer than females [@crawford1987, @marsal1996].
+
+```{r}
+#| label: fig-schematic-4
+#| fig-cap: Scatterplot of birth weight by length stratified by sex
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","length_weight_sex.png"))
+```
 
@@
 
+Given that this study is an inferential, hypothesis-supporting approach that is exploring how specific maternal predictors affect the outcome of interest, birth weight, please see the `fullanalysis-v1.qmd` for all predictive metrics, where test/train and cross-validation were applied to test performance.
+Model performance did not significantly improve with these additional tests.
+When run on testing data, most results were insignificant.
+
 Test/train (75-25 split) and cross-validation were performed on all models.
@@
 Model performance resulted in insignificant results when run on the testing data given the smaller sample size.
-All code for the train/test analysis and cross-validation analysis can be found in `fullanalysis-v1.qmd`.
 

### 2024-06-29 code/analysis-code/fullanalysis-v1.qmd
Merged GT tables; updated supplement
Revisions

@@
 #load needed packages. make sure they are installed.
+#install.packages("glue")
 library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
@@
 library(tidymodels) #for modeling
-
+library(glue)
 rm(list=ls()) #clear the environment
@@
 
-whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
+whereas BW is birth weight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
 
@@
 ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
-  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
-  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") #%>% # Set reference level for ModifiedColor
+  #step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable. Cannot do this if I am merging tables
 
@@
 # Create a gtsummary table directly from the model
-table2 <-
-  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
-  bold_p() %>% 
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
-  modify_column_unhide(columns = c(statistic, std.error)) %>%
-  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table2
+table_model1 <-
+  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
+  add_glance_table(include = c(r.squared, nobs))
+table_model1
 
 #Print table
-table2
-
-# Save regression output into table
-saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
+table_model1
+
+# Save regression output into table
+saveRDS(table_model1, file = here::here("results", "tables", "table_model1_final.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table2_train <-
+table_model1_train <-
   tbl_regression(ML_fit1_train, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
   bold_p() %>% 
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table2_train
-
-# Save regression output into table
-saveRDS(table2_train, file = here::here("results", "tables", "table2_train.rds"))
+table_model1_train
+
+# Save regression output into table
+saveRDS(table_model1_train, file = here::here("results", "tables", "table_model1_train.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table2_test <-
+table_model1_test <-
   tbl_regression(ML_fit1_test, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
   bold_p() %>% 
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table2_test
-
-# Save regression output into table
-saveRDS(table2_test, file = here::here("results", "tables", "table2_test.rds"))
+table_model1_test
+
+# Save regression output into table
+saveRDS(table_model1_test, file = here::here("results", "tables", "table_model1_test.rds"))
 ```
@@
 ML_recipe2 <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>% 
-  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
-  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
-  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") #%>% # Set reference level for ModifiedColor
+  #step_dummy(ModifiedColor) #%>% # Convert all nominal variables to dummy variables. Cannot do this if I am merging tables in gtsummary
+  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors. Cannot do this if I am merging tables in gtsummary
 
@@
 # Create a gtsummary table directly from the model
-table3 <- pull_workflow_fit(ML_fit2) %>%
-  tbl_regression(exponentiate = FALSE, intercept = TRUE) %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
-  modify_column_unhide(columns = c(statistic, std.error)) %>%
-  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
-  bold_p()
-table3
-
-# Save regression output into table
-saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
+table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
+  add_glance_table(include = c(r.squared, nobs))
+table_model2
+
+# Save regression output into table
+saveRDS(table_model2, file = here::here("results", "tables", "table_model2_final.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table3_train <- pull_workflow_fit(ML_fit2_train) %>%
+table_model2_train <- pull_workflow_fit(ML_fit2_train) %>%
   tbl_regression(exponentiate = FALSE, intercept = TRUE) %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
-  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
   bold_p()
-table3_train
-
-# Save regression output into table
-saveRDS(table3_train, file = here::here("results", "tables", "table3_train.rds"))
+table_model2_train
+
+# Save regression output into table
+saveRDS(table_model2_train, file = here::here("results", "tables", "table_model2_train.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table3_test <- pull_workflow_fit(ML_fit2_test) %>%
+table_model2_test <- pull_workflow_fit(ML_fit2_test) %>%
   tbl_regression(exponentiate = FALSE, intercept = TRUE) %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
-  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
   bold_p()
-table3_test
-
-# Save regression output into table
-saveRDS(table3_test, file = here::here("results", "tables", "table3_test.rds"))
+table_model2_test
+
+# Save regression output into table
+saveRDS(table_model2_test, file = here::here("results", "tables", "table_model2_test.rds"))
 ```
 Running the model on our test data, we see no significant results at any alpha level.
+
+# Merging Model 1 and 2 into Table
+
+```{r}
+# Combine the two tables using tbl_merge
+table3 <- tbl_merge(list(table_model1, table_model2), tab_spanner = c("**Model 1**", "**Model 2**")) %>%
+  modify_table_body(fun = ~.x %>% arrange(variable))
+table3
+
+# Save regression output into table
+saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
+```
 
@@
 ML_recipe3 <- recipe(Weightgrams ~ Color, data = ML_linear) %>%
-  step_relevel(Color, ref_level = "White")
+  step_relevel(Color, ref_level = "White") #%>% # Set reference level for White
+  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
 
@@
 # Create a gtsummary table from model
-table4 <- tbl_regression(ML_fit3, exponentiate = FALSE, intercept = TRUE) %>%
-  bold_p() %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+table_model3 <- tbl_regression(ML_fit3, exponentiate = FALSE, intercept = TRUE) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
-  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table4
-
-# Save regression output into table
-saveRDS(table4, file = here("results", "tables", "table4_final.rds"))
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
+  add_glance_table(include = c(r.squared, nobs))
+table_model3
+
+# Save regression output into table
+saveRDS(table_model3, file = here("results", "tables", "table_model3_final.rds"))
 ```
@@
 # Create a gtsummary table from model
-table4_train <- tbl_regression(ML_fit3_train, exponentiate = FALSE, intercept = TRUE) %>%
-  bold_p() %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+table_model3_train <- tbl_regression(ML_fit3_train, exponentiate = FALSE, intercept = TRUE) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
-  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table4_train
-
-# Save regression output into table
-saveRDS(table4_train, file = here("results", "tables", "table4_train.rds"))
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE)
+table_model3_train
+
+# Save regression output into table
+saveRDS(table_model3_train, file = here("results", "tables", "table_model3_train.rds"))
 ```
@@
 # Create a gtsummary table from model
-table4_test <- tbl_regression(ML_fit3_test, exponentiate = FALSE, intercept = TRUE) %>%
+table_model3_test <- tbl_regression(ML_fit3_test, exponentiate = FALSE, intercept = TRUE) %>%
   bold_p() %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table4_test
-
-# Save regression output into table
-saveRDS(table4_test, file = here("results", "tables", "table4_test.rds"))
+table_model3_test
+
+# Save regression output into table
+saveRDS(table_model3_test, file = here("results", "tables", "table_model3_test.rds"))
 ```
@@
   step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
-  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
-  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
-  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") #%>% # Set reference level for ModifiedStatus
+  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
 
@@
 # Create a gtsummary table from model
-table5 <- tbl_regression(ML_fit4, exponentiate = FALSE, intercept = TRUE) %>%
-  bold_p() %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+table_model4 <- tbl_regression(ML_fit4, exponentiate = FALSE, intercept = TRUE) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
-  add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table5
-
-# Save regression output into table
-saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>% # Hide confidence intervals and p-values
+    add_glance_table(include = c(r.squared, nobs)) # Add R-squared and number of observations
+table_model4
+
+# Save regression output into table
+saveRDS(table_model4, file = here("results", "tables", "table_model4_final.rds"))
 ```
@@
 # Create a gtsummary table from model
-table5_train <- tbl_regression(ML_fit4_train, exponentiate = FALSE, intercept = TRUE) %>%
+table_model4_train <- tbl_regression(ML_fit4_train, exponentiate = FALSE, intercept = TRUE) %>%
   bold_p() %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table5_train
-
-# Save regression output into table
-saveRDS(table5_train, file = here("results", "tables", "table5_train.rds"))
+table_model4_train
+
+# Save regression output into table
+saveRDS(table_model4_train, file = here("results", "tables", "table_model4_train.rds"))
 ```
@@
 # Create a gtsummary table from model
-table5_test <- tbl_regression(ML_fit4_test, exponentiate = FALSE, intercept = TRUE) %>%
+table_model4_test <- tbl_regression(ML_fit4_test, exponentiate = FALSE, intercept = TRUE) %>%
   bold_p() %>%
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = FALSE, hide_p = FALSE)
-table5_test
-
-# Save regression output into table
-saveRDS(table5_test, file = here("results", "tables", "table5_test.rds"))
+table_model4_test
+
+# Save regression output into table
+saveRDS(table_model4_test, file = here("results", "tables", "table_model4_test.rds"))
 ```
 Again, most of our results are not significant except the intercept given the small sample size.
+
+## Merge Models 3 and 4
+```{r}
+# Combine the two tables using tbl_merge
+table4 <- tbl_merge(list(table_model3, table_model4), tab_spanner = c("**Model 3**", "**Model 4**")) %>%
+  modify_table_body(fun = ~.x %>% arrange(variable))
+table4
+
+# Save regression output into table
+saveRDS(table4, file = here("results", "tables", "table4_final.rds"))
+```

### 2024-06-29 code/analysis-code/introanalysis-v1.qmd
Merged GT tables; updated supplement
Revisions

@@
 # Create a gtsummary table directly from the model
-table2 <-
+table_fit1 <-
   tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
@@
 #Print table
-table2
+table_fit1
 
 # Save regression output into table
-saveRDS(table2, file = here::here("results", "tables", "table2.rds"))
+saveRDS(table_fit1, file = here::here("results", "tables", "table_fit1.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table_age <-
+table_fit2 <-
   tbl_regression(ML_fit_age, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
   bold_p() %>% 
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
   modify_column_unhide(columns = c(statistic, std.error))
@@
 #Print table
-table_age
+table_fit2
 
 # Save regression output into table
-saveRDS(table_age, file = here::here("results", "tables", "tableage_final.rds"))
+saveRDS(table_fit2, file = here::here("results", "tables", "table_fit2.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table3 <- tbl_regression(ML_lm3, exponentiate = FALSE, intercept = TRUE) %>% 
+table_fit3 <- tbl_regression(ML_lm3, exponentiate = FALSE, intercept = TRUE) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
-table3
+table_fit3
 
 # Save regression output into table
-saveRDS(table3, file = here::here("results", "tables", "table3.rds"))
+saveRDS(table_fit3, file = here::here("results", "tables", "table_fit3.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table4 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
+table_fit4 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
-table4
+table_fit4
 
 # Save regression output into table
-saveRDS(table4, file = here::here("results", "tables", "table4.rds"))
+saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))
 
@@
 # Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
+ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
 
 # Summarize model fit with tidiers
-ML_lm5_summary <- summary(ML_lm5) # Summary of fit
-tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm5) # Model-level statistics
+ML_lm4_summary <- summary(ML_lm4) # Summary of fit
+tidy(ML_lm4, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_lm4) # Model-level statistics
 
 # Create a gtsummary table directly from the model
-table5 <- tbl_regression(ML_lm5, exponentiate = FALSE, intercept = TRUE) %>% 
+table_fit4 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
-table5
+table_fit4
 
 # Save regression output into table
-saveRDS(table5, file = here::here("results", "tables", "table5.rds"))
+saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))
 ```
@@
 # Create a gtsummary table directly from the model
-table6 <- tbl_regression(log1_fitted, exponentiate = TRUE) # use exponentiate true for ORs
-table6
+table_log1 <- tbl_regression(log1_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table_log1
 
 # Save regression output into table
-saveRDS(table6, file = here::here("results", "tables", "table6.rds"))
+saveRDS(table_log1, file = here::here("results", "tables", "table_log1.rds"))
 
@@
 # Create a gtsummary table directly from the model
-table7 <- tbl_regression(log2_fitted, exponentiate = TRUE) # use exponentiate true for ORs
-table7
+table_log2 <- tbl_regression(log2_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table_log2
 
 # Save regression output into table
-saveRDS(table7, file = here::here("results", "tables", "table7.rds"))
+saveRDS(table_log2, file = here::here("results", "tables", "table_log2.rds"))
 
@@
 # Create a gtsummary table directly from the model
-table8 <- tbl_regression(log3_fitted, exponentiate = TRUE) # use exponentiate true for ORs
-table8 <- bold_p(table8, t = 0.05)
-table8
+table_log3 <- tbl_regression(log3_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table_log3 <- bold_p(table_log3, t = 0.05)
+table_log3
 
 # Save regression output into table
-saveRDS(table8, file = here::here("results", "tables", "table8.rds"))
+saveRDS(table_log3, file = here::here("results", "tables", "table_log3.rds"))
 
@@
 # Create a gtsummary table directly from the model
-table9 <- tbl_regression(log4_fitted, exponentiate = TRUE) # exponentiate true for ORs
+table_log4 <- tbl_regression(log4_fitted, exponentiate = TRUE) # exponentiate true for ORs
 
 #Bold significant p-values
-table9 <- bold_p(table9, t = 0.05)
-table9
+table_log4 <- bold_p(table_log4, t = 0.05)
+table_log4
 
 # Save regression output into table
-saveRDS(table9, file = here::here("results", "tables", "table9.rds"))
+saveRDS(table_log4, file = here::here("results", "tables", "table_log4.rds"))
 

### 2024-06-29 code/eda-code/eda-v1.qmd
Merged GT tables; updated supplement
Revisions

@@
 #load needed packages. make sure they are installed.
+#install.packages("hrbrthemes")
 library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
@@
   geom_histogram(binwidth = 25, fill="#69b3a2") +
-  geom_vline(xintercept = c(1500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
+  geom_vline(xintercept = c(2500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
   theme_ipsum() +
@@
 # Combine the two tables
-table1.1 <- tbl_merge(
+table2 <- tbl_merge(
   tbls = list(stratified_color, stratified_sex),
@@
 # Display the modified table
-table1.1
+table2
 
 #Save table
-saveRDS(table1.1, file = here::here("results", "tables", "table1.1_final.rds"))
+saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
 ```

### 2024-06-29 products/manuscript/Manuscript_Draft.qmd
Merged GT tables; updated supplement
Revisions

@@
 #install.packages("gdtools")
+#install.packages("tidyverse")
 library(tidyverse) #for data processing/cleaning

### 2024-06-29 products/manuscript/Manuscript_Final.qmd
Merged GT tables; updated supplement
Revisions

@@
 ## Data: Description and Sources
-MAYBE ADD IN HERE ON COVARIATES AND TAKE OUT SOME OTHER PARTS.
+
 In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
-[For information on the data processing, exploratory analysis, and final analysis, including all code in R and the open-access dataset, see my Github repository at BLANK.
+I extracted the sample from Brazil’s first major obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO).
+The RGO was associated with the country’s national medical association and obstetrics and gynecological society, both based in Rio de Janeiro.
+The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings.
+Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras.
+[@roth2020; @roth2020a.
+For information on the data processing, exploratory analysis, and final analysis, including all code in R and the open-access dataset, see my Github repository.
 The `Supplemental_Materials` includes information on data processing, exploratory analysis, and model fitting.
-All materials, including the dataset, are open access and reproducible, with instructions on the schematic of the workflow.] I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO).
-The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro.
-The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*.
-In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*.
-In 1922, it became the *Revista de Gynecologia e D’Obstetricia*.
-
-The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings.
-Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras.
-From the available issues, I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery ("natural" or vaginal, "interventionist" or requiring some intervention most often forceps, and "operative," or a cesarean section), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth or neonatal death, or live birth), and the mother’s reproductive history [I was unable to locate vol.
-18, nos.
-4, 5, 6, 8 (1924) and vol.
-20, no.
-4 (1926).
-For full citations on the journals from which these data came, see @roth2020; @roth2020a].
+All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available in the README.
+@roth2024.]
 
@@
 
-I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil?
-I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, the Maternidade Laranjeiras, in the 1920s.
+I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil measured using birth weight as a proxy for general fetal and neonatal health?
+I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro’s first public maternity hospital.
 I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), had lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.
 
-# Methods
-
-I use linear models to estimate the relationship of maternal variables on infant birth weight.
-
-## Study Measures
-
-The outcome I will measure is infant birth weight.
+To explore this relationship, I use linear models to estimate the relationship of maternal variables on infant birth weight.
+In the supplemental analysis, I include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable.
+Most results were insignificant, and thus I exclude them here.
 The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
-I employ birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data.
-For my main models, I measured birthweight as a continuous variable in line with current studies [CITE].
+In my main models, employ birthweight as a continuous variable in line with both current and historical studies [CITE], so these categories are not relevant for the linear models, although they can help explain the skew of the data.
 Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017].
@@
 
-Spontaneous abortion data points are included in the exploratory analysis but excluded from the statistical analysis.
-I also excluded any live births weighing \<500 grams to exclude any possible stillbirths not recorded as such [@nyarko2013; @wehby2015; @wehby2017].
+In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births.
+For the statistical analysis I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid any possible stillbirths [@nyarko2013; @wehby2015; @wehby2017].
 
@@
 ```
+
+ADD IN SECOND SUMMARY TABLE HERE
 
@@
 I ran four linear models to understand the relationship between maternal factors and infant birth weight.
-The first two models are a simple and multiple linear regression, respectively, with outcome variable as birthweight in grams.
+The first two models are a simple and multiple linear regression, respectively, with outcome variable as birth weight in grams.
 In these two models, I recoded maternal skin color as maternal ancestry (`Euro-descent` or `Afro-descent`).
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
-Results are displayed in Table 2.
-
-{{< pagebreak >}}
-
-```{r}
-#| label: tbl-table2
-#| tbl-cap: "Model 1 Results"
-#| echo: FALSE
-table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
-flex_table2 <- as_flex_table(table2)
-flex_table2 <- flex_table2 %>%
-  flextable::set_table_properties(width = .5, align = "center")
-flex_table2
-```
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
 
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term.
-Results are displayed in Table 3.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term.
+Results for both models are displayed in Table 3.
 
@@
 #| label: tbl-table3
-#| tbl-cap: "Model 2 Results"
+#| tbl-cap: "Model 1 and 2 Results"
 #| echo: FALSE
-i = 16.5
-w = i*0.3937
-font_size <- 10
 table3 <- readRDS(here::here("results", "tables", "table3_final.rds"))
 flex_table3 <- as_flex_table(table3)
-flex_table3 <- autofit(flex_table3)
-auto_widths <- dim(flex_table3)$widths/sum(dim(flex_table3)$widths) # calculate column widths
-flex_table3 <- width(flex_table3, width=w*auto_widths) # adjust column widths
-flex_table3 <- fontsize(flex_table3, size=font_size) # set font size
+flex_table3 <- flex_table3 %>%
+  flextable::set_table_properties(width = .5, align = "center")
 flex_table3
@@
 
-The second two models are also simple and multiple linear regressions with outcome variable as birthweight in grams.
-However, in these two models, I maintained maternal skin color as three categories, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women.
+The second two models are also simple and multiple linear regressions with outcome variable as birth weight in grams.
+However, in these two models, I maintained maternal skin color as the original three categories recorded in the hospital records, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women.
 The first simple linear regression model is as follows:
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
 Results are displayed in Table 4.
+
+The second multiple linear regression model is as follows:
+
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+Results are displayed in Table 5.
 
@@
 #| label: tbl-table4
-#| tbl-cap: "Model 3 Results"
+#| tbl-cap: "Model 3 and 4 Results"
 #| echo: FALSE
@@
 flex_table4
-```
-
-The second multiple linear regression model is as follows:
-
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
-Results are displayed in Table 5.
-
-{{< pagebreak >}}
-
-```{r}
-#| label: tbl-table5
-#| tbl-cap: "Model 4 Results"
-#| echo: FALSE
-i = 16.5
-w = i*0.3937
-font_size <- 10
-table5 <- readRDS(here::here("results", "tables", "table5_final.rds"))
-flex_table5 <- as_flex_table(table5)
-flex_table5 <- autofit(flex_table5)
-auto_widths <- dim(flex_table5)$widths/sum(dim(flex_table5)$widths) # calculate column widths
-flex_table5 <- width(flex_table5, width=w*auto_widths) # adjust column widths
-flex_table5 <- fontsize(flex_table5, size=font_size) # set font size
-flex_table5
 ```
@@
 In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
-This means that the birthweight variable is capturing both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation.
+Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation.
 This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017].

### 2024-06-29 products/manuscript/supplement/Supplemental_Materials.qmd
Merged GT tables; updated supplement
Revisions

@@
 
+## Models
+
+### Linear Models
+
 For the basic statistical analysis, I ran three linear models to understand the relationship between maternal factors and infant birth weight.
@@
 ```{r}
-#| label: tbl-table2
-#| tbl-cap: "Supplemental Table"
+#| label: tbl-table1
+#| tbl-cap: "Model 1"
 #| echo: FALSE
-table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
-table_gt <- gtsummary::as_gt(table2)
+table1 <- readRDS(here::here("results", "tables", "table_fit1.rds"))
+table_gt <- gtsummary::as_gt(table_fit1)
 table_gt |> gt::as_raw_html()
@@
 ```{r}
-#| label: tbl-table_age
-#| tbl-cap: "Supplemental Table"
+#| label: tbl-table2
+#| tbl-cap: "Model 2"
 #| echo: FALSE
-table_age <- readRDS(here::here("results", "tables", "tableage_final.rds"))
-table_gt <- gtsummary::as_gt(table_age)
+table2 <- readRDS(here::here("results", "tables", "table_fit2.rds"))
+table_gt <- gtsummary::as_gt(table_fit2)
 table_gt |> gt::as_raw_html()
@@
 
-The third and final exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, nationality, and gestational status on infant birth weight.
+The third exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, nationality, and gestational status on infant birth weight.
 
 ```{r}
-#| label: tbl-table4
-#| tbl-cap: "Supplemental Table"
+#| label: tbl-table3
+#| tbl-cap: "Model 3"
 #| echo: FALSE
-table4 <- readRDS(here::here("results", "tables", "table4.rds"))
-table_gt <- gtsummary::as_gt(table4)
+table3 <- readRDS(here::here("results", "tables", "table_fit3.rds"))
+table_gt <- gtsummary::as_gt(table_fit3)
 table_gt |> gt::as_raw_html()
 ```
-
 In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women.
@@
 This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
+
+The fourth exploratory model again runs a multivariate linear regression, this time using skin color as it was differentiated in the original data, or with three categories (White, Black, and Mixed Race).
+
+```{r}
+#| label: tbl-table4
+#| tbl-cap: "Model 4"
+#| echo: FALSE
+table4 <- readRDS(here::here("results", "tables", "table_fit4.rds"))
+table_gt <- gtsummary::as_gt(table_fit4)
+table_gt |> gt::as_raw_html()
+```
+
+### Logistic Models
+
+Now we will still consider `Birthweight` as the outcome of interest, but we will transform it into a categorical outcome of <=2500 (low birth weight, LBW) and >2500 (normal birth weight, NBW) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
+
+The first model is a simple logistic regression, with modified maternal skin color as the predictor variable.
+
+```{r}
+#| label: tbl-table5
+#| tbl-cap: "Model 5"
+#| echo: FALSE
+table5 <- readRDS(here::here("results", "tables", "table_log1.rds"))
+table_gt <- gtsummary::as_gt(table_log1)
+table_gt |> gt::as_raw_html()
+```
+
+The second model is also a simple logistic regression, this time using `Color` as the predictor variable (so three categories, Black, mixed race, and White).
+
+```{r}
+#| label: tbl-table6
+#| tbl-cap: "Model 6"
+#| echo: FALSE
+table6 <- readRDS(here::here("results", "tables", "table_log2.rds"))
+table_gt <- gtsummary::as_gt(table_log2)
+table_gt |> gt::as_raw_html()
+```
+
+Then, I ran two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.
+The results are below.
+
+```{r}
+#| label: tbl-table7
+#| tbl-cap: "Model 7"
+#| echo: FALSE
+table7 <- readRDS(here::here("results", "tables", "table_log3.rds"))
+table_gt <- gtsummary::as_gt(table_log3)
+table_gt |> gt::as_raw_html()
+```
+
+```{r}
+#| label: tbl-table8
+#| tbl-cap: "Model 8"
+#| echo: FALSE
+table8 <- readRDS(here::here("results", "tables", "table_log4.rds"))
+table_gt <- gtsummary::as_gt(table_log4)
+table_gt |> gt::as_raw_html()
+```
+In these two models, the only significant finding is the final logistic model, in which infants born to mixed-race mothers had 21% higher odds of being born with low birth weight compared to infants born to white mothers.
+The other models did not show significant results.
+
+Best practice is to not categorize continuous variables, as this can include arbitrary cutoffs for variables that are best understood continuously [@Senn2005].
+Thus, I will not include these logistic regressions in the final analysis.
 
@@
 
-Here I include the cross-validation results for each model.
+Here I include the cross-validation results for the linear models.
 
@@
 
+The performance metrics for the logistic models are available in the `introanalysis-v1.qmd` file.
+
 {{< pagebreak >}}

### 2024-07-05 code/eda-code/eda-v1.qmd
Draft 1 for advisor
Final changes for draft to advisor

--- 2024-04-30 code/eda-code/eda-v1.qmd
@@
 library(hrbrthemes) #for nice themes
+library(flextable) #for nice tables
 hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
@@
 ```
+# Feature Engineering
+
 Now let's make some tables about the data to get a better idea of the overall numbers.
@@
 
+Finally, for the summary table, in which I want to include infant birth weight stratified by maternal skin color and by fetal sex, I will create new dummy variables here for that purpose.
+
 ```{r}
@@
 
-# Recode factor using dplyr (easier)
+# Recode factor using dplyr
 ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
@@
 ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
+
+# Create dummy variables for birth weight for infants born to Black, mixed-race, and White mothers for summary table. Create dummy variables for birth weight according to fetal sex.
+ML_summary <- ML_summary %>%
+  mutate(
+    Weightblack = factor(if_else(Color == "Preta", "Yes", "No"), levels = c("No", "Yes")),
+    Weightmixed = factor(if_else(Color == "Parda", "Yes", "No"), levels = c("No", "Yes")),
+    Weightwhite = factor(if_else(Color == "Branca", "Yes", "No"), levels = c("No", "Yes")),
+    Weightfemale = factor(if_else(Sex == "F", "Yes", "No"), levels = c("No", "Yes")),
+    Weightmale = factor(if_else(Sex == "M", "Yes", "No"), levels = c("No", "Yes")),
+    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
+    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
+    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
+    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
+    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
+  )
+
+# Verify new variables
+str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale")])
 
@@
 ```{r, output="html"}
-# Create gtsummary table 
+# Create gtsummary table full table for supplemental analysis
+table_supplemental <- 
+  ML_summary %>%
+  tbl_summary(
+    include = c(Color, ModifiedColor, Status, ModifiedStatus, Age, Nationality, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams),
+    missing = "no",
+    label = list(ModifiedColor ~ "Ancestry", Status ~ "Parity or Gravidity", ModifiedStatus ~ "Parity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Lengthcentimeters ~ "Birth Length (cms)", Weightgrams ~ "Birth Weight (grms)"),
+    statistic = list(all_continuous() ~ "{mean} ({sd})")
+  ) %>%
+  add_n()
+table_supplemental
+
+#Save table
+saveRDS(table_supplemental, file = here::here("results", "tables", "table_supplemental_final.rds"))
+
+# Create gtsummary table for final analysis
 table1 <- 
   ML_summary %>%
+  mutate(
+    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
+    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
+    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
+    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
+    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
+  ) %>%
   tbl_summary(
-    include = c(Color, ModifiedColor, Status, ModifiedStatus, Age, Nationality, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Weightgrams, Lengthcentimeters),
+    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale),
     missing = "no",
-    label = list(ModifiedColor ~ "Ancestry", Status ~ "Parity or Gravidity", ModifiedStatus ~ "Parity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Weightgrams ~ "Infant Birthweight (grms)", Lengthcentimeters ~ "Infant Birth Length (cms)")
+    label = list(
+      ModifiedColor ~ "Ancestry", 
+      ModifiedStatus ~ "Parity", 
+      Age ~ "Maternal Age", 
+      ModifiedNationality ~ "Combined Nationality", 
+      Birth ~ "Birth Outcome", 
+      MaternalOutcome ~ "Maternal Outcome", 
+      FetalOutcome ~ "Fetal Outcome", 
+      Lengthcentimeters ~ "Birth Length (cms)", 
+      Weightgrams ~ "Birth Weight (grms)",
+      WeightgramsBlack ~ "Black (grms)",
+      WeightgramsMixed ~ "Mixed-Race (grms)",
+      WeightgramsWhite ~ "White (grms)",
+      WeightgramsFemale ~ "Female (grms)",
+      WeightgramsMale ~ "Male (grms)"
+    ),
+    statistic = list(
+       all_continuous() ~ "{mean} ({sd})",
+      WeightgramsBlack ~ "{mean} ({sd})",
+      WeightgramsMixed ~ "{mean} ({sd})",
+      WeightgramsWhite ~ "{mean} ({sd})",
+      WeightgramsFemale ~ "{mean} ({sd})",
+      WeightgramsMale ~ "{mean} ({sd})"
+    ),
+    digits = list(WeightgramsBlack ~ 0, WeightgramsMixed ~ 0, WeightgramsWhite ~ 0, WeightgramsFemale ~ 0, WeightgramsMale ~ 0)
   ) %>%
-  add_n()
+  add_n() %>%
+  modify_table_body( #indent these three rows
+    ~.x %>%
+      mutate(label = case_when(
+        variable == "WeightgramsBlack" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsWhite" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsMixed" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsFemale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsMale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        TRUE ~ label
+      ))
+  )
+
 table1
@@
 saveRDS(table1, file = here::here("results", "tables", "table1_final.rds"))
+```
+
+Now we will create the summary table for mean birthweights across time and place.
+
+```{r}
+#Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")
+
+#load data
+Mean_Birthweights <- readRDS(data_location)
+
+# Replace NA with a dash for each column
+Mean_Birthweights <- Mean_Birthweights %>%
+  mutate(across(where(is.numeric), ~ ifelse(is.na(.), "-", as.character(.)))) %>%
+  mutate(across(where(is.character), ~ replace_na(., "-")))
+
+# Display the table using flextable
+table2 <- flextable(Mean_Birthweights) %>%
+  set_caption("Mean Birth Weight Comparison") %>%
+  align(j = 2, align = "left", part = "all") %>%
+  align(j = 4, align = "right", part = "all") %>%
+  set_header_labels(
+    Sample = "Sample",
+    Year = "Year", 
+    Mean = "Mean (grams)",
+    Source = "Source"
+  ) %>%
+  bold(part = "header") %>%
+  #autofit() %>%
+  set_table_properties(layout = "autofit", width = 1, align = "center")
+table2
+
+#Save table
+saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
 ```
@@
 ```
-
 # Notes

### 2024-07-05 code/processing-code/processingfile-v1.qmd
Merged GT tables; updated supplement
Revisions
Draft 1 for advisor
Final changes for draft to advisor

@@
 
+# Data loading for mean birth weights table
+```{r}
+# path to data
+# note the use of the here() package and not absolute paths
+data_location <- here::here("data","raw-data","Mean_Birthweights.csv")
+Mean_Birthweights <- read_csv(data_location)
+
+# Structure of Data
+glimpse(Mean_Birthweights)
+
+# Remove median and maternal age variables given missing data
+Mean_Birthweights <- Mean_Birthweights %>% dplyr::select(-Median, -`Maternal Age (mean)`, -"<2500gm (%)")
+
+# Structure of data
+glimpse(Mean_Birthweights)
+
+# Save Mean_Birthweight as RDS
+save_data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")
+saveRDS(Mean_Birthweights, file = save_data_location)
+```
 # Notes
-
-This script is a work in progress. I will be adding more to it as I continue to work on the data.

### 2024-07-05 products/manuscript/Manuscript_Final.qmd
Draft 1 for advisor
Final changes for draft to advisor

@@
 ```
-
 # Abstract
 
-This cross-sectional study uses linear regression models to understand disparities in infant birth weight in relation to maternal skin color (ancestry), age, gestational status (parity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
-Infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant.
-I hypothesize these differences were related to the historical legacies of slavery, only abolished in 1888, including intergenerational BLANK caused by racism, on population health in Brazil.
+This cross-sectional study uses linear models to understand disparities in infant birth weight in relation to maternal skin color, age, parity, and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s.
+It finds that infants born to women of color (either Black, "*preta,*" or mixed-race, "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant.
+I hypothesize these differences were related to the historical legacies of slavery, only abolished in 1888, including race-based socioeconomic inequality and possible epigenetic effects from the lived experience of racism, on population health in Brazil.
 
@@
 
-At 4 in the morning on November 17, 1922, Virginia Maria da Conceição, a twenty-six-year-old Black ("*preta*") woman gave birth to a stillborn daughter in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now called Maternidade Escola).
-This was Conceição's fifth birth; prior, she had delivered three full-term infants vaginally (described as "naturally" in the clinical notes) and one premature infant, born at seven months gestational age.
-According to the notes, Conceição had had a "good" pregnancy.
-After arriving at the hospital, she remained in the first stage of labor ("dilatação") for five hours.
-The second stage of labor ("expulsão") lasted only thirty minutes.
-The placenta was delivered ten minutes later.
-The unnamed stillborn girl weighed 3,180 grams at birth, a normal weight for a full-term infant.
-Despite the apparently easy pregnancy and delivery, Conceição would not take home a healthy baby; rather, "hereditary syphilis," as the fetal autopsy noted, was the cause of the stillbirth [@rolindo1923, pp.
+At four in the morning on November 17, 1922, Virginia Maria da Conceição, a twenty-six-year-old Black ("*preta*") woman gave birth to a stillborn daughter in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now called Maternidade Escola).
+This was Conceição's fifth birth; prior, she had delivered three full-term infants vaginally and one premature infant, born at seven months gestational age.
+According to the hospital records, Conceição had had a "good" pregnancy, with a relatively easy delivery.
+After arriving, she remained in the first stage of labor ("dilatação") for five hours.
+The second stage of labor ("expulsão") lasted thirty minutes.
+She delivered the placenta ten minutes later.
+The unnamed stillborn girl weighed 3180 grams at birth, a normal weight for a full-term infant and around 150 grams higher than the average weight for female infants and for Black infants born at the hospital in the 1920s.
+But, as the hospital records detailed, Conceição did not take home a healthy baby; rather, "hereditary syphilis," as the fetal autopsy noted, was the cause of intrauterine stillbirth [@rolindo1923, pp.
 542-44].
 
-Virginia Maria da Conceição's experience with stillbirth demonstrates the precarious state of maternal-infant health in early twentieth-century Brazil.
-I use Conceição's tragic but all too common story of poor perinatal and neonatal outcomes as a jumping off point to explore the historical relationship between social determinants of health and infant birth weight in a former slave society.
+Virginia Maria da Conceição's experience with syphilis and stillbirth demonstrates the precarious state of maternal-infant health in early twentieth-century Brazil.
+I use Conceição's tragic but all too common story of poor perinatal and neonatal outcomes as a jumping off point to explore the historical legacy of slavery on maternal-infant health in Brazil.
 
-Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality [@travassos2004].
-Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts [@nyarko2013; @wehby2015; @vilanova2019].
-Mothers with lower education levels and lower socioeconomic status also give birth to infants at lower birth weights [@wehby2015; @vilanova2019].
-Understanding these patterns is important as low birth weight is associated with higher rates of infant mortality in the first year of life [@cardoso2005; @vilanova2019].
+Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
+Contemporary studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts [@nyarko2013; @wehby2015a; @vilanova2019].
+Mothers with lower education levels and lower socioeconomic status also give birth to infants at lower birth weights [@wehby2015a; @wehby2017a; @vilanova2019].
 
-These trends hold true in the United States, as well, where Black newborns have lower birth weights than White newborns, and these trends have worsened over time [@mangold1991; @pollock2021].
-Racism is a major factor in these outcomes [@ifatunji2022], as studies demonstrate that U.S.-born Black women have infants with lower birth weights and higher rates of pre-term birth than Black immigrant women [@braveman2021; @clay2023; @mckenzie-sampson2021].
+These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns have lower birth weights than White newborns, with trends worsening over time [@pollock2021].
+Racism is a major factor in these outcomes, as studies demonstrate that U.S.-born Black women have infants with lower birth weights and higher rates of pre-term birth than Black immigrant women [@braveman2021a; @clay2023a; @ifatunji2022a; @mckenzie-sampson2021].
 Other racial groups, whether U.S.
 or foreign-born, also face similar trends, although more research is needed to understand confounding factors [@montoya-williams2022].
+Understanding these patterns within and across communities is important as low birth weight is associated with higher rates of infant mortality in the first year of life and possible health problems over the life course [@cardoso2005; @vilanova2019].
 
-Scholars across disciplines have argued that health disparities in countries with histories of race-based chattel slavery, from the United States to Brazil, are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @steckel1986].
-In Brazil, most historical studies remain descriptive in nature and have not tested any quantitative associations between the legacies of slavery, including race-based inequality and its consequences and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2018a; @roth2018; @roth2020; @telles2022].
+Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery, including the United States, are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @steckel1986].
+In Brazil, most historical studies of health under slavery and in its aftermath remain descriptive in nature and have not tested any quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2018a; @roth2018; @roth2020; @telles2022].
 
-Here, I am to fill this gap by examining infant birth weight in relation to maternal skin color or ancestry, age, gestational status (parity), and nationality in Rio de Janeiro, Brazil's public maternity hospital, the Maternidade Laranjeiras, which opened in 1904 and provided free gynecological and obstetric care to the then-capital city’s poor and working-class women.
-It relies on a unique sample of infant birth weight data (n=2845).
+Here, I aim to fill this gap by examining infant birth weight in relation to maternal skin color, age, parity, and nationality in the Maternidade Laranjeiras, which opened in 1904 and provided free gynecological and obstetric care to the then-capital city’s poor and working-class women.
+
+In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon.
+In our current postgenomic era, in which "race" as a biological category has been thoroughly debunked, the possibility for centuries-old patterns of hierarchical scientific thinking remains very real [@alam2024; @desalle2018; @roberts2011].
+Discussions of the potential effects of racism on epigenetic markers is one area where this essentialization can occur.
+Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022].
+Poor Black women like Virginia Maria da Conceição and her children were not "pathologically" different or part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].
+
+This paper aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine the lives of people like Virginia Maria da Conceição.
 
@@
 
-Brazil imported over 4 million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009].
+Brazil imported over four million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009].
 It was also the last country to abolish slavery (1888) in the Western Hemisphere.
-It had, and still has, the largest number of African-descended peoples in the world outside of Nigeria [@michelle2023].
-Around 55% of country's population today is of African-descent [@belandi2023].
+In the nineteenth century, population health was poor for all people, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from severe nutritional deficiencies [@karasch1987; @kiple1989].
 
-Population health was poor for all people in the nineteenth century, but particularly so for enslaved individuals, who not only experienced epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from severe nutritional deficiencies (@kiple1989).
-
-In the immediate aftermath of the abolition of slavery, which coincided with the implementation of a republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into civil and political life occurred [@carvalho2004].
+In the immediate aftermath of the abolition of slavery, which coincided with the implementation of the country's first republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into civil and political life occurred [@carvalho2004].
 Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016; @roth2020].
@@
 Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986; @benchimol1999; @hochman2013].
-INFANT MORTALITY RATES IMPROVED.
-But the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s.
+Although infant mortality rates declined from the nineteenth century, the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@roth2020].
 Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].
 
-By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to hospitalize labor and delivery [@barreto2015; @barreto2016; @brenes1991; @martins2004; @mott2002; @roth2020].
+By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2015; @barreto2016; @brenes1991; @martins2004; @mott2002; @roth2020].
 Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
-In hospitals, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine, including new cesarean section techniques [@martins2004; @rohden2009; @roth2021].
-Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which harnessed advances in medical knowledge and public health infrastructure to support a technocratic model towards pregnancy, delivery, and motherhood.
+There, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine [@martins2004; @rohden2009; @roth2021].
+Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which harnessed advances in medical knowledge and public health infrastructure to support a technocratic model towards pregnancy and motherhood.
 This ideology relied on essentialized notions of gender and race, foregrounding physicians’ scientific authority [@besse1996; @freire2009; @martins2008; @otovo2016].
 
-Laranjeiras, which was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces.
-Maternidade Laranjeiras provided gynecological and obstetric care free of charge; thus, its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
+At the same time, old racial hierarchies based on legal status morphed into new "scientific" ways to categorize humans based on skin color.
+In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than coercively restricting reproduction through sterilization [@schwarcz1993; @stepan1991].
+Although some eugenicists called for sterilization of "unfit" citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013].
+In the end, however, whiteness still remained a cultural and social "goal" even if the state did not engage in coercive measures of reproductive control.
+
+Maternidade Laranjeiras, which was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces.
+Laranjeiras provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
 The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants.
+Because most clinical notes were short – only those that described surgical births such as cesarean sections, or in the case of Virginia Maria da Conceição included the fetal autopsy report, had longer notes – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
+Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
 
-This study builds on an important body of work in economic history that uses hospital records to explore trends in birth weights in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998].
-It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020].
+This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998].
+It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].
 
-## Data: Description and Sources
+## Analysis
 
-In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
-I extracted the sample from Brazil’s first major obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO).
-The RGO was associated with the country’s national medical association and obstetrics and gynecological society, both based in Rio de Janeiro.
+I analyzed a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
+I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO).
 The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings.
-Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras.
-[@roth2020; @roth2020a.
-For information on the data processing, exploratory analysis, and final analysis, including all code in R and the open-access dataset, see my Github repository.
-The `Supplemental_Materials` includes information on data processing, exploratory analysis, and model fitting.
-All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available in the README.
-@roth2024.]
+Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras.[^1]
 
-## Questions and Hypotheses
+[^1]: For complete information on the source, see @roth2020; @roth2020a.
+My Github repository, in particular, the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting.
+All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online.
+See @roth2024.
 
-I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil measured using birth weight as a proxy for general fetal and neonatal health?
-I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro’s first public maternity hospital.
-I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), had lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.
+Relying on the dataset, I ask whether the legacy of slavery, both as possible intergenerational inheritance patterns or as contemporary experiences of discrimination in a post-abolition society, affected the health of infants born in the first public maternity hospital in 1920s Brazil using birth weight as a proxy for general fetal and neonatal health.
+I quantify how maternal race, nationality, age, and parity explain racial disparities in infant birth weight.
 
-To explore this relationship, I use linear models to estimate the relationship of maternal variables on infant birth weight.
-In the supplemental analysis, I include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable.
-Most results were insignificant, and thus I exclude them here.
-The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022].
-In my main models, employ birthweight as a continuous variable in line with both current and historical studies [CITE], so these categories are not relevant for the linear models, although they can help explain the skew of the data.
-Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017].
-I discuss this constraint in the limitations section below.
-Infant length was recorded in centimeters.
-
-The covariates include maternal age (recorded as a continuous variable in years), maternal skin color, gestational status, and maternal nationality.
-
-I ran two multivariate linear models: one in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent).
-In the second, I categorized maternal skin color into the dataset's original racial groups: White (the reference group), Black, and mixed race.
-Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011; @travassos2004].
-Racial categories and skin color exist on a spectrum rather than a Black-White binary as in the United States [@telles2006].
-Analyzing both specific categories of skin color and more general categories of ancestry is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015].
-
-In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births.
-For the statistical analysis I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid any possible stillbirths [@nyarko2013; @wehby2015; @wehby2017].
-
-## Descriptive analysis
-
-I performed the exploratory analysis on the ML_summary dataset with n = 2845 observations and 14 variables. Table 1 demonstrates the distribution of the data.
+{{< pagebreak >}}
 
@@
 ```
+{{< pagebreak >}}
 
-ADD IN SECOND SUMMARY TABLE HERE
+Table 1 provides summary statistics for the sample.
+When comparing White patients to all patients of color by combining Black and mixed-race categories, 1551 (58%) were of African descent (defined as *preta* or *parda*) and 1144 (42%) were of European descent (*branca*).
+First-time mothers ("nulliparous") were 41% of the sample (n = 2836).
+Perhaps they would return to Laranjeiras, or another maternity hospital, to give birth to their future children, demonstrating the process of hospitalization of childbirth in real time.
+Most mothers were Brazilian.
+Of the total sample (n = 2773), 84% were Brazilian.
+Most immigrants were European.
 
-For maternal race (sample size n = 2695), the majority of the sample was White (*branca*) (42.45%).
-Mixed-race (*parda*) comprised 29.2%, and Black (*preta*) comprised 28.32%.
-If comparing White patients to all patients of color by combining Black and mixed-race categories, 1551 (57.6%) were of African descent (defined as *preta* or *parda*) and 1144 (42.4%) were of European descent (*branca*).
+Of all reproductive outcomes (n = 2761), 88% were natural deliveries.
+If we exclude spontaneous abortion from the sample (n = 2672), 91% of outcomes were natural deliveries.
+The remaining 9% were interventionist or operative deliveries.
+These included forceps and cesarean sections [@roth2020; @roth2020a].
 
-Of all reproductive outcomes (sample size n = 2761), 88% were natural deliveries.
-Yet, this also included spontaneous abortions.
-If we exclude spontaneous abortion (sample size n = 2672), 91% of outcomes were natural deliveries.
-The remaining 9% were interventionist or operative deliveries.
-These included cesarean sections or hysterectomies after a difficult delivery [@roth2020; @roth2020a]
-
-For patients who went to the clinic to deliver their infant (thus excluding those suffering from spontaneous abortions or receiving postpartum care after an at-home delivery), 23 died.
-The Maternal Mortality Ratio (MMR), calculated as:
+For patients who went to the clinic to deliver their infant (thus excluding those suffering from spontaneous abortions), 23 died. The Maternal Mortality Ratio (MMR), calculated as:
 
@@
 For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018; @roth2020].
-But differential levels of recording must be taken into account.
-These rates could reflect a registration effect, as all births and deaths were recorded in the hospital, whereas accurate reporting at the city level was less reliable [@loudon1992].
-Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s.
+This is a stark difference, but we must take into account differential levels of recording.
+Hospital rates could have reflected a registration effect, as all births and deaths were recorded in the hospital, whereas reporting at the city level was less reliable [@loudon1992].
+Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s [@roth2020].
 Thus, the city’s rates were probably higher.
-Nonetheless, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures could be detrimental to maternal-infant health [@goldin1989].
+Nonetheless, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures could be detrimental to maternal-infant health before the rise of antibiotics and blood transfusions [@goldin1989].
 
@@
 
-$$SBR = \frac{SB}{TB} \times 1,000$$ where the total number of stillbirths (SB) is divided by 1000 total births (TB).
-The SBR is 84.8% in this specific clinic.
+$$SBR = \frac{SB}{TB} \times 1,000$$ 
+where the total number of stillbirths (SB) is divided by 1000 total births (TB), was 84.8%.
 We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% [@roth2020; @roth2018].
-Again, registration effects could explain the higher rates in the hospital [@loudon1992].
+Again, registration effects could explain the higher hospital rates.
 
@@
 
-where the total number of live male births (M) is divided by by 100 live female births (F) in a given period.
+where the total number of live male births (M) is divided by 100 live female births (F) in a given period.
 
-The sex ratio is 1.2: there were 120 male live births per 100 female live births at this specific clinic.
+The sex ratio was 1.2: there were 120 male live births per 100 female live births at this specific clinic.
 This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
@@
 
+Table 2 provides the mean birth weights for the Maternidade Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight across the Americas and in New Zealand.
+
+```{r}
+#| label: tbl-table2
+#| tbl-cap: "Mean Birth Weights"
+#| echo: FALSE
+table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
+table2
+```
+
+As we can see, Black infants born at the Maternidade Laranjeiras had the lowest mean birth weight (3037 grams) compared to all other historical samples except for estimated enslaved birth weights in the antebellum US South, which scholars have since called into question for being too low [@costa2004].
+Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent time period.
+Yet, they were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
+White infants born at Laranjeiras had a mean birth weight of 3133 grams, slightly more than 50 grams lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
+Yet they were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent time period (3423 grams).
+As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same time period [@costa1998; @costa2004; @goldin1989].
+
 ## Statistical analysis
 
-I ran four linear models to understand the relationship between maternal factors and infant birth weight.
-The first two models are a simple and multiple linear regression, respectively, with outcome variable as birth weight in grams.
-In these two models, I recoded maternal skin color as maternal ancestry (`Euro-descent` or `Afro-descent`).
+I use Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight.
+In the supplemental analysis available online, I include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable (either normal birth weight, NBW, or low birth weight, LBW, according to current WHO classifications), and in line with some studies on birth weight [@worldhealthorganization2022; @falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
+Most logistic model results were insignificant, and thus I exclude them here.
+In my main models, I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
+Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is best practice for disentangling pre-term birth (PTB) from low birth weight (LBW) [@cutland2017].
+I discuss this constraint in the discussion below.
+The covariates include maternal age (recorded as a continuous variable in years), maternal skin color, gestational status, and maternal nationality.
 
-The first simple linear regression model is as follows:
+I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race.
+Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011; @travassos2004].
+Racial categories and skin color exist on a spectrum rather than a Black-White binary [@telles2006].
+Analyzing both specific categories of skin color and more general categories of ancestry can help address these different understandings of race and is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].
+
+Yet, neither the records' reporting of maternal skin color nor my own reclassification of the skin color of the mothers in this study can fully serve as a proxy for maternal health.
+Assuming so runs the potential of reifying intra-group differences at the expense of inter-group ones.
+Yet, given the limited historical data available, I run these models understanding the inherent limitations of using skin color as a category at all and as a proxy for health status in particular.
+
+The simple linear regression model is as follows:
 
@@
 
-The second multiple linear regression model is as follows:
+The multiple linear regression model is as follows:
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
 Results for both models are displayed in Table 3.
@@
 
-The second two models are also simple and multiple linear regressions with outcome variable as birth weight in grams.
+The second two models are also simple and multiple linear regressions.
 However, in these two models, I maintained maternal skin color as the original three categories recorded in the hospital records, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women.
-The first simple linear regression model is as follows:
+The simple linear regression model is as follows:
 
@@
 
-The second multiple linear regression model is as follows:
+The multiple linear regression model is as follows:
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
-Results are displayed in Table 5.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+Results are displayed in Table 4.
 
@@
 
-## Results
-
 All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference group) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race).
 Age and nationality were not significant.
-Women who were nulliparous (delivering their first child) had infants with lower birth weights.
+Women who were nulliparous (delivering their first child) had infants with lower birth weights, which is an association still found today [@souza2020].
 
-paradox between taller people and racial disparities in infant birth weight 
-catch up if surviving childhood (Steckel)
+# Discussion and Conclusion
 
-# Discussion
+The intrauterine environment is a critical determinant of infant birth weight.
+Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care can all influence birth weight.
+Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother [@costa1998; @steckel1986].
+A threshold of malnutrition must be met before there is an adverse effect on fetal growth.
+Other factors include maternal age, parity, and birth spacing, as well as gestational age.
+We can imagine that poorer women performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
+Generalized poverty might have influenced maternal nutrition, as well as the ability to access prenatal care.
 
-Syphilis
-prenatal care had an 8.22% positive rate for prenatal syphilis in 1935 and an 8.66% positive rate in 1936 (Fontenelle 326).
+Because the hospital records excluded information on maternal health factors, we don't know which contributed to racial disparities in infant birth weight.
+One possible cause that we do have some data for is syphilis.
+Like Costa found for early twentieth-century Baltimore, higher syphilis rates among African-descended women in Rio de Janeiro might have explained repeated pregnancy loss, whether a miscarriage or stillbirth, as was the case of Virginia Maria da Conceição.
+Studies demonstrate that syphilis increases adverse birth outcomes including increased risk of LBW and PTB [@dasilva2024].
+The clinical records that comprise this dataset did not consistently record syphilis rates.
+Scholars have shown how medical and public health professionals viewed it as an important cause of adverse health outcomes and even a possible "scourge" on the future of the Brazilian race [@carrara1996a].
+Although I don't have accurate rates for Laranjeiras, city-wide rates from the 1930s can provide some context for the prevalence of the disease among pregnant women.
+Of all prenatal syphilis exams performed between 1935 and 1938 at public clinics or hospitals across the city, not stratified by maternal skin color, 7.92% came back positive [@fontenelle, p.
+275].
 
-Between 1935 and 1938 7.92% positive tests (Fontenelle, 275 second Fontenelle).
-
-Racism
-Malnutrition during pregnancy (not linear)
-
-## Strengths and Limitations
+New discussions of epigenetic markers associated with the lived experience of racism might also be applicable here.
+Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors), mediated through epigenetic mechanisms.
+Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@leimert2020].
+Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health [@meloni2022].
 
@@
 In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
-Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation.
-This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017].
-Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989].
+Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks' gestation.
+We are thus unable to know how many of these infants were low birth weight at term [@falcao2020].
+This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@cutland2017].
+Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989].
+I assume that infants at risk for pre-term birth (PTB) were also at-risk for LBW.
 
-A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
-The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009].
-But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
-Research on the Dutch famine has shown that the later health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
-Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study and in all that look at these issues in the past [@loudon1992; @woods2009].
-
-For all models in this study, the $(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight.
-This is likely due to the fact that birth weight is a complex trait influenced by genetic, environmental, and social factors [@leimert2020].
-Yet, as Claudia Goldin and Robert A.
-Margo have shown for nineteenth-century Philadelphia, the inclusion of gestational age greatly improved the regression fit in their studies, demonstrating that gestational age could explain much of the variance of birth weight in this study [@goldin1989].
-Nonetheless, other scholars have shown that even though differential rates of PTB explain some of the variance in birth weight according to maternal skin color in the United States, up to 91 percent of variance remained unexplained [@costa2004].
-
-This study is not trying to explain which general factors are influencing birth weight.
-Rather, it is exploring if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil.
-Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color, however, is also probably associated with premature birth, which is correlated to infant birth weight.
-
-## Conclusions
-
-Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between skin color and health in the early twentieth century.
+Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study.
+Given the paucity of historical understandings of current-day health disparities in maternal-infant health in post-abolition Brazil, this paper demonstrates the need for more research into quantitative associations between skin color and health in the past.

### 2024-07-05 products/manuscript/supplement/Supplemental_Materials.qmd
Draft 1 for advisor
Final changes for draft to advisor

@@
 
+In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births.
+For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births [@nyarko2013; @wehby2015; @wehby2017].
+
+A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
+The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009].
+But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
+Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
+
 ### Maternal Skin Color
@@
 I classified clinical outcomes as spontaneous abortion, natural (normal delivery), interventionist (interventionist delivery), and operative (surgical delivery).
+
+### Full Summary Table
+
+Here is the code for the full summary table of all variables.
+
+```{r}
+#| label: tbl-tablesupplemental
+#| tbl-cap: "Supplemental Table"
+#| echo: FALSE
+table_supplemental <- readRDS(here::here("results", "tables", "table_supplemental_final.rds"))
+table_gt <- gtsummary::as_gt(table_supplemental)
+table_gt |> gt::as_raw_html()
+```
 
@@
 ```
+
+For all models in this study, the $(R^2)$ is small, indicating that the model does not explain much of the variance in birth weight.
+This is likely due to the fact that birth weight is a complex trait influenced by genetic (not related to skin color), environmental, and social factors [@leimert2020].
+Yet, as Claudia Goldin and Robert A.
+Margo have shown for nineteenth-century Philadelphia, the inclusion of gestational age greatly improved the regression fit in their studies, demonstrating that gestational age could explain much of the variance of birth weight in this study [@goldin1989].
+Nonetheless, other scholars have shown that even though differential rates of PTB explain some of the variance in birth weight according to maternal skin color in the United States, up to 91 percent of influenced birth weight, as the lack of historical data precludes this line of inquiry.
+Rather, it is exploring if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro.
+Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color is also probably associated with premature birth, which is correlated to infant birth weight.
 

### 2024-07-05 code/analysis-code/fullanalysis.R

@@ -1,5 +1,6 @@
 ## ---- packages --------
 #load needed packages. make sure they are installed.
+#install.packages("glue")
 library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
 library(skimr) #for nice visualization of data 
 library(here) #to set paths
@@ -10,13 +11,12 @@ library(gt) #for nice tables
 library(gtsummary) #for summary tables
 library(ggplot2) #for plotting
 library(tidymodels) #for modeling
-
-#rm(list=ls()) #clear the environment
+library(glue)
+rm(list=ls()) #clear the environment
 
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear.rds")
-
 #load data
 ML_linear <- readRDS(data_location)
 
@@ -25,11 +25,9 @@ ML_linear <- readRDS(data_location)
 # This enables the analysis to be reproducible when random numbers are used 
 rngseed = 1234
 set.seed(rngseed)
-
 # Create training data
 # Put 3/4 of the data into the training set 
 data_split <- initial_split(ML_linear, prop = 3/4)
-
 # Create data frames for the two sets:
 train_data <- training(data_split)
 test_data  <- testing(data_split)
@@ -39,34 +37,27 @@ test_data  <- testing(data_split)
 # Model using recipes and workflows
 # Define recipe for model
 ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
-  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
-  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
-
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") #%>% # Set reference level for ModifiedColor
+#step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable. Cannot do this if I am merging tables
 # Specify model
 ML_spec1 <- linear_reg() %>% 
   set_engine("lm")
-
 # Create work flow
 ML_wf1 <- workflow() %>%
   add_recipe(ML_recipe1) %>%
   add_model(ML_spec1)
-
 # Fit the work flow
[…]

### 2024-07-05 code/analysis-code/introanalysis.R

@@ -10,12 +10,11 @@ library(gt) #for nice tables
 library(gtsummary) #for summary tables
 library(ggplot2) #for plotting
 library(tidymodels) #for modeling
-#rm(list=ls()) #clear the environment
+rm(list=ls()) #clear the environment
 
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear.rds")
-
 #load data
 ML_linear <- readRDS(data_location)
 
@@ -25,32 +24,25 @@ ML_linear <- readRDS(data_location)
 ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
   step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
   step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
-
 # Specify model
 ML_spec1 <- linear_reg() %>% 
   set_engine("lm")
-
 # Create work flow
 ML_wf1 <- workflow() %>%
   add_recipe(ML_recipe1) %>%
   add_model(ML_spec1)
-
 # Fit the work flow
 ML_fit1 <- fit(ML_wf1, data = ML_linear)
-
 # Summarize model fit with tidiers
 ML_summary1 <- summary(ML_fit1) # Summary of fit
 tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
 glance(ML_fit1) # Model-level statistics
-
 # Produce predictions for R-squared
 ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
 ML_fit1_pred
-
 # Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
 ML_fit1_pred <- bind_cols(ML_fit1_pred, ML_linear %>% select(Weightgrams))
 ML_fit1_pred
-
 # Plot observed versus predicted
 plot1 <- ggplot(ML_fit1_pred, aes(x = Weightgrams, y = .pred)) +
   geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
@@ -58,60 +50,48 @@ plot1 <- ggplot(ML_fit1_pred, aes(x = Weightgrams, y = .pred)) +
   labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
   coord_obs_pred()
 plot1
-
 # Create metric set including RMSE and R-squared
[…]

### 2024-07-05 code/eda-code/edacode-extra.R

@@ -1,5 +1,25 @@
 ## ---- Extracode --------
 # Don't use this code in future
+# Create summary stats for birth weight by maternal modifiedcolor
+birthweight_color <- ML_summary %>%
+  filter(!is.na(ModifiedColor) & !is.na(Weightgrams)) %>%
+  group_by(ModifiedColor) %>%
+  summarize(
+    mean_weight = mean(Weightgrams, na.rm = TRUE),
+    sd_weight = sd(Weightgrams, na.rm = TRUE)
+  )
+print(birthweight_color)
+
+# Create summary stats for birth weight by fetal sex
+birthweight_sex <- ML_summary %>%
+  filter(!is.na(Sex) & !is.na(Weightgrams)) %>%
+  group_by(Sex) %>%
+  summarize(
+    mean_weight = mean(Weightgrams, na.rm = TRUE),
+    sd_weight = sd(Weightgrams, na.rm = TRUE)
+  ) 
+print(birthweight_sex)
+
 # Summary stats of continuous variables
 summary_stats <- ML_summary %>%
   summarize(
@@ -10,8 +30,41 @@ summary_stats <- ML_summary %>%
     mean_length = mean(Lengthcentimeters, na.rm = TRUE),
     sd_length = sd(Lengthcentimeters, na.rm = TRUE)
   )
-
 print(summary_stats)
+
+#Create gtsummary table on infant birth weight stratified by sex and maternal skin color
+stratified_color <- 
+  ML_summary %>%
+  filter(!is.na(Weightgrams) & !is.na(ModifiedColor)) %>% # Exclude rows with NA in weight & MC variable
+  rename("Infant Weight (grms)" = Weightgrams) %>% # Rename Weightgrams to Infant Weight (grms)
+  tbl_summary(
+    by = ModifiedColor,  # Grouping by ModifiedColor
+    include = "Infant Weight (grms)",
+    statistic = list("Infant Weight (grms)" ~ "{mean} ({sd})")
+  )
+stratified_color
+
+stratified_sex <- 
+  ML_summary %>%
+  filter(!is.na(Weightgrams) & !is.na(Sex)) %>%
+  rename("Infant Weight (grms)" = Weightgrams) %>% # Rename Weightgrams to Infant Weight (grms)
+  tbl_summary(
+    by = Sex,  # Grouping by Sex
+    include = "Infant Weight (grms)",
+    statistic = list("Infant Weight (grms)" ~ "{mean} ({sd})")
+  )
+stratified_sex
+
+# Combine the two tables
[…]

### 2024-07-05 code/eda-code/edacode.R

@@ -1,5 +1,6 @@
 ## ---- packages --------
 #load needed packages. make sure they are installed.
+#install.packages("hrbrthemes")
 library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
 library(skimr) #for nice visualization of data 
 library(here) #to set paths
@@ -10,21 +11,20 @@ library(gt) #for nice tables
 library(gtsummary) #for summary tables
 library(ggplot2) #for plotting
 library(hrbrthemes) #for nice themes
+library(flextable) #for nice tables
 hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
 rm(list=ls()) #clear the environment
 
 ## ---- loaddata --------
 #Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_summary_processed.rds")
-
 #load data
 ML_summary <- readRDS(data_location)
 
-## ---- table1 --------
+## ---- createsummary --------
 # Create summary
 summary = skimr::skim(ML_summary)
 print(summary)
-
 # save to file
 summarytable_file = here("results", "tables", "summarytable.rds")
 saveRDS(summary, file = summarytable_file)
@@ -47,7 +47,7 @@ ggsave(filename = figure_file, plot=age_plot, bg="white") #background is white
 weight_plot <- ML_summary %>% filter(!is.na(Weightgrams)) %>% #filter out NAs
   ggplot(aes(x = Weightgrams)) +
   geom_histogram(binwidth = 25, fill="#69b3a2") +
-  geom_vline(xintercept = c(1500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
+  geom_vline(xintercept = c(2500, 4000), linetype = "dashed", color = "red") +  # Add vertical lines that mark normal birthweight
   theme_ipsum() +
   theme(plot.title = element_text(size=15))
 plot(weight_plot)
@@ -64,7 +64,6 @@ length_plot <- ML_summary %>% filter(!is.na(Lengthcentimeters)) %>% #filter out 
   theme_ipsum() +
   theme(plot.title = element_text(size=15))
 plot(length_plot)
-
 # Save plot to file
[…]

### 2024-07-05 code/processing-code/processingcode.R

@@ -18,6 +18,7 @@ library(here) #to set paths
 library(renv) #for package management
 library(knitr) #for nice tables
 library(kableExtra) #for nice tables
+rm(list=ls()) #clear the environment
 
 ## ---- loaddata --------
 # path to data
@@ -50,37 +51,48 @@ skimr::skim(ML_raw)
 ## ---- cleandata --------
 # Dataset created for multiple reasons. Need to narrow down relevant variables
 # Remove unnecessary variables
-ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Birth_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
+ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
 #Summarize missing data using dplyr
-missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
+missing_data_summary <- ML_1 %>% dplyr::summarise_all(list(~sum(is.na(.))))
+missing_data_summary
+# Extract information on gestation from Birth_Notes and create new variable gestation with singleton and twin births
+ML_2 <- ML_1 %>%
+  mutate(Birth_Notes = str_trim(Birth_Notes)) %>% # removes any leading or trailing whitespace from Birth_Notes column
+  mutate(Gestation = if_else(is.na(Birth_Notes) | Birth_Notes == "", "singleton", # checks if Birth_Notes is NA or empty strings and assigns singleton to Gestation column
+                             if_else(str_detect(tolower(Birth_Notes), "gemelar"), "twin", "singleton"))) # checks for gemelar in case-insensitive manner and assigns twin accordingly or singleton otherwise
+ML_2
 # Remove variables with missing data not relevant to hypothesis
-ML_2 <- ML_1 %>% dplyr::select(-CivilStatus, -GestationalAge_Months)
+ML_3 <- ML_2 %>% dplyr::select(-CivilStatus, -GestationalAge_Months, -Birth_Notes)
 #Summarize missing data using dplyr
-missing_data_summary <- ML_2 %>% dplyr::summarise_all(funs(sum(is.na(.))))
-View(missing_data_summary)
[…]

### 2024-07-05 products/manuscript/supplement/Supplemental_Materials.qmd
Updated supplemental
For Draft to advisor

@@
 table5 <- readRDS(here::here("results", "tables", "table_log1.rds"))
-table_gt <- gtsummary::as_gt(table_log1)
+table_gt <- gtsummary::as_gt(table5)
 table_gt |> gt::as_raw_html()
@@
 table6 <- readRDS(here::here("results", "tables", "table_log2.rds"))
-table_gt <- gtsummary::as_gt(table_log2)
+table_gt <- gtsummary::as_gt(table6)
 table_gt |> gt::as_raw_html()
@@
 table7 <- readRDS(here::here("results", "tables", "table_log3.rds"))
-table_gt <- gtsummary::as_gt(table_log3)
+table_gt <- gtsummary::as_gt(table7)
 table_gt |> gt::as_raw_html()
@@
 table8 <- readRDS(here::here("results", "tables", "table_log4.rds"))
-table_gt <- gtsummary::as_gt(table_log4)
+table_gt <- gtsummary::as_gt(table8)
 table_gt |> gt::as_raw_html()
@@
 ```{r}
-#| label: tbl-table5
+#| label: tbl-table9
 #| tbl-cap: "Cross-Validation Model 1"
@@
 ```{r}
-#| label: tbl-table6
+#| label: tbl-table10
 #| tbl-cap: "Cross-Validation Model 2"
@@
 ```{r}
-#| label: tbl-table7
+#| label: tbl-table11
 #| tbl-cap: "Cross-Validation Model 3"
@@
 ```{r}
-#| label: tbl-table8
+#| label: tbl-table12
 #| tbl-cap: "Cross-Validation Model 4"

### 2024-07-08 products/manuscript/Manuscript_Final.qmd
Final update for draft
Draft for review

@@
 
-This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998].
+This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993].
 It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].
@@
 See @roth2024.
+In the supplemental analysis, I further include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable (either normal birth weight, NBW, or low birth weight, LBW, according to current WHO classifications), and in line with some studies on birth weight @worldhealthorganization2022; @falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a.
 
 Relying on the dataset, I ask whether the legacy of slavery, both as possible intergenerational inheritance patterns or as contemporary experiences of discrimination in a post-abolition society, affected the health of infants born in the first public maternity hospital in 1920s Brazil using birth weight as a proxy for general fetal and neonatal health.
-I quantify how maternal race, nationality, age, and parity explain racial disparities in infant birth weight.
+I quantify how maternal skin color, nationality, age, and parity were associated with infant birth weight.
 
@@
 I use Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight.
-In the supplemental analysis available online, I include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable (either normal birth weight, NBW, or low birth weight, LBW, according to current WHO classifications), and in line with some studies on birth weight [@worldhealthorganization2022; @falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
-Most logistic model results were insignificant, and thus I exclude them here.
-In my main models, I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
+I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
 Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is best practice for disentangling pre-term birth (PTB) from low birth weight (LBW) [@cutland2017].

### 2024-07-08 products/manuscript/supplement/Supplemental_Materials.qmd
Final update for draft
Draft for review

@@
     csl: ../../../assets/chicago-fullnote-bibliography.csl
+    suppress-bibliography: true
 ---
@@
 
+```{r, echo=FALSE, message=FALSE}
+#load packages
+library(tidyverse) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+library(renv) #for package management
+library(gtsummary) #for summary tables
+library(flextable) #for nice tables
+```
 ## Schematic of workflow
@@
 In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births.
-For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births [@nyarko2013; @wehby2015; @wehby2017].
+For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births [@nyarko2013; @wehby2015a; @wehby2017a].
 
@@
 table_supplemental <- readRDS(here::here("results", "tables", "table_supplemental_final.rds"))
-table_gt <- gtsummary::as_gt(table_supplemental)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table_supplemental)
+flex_table
 ```
@@
 table1 <- readRDS(here::here("results", "tables", "table_fit1.rds"))
-table_gt <- gtsummary::as_gt(table_fit1)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table1)
+flex_table
 ```
@@
 table2 <- readRDS(here::here("results", "tables", "table_fit2.rds"))
-table_gt <- gtsummary::as_gt(table_fit2)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table2)
+flex_table
 ```
@@
 table3 <- readRDS(here::here("results", "tables", "table_fit3.rds"))
-table_gt <- gtsummary::as_gt(table_fit3)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table3)
+flex_table
 ```
@@
 table4 <- readRDS(here::here("results", "tables", "table_fit4.rds"))
-table_gt <- gtsummary::as_gt(table_fit4)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table4)
+flex_table
 ```
@@
 table5 <- readRDS(here::here("results", "tables", "table_log1.rds"))
-table_gt <- gtsummary::as_gt(table5)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table5)
+flex_table
 ```
@@
 table6 <- readRDS(here::here("results", "tables", "table_log2.rds"))
-table_gt <- gtsummary::as_gt(table6)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table6)
+flex_table
 ```
@@
 table7 <- readRDS(here::here("results", "tables", "table_log3.rds"))
-table_gt <- gtsummary::as_gt(table7)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table7)
+flex_table
 ```
@@
 table8 <- readRDS(here::here("results", "tables", "table_log4.rds"))
-table_gt <- gtsummary::as_gt(table8)
-table_gt |> gt::as_raw_html()
+flex_table <- as_flex_table(table8)
+flex_table
 ```
@@
 
-Best practice is to not categorize continuous variables, as this can include arbitrary cutoffs for variables that are best understood continuously [@Senn2005].
+Best practice is to not categorize continuous variables, as this can include arbitrary cutoffs for variables that are best understood continuously [@senn2005].
 Thus, I will not include these logistic regressions in the final analysis.
@@
 The performance metrics for the logistic models are available in the `introanalysis-v1.qmd` file.
-
-{{< pagebreak >}}
-
-# References

### 2024-07-24 code/analysis-code/fullanalysis-v1.qmd
Final Draft Updates
After meeting with Swartzendruber

@@
   tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
@@
 table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%

### 2024-07-24 code/analysis-code/introanalysis-v1.qmd
Draft 1 for advisor
Final changes for draft to advisor
Final Draft Updates
After meeting with Swartzendruber

@@
 ```{r}
+#set seed
 set.seed(123)
@@
 # Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
-log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) %>%
-  step_dummy(all_nominal(), -all_outcomes()) %>%
-  step_normalize(all_predictors())
+log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
+  #step_dummy(all_nominal(), -all_outcomes()) %>%
+  #step_normalize(all_predictors())
 
@@
 # Create a gtsummary table directly from the model
-table_log1 <- tbl_regression(log1_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table_log1 <- tbl_regression(log1_fitted, exponentiate = TRUE) %>% # use exponentiate true for ORs
+  bold_p(t = 0.05) %>%
+  modify_table_styling(
+    columns = label,
+    rows = label == "ModifiedColor",
+    text_format = "bold"
+  )
 table_log1
@@
 ```{r}
-# Simple logistic regression model with ModifiedColor as the predictor
+# Simple logistic regression model with Color as the predictor
 # Check factor levels
@@
 # Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
-log2_recipe <- recipe(BirthweightCategory ~ Color, data = ML_linear) %>%
-  step_dummy(all_nominal(), -all_outcomes()) %>%
-  step_normalize(all_predictors())
+log2_recipe <- recipe(BirthweightCategory ~ Color, data = ML_linear) #%>%
+  #step_dummy(all_nominal(), -all_outcomes()) %>%
+  #step_normalize(all_predictors())
 
@@
 # Create a gtsummary table directly from the model
-table_log2 <- tbl_regression(log2_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table_log2 <- tbl_regression(log2_fitted, exponentiate = TRUE) %>% # use exponentiate true for ORs
+  bold_p(t = 0.05) %>%
+  modify_table_styling(
+    columns = label,
+    rows = label == "Color",
+    text_format = "bold"
+  )
 table_log2
@@
 # Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
-log3_recipe <- recipe(BirthweightCategory ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
-  step_dummy(all_nominal(), -all_outcomes()) %>%
-  step_normalize(all_predictors())
+log3_recipe <- recipe(BirthweightCategory ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) #%>%
+  #step_dummy(all_nominal(), -all_outcomes()) %>%
+  #step_normalize(all_predictors())
 
@@
 In our final model, our ROC_AUC is 0.477, still worse than random guessing. Our pseudo-R2 is 0.007.
+
+# Merge unadjusted log models 1 and 2
+
+```{r}
+# Combine the two tables using tbl_stack for unadjusted ORs
+table5 <- tbl_stack(list(table_log1, table_log2), group_header = c()) %>%
+  as_gt() %>%
+  gt::tab_style(style = gt::cell_text(weight = "bold"),
+  locations = gt::cells_row_groups(groups = everything()))
+table5
+
+# Save regression output into table
+saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
+```
+```

### 2024-07-24 code/eda-code/eda-v1.qmd
Final Draft Updates
After meeting with Swartzendruber

@@
 saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
+
+#Summarize missing data using dplyr
+missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
+missing_data_summary
 ```
@@
     label = list(
-      ModifiedColor ~ "Ancestry", 
-      ModifiedStatus ~ "Parity", 
-      Age ~ "Maternal Age", 
-      ModifiedNationality ~ "Combined Nationality", 
-      Birth ~ "Birth Outcome", 
-      MaternalOutcome ~ "Maternal Outcome", 
-      FetalOutcome ~ "Fetal Outcome", 
-      Lengthcentimeters ~ "Birth Length (cms)", 
-      Weightgrams ~ "Birth Weight (grms)",
-      WeightgramsBlack ~ "Black (grms)",
-      WeightgramsMixed ~ "Mixed-Race (grms)",
-      WeightgramsWhite ~ "White (grms)",
-      WeightgramsFemale ~ "Female (grms)",
-      WeightgramsMale ~ "Male (grms)"
+      Color ~ "Color (n = 2,695)",
+      ModifiedColor ~ "Ancestry (n = 2,695)", 
+      ModifiedStatus ~ "Parity (n = 2,836)", 
+      Age ~ "Maternal Age (n = 2,783)", 
+      ModifiedNationality ~ "Combined Nationality (n = 2,773)", 
+      Birth ~ "Birth Outcome (n = 2,761)", 
+      MaternalOutcome ~ "Maternal Outcome (n = 2,829)", 
+      FetalOutcome ~ "Fetal Outcome (n = 2,666)", 
+      Sex ~ "Sex (n = 2,534)",
+      Lengthcentimeters ~ "Birth Length (cm) (n = 2,405)", 
+      Weightgrams ~ "Birth Weight (g) (n = 2,384)",
+      WeightgramsBlack ~ "Black (g) (n = 643)",
+      WeightgramsMixed ~ "Mixed-Race (g) (n = 675)",
+      WeightgramsWhite ~ "White (g) (n = 950)",
+      WeightgramsFemale ~ "Female (g) (n = 1,074)",
+      WeightgramsMale ~ "Male (g) (n = 1,270)"
     ),
@@
   ) %>%
-  add_n() %>%
+  #add_n() #%>%
   modify_table_body( #indent these three rows

### 2024-07-24 products/manuscript/Manuscript_Final_Lancet.qmd
Final Draft Updates
After meeting with Swartzendruber

---
title: "Low Infant Birth Weight in Brazil"
subtitle: "A Cross-Sectional Historical Approach"
author: Cassia Roth
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
    bibliography: ../../assets/Mada_Bib.bib
    csl: ../../assets/vancouver.csl
    suppress-bibliography: false
---

```{r, echo=FALSE, message=FALSE}
#Loading packages
#install.packages("flextable")
#install.packages("gdtools")
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(gtsummary) #for summary tables
library(flextable) #for nice tables
library(gt) #for nice tables
```
# Abstract
## Background
In 1888, after nearly four centuries, Brazil became the last country in the Western Hemisphere to abolish slavery. After abolition, the state did not create programs to incorporate the formerly enslaved and their descendants into economic, civil, or political life. Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century. Contemporary public health studies have shown that health disparities continue to mirror patterns from over a century ago. The goal of this study is to quantify health disparities in the post-abolition period in the realm of maternal-infant health by analyzing infant birth weight in relation to maternal variables in one of Brazil's first public maternity hospitals.

## Methods
This cross-sectional analysis uses hospital-level data collected between 1922-1926 from the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil. It was the city's first public maternity hospital. I use linear models to assess disparities in infant birth weight in relation to maternal skin color, age, parity, and nationality.

## Findings
Infants born to women of color (either Black, "*preta,*" or mixed-race, "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant.

## Interpretations
This study hypothesizes that these differences were related to the historical legacies of slavery, including race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments -- and possible epigenetic effects from the lived experience of racism. The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.

## Funding
This archival research for this project was funded by the National Science Foundation (NSF) and the Fulbright IIE Student Program.
{{< pagebreak >}}

# Introduction
Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research, as low birth weight is associated with higher rates of neonatal and infant mortality and possible health problems over the life course [@cardoso2005; @paixao2021; @vilanova2019].

Contemporary studies in Brazil have shown that non-White mothers have higher rates of LBW [@nyarko2013; @wehby2015a; @vilanova2019]. These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns have higher rates of LBW than White newborns, with trends worsening over time [@pollock2021]. Racism is a major factor in these outcomes, as studies demonstrate that U.S.-born Black women have higher rates of LBW and PTB than Black immigrant women [@braveman2021a; @clay2023a; @david1997].

Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @martins2024; @steckel1986]. In Brazil, most historical studies of health under slavery and in its aftermath remain descriptive in nature and have not tested any quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2020; @telles2022].

Brazil imported over four million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009]. It was also the last country to abolish slavery (1888) in the Western Hemisphere. In the nineteenth century, population health was poor for all people, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from severe nutritional deficiencies [@karasch1987; @kiple1989].

In the immediate aftermath of the abolition of slavery, which coincided with the implementation of the country's first republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into economic, civic, or political life occurred [@carvalho2004]. Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016; @roth2020].

Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019]. The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe. Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986; @benchimol1999; @hochman2013]. Although infant mortality rates declined from the nineteenth century, the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@roth2020]. Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].

By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @martins2004; @mott2002; @roth2020]. Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process. There, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine [@martins2004; @rohden2009; @roth2021]. Underlying these structural and technological changes was an ideology scholars have described as "scientific motherhood," supported by both physicians and elite women, which relied on a technocratic model towards pregnancy and motherhood and, in doing so, essentialized notions of gender and foregrounded physicians’ scientific authority [@freire2009; @martins2004; @otovo2016].

At the same time, longstanding racial hierarchies forged under racial slavery morphed into "scientific" ways to categorize humans based on skin color in post-abolition contexts. In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than coercively restricting reproduction through sterilization [@schwarcz1993; @stepan1991]. Although some eugenicists called for sterilization of "unfit" citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013]. Whiteness still remained a cultural and social "goal" even if the state did not engage in coercive measures of reproductive control.

Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces. Laranjeiras provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020]. The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants. Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021]. Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.

This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993]. It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].

In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon. In our current postgenomic era, in which "race" as a biological category has been thoroughly debunked, the possibility for centuries-old patterns of hierarchical scientific thinking remains very real [@roberts2011]. Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022]. The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not "pathologically" different or part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].

This study analyzes infant birth weight as a proxy for general fetal and neonatal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras. It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.

# Methods

## Data

This analysis draws on an unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926. I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). In the initial months of publication, the reports included more complete information, including labor and delivery time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a].

From the available issues, I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.

The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca de Biomedicina-A, Universidade Estadual do Rio de Janeiro (BBA-UERJ), all in Rio de Janeiro, Brazil. Between January 2012 and July 2013, I manually digitized the journal by photographing each volume. Then, between January and August 2017, I manually input the data into Excel from the digital reproductions. I then converted this into a .csv file for upload into R version 4.4.1 (2024-06-14).

The project's Github repository, in particular, the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting. All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online [@roth2024].

## Variables
### Outcome

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams. In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW, <p>&ge;</p>2500 g or low birth weight, LBW, <2500 g, according to current WHO classifications) [@worldhealthorganization2022].

### Key Independent Variables

*Maternal skin color*. The original records categorized maternal racial categories as White (*branca*), the reference group; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and mixed-race patients into one category, `Afro-Descent`, and White patients into a `Euro-Descent` category.

Racial terminology in Brazil was, and continues to be, complex and dynamic [@telles2006; @travassos2004]. Racial categories and skin color exist on a spectrum rather than a Black-White binary [@telles2006]. Analyzing both specific categories of skin color and more general categories of ancestry can help address these different understandings of race and is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

*Maternal age*. I maintained the original data of maternal age in years.

*Gestational status*. The original hospital records divided gestational status into eight categories of parity and gravidity: nulliparous, primiparous, secundiparous, multiparous; and primigravida, secundigravida, trigravida, and multigravida. Due to issues of statistical power, I combined these into two categories: 1) nulliparous, primigravida, and primiparous into `nulliparous` (or any woman who is giving birth for the first time); and 2) secundiparous, multiparous, secundigravida, trigravida, and multigravida into `multiparous` (or any woman who has given birth more than once).

*Maternal nationality*. The original hospital records recorded maternal nationality by country, which are available in the supplemental file. For analysis here, I created categories based on individual country (`nationality`) and a dummy variable based on region (`modifiednationality`). I recategorized the latter as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian).

### Key Descriptive Variables

*Delivery type*. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy. However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries), coded as operatory or natural by hospital records, as interventionist. I classified spontaneous abortion as a separate category.

*Infant length*. Infant length was recorded in centimeters in the original records. A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021]. The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009]. But other health problems related to maternal undernutrition during pregnancy depend on gestational period. Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].

*Maternal outcome*. Hospital records recorded maternal outcome was recorded as death, discharge, or hospital transferal, which I maintain here.

*Fetal outcome*. In the summary statistics, I included both spontaneous abortions and all live births. For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data [@nyarko2013; @wehby2015a; @wehby2017a].

*Fetal sex*. I maintained the original hospital categories of fetal sex as male and female.

## Descriptive statistics

I calculated summary statistics for the sample, including sample proportions for maternal skin color, maternal ancestry, parity, maternal nationality, birth outcome, maternal outcome, fetal outcome, and fetal sex. I also calculated mean maternal age, infant birth length, and infant birth weight, as a whole and stratified by fetal sex and maternal skin color.

I also calculated the maternal mortality ratio (MMR), the mean stillbirth rate (SBR), and the sex ratio at birth (SRaB).

The (MMR) is:

$$MMR = \frac{MD}{LB} \times 10,000$$

where maternal deaths (MD) are divided by 10,000 live births (LB). In current studies, MD are divided by 100,000 live births, but I use 10,000 in the denominator, which is in line with historical studies given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].

The SBR (excluding spontaneous abortion but including intrapartum death defined here as stillbirth) for the hospital, is:

$$SBR = \frac{SB}{TB} \times 1,000$$

where the total number of stillbirths (SB) is divided by 1000 total births (TB).

The SRaB is :

$$SRaB = \frac{M}{F} \times 100$$

where the total number of live male births (M) is divided by 100 live female births (F) in a given period.

## Statistical analysis

I use Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight. I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is best practice for disentangling PTB from LBW [@cutland2017]. I discuss this constraint in the discussion below.

I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a]. Because most results were insignificant in these models, I only include unadjusted odds ratios (ORs) in the results. The full logistic models and results are included in the supplemental analysis.

I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race to see if there were differential outcomes for Black and mixed-race women. The independent variables include maternal age, maternal skin color, gestational status, and maternal nationality.

The simple linear regression model for modified ancestry is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.

The multiple linear regression model for modified ancestry is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.

The simple linear regression model for original color is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.

The multiple linear regression model for original color is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.

# Results

Table 1 provides summary statistics for the sample. In the hospital records, 58% were of African descent (defined as *preta* or *parda*) and 42% were of European descent (*branca*). First-time mothers ("nulliparous") were 41% of the sample. Of the total sample, 84% were Brazilian. Of the nearly 16% immigrant patients, 15% were from Europe.

Of all reproductive outcomes, excluding spontaneous abortion, 91% were natural deliveries. The remaining 9% were interventionist or operative deliveries. For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94.26%. For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2020]. This is a stark difference. Nonetheless, hospital rates probably reflected a registration effect, as medical personnel recorded all births and deaths, whereas reporting at the city level was less reliable [@loudon1992]. Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s [@roth2020]. Thus, the city’s rates were probably higher. However, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures before the rise of antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].

The SBR was 84.8%. We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% [@roth2020]. Again, registration effects could explain the higher hospital rates.

The sex ratio was 1.2: there were 120 male live births per 100 female live births. This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024]. Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was a female. Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016]. However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants. The skewed sex ratio deserves further study.

Results for all linear models are displayed in Tables 2 and 3. All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race). When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 83 grams lighter than their Euro-descended counterparts (Table 2). In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 96 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers.

In all models, women who were nulliparous (delivering their first child) had infants weighing on average 111 grams less than multiparous women, which is an association found both in other historical studies and today [@costa1998; @souza2020]. Age and nationality were not significant.

Unadjusted ORs (Table 4) demonstrate that women of African-descent had a higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019). When using the hospital records' original skin color categories, mixed-race women had a higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were insignificant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).

# Discussion

This study found that women of African descent in early twentieth-century Brazil had infants with lower birth weights than their White counterparts. When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today. Although the mean birthweight was still within today's normal range, the differences between racialized groups demonstrate differential health outcomes that have persisted until today.

Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight across the Americas and in New Zealand.

Black infants born at Laranjeiras had the lowest mean birth weight (3037 grams) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004]. Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent time period. Yet, they were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. White infants born at Laranjeiras had a mean birth weight of 3133 grams, slightly more than 50 grams lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. Yet they were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent time period (3423 grams). As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same time period [@costa1998; @costa2004; @goldin1989].

What can explain the differences in birth weight both historically and today? The intrauterine environment is a critical determinant of infant birth weight. Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care can all influence birth weight. Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother [@costa1998; @steckel1986]. A threshold of malnutrition must be met before there is an adverse effect on fetal growth. Other factors include maternal age, parity, birth spacing, and gestational age. We can imagine that poorer women performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992]. These same women might have lived in crowded housing arrangements with lack of access to running water [@graham1992; @fischer2008; @roth2020]. Generalized poverty might have influenced maternal nutrition, as well as the ability to access prenatal care.

Because the hospital records excluded information on maternal health factors, we don't know which contributed to racial disparities in infant birth weight. One possible cause that we do have some data for is syphilis. Like Costa found for early twentieth-century Baltimore, higher syphilis rates among African-descended women in Rio de Janeiro might have explained lower birth weights. Studies demonstrate that syphilis increases adverse birth outcomes including increased risk of LBW and PTB [@dasilva2024]. The clinical records that comprise this dataset did not consistently record syphilis rates. Scholars have shown how medical and public health professionals viewed it as an important cause of adverse health outcomes and even a possible "scourge" on the future of the Brazilian race [@carrara1996a]. Although I don't have accurate rates for Laranjeiras, city-wide rates from the 1930s can provide some context for the prevalence of infection among pregnant women. Of all prenatal syphilis exams performed between 1935 and 1938 at public clinics or hospitals across the city, not stratified by maternal skin color, 7.92% came back positive [@fontenelle, p. 275].

Discussions of epigenetic markers associated with the lived experience of racism might also be applicable here. Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors), mediated through epigenetic mechanisms. Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@leimert2020]. Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health [@meloni2022].

*Strengths and Limitations*
This study provides heretofore unpublished hospital records on one of Brazil's most important public maternity hospitals in the early twentieth century, a hospital that remains in operation. It provides quantitative evidence of racialized disparities in maternal-infant health in early twentieth-century Brazil. These findings are thus useful to compare to present-day research to assess the extent of progress in improved outcomes over time. They also complement descriptive historical studies and provide a base for qualitative research to analyze individual experience in the past.

All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014]. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age, SGA) and pre-term infants, or those born before 37 weeks' gestation. We are thus unable to know how many of these infants were low birth weight at term [@falcao2020]. This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989]. I assume that infants at risk for PTB were also at-risk for LBW.

*Conclusion*
Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study. Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.

{{< pagebreak >}}

```{r}
#| label: tbl-table1
#| tbl-cap: "Summary Statistics"
#| echo: FALSE
table1 <- readRDS(here::here("results", "tables", "table1_final.rds"))
flex_table <- as_flex_table(table1)
flex_table
```
{{< pagebreak >}}

```{r}
#| label: tbl-table2
#| tbl-cap: "Model 1 and 2 Results"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table3_final.rds"))
flex_table2 <- as_flex_table(table2)
flex_table2 <- flex_table2 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table2
```
{{< pagebreak >}}

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 3 and 4 Results"
#| echo: FALSE
table3 <- readRDS(here::here("results", "tables", "table4_final.rds"))
flex_table3 <- as_flex_table(table3)
flex_table3 <- flex_table3 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table3
```
{{< pagebreak >}}

```{r}
#| label: tbl-table4
#| tbl-cap: "Unadjusted Odds Ratios"
#| echo: FALSE
table4 <- readRDS(here::here("results", "tables", "table5_final.rds"))
table4
```
{{< pagebreak >}}

```{r}
#| label: tbl-table5
#| tbl-cap: "Mean Birth Weights"
#| echo: FALSE
table5 <- readRDS(here::here("results", "tables", "table2_final.rds"))
table5
```
{{< pagebreak >}}

### 2024-07-24 products/manuscript/supplement/Supplemental_Materials.qmd
Final Draft Updates
After meeting with Swartzendruber

@@
 There are three continuous variables: maternal age, infant birth weight, and infant length.
-There are nine categorical variables: skin color; ancestry; parity or gravidity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.
-
-In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births.
-For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births [@nyarko2013; @wehby2015a; @wehby2017a].
-
-A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
-The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009].
-But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
-Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
-
-### Maternal Skin Color
-
-The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*).
-I maintained this categorization.
-However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category.
-Please see the `eda-v1.qmd` file for more information on the variable manipulation process.
-
-### Gestational Status
-
-Gestational status was divided into parity and gravidity.
-Originally, the data had eight categories: nullipara, primipara, secundipara, multipara; and primigravida, secundigravida, trigravida, and multigravida.
-I combined categories into four final categories: 1) nullipara and primigravida into `nullipara` (or any woman who has never given birth to a viable fetus); 2) primipara into `primipara` (or any woman who has given birth to one viable fetus); 3) secundipara and multipara into `multipara` (or any woman who has given birth to more than one viable fetus); and 4) secundigravida, trigravida, and multigravida into `multigravida` (or any woman who has been pregnant more than one but never given birth to a viable fetus).
-
-### Maternal Nationality
-
-For maternal nationality, I created categories based on individual country (`nationality`) and categories based on region (`modifiednationality`).
-Recategorization occurred as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian).
-
-### Clinical Outcomes
-
-The clinical notes upon which I based this dataset only sporadically included infant demographic information for spontaneous abortions or stillbirths.
-For spontaneous abortions, stillbirths, and live births, I recorded, when available, infant sex, weight, length.
-I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy.
-However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraud1997), coded as operatory or natural by physicians, as interventionist.
-I classified clinical outcomes as spontaneous abortion, natural (normal delivery), interventionist (interventionist delivery), and operative (surgical delivery).
+There are nine categorical variables: skin color; ancestry; parity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.
 
@@
 ```
+
 In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women.

### 2024-07-24 code/analysis-code/fullanalysis.R

@@ -76,6 +76,7 @@ metrics_fit1
 # Create a gtsummary table directly from the model
 table_model1 <-
   tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
   add_glance_table(include = c(r.squared, nobs))
 table_model1
@@ -262,6 +263,7 @@ options(scipen = 999) # set opens to avoid scientific notation
 metrics_fit2
 # Create a gtsummary table directly from the model
 table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
   add_glance_table(include = c(r.squared, nobs))
 table_model2

### 2024-07-24 code/analysis-code/introanalysis.R

@@ -187,9 +187,9 @@ ML_linear <- ML_linear %>%
   mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))
 # Create recipe
 # Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
-log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) %>%
-  step_dummy(all_nominal(), -all_outcomes()) %>%
-  step_normalize(all_predictors())
+log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
+  #step_dummy(all_nominal(), -all_outcomes()) %>%
+  #step_normalize(all_predictors())
 # Define model specification
 log1_spec <- logistic_reg() %>%
   set_engine("glm") %>%
@@ -204,7 +204,8 @@ log1_fit <- log1_wf %>%
 # Extract fitted model from workflow
 log1_fitted <- extract_fit_parsnip(log1_fit)$fit
 # Create a gtsummary table directly from the model
-table_log1 <- tbl_regression(log1_fitted, exponentiate = TRUE) # use exponentiate true for ORs
+table_log1 <- tbl_regression(log1_fitted, exponentiate = TRUE) %>% # use exponentiate true for ORs
+  bold_p(t = 0.05)
 table_log1
 # Save regression output into table
 saveRDS(table_log1, file = here::here("results", "tables", "table_log1.rds"))
@@ -240,7 +241,7 @@ roc_plot1 <- ggplot(roc_curve_data1, aes(x = 1 - specificity, y = sensitivity)) 
 print(roc_plot1)
 
 ## ---- log2 --------
-# Simple logistic regression model with ModifiedColor as the predictor
+# Simple logistic regression model with Color as the predictor
 # Check factor levels
 levels(ML_linear$BirthweightCategory) # Reference is LBW
 # Set the reference level explicitly to NBW
@@ -248,9 +249,9 @@ ML_linear <- ML_linear %>%
   mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))
 # Create recipe
 # Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
-log2_recipe <- recipe(BirthweightCategory ~ Color, data = ML_linear) %>%
-  step_dummy(all_nominal(), -all_outcomes()) %>%
-  step_normalize(all_predictors())
[…]

### 2024-07-24 code/eda-code/edacode.R

@@ -128,6 +128,9 @@ ML_summary <- ML_summary %>%
   )
 # Save new summary data with recoded factor variables
 saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
+#Summarize missing data using dplyr
+missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
+missing_data_summary
 
 ## ---- Summarystats --------
 # Create gtsummary table full table for supplemental analysis
@@ -157,20 +160,22 @@ table1 <-
     include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale),
     missing = "no",
     label = list(
-      ModifiedColor ~ "Ancestry", 
-      ModifiedStatus ~ "Parity", 
-      Age ~ "Maternal Age", 
-      ModifiedNationality ~ "Combined Nationality", 
-      Birth ~ "Birth Outcome", 
-      MaternalOutcome ~ "Maternal Outcome", 
-      FetalOutcome ~ "Fetal Outcome", 
-      Lengthcentimeters ~ "Birth Length (cms)", 
-      Weightgrams ~ "Birth Weight (grms)",
-      WeightgramsBlack ~ "Black (grms)",
-      WeightgramsMixed ~ "Mixed-Race (grms)",
-      WeightgramsWhite ~ "White (grms)",
-      WeightgramsFemale ~ "Female (grms)",
-      WeightgramsMale ~ "Male (grms)"
+      Color ~ "Color (n = 2,695)",
+      ModifiedColor ~ "Ancestry (n = 2,695)", 
+      ModifiedStatus ~ "Parity (n = 2,836)", 
+      Age ~ "Maternal Age (n = 2,783)", 
+      ModifiedNationality ~ "Combined Nationality (n = 2,773)", 
+      Birth ~ "Birth Outcome (n = 2,761)", 
+      MaternalOutcome ~ "Maternal Outcome (n = 2,829)", 
+      FetalOutcome ~ "Fetal Outcome (n = 2,666)", 
+      Sex ~ "Sex (n = 2,534)",
+      Lengthcentimeters ~ "Birth Length (cm) (n = 2,405)", 
[…]

### 2024-07-24 code/eda-code/eda-v1.qmd
Final draft updates
LBW/NBW

@@
 
+# Create dummy variables for birth weight category by maternal skin color for summary table
+ML_summary <- ML_summary %>%
+  mutate(
+   Afro_LBW = ifelse(BirthweightCategory == "LBW" & (!is.na(WeightgramsBlack) | !is.na(WeightgramsMixed)), 1, 0),
+   Afro_NBW = ifelse(BirthweightCategory == "NBW" & (!is.na(WeightgramsBlack) | !is.na(WeightgramsMixed)), 1, 0),
+   Euro_LBW = ifelse(BirthweightCategory == "LBW" & !is.na(WeightgramsWhite), 1, 0),
+   Euro_NBW = ifelse(BirthweightCategory == "NBW" & !is.na(WeightgramsWhite), 1, 0),
+ )
+
+# Change value to missing for opposite skin color infants for summary statistics
+ML_summary <- ML_summary %>%
+  mutate(
+    Afro_LBW = case_when(
+      ModifiedColor == "EuroDescent" ~ NA_real_,
+      TRUE ~ Afro_LBW
+    ),
+    Euro_LBW = case_when(
+      ModifiedColor == "AfroDescent" ~ NA_real_,
+      TRUE ~ Euro_LBW
+    )
+  )
+
 # Verify new variables
-str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale")])
+str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale", "Afro_LBW", "Euro_LBW")])
 
@@
   tbl_summary(
-    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale),
+    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale, BirthweightCategory, Afro_LBW, Euro_LBW),
     missing = "no",
@@
       WeightgramsFemale ~ "Female (g) (n = 1,074)",
-      WeightgramsMale ~ "Male (g) (n = 1,270)"
+      WeightgramsMale ~ "Male (g) (n = 1,270)",
+      BirthweightCategory ~ "Birth Weight Category (n = 2,231)",
+      Afro_LBW ~ "LBW Afro-Descent",
+      Euro_LBW ~ "LBW Euro-Descent"
     ),
@@
   )
-
 table1
@@
 
-Now we will create the summary table for mean birthweights across time and place.
+Now we will create the summary table for mean birth weights across time and place.
 

### 2024-07-24 code/processing-code/processingfile-v1.qmd
Final draft updates
LBW/NBW

@@
 ML_4 <- ML_3 %>%
-  mutate(BirthweightCategory = if_else(Weightgrams <= 2500, "LBW", "NBW"))
+  mutate(BirthweightCategory = case_when(
+    Weightgrams >= 500 & FetalOutcome != "morto" & Gestation != "twin" & Weightgrams <= 2500 ~ "LBW",
+    Weightgrams >= 500 & FetalOutcome != "morto" & Gestation != "twin" & Weightgrams > 2500 ~ "NBW",
+    TRUE ~ NA_character_
+  ))
 ML_4

### 2024-07-24 products/manuscript/Manuscript_Final_Lancet.qmd
Final draft updates
LBW/NBW

@@
 ---
-title: "Low Infant Birth Weight in Brazil"
+title: "Infant Birth Weight in Brazil"
 subtitle: "A Cross-Sectional Historical Approach"
@@
 ## Background
-In 1888, after nearly four centuries, Brazil became the last country in the Western Hemisphere to abolish slavery.
+In 1888, after nearly four centuries of its existence, Brazil became the last country in the Western Hemisphere to abolish slavery.
 After abolition, the state did not create programs to incorporate the formerly enslaved and their descendants into economic, civil, or political life.
@@
 ## Interpretations
-This study hypothesizes that these differences were related to the historical legacies of slavery, including race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments -- and possible epigenetic effects from the lived experience of racism.
+This study hypothesizes that these differences were related to the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments -- and possible epigenetic effects from the lived experience of racism.
 The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.
@@
 Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
-In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research, as low birth weight is associated with higher rates of neonatal and infant mortality and possible health problems over the life course [@cardoso2005; @paixao2021; @vilanova2019].
+In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research, as LBW is associated with higher rates of neonatal and infant mortality and possible health problems over the life course [@cardoso2005; @paixao2021; @vilanova2019].
 
@@
 
+The overall rate of LBW for infants born to Afro-descended mothers was 9.5% compared to 7% for Euro-descended mothers (Table 1).
 Results for all linear models are displayed in Tables 2 and 3.
@@
 
-This study found that women of African descent in early twentieth-century Brazil had infants with lower birth weights than their White counterparts.
+This study found that women of African descent in early twentieth-century Brazil both had infants with lower birth weights and higher rates of LBW than their White counterparts.
 When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today.

### 2024-07-24 code/eda-code/edacode.R

@@ -108,8 +108,28 @@ ML_summary <- ML_summary %>%
     WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
     WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
     WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_))
+# Create dummy variables for birth weight category by maternal skin color for summary table
+ML_summary <- ML_summary %>%
+  mutate(
+    Afro_LBW = ifelse(BirthweightCategory == "LBW" & (!is.na(WeightgramsBlack) | !is.na(WeightgramsMixed)), 1, 0),
+    Afro_NBW = ifelse(BirthweightCategory == "NBW" & (!is.na(WeightgramsBlack) | !is.na(WeightgramsMixed)), 1, 0),
+    Euro_LBW = ifelse(BirthweightCategory == "LBW" & !is.na(WeightgramsWhite), 1, 0),
+    Euro_NBW = ifelse(BirthweightCategory == "NBW" & !is.na(WeightgramsWhite), 1, 0),
+  )
+# Change value to missing for opposite skin color infants for summary statistics
+ML_summary <- ML_summary %>%
+  mutate(
+    Afro_LBW = case_when(
+      ModifiedColor == "EuroDescent" ~ NA_real_,
+      TRUE ~ Afro_LBW
+    ),
+    Euro_LBW = case_when(
+      ModifiedColor == "AfroDescent" ~ NA_real_,
+      TRUE ~ Euro_LBW
+    )
+  )
 # Verify new variables
-str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale")])
+str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale", "Afro_LBW", "Afro_NBW", "Euro_LBW", "Euro_NBW")])
 # View the updated dataframe
 print(head(ML_summary))
 
@@ -157,7 +177,7 @@ table1 <-
[…]

### 2024-07-24 code/processing-code/processingcode.R

@@ -68,7 +68,11 @@ missing_data_summary <- ML_3 %>% dplyr::summarise_all(list(~sum(is.na(.))))
 missing_data_summary
 # Create new variable of NBW and LBW for future logistic regression
 ML_4 <- ML_3 %>%
-  mutate(BirthweightCategory = if_else(Weightgrams <= 2500, "LBW", "NBW"))
+  mutate(BirthweightCategory = case_when(
+    Weightgrams >= 500 & FetalOutcome != "morto" & Gestation != "twin" & Weightgrams <= 2500 ~ "LBW",
+    Weightgrams >= 500 & FetalOutcome != "morto" & Gestation != "twin" & Weightgrams > 2500 ~ "NBW",
+    TRUE ~ NA_character_
+  ))
 ML_4
 # Transform character variables into factors
 ML_5 <- ML_4 %>%

### 2024-07-25 code/eda-code/eda-v1.qmd
Final addition
Added new LBW/NBW stats

--- 2024-07-24 code/eda-code/eda-v1.qmd
@@
 
+# Create dummy variables for birth weight category by maternal skin color for summary table
+ML_summary <- ML_summary %>%
+  mutate(
+   Afro_LBW = ifelse(BirthweightCategory == "LBW" & (!is.na(WeightgramsBlack) | !is.na(WeightgramsMixed)), 1, 0),
+   Afro_NBW = ifelse(BirthweightCategory == "NBW" & (!is.na(WeightgramsBlack) | !is.na(WeightgramsMixed)), 1, 0),
+   Euro_LBW = ifelse(BirthweightCategory == "LBW" & !is.na(WeightgramsWhite), 1, 0),
+   Euro_NBW = ifelse(BirthweightCategory == "NBW" & !is.na(WeightgramsWhite), 1, 0),
+ )
+
+# Change value to missing for opposite skin color infants for summary statistics.
+This uses the wrong denominator (all mothers regardless of whether we have their infants' birth weights, so I will use code below.)
+#ML_summary <- ML_summary %>%
+  #mutate(Afro_LBW = case_when(ModifiedColor == "EuroDescent" ~ NA_real_, TRUE ~ Afro_LBW),
+    #Euro_LBW = case_when(ModifiedColor == "AfroDescent" ~ NA_real_,TRUE ~ Euro_LBW))
+
 # Verify new variables
@@
 print(head(ML_summary))
+
+# See LBW and NBW with true denominator
+LBW_percentage <- ML_summary %>%
+# Remove NAs from both variables
+filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
+# Group by ModifiedColor
+group_by(ModifiedColor) %>%
+# Calculate the percentage for each group
+summarise(
+  total_count = n(),
+  lbw_count = sum(BirthweightCategory == "LBW"),
+  lbw_percentage = (lbw_count / total_count) * 100
+  )
+
+# Print the result
+print(LBW_percentage)
 ```
@@
   tbl_summary(
-    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale),
+    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale, BirthweightCategory),
     missing = "no",
@@
       WeightgramsFemale ~ "Female (g) (n = 1,074)",
-      WeightgramsMale ~ "Male (g) (n = 1,270)"
+      WeightgramsMale ~ "Male (g) (n = 1,270)",
+      BirthweightCategory ~ "Birth Weight Category (n = 2,231)"
     ),

### 2024-07-25 products/manuscript/Manuscript_Final_Lancet.qmd
Final addition
Added new LBW/NBW stats

@@
 
-The overall rate of LBW for infants born to Afro-descended mothers was 9.5% compared to 7% for Euro-descended mothers (Table 1).
+The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers.
 Results for all linear models are displayed in Tables 2 and 3.
@@
 In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 96 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers.
+Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent.
 

### 2024-07-25 code/eda-code/edacode.R

@@ -116,22 +116,28 @@ ML_summary <- ML_summary %>%
     Euro_LBW = ifelse(BirthweightCategory == "LBW" & !is.na(WeightgramsWhite), 1, 0),
     Euro_NBW = ifelse(BirthweightCategory == "NBW" & !is.na(WeightgramsWhite), 1, 0),
   )
-# Change value to missing for opposite skin color infants for summary statistics
-ML_summary <- ML_summary %>%
-  mutate(
-    Afro_LBW = case_when(
-      ModifiedColor == "EuroDescent" ~ NA_real_,
-      TRUE ~ Afro_LBW
-    ),
-    Euro_LBW = case_when(
-      ModifiedColor == "AfroDescent" ~ NA_real_,
-      TRUE ~ Euro_LBW
-    )
-  )
+# Change value to missing for opposite skin color infants for summary statistics. This uses the wrong denominator (all mothers regardless of whether we have their infants' birth weights, so I will use code below.)
+#ML_summary <- ML_summary %>%
+#mutate(Afro_LBW = case_when(ModifiedColor == "EuroDescent" ~ NA_real_, TRUE ~ Afro_LBW),
+#Euro_LBW = case_when(ModifiedColor == "AfroDescent" ~ NA_real_,TRUE ~ Euro_LBW))
 # Verify new variables
-str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale", "Afro_LBW", "Afro_NBW", "Euro_LBW", "Euro_NBW")])
+str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale")])
 # View the updated dataframe
 print(head(ML_summary))
+# See LBW and NBW with true denominator
+LBW_percentage <- ML_summary %>%
+  # Remove NAs from both variables
+  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
+  # Group by ModifiedColor
+  group_by(ModifiedColor) %>%
+  # Calculate the percentage for each group
+  summarise(
+    total_count = n(),
[…]

### 2024-08-04 products/manuscript/Manuscript_Final_Lancet.qmd
Final Draft
Final revisions

@@
 In 1888, after nearly four centuries of its existence, Brazil became the last country in the Western Hemisphere to abolish slavery.
-After abolition, the state did not create programs to incorporate the formerly enslaved and their descendants into economic, civil, or political life.
+After abolition, the state did not create programs to incorporate people formerly enslaved and their descendants into economic, civil, or political life.
 Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century.
 Contemporary public health studies have shown that health disparities continue to mirror patterns from over a century ago.
-The goal of this study is to quantify health disparities in the post-abolition period in the realm of maternal-infant health by analyzing infant birth weight in relation to maternal variables in one of Brazil's first public maternity hospitals.
+The goal of this study is to quantify health disparities in infant birth weights by maternal characteristics in the post-abolition period in one of Brazil's first public maternity hospitals.
 
 ## Methods
-This cross-sectional analysis uses hospital-level data collected between 1922-1926 from the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil.
-It was the city's first public maternity hospital.
-I use linear models to assess disparities in infant birth weight in relation to maternal skin color, age, parity, and nationality.
+This cross-sectional analysis uses hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital.
+I used linear models to assess differences in infant birth weight by maternal skin color, age, parity, and nationality.
 
 ## Findings
-Infants born to women of color (either Black, "*preta,*" or mixed-race, "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant.
+Infants born to Afro-descendant women had birth weights estimated to be 83 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant.
+Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 96 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.
 
 ## Interpretations
-This study hypothesizes that these differences were related to the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments -- and possible epigenetic effects from the lived experience of racism.
+The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments -- and possible epigenetic effects from the lived experience of racism.
 The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.
@@
 Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @martins2024; @steckel1986].
-In Brazil, most historical studies of health under slavery and in its aftermath remain descriptive in nature and have not tested any quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2020; @telles2022].
+In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2020; @telles2022].
 
-Brazil imported over four million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009].
+Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country, compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
 It was also the last country to abolish slavery (1888) in the Western Hemisphere.
@@
 Although some eugenicists called for sterilization of "unfit" citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013].
-Whiteness still remained a cultural and social "goal" even if the state did not engage in coercive measures of reproductive control.
+Whiteness remained a cultural and social "goal" even if the state did not engage in coercive measures of reproductive control.
 
@@
 Laranjeiras provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
-The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants.
+Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants.
 Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
@@
 
-This analysis draws on an unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926.
+This analysis draws on a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926.
 I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO).
@@
 
-The project's Github repository, in particular, the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting.
+The project's Github repository, particularly the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting.
 All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online [@roth2024].
@@
 The outcome was birth weight measured as a continuous variable in grams.
-In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW, <p>&ge;</p>2500 g or low birth weight, LBW, <2500 g, according to current WHO classifications) [@worldhealthorganization2022].
+In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW (<p>&ge;</p>2500 g) or LBW (<2500 g), according to current WHO classifications) [@worldhealthorganization2022].
 
@@
 *Delivery type*.
-I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy.
+I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention using forceps; and operatory, indicating a cesarean section or embryotomy.
 However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries), coded as operatory or natural by hospital records, as interventionist.
@@
 In the summary statistics, I included both spontaneous abortions and all live births.
-For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data [@nyarko2013; @wehby2015a; @wehby2017a].
+For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data, which is standard methodology for looking at infant birth weight in a contemperaneous context [@nyarko2013; @wehby2015a; @wehby2017a].
 
@@
 
-The SRaB is :
+The SRaB is:
 
@@
 
-I use Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight.
+I used Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight.
 I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
-Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is best practice for disentangling PTB from LBW [@cutland2017].
+Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age, as is best practice for disentangling PTB from LBW [@cutland2017].
 I discuss this constraint in the discussion below.
@@
 I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
-Because most results were insignificant in these models, I only include unadjusted odds ratios (ORs) in the results.
+Because most results were not statistically significant in these models, no unadjusted models were conducted.
 The full logistic models and results are included in the supplemental analysis.
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
 
@@
 For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94.26%.
-For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2020].
-This is a stark difference.
-Nonetheless, hospital rates probably reflected a registration effect, as medical personnel recorded all births and deaths, whereas reporting at the city level was less reliable [@loudon1992].
-Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s [@roth2020].
-Thus, the city’s rates were probably higher.
-However, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures before the rise of antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
-
 The SBR was 84.8%.
-We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% [@roth2020].
-Again, registration effects could explain the higher hospital rates.
-
 The sex ratio was 1.2: there were 120 male live births per 100 female live births.
 This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
-Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was a female.
-Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
-However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
-The skewed sex ratio deserves further study.
 
@@
 
-Unadjusted ORs (Table 4) demonstrate that women of African-descent had a higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019).
-When using the hospital records' original skin color categories, mixed-race women had a higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were insignificant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
+Unadjusted ORs (Table 4) demonstrate that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019).
+When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were insignificant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
 
 # Discussion
+
+In general, maternal and infant outcomes were not better in the Maternidade Laranjeiras.
+For example, between 1922 and 1926, the clinic's MMR was 94.26%.
+For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2020].
+This is a stark difference.
+Nonetheless, hospital rates probably reflected a registration effect, as medical personnel recorded all births and deaths, whereas reporting at the city level was less reliable [@loudon1992].
+Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s [@roth2020].
+Thus, the city’s rates were probably higher.
+The same pattern holds true for the SBR, which was 84.8% for Laranjeiras and 73.68% for the city of Rio de Janeiro for the same period [@roth2020].
+Again, registration effects could explain the higher hospital rates.
+However, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures before the rise of antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
+
+Laranjeiras also had a high sex ratio (1.2).
+Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was a female.
+Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
+However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
+The skewed sex ratio deserves further study.
 
@@
 Black infants born at Laranjeiras had the lowest mean birth weight (3037 grams) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004].
-Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent time period.
+Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period.
 Yet, they were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
 White infants born at Laranjeiras had a mean birth weight of 3133 grams, slightly more than 50 grams lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
-Yet they were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent time period (3423 grams).
-As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same time period [@costa1998; @costa2004; @goldin1989].
+Yet they were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 grams).
+As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same historical period [@costa1998; @costa2004; @goldin1989].
 

### 2024-08-12 code/eda-code/eda-v1.qmd
Revisions for submission
Revisions 1

@@
     MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
-    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death"),
+    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth"),
     Status = recode_factor(Status, "Multigesta" = "Multigravida", "Trigesta" = "Trigravida", "Secundigesta" = "Secundigravida", "Secundipara" = "Secundiparous", "Multipara" = "Multiparous", "Primigesta" = "Primigravida", "Nulipara" = "Nulliparous", "Primipara" = "Primiparous")
@@
 maternal_deaths <-23
-live_births <- 2440
+live_births <- 2445
 MMR <- (maternal_deaths / live_births) * 10000
@@
 #Stillbirth rate SBR
-still_births <- 226
+still_births <- 221
 total_births <- 2666
@@
 sex_ratio <- ML_summary %>%
-  filter(!Birth %in% c("Abortion") & !FetalOutcome %in% c("Stillbirth or Neonatal Death")) %>% # Exclude rows with abortion and stillbirth
+  filter(!Birth %in% c("Abortion") & !FetalOutcome %in% c("Stillbirth")) %>% # Exclude rows with abortion and stillbirth
  filter(!is.na(Sex) & !is.na(Birth) & !is.na(FetalOutcome)) %>%
@@
 ```
-The MMR is 94.3% in this specific clinic.
-The SBR is 84.8% in this specific clinic.
+The MMR is 94.1% in this specific clinic.
+The SBR is 82.9% in this specific clinic.
 The sex ratio is 1.2 or there were 120 male live births per 100 female live births at this specific clinic.
@@
     MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
-    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
+    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth")
     )

### 2024-08-12 code/processing-code/processingfile-v1.qmd
Revisions for submission
Revisions 1

--- 2024-07-05 code/processing-code/processingfile-v1.qmd
@@
 These variables include qualitative information that can be excluded from our processed data, since we are not looking at qualitative notes on cases.
+Nonetheless, for `Fetal_Notes`, we need to keep the Fetal_Notes variable at the moment as it qualifies the `FetalOutcome` variable.
+
 We also don't need the citation information to run regressions.
-So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Maternal_Notes`, `Fetal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.
+So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Maternal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.
 Further, we will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).
@@
 # Remove unnecessary variables
-ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
+ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Maternal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
 ```
@@
 
-We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations.
+We can see that `CivilStatus` has 2834 missing observations, `GestationalAge_Months` has 2789 missing observations, `Fetal_Notes` has 2823 missing observations, and `Birth_Notes` also has 2330 missing values.
 This is a lot.
-Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.
-`Birth_Notes` also has 2330 missing values, but first I need to extract information on gestation (twin versus singleton birth) from this variable before I exclude it.
+Although civil status can serve as a proxy for income and class, I have too many missing values here for the variable to be meaningful, so I will exclude `CivilStatus`.
+Gestational age in months is also a valuable insight in relation to LBW and PTB, but my data set has too many missing values to be of use here, we can also exclude `GestationalAge_Months`.
+Some gestational age entries were also in relation to spontaneous abortions.
+`Birth_Notes` helped me initially decided on how to code the `Birth` variable, which was done when I engaged in manual data entry, so I will exclude here.
+
+Before I exclude any variables, I need to make sure that the original hospital records did not misrecord stillbirths as neonatal deaths and vice versa before I exclude them.
+First, we will look at the values for `FetalOutcome` and `Fetal_Notes` that are not missing any values for both variables.
+This gives us 22 entries.
+Most values in the `Fetal_Notes` variable refer to either a macerated (*macerado*) stillbirth, perhaps implying maternal syphilis or a neonatal death (*faleceu*) with a given time frame.
+We can see that the `FetalOutcome` variable is not always consistent with the `Fetal_Notes` variable.
+For example, a live birth and subsequent neonatal death is marked as a stillbirth in the `FetalOutcome` variable.
+I will mark all live births as live, even if death occurred immediately after birth and I will mark all stillbirths as stillbirths.
+There were 20 recorded neonatal deaths.
+
+```{r}
+# See values for `FetalOutcome` and `Fetal_Notes` that are not missing any values for both variables
+ML_1 %>% filter(!is.na(FetalOutcome) & !is.na(Fetal_Notes))
+
+# Change 'FetalOutcome' to vivo if 'Fetal_Notes' indicates neonatal death (via Faleceu or faleceu in 'Fetal_Notes' column)
+ML_2 <- ML_1 %>%
+  mutate(FetalOutcome = case_when(
+    FetalOutcome == "morto" & 
+      str_detect(tolower(Fetal_Notes), "faleceu") ~ "vivo",
+    TRUE ~ FetalOutcome
+  ))
+ML_2
+
+# Check values
+ML_2 %>% filter(!is.na(FetalOutcome) & !is.na(Fetal_Notes))
+
+# See how many neonatal deaths were recorded
+ML_neonatal <- ML_2 %>%
+  filter(
+    FetalOutcome == "vivo" &
+    str_detect(tolower(Fetal_Notes), "faleceu")
+  )
+
+# View the filtered data
+print(ML_neonatal)
+```
+
+I also need to extract information on gestation (twin versus singleton birth) from this `GestationalAge_Months` before I exclude it.
+My data set will only include singleton births for the statistical analysis.
 
@@
 # Extract information on gestation from Birth_Notes and create new variable gestation with singleton and twin births
-ML_2 <- ML_1 %>%
-  mutate(Birth_Notes = str_trim(Birth_Notes)) %>% # removes any leading or trailing whitespace from Birth_Notes column
+ML_3 <- ML_2 %>%
+  mutate(Birth_Notes = str_trim(Birth_Notes)) %>% # removes any leading or trailing white space from Birth_Notes column
   mutate(Gestation = if_else(is.na(Birth_Notes) | Birth_Notes == "", "singleton", # checks if Birth_Notes is NA or empty strings and assigns singleton to Gestation column
                              if_else(str_detect(tolower(Birth_Notes), "gemelar"), "twin", "singleton"))) # checks for gemelar in case-insensitive manner and assigns twin accordingly or singleton otherwise
-ML_2
+ML_3
 
 # Remove variables with missing data not relevant to hypothesis
-ML_3 <- ML_2 %>% dplyr::select(-CivilStatus, -GestationalAge_Months, -Birth_Notes)
+ML_4 <- ML_3 %>% dplyr::select(-CivilStatus, -GestationalAge_Months, -Birth_Notes, -Fetal_Notes)
 ```
@@
 #Summarize missing data using dplyr
-missing_data_summary <- ML_3 %>% dplyr::summarise_all(list(~sum(is.na(.))))
+missing_data_summary <- ML_4 %>% dplyr::summarise_all(list(~sum(is.na(.))))
 missing_data_summary
@@
 # Create new variable of NBW and LBW for future logistic regression
-ML_4 <- ML_3 %>%
-  mutate(BirthweightCategory = if_else(Weightgrams <= 2500, "LBW", "NBW"))
-ML_4
+ML_5 <- ML_4 %>%
+  mutate(BirthweightCategory = case_when(
+    Weightgrams >= 500 & FetalOutcome != "morto" & Gestation != "twin" & Weightgrams <= 2500 ~ "LBW",
+    Weightgrams >= 500 & FetalOutcome != "morto" & Gestation != "twin" & Weightgrams > 2500 ~ "NBW",
+    TRUE ~ NA_character_
+  ))
+ML_5
 ```
@@
 # Transform character variables into factors
-ML_5 <- ML_4 %>%
+ML_6 <- ML_5 %>%
   mutate(across(c(Color, Status, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex, Gestation, BirthweightCategory), as.factor))
@@
 # Check variables
-glimpse(ML_5)
+glimpse(ML_6)
 
 # Check levels of factor variables
-map(ML_5[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex', 'Gestation', 'BirthweightCategory')], levels)
+map(ML_6[c('Color', 'Status', 'Nationality', 'Birth', 'MaternalOutcome', 'FetalOutcome', 'Sex', 'Gestation', 'BirthweightCategory')], levels)
 
@@
 # Check frequencies of factor variables
-frequency_table <- ML_5 %>%
+frequency_table <- ML_6 %>%
   count(across(all_of(factor_variables)))
@@
 # Create dataset for summary statistics (includes missing values)
-ML_summary <- ML_5
+ML_summary <- ML_6
 
 # Creating dataset for linear regression
-ML_linear <- ML_5 %>% tidyr::drop_na() %>%
+ML_linear <- ML_6 %>% tidyr::drop_na() %>%
   filter(Weightgrams >= 500, FetalOutcome != "morto", Gestation != "twin") # Drop observations of < 500 grams, all non-live births, and all twin births for regression dataset

### 2024-08-12 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission
Revisions 1

@@
 ## Background
-In 1888, after nearly four centuries of its existence, Brazil became the last country in the Western Hemisphere to abolish slavery.
+In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence.
 After abolition, the state did not create programs to incorporate people formerly enslaved and their descendants into economic, civil, or political life.
 Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century.
-Contemporary public health studies have shown that health disparities continue to mirror patterns from over a century ago.
+Public health studies have shown that health disparities continue to mirror patterns from over a century ago.
 The goal of this study is to quantify health disparities in infant birth weights by maternal characteristics in the post-abolition period in one of Brazil's first public maternity hospitals.
@@
 ## Findings
-Infants born to Afro-descendant women had birth weights estimated to be 83 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant.
+Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant.
 Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 96 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.
@@
 ## Interpretations
-The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments -- and possible epigenetic effects from the lived experience of racism.
+The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for women of color -- and possible epigenetic effects from the lived experience of racism.
 The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.
@@
 
-Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @martins2024; @steckel1986].
+Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@jasienska2009; @owens2019; @martins2024; @steckel1986].
 In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2020; @telles2022].
@@
 Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country, compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
-It was also the last country to abolish slavery (1888) in the Western Hemisphere.
-In the nineteenth century, population health was poor for all people, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from severe nutritional deficiencies [@karasch1987; @kiple1989].
+Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission or the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
+
+Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere.
+In the nineteenth century, population-level health was poor for all Brazilians, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from nutritional deficiencies [@karasch1987; @kiple1989].
 
@@
 The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe.
-Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986; @benchimol1999; @hochman2013].
+Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@benchimol1999; @hochman2013].
 Although infant mortality rates declined from the nineteenth century, the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@roth2020].
@@
 Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
-There, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine [@martins2004; @rohden2009; @roth2021].
+There, obstetricians and gynecologists began implementing surgical advances in labor and delivery [@martins2004; @roth2021].
 Underlying these structural and technological changes was an ideology scholars have described as "scientific motherhood," supported by both physicians and elite women, which relied on a technocratic model towards pregnancy and motherhood and, in doing so, essentialized notions of gender and foregrounded physicians’ scientific authority [@freire2009; @martins2004; @otovo2016].
@@
 At the same time, longstanding racial hierarchies forged under racial slavery morphed into "scientific" ways to categorize humans based on skin color in post-abolition contexts.
-In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than coercively restricting reproduction through sterilization [@schwarcz1993; @stepan1991].
-Although some eugenicists called for sterilization of "unfit" citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013].
-Whiteness remained a cultural and social "goal" even if the state did not engage in coercive measures of reproductive control.
+In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than restricting reproduction through sterilization [@schwarcz1993; @stepan1991].
+Although some eugenicists called for sterilization of citizens they deemed unfit, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013].
+Whiteness remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.
 
@@
 In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon.
-In our current postgenomic era, in which "race" as a biological category has been thoroughly debunked, the possibility for centuries-old patterns of hierarchical scientific thinking remains very real [@roberts2011].
+In our current postgenomic era, in which race as a biological category has been rejected, the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking remains very real [@roberts2011].
 Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022].
@@
 
-This analysis draws on a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926.
+This analysis draws on a unique sample of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926.
 I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO).
@@
 
-The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca de Biomedicina-A, Universidade Estadual do Rio de Janeiro (BBA-UERJ), all in Rio de Janeiro, Brazil.
+The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.
 Between January 2012 and July 2013, I manually digitized the journal by photographing each volume.
@@
 *Birth weight*.
-The outcome was birth weight measured as a continuous variable in grams.
+The outcome was birth weight measured as a continuous variable in grams for live births only.
 In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW (<p>&ge;</p>2500 g) or LBW (<2500 g), according to current WHO classifications) [@worldhealthorganization2022].
@@
 
-*Delivery type*.
-I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention using forceps; and operatory, indicating a cesarean section or embryotomy.
+*Birth*.
+I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention and a vaginal delivery; interventionist, indicating medium medical intervention using forceps and a vaginal delivery; and operatory, indicating a cesarean section or embryotomy.
 However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries), coded as operatory or natural by hospital records, as interventionist.
-I classified spontaneous abortion as a separate category.
+I classified spontaneous abortion as a separate category within this variable.
 
@@
 
-*Maternal outcome*. Hospital records recorded maternal outcome was recorded as death, discharge, or hospital transferal, which I maintain here.
+*Maternal outcome*.
+Hospital records recorded maternal outcome was recorded as death, discharge, or hospital transferal, which I maintain here.
+I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality below.
 
 *Fetal outcome*.
-In the summary statistics, I included both spontaneous abortions and all live births.
-For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data, which is standard methodology for looking at infant birth weight in a contemperaneous context [@nyarko2013; @wehby2015a; @wehby2017a].
+I recorded fetal outcome as live birth or stillbirth.
+For the statistical models, I excluded both spontaneous abortions (recorded in the birth variable) and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data, which is standard methodology for looking at infant birth weight today [@nyarko2013; @wehby2015a; @wehby2017a].
+The clinic records recorded 20 neonatal deaths, which I included in the live birth category.
+These numbers could be higher, as the clinic only recorded deaths within the first week of life, probably because those infants died in the hospital rather than at home.
 
@@
 I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
-Because most results were not statistically significant in these models, no unadjusted models were conducted.
-The full logistic models and results are included in the supplemental analysis.
+Because most results were not statistically significant in these models, I do not include adjusted models here.
+Unadjusted and adjusted logistic models and results are included in the supplemental analysis.
 
@@
 The remaining 9% were interventionist or operative deliveries.
-For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94.26%.
-The SBR was 84.8%.
+For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94.1%.
+The SBR was 82.9%.
 The sex ratio was 1.2: there were 120 male live births per 100 female live births.
@@
 All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race).
-When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 83 grams lighter than their Euro-descended counterparts (Table 2).
-In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 96 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers.
+When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter than their Euro-descended counterparts (Table 2).
+In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers.
 Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent.
 
-In all models, women who were nulliparous (delivering their first child) had infants weighing on average 111 grams less than multiparous women, which is an association found both in other historical studies and today [@costa1998; @souza2020].
+In models 1 and 2 (Table 2), women who were nulliparous (delivering their first child) had infants weighing on average 113 grams less than multiparous women.
+In models 3 and 4 (Table 3), the average was 114 grams lighter.
+This is an association found both in other historical studies and today [@costa1998; @souza2020].
 Age and nationality were not significant.
@@
 Unadjusted ORs (Table 4) demonstrate that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019).
-When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were insignificant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
+When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were not significant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
 
@@
 In general, maternal and infant outcomes were not better in the Maternidade Laranjeiras.
-For example, between 1922 and 1926, the clinic's MMR was 94.26%.
+Between 1922 and 1926, the clinic's MMR was 94.26%.
 For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2020].
-This is a stark difference.
-Nonetheless, hospital rates probably reflected a registration effect, as medical personnel recorded all births and deaths, whereas reporting at the city level was less reliable [@loudon1992].
+Hospital rates probably reflected a registration effect, as medical personnel recorded all births and deaths, whereas reporting at the city level was less reliable [@loudon1992].
 Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s [@roth2020].
@@
 When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today.
-Although the mean birthweight was still within today's normal range, the differences between racialized groups demonstrate differential health outcomes that have persisted until today.
+Although the mean birth weight was still within today's normal range, the differences between racialized groups demonstrate differential health outcomes that have persisted until today.
 
@@
 Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period.
-Yet, they were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
+They were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
 White infants born at Laranjeiras had a mean birth weight of 3133 grams, slightly more than 50 grams lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
-Yet they were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 grams).
+They were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 grams).
 As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same historical period [@costa1998; @costa2004; @goldin1989].
@@
 Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care can all influence birth weight.
-Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother [@costa1998; @steckel1986].
-A threshold of malnutrition must be met before there is an adverse effect on fetal growth.
+Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986].
 Other factors include maternal age, parity, birth spacing, and gestational age.
-We can imagine that poorer women performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
+We can imagine that poorer women in early twentieth-century Rio de Janeiro performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
 These same women might have lived in crowded housing arrangements with lack of access to running water [@graham1992; @fischer2008; @roth2020].
-Generalized poverty might have influenced maternal nutrition, as well as the ability to access prenatal care.
+Generalized poverty might have influenced both maternal nutrition and the ability to access prenatal care.
 
@@
 The clinical records that comprise this dataset did not consistently record syphilis rates.
-Scholars have shown how medical and public health professionals viewed it as an important cause of adverse health outcomes and even a possible "scourge" on the future of the Brazilian race [@carrara1996a].
+Scholars have shown how medical and public health professionals viewed it as an important cause of adverse health outcomes and even a possible "scourge" on the future of the Brazilian race during this historical period [@carrara1996a].
 Although I don't have accurate rates for Laranjeiras, city-wide rates from the 1930s can provide some context for the prevalence of infection among pregnant women.
-Of all prenatal syphilis exams performed between 1935 and 1938 at public clinics or hospitals across the city, not stratified by maternal skin color, 7.92% came back positive [@fontenelle, p.
+Of all prenatal syphilis exams performed between 1935 and 1938 at public clinics or hospitals across the city, unadjusted for maternal skin color, 7.92% came back positive [@fontenelle, p.
 275].

### 2024-08-15 code/analysis-code/fullanalysis-v1.qmd
Revisions for submission 2
After call w/Bill

--- 2024-06-29 code/analysis-code/fullanalysis-v1.qmd
@@
 
-whereas BW is birth weight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
+whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
 
@@
   tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
-  add_glance_table(include = c(r.squared, nobs))
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
 table_model1
@@
 
-whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), Gest is gestational status (nulliparous or multiparous), Age is maternal age in years, and $\epsilon$ is the error term.
+whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
 Results are displayed in Table 3.
@@
 ```{r}
-# Final model 2 for paper using all data
+# Model 2: Final model for paper using all data
 # Using recipes and workflows
@@
 table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
-  add_glance_table(include = c(r.squared, nobs))
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
 table_model2
@@
 # Combine the two tables using tbl_merge
-table3 <- tbl_merge(list(table_model1, table_model2), tab_spanner = c("**Model 1**", "**Model 2**")) %>%
+table3 <- tbl_merge(
+  list(table_model1, table_model2),
+  tab_spanner = c("**Model 1**", "**Model 2**")) %>%
   modify_table_body(fun = ~.x %>% arrange(variable))
@@
 $$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$
-whereas BW is birthweight in grams, MC is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.
+whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
+Results are displayed in Table 4.
 
 ```{r}
-# Final model 3 for paper using all data
+# Model 3: Final model for paper using all data
 # Using recipes and workflows
@@
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
-  add_glance_table(include = c(r.squared, nobs))
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
 table_model3
@@
 
-whereas BW is birthweight in grams, MC is maternal skin color (White, Black and mixed race), Gest is gestational status (nullipara, primipara, multipara, multigravida), Nat is maternal nationality (Brazilian, Latin American, European, Middle Eastern), Age is maternal age in years, and $\epsilon$ is the error term.
+whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
 
 ```{r}
-# Final model 4 for paper using all data
+# Model 4: Final model for paper using all data
 # Using recipes and workflows
@@
   add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>% # Hide confidence intervals and p-values
-    add_glance_table(include = c(r.squared, nobs)) # Add R-squared and number of observations
+  add_glance_table(include = c(r.squared, nobs)) %>% # Add R-squared and number of observations
+  modify_header(statistic = "**T-Statistic**")
 table_model4

### 2024-08-15 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 2
After call w/Bill

@@
 
-For all three models, the r-squared is very small (0.009, 0.008, and 0.026), indicating that the model does not explain much of the variance in birth weight.
+For all three models, the r-squared is very small (0.009, 0.008, and 0.027), indicating that the model does not explain much of the variance in birth weight.
 This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
@@
 
-Now let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race).
-We'll start with a simple model of infant birth weight as a function of maternal skin color.
-Then we will run a model with all the predictors.
+There is potentially a (partial) multi-collinearity problem with maternal skin color and maternal nationality. So let's run a regression with just maternal nationality as the covariate.
 
@@
 # Set reference level for Color
-ML_linear$Color <- relevel(ML_linear$Color, ref = "White")
+ML_linear$ModifiedNationality <- relevel(ML_linear$ModifiedNationality, ref = "Brazilian")
 
 # Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)
-
-# Summarize model fit with tidiers
-ML_lm4_summary <- summary(ML_lm4) # Summary of fit
-tidy(ML_lm4, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm4) # Model-level statistics
-
-# Create a gtsummary table directly from the model
-table_fit4 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
-  bold_p() %>% 
-  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
-  modify_column_unhide(columns = c(statistic, std.error))
-table_fit4
-
-# Save regression output into table
-saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))
-
-# Complex linear model
-# Set reference level for Color
-ML_linear$Color <- relevel(ML_linear$Color, ref = "White")
-
-# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
+ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedNationality, data = ML_linear)
 
@@
 ```
+From our results, we can see that Europeans had heavier infants than Brazilians.
+Since all women of color in this sample were Brazilian, we have a problem with multicollinearity.
+So let's try a couple things to understand this issue.
+We will create a `Brazilian` (yes, no) variable and then run a regression with this variable.
+Then, we will drop all non-Brazilians from the study and run the original regression with the covariates.
+
+```{r}
+# Create dummy variables for new category of Brazilian or not
+# Create dummy variables for status and change to English
+ML_linear$Brazilian <- recode_factor(ML_linear$ModifiedNationality, "European" = "Non-Brazilian", "Middle Eastern" = "Non-Brazilian", "Latin American" = "Non-Brazilian", "Brazilian" = "Brazilian")
+
+str(ML_linear)
+```
+Now, we will run a regression with only this new variable.
+
+```{r}
+# Simple linear model
+# Set reference level for Brazilian
+ML_linear$Brazilian <- relevel(ML_linear$Brazilian, ref = "Brazilian")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Brazilian, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_lm5_summary <- summary(ML_lm5) # Summary of fit
+tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_lm5) # Model-level statistics
+
+# Create a gtsummary table directly from the model
+table_fit5 <- tbl_regression(ML_lm5, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table_fit5
+
+# Save regression output into table
+saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))
+```
+Here, we can see that non-Brazilians have babies on average weighing 100 grams more than Brazilians (p-value 0.001).
+So we will now run a regression of only Brazilian women using Afro-Descent and Euro-descent categories.
+
+```{r}
+# Filter data based on desired conditions
+filtered_data <- ML_linear %>%
+  filter(Brazilian == "Brazilian")
+
+# Simple linear model
+# Set reference level for Brazilian
+ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm6 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = filtered_data)
+
+# Summarize model fit with tidiers
+ML_lm6_summary <- summary(ML_lm6) # Summary of fit
+tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_lm6) # Model-level statistics
+
+# Create a gtsummary table directly from the model
+table_fit6 <- tbl_regression(ML_lm6, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table_fit6
+
+# Save regression output into table
+saveRDS(table_fit6, file = here::here("results", "tables", "table_fit6.rds"))
+```
+
+Moving on from issues of collinearity, let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race).
+We'll start with a simple model of infant birth weight as a function of maternal skin color.
+Then we will run a model with all the predictors.
+
+```{r}
+# Simple linear model
+# Set reference level for Color
+ML_linear$Color <- relevel(ML_linear$Color, ref = "White")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_lm5_summary <- summary(ML_lm5) # Summary of fit
+tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_lm5) # Model-level statistics
+
+# Create a gtsummary table directly from the model
+table_fit5 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table_fit5
+
+# Save regression output into table
+saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))
+
+# Complex linear model
+# Set reference level for Color
+ML_linear$Color <- relevel(ML_linear$Color, ref = "White")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm6 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_lm6_summary <- summary(ML_lm6) # Summary of fit
+tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+glance(ML_lm6) # Model-level statistics
+
+# Create a gtsummary table directly from the model
+table_fit6 <- tbl_regression(ML_lm6, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table_fit6
+
+# Save regression output into table
+saveRDS(table_fit6, file = here::here("results", "tables", "table_fit6.rds"))
+```
 
@@
 ```{r}
+# Check factor levels
+levels(ML_linear$BirthweightCategory) # Reference is LBW, so need to change
+
+# Set the reference level explicitly to NBW
+ML_linear <- ML_linear %>%
+  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))
+
 #set seed
@@
 ```{r}
-# Simple logistic regression model with ModifiedColor as the predictor
-# Check factor levels
-levels(ML_linear$BirthweightCategory) # Reference is LBW
-
-# Set the reference level explicitly to NBW
-ML_linear <- ML_linear %>%
-  mutate(BirthweightCategory = relevel(BirthweightCategory, ref = "NBW"))
-
 # Create recipe

### 2024-08-15 code/eda-code/eda-v1.qmd
Revisions for submission 2
After call w/Bill

@@
 # Recode factor using dplyr
-ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")
+ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "Afro-Descent", "Preta" = "Afro-Descent", "Branca" = "Euro-Descent")
 
@@
 # Create dummy variables for nationality and change to English
-ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
+ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "Latin American", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "Latin American", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "Middle Eastern", "Uruguaya" = "Latin American")
 
@@
 #ML_summary <- ML_summary %>%
-  #mutate(Afro_LBW = case_when(ModifiedColor == "EuroDescent" ~ NA_real_, TRUE ~ Afro_LBW),
-    #Euro_LBW = case_when(ModifiedColor == "AfroDescent" ~ NA_real_,TRUE ~ Euro_LBW))
+  #mutate(Afro_LBW = case_when(ModifiedColor == "Euro-Descent" ~ NA_real_, TRUE ~ Afro_LBW),
+    #Euro_LBW = case_when(ModifiedColor == "Afro-Descent" ~ NA_real_,TRUE ~ Euro_LBW))
 
@@
 print(LBW_percentage)
+
+# Perform difference in means test (t-test) for birth weight category
+# Modify code to create dataset for t-test
+LBW_data <- ML_summary %>%
+  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
+  mutate(is_LBW = as.numeric(BirthweightCategory == "LBW"))  # 1 if LBW, 0 if not
+
+# Print the first few rows to check the structure
+print(head(LBW_data))
+
+# Perform t-test
+t_test_result <- t.test(is_LBW ~ ModifiedColor, data = LBW_data, subset = ModifiedColor %in% c("Euro-Descent", "Afro-Descent"))
+
+# Print the result
+print(t_test_result)
 ```
@@
 ```{r}
-# Change observations from Portuguese to English
+# Change rest of observations from Portuguese to English
 ML_summary <- ML_summary %>%
   mutate(
-    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
-    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
     Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
@@
     label = list(ModifiedColor ~ "Ancestry", Status ~ "Parity or Gravidity", ModifiedStatus ~ "Parity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Lengthcentimeters ~ "Birth Length (cms)", Weightgrams ~ "Birth Weight (grms)"),
-    statistic = list(all_continuous() ~ "{mean} ({sd})")
+    statistic = list(all_continuous() ~ "{mean} ({p25}, {p75})")
   ) %>%
@@
     statistic = list(
-       all_continuous() ~ "{mean} ({sd})",
-      WeightgramsBlack ~ "{mean} ({sd})",
-      WeightgramsMixed ~ "{mean} ({sd})",
-      WeightgramsWhite ~ "{mean} ({sd})",
-      WeightgramsFemale ~ "{mean} ({sd})",
-      WeightgramsMale ~ "{mean} ({sd})"
+       all_continuous() ~ "{mean} ({p25}, {p75})",
+      WeightgramsBlack ~ "{mean} ({p25}, {p75})",
+      WeightgramsMixed ~ "{mean} ({p25}, {p75})",
+      WeightgramsWhite ~ "{mean} ({p25}, {p75})",
+      WeightgramsFemale ~ "{mean} ({p25}, {p75})",
+      WeightgramsMale ~ "{mean} ({p25}, {p75})"
     ),
@@
 
-# Create dummy variables using recode factor from dplyr
+# Create dummy variables for color and change to English using recode factor from dplyr
 ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Preta" = "Afro-Descent", "Parda" = "Afro-Descent", "Branca" = "Euro-Descent")
@@
 # Create dummy variables for nationality and change to English
-ML_linear$ModifiedNationality <- recode_factor(ML_linear$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")
+ML_linear$ModifiedNationality <- recode_factor(ML_linear$Nationality, "Alema" = "European", "Argentina" = "Latin American", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "Latin American", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "Middle Eastern", "Uruguaya" = "Latin American")
 
-# Change observations from Portuguese to English
+# Change remaining observations from Portuguese to English
 ML_linear <- ML_linear %>%
   mutate(
-    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
     Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),

### 2024-08-15 code/processing-code/processingfile-v1.qmd
Revisions for submission 2
After call w/Bill

@@
 # Remove median and maternal age variables given missing data
-Mean_Birthweights <- Mean_Birthweights %>% dplyr::select(-Median, -`Maternal Age (mean)`, -"<2500gm (%)")
+Mean_Birthweights <- Mean_Birthweights %>% dplyr::select(-Median, -`Maternal Age (mean)`, -"<2500gm (%)", -`Page #`)
 

### 2024-08-15 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission 2
After call w/Bill

@@
 This cross-sectional analysis uses hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital.
-I used linear models to assess differences in infant birth weight by maternal skin color, age, parity, and nationality.
+I used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.
 
 ## Findings
+African ancestry was correlated with lower birth weights.
 Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant.
-Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 96 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.
+Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.
 
@@
 In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research, as LBW is associated with higher rates of neonatal and infant mortality and possible health problems over the life course [@cardoso2005; @paixao2021; @vilanova2019].
+BILL COMMENT: ARE THERE LONGITUDINAL STUDIES OF LATER-LIFE MORTALITY OR LIFE EXPECTANCY WITH BIRTH WEIGHT?
 
@@
 Racism is a major factor in these outcomes, as studies demonstrate that U.S.-born Black women have higher rates of LBW and PTB than Black immigrant women [@braveman2021a; @clay2023a; @david1997].
+BILL COMMENT: RELATED TO RACISM, WHAT DOES LITERATURE ON BIRTH WEIGHT SAY ABOUT ITS CAUSES?
+THESE ARE CAUSAL STATEMENTS.
 
@@
 
-Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019].
+Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019].
 The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe.
@@
 
+BILL COMMENT: I AM CURIOUS, DO WHITE SOUTH AFRICANS HAVE HIGHER BIRTH WEIGHT, CONTROLLING FOR INCOME? ANY STUDY OF THIS?
+
 This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993].
@@
 In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon.
-In our current postgenomic era, in which race as a biological category has been rejected, the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking remains very real [@roberts2011].
+In our current postgenomic era, in which race as a biological category has been rejected, the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking remains very real [@meloni2022].
 Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022].
@@
 
-This study analyzes infant birth weight as a proxy for general fetal and neonatal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras.
+This study analyzes infant birth weight as a proxy for general fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras.
 It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
+BILL COMMENT: PUT THIS PARAGRAPH MUCH EARLIER.
 
@@
 
-This analysis draws on a unique sample of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926.
+This analysis draws on a unique data set of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926.
 I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO).
@@
 When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a].
+BILL COMMENT: FOR OBSTETRICIANS WHO APPEAR MORE THAN ONCE, IS THERE A RELATIONSHIP WITH BIRTH WEIGHT?
 
@@
 In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW (<p>&ge;</p>2500 g) or LBW (<2500 g), according to current WHO classifications) [@worldhealthorganization2022].
+BILL COMMENT: THESE COME FROM WHERE AND WHEN?
 
@@
 
-Racial terminology in Brazil was, and continues to be, complex and dynamic [@telles2006; @travassos2004].
+Except at the extremes, racial classification in Brazil was, and continues to be, imprecise [@travassos2004].
 Racial categories and skin color exist on a spectrum rather than a Black-White binary [@telles2006].
-Analyzing both specific categories of skin color and more general categories of ancestry can help address these different understandings of race and is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].
+Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category.
+This method is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].
 
@@
 *Maternal outcome*.
-Hospital records recorded maternal outcome was recorded as death, discharge, or hospital transferal, which I maintain here.
+Hospital records recorded maternal outcome as death, discharge, or hospital transferal, which I maintain here.
 I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality below.
@@
 
-I used Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight.
+I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight.
 I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
@@
 Unadjusted and adjusted logistic models and results are included in the supplemental analysis.
+BILL COMMENT: WHY WEREN'T MODELS SIGNIFICANT?
 
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
 
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
+Results for both models are displayed in Table 2.
 
-The simple linear regression model for original color is as follows:
+The simple linear regression model for original skin color is as follows:
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
-Results are displayed in Table 4.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
 
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+Results for both models are displayed in Table 3.
 
@@
 The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers.
+A difference in means test shows borderline significance between the means (t-statistic = 1.9119, df.
+2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).
+
 Results for all linear models are displayed in Tables 2 and 3.
@@
 Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent.
+AT THIS POINT I NEED TO SEE A LITERATURE THAT TELLS ME 100 GRAMS MATTERS FOR SOMETHING, HEALTH-WISE, IS 100 GRAMS MORE THAN 2 STANDARD DEVIATIONS?
 
@@
 When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were not significant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
+MAKING CONTINUOUS VARIBLE CATEGORICAL.
+LBW NOT A CATEGORY YET PER WHO.
 
@@
 Discussions of epigenetic markers associated with the lived experience of racism might also be applicable here.
-Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors), mediated through epigenetic mechanisms.
+Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by a variety of biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms.
 Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@leimert2020].
@@
 
-*Conclusion*
 Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study.
 Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.
+
+*Conclusion*
+This study has shown that birth weight differed by maternal skin color.
+This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms.
+These findings, although not generalizable beyond this specific hospital, are consistent with present-day studies, demonstrating how racial inequality in health outcomes has persisted over time.
 
@@
 #| label: tbl-table2
-#| tbl-cap: "Model 1 and 2 Results"
+#| tbl-cap: "Model 1 and 2 Results (Dependent variable: Birth weight in grams)"
 #| echo: FALSE
@@
 #| label: tbl-table3
-#| tbl-cap: "Model 3 and 4 Results"
+#| tbl-cap: "Model 3 and 4 Results (Dependent variable: Birth weight in grams)"
 #| echo: FALSE

### 2024-08-15 code/processing-code/processingcode.R

@@ -51,48 +51,68 @@ skimr::skim(ML_raw)
 ## ---- cleandata --------
 # Dataset created for multiple reasons. Need to narrow down relevant variables
 # Remove unnecessary variables
-ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Maternal_Notes, -Fetal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
+ML_1 <- ML_raw %>% dplyr::select(-VolN, -Page, -Number, -Nationality_Notes, -Maternal_Notes, -CauseofDeath, -PreviousHistory, -Notes)
 #Summarize missing data using dplyr
 missing_data_summary <- ML_1 %>% dplyr::summarise_all(list(~sum(is.na(.))))
 missing_data_summary
-# Extract information on gestation from Birth_Notes and create new variable gestation with singleton and twin births
+# See values for `FetalOutcome` and `Fetal_Notes` that are not missing any values for both variables
+ML_1 %>% filter(!is.na(FetalOutcome) & !is.na(Fetal_Notes))
+# Change 'FetalOutcome' to vivo if 'Fetal_Notes' indicates neonatal death (via Faleceu or faleceu in 'Fetal_Notes' column)
 ML_2 <- ML_1 %>%
-  mutate(Birth_Notes = str_trim(Birth_Notes)) %>% # removes any leading or trailing whitespace from Birth_Notes column
+  mutate(FetalOutcome = case_when(
+    FetalOutcome == "morto" & 
+      str_detect(tolower(Fetal_Notes), "faleceu") ~ "vivo",
+    TRUE ~ FetalOutcome
+  ))
+ML_2
+# Check values
+ML_2 %>% filter(!is.na(FetalOutcome) & !is.na(Fetal_Notes))
+# See how many neonatal deaths were recorded
+ML_neonatal <- ML_2 %>%
+  filter(
+    FetalOutcome == "vivo" &
+      str_detect(tolower(Fetal_Notes), "faleceu")
+  )
+# View the filtered data
+print(ML_neonatal)
+# Extract information on gestation from Birth_Notes and create new variable gestation with singleton and twin births
+ML_3 <- ML_2 %>%
+  mutate(Birth_Notes = str_trim(Birth_Notes)) %>% # removes any leading or trailing white space from Birth_Notes column
[…]

### 2024-08-16 code/analysis-code/fullanalysis-v1.qmd
Revisions for submission 3
Clean up of code

@@
 ## Merge Models 3 and 4
+
 ```{r}

### 2024-08-16 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 3
Clean up of code

@@
 date: "`r Sys.Date()`"
-output: html_document
+format: 
+  html:
+    code-fold: true
+editor: visual
 ---
@@
 ```{r}
-#Path to summary data. Note the use of the here() package and not absolute paths
+#Path to linear data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear.rds")
@@
 ML_linear <- readRDS(data_location)
+
+#Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","ML_summary.rds")
+#load data
+ML_summary <- readRDS(data_location)
 ```
 
-## Linear Models
+# LBW and NBW percentages
 
-Let's run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry. I will use the recipe and workflow features of `tidymodels`.
+First, I will see how many infants were LBW or NBW for each maternal descendant category.
+Then, I will perform a simple difference in means (t-test) to see if the LBW and NBW means stratified by maternal skin color are statistically significant.
 
 ```{r}
-# Model using recipes and workflows
+# See LBW and NBW with true denominator
+LBW_percentage <- ML_summary %>%
+  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>% # Remove NAs from both variables
+
+group_by(ModifiedColor) %>% # Group by ModifiedColor
+# Calculate the percentage for each group
+summarise(
+  total_count = n(),
+  lbw_count = sum(BirthweightCategory == "LBW"),
+  lbw_percentage = (lbw_count / total_count) * 100
+  )
+
+# Print the result
+print(LBW_percentage)
+
+# Perform difference in means test (t-test) for birth weight category
+# Modify code to create dataset for t-test
+LBW_data <- ML_summary %>%
+  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
+  mutate(is_LBW = as.numeric(BirthweightCategory == "LBW"))  # 1 if LBW, 0 if not
+
+# Print the first few rows to check the structure
+print(head(LBW_data))
+
+# Perform t-test
+t_test_result <- t.test(is_LBW ~ ModifiedColor, data = LBW_data, subset = ModifiedColor %in% c("Euro-Descent", "Afro-Descent"))
+
+# Print the result
+print(t_test_result)
+
+# Extract specific values for inline text
+p_value <- t_test_result$p.value
+t_statistic <- t_test_result$statistic
+df <- t_test_result$parameter
+mean_diff <- diff(t_test_result$estimate)
+ci_lower <- t_test_result$conf.int[1]
+ci_upper <- t_test_result$conf.int[2]
+```
+
+After performing a Welch Two Sample t-test to compare the rates of LBW between Euro- and Afro-descent mothers, the results were as follows: t = `r round(t_statistic, 2)`, df = `r round(df, 1)`, p-value = `r round(p_value, digits = 3)`, mean difference = `r round(mean_diff, 4)`, 95% CI = [`r round(ci_lower, 4)`, `r round(ci_upper, 4)`].
+
+# Linear Models
+
+To be able include metrics like $(R^2)$ dynamically in the Quarto document, I first need to store the values in a list.
+I will create a list to store the values.
+In each code chunk with a fitted model, I will add the $(R^2)$ to the list.
+Then I can use the stored values inline in the document.
+
+```{r}
+# Create a list to store R-squared values
+r_squared_values <- list()
+```
+
+I will now run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry. I will use the recipe and workflow features of `tidymodels`.
+
+```{r}
+# Model 1 using recipes and workflows
 # Define recipe for model
 ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
-  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level
   step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
@@
 tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
-glance(ML_fit1) # Model-level statistics
+model_stats1 <- glance(ML_fit1) # Extract model-level statistics
 
+# This code is to write dynamic inline text in Quarto
+# Extract R-squared value
+r_squared_values$model1 <- model_stats1$r.squared
+
+# This code is for performance measures that aren't necessary in this study
 # Produce predictions for R-squared
@@
 
-#Print table
+# Print table
 table_fit1
@@
 ```
-Now let's run a second simple bivariate analysis, looking at infant birth weight as a function of maternal age. This time, when plotting weight as a function of maternal age, we can use a scatterplot.
+
+Now I will run a second simple bivariate analysis, looking at infant birth weight as a function of maternal age.
+This time, when plotting weight as a function of maternal age, I will use a scatterplot.
 
 ```{r}
-# Model using recipes and workflows
+# Model Age using recipes and workflows
 # Create recipe
@@
 tidy(ML_fit_age, conf.int = TRUE) # Coefficients and confidence intervals
-glance(ML_fit_age) # Model-level statistics
+model_stats_age <- glance(ML_fit_age) # Extract model-level statistics
 
+# This code is to write dynamic inline text in Quarto
+# Extract R-squared value
+r_squared_values$model2 <- model_stats_age$r.squared
+
+# This code is for performance measures that aren't necessary in this study
 # Produce predictions for RMSE and R-squared
@@
 
-Now let's run a slightly more complicated model that includes maternal skin color, age, nationality, and gestational status.
+Now I will run a slightly more complicated model that includes maternal skin color, age, nationality, and gestational status.
 
@@
 tidy(ML_lm3, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm3) # Model-level statistics
+model_stats3 <- glance(ML_lm3) # extract model-level statistics
+
+# This code is to write dynamic inline text in Quarto
+# Extract R-squared value
+r_squared_values$model3 <- model_stats3$r.squared
 
@@
 
-For all three models, the r-squared is very small (0.009, 0.008, and 0.027), indicating that the model does not explain much of the variance in birth weight.
+For all three models, the r-squared is small `r round(r_squared_values$model1, 3)`, `r round(r_squared_values$model2, 3)`, and `r round(r_squared_values$model3, 3)`, indicating that the model does not explain much of the variance in birth weight.
 This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
@@
 
+## Sub-Sample Modeling
+
 There is potentially a (partial) multi-collinearity problem with maternal skin color and maternal nationality. So let's run a regression with just maternal nationality as the covariate.
@@
 ```{r}
-# Simple linear model
+# Create a list to store R-squared values
+r_squared_values2 <- list()
+```
+
+```{r}
+# Model 4 sub-sample simple model
 # Set reference level for Color
@@
 tidy(ML_lm4, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm4) # Model-level statistics
+model_stats4 <- glance(ML_lm4) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values2$model4 <- model_stats4$r.squared
 
@@
 ```
-From our results, we can see that Europeans had heavier infants than Brazilians.
-Since all women of color in this sample were Brazilian, we have a problem with multicollinearity.
-So let's try a couple things to understand this issue.
+
+These results imply that Europeans had heavier infants than Brazilians.
+Since all women of color in this sample were Brazilian, there is a problem with multicollinearity.
+So I will perform some subsample modeling to understand this issue.
 We will create a `Brazilian` (yes, no) variable and then run a regression with this variable.
-Then, we will drop all non-Brazilians from the study and run the original regression with the covariates.
+Then, I will drop all non-Brazilians from the study and run the original regression with the covariates.
 
@@
 ```
-Now, we will run a regression with only this new variable.
+
+Now, I will run a regression with only this new variable.
 
 ```{r}
-# Simple linear model
+# Model 5 sub-sample simple model
 # Set reference level for Brazilian
@@
 tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm5) # Model-level statistics
+model_stats5 <- glance(ML_lm5) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values2$model5 <- model_stats5$r.squared
 
@@
 ```
-Here, we can see that non-Brazilians have babies on average weighing 100 grams more than Brazilians (p-value 0.001).
-So we will now run a regression of only Brazilian women using Afro-Descent and Euro-descent categories.
+
+Here, non-Brazilians had babies on average weighing 100 grams more than Brazilians (p-value 0.001).
+So I will now run a regression of only Brazilian women using `Afro-Descent` and `Euro-descent` categories.
 
@@
 
-# Simple linear model
+# Model 6 sub-sample simple linear model
 # Set reference level for Brazilian
@@
 tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm6) # Model-level statistics
+model_stats6 <- glance(ML_lm6) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values2$model6 <- model_stats6$r.squared
 
@@
 
-Moving on from issues of collinearity, let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race).
-We'll start with a simple model of infant birth weight as a function of maternal skin color.
-Then we will run a model with all the predictors.
+The $R^2$ for these models were `r round(r_squared_values2$model4, 3)`, `r round(r_squared_values2$model5, 3)`, and `r round(r_squared_values2$model6, 3)`.
+
+## Multivariate analysis
+
+Moving on from issues of collinearity, I will run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race).
+I'll start with a simple model of infant birth weight as a function of maternal skin color.
+Then I will run a model with all the predictors.
 
 ```{r}
-# Simple linear model
+# Create a list to store R-squared values
+r_squared_values3 <- list()
+```
+
+```{r}
+# Model 7 simple linear model
 # Set reference level for Color
@@
 # Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)
+ML_lm7 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)
 
 # Summarize model fit with tidiers
-ML_lm5_summary <- summary(ML_lm5) # Summary of fit
-tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm5) # Model-level statistics
+ML_lm7_summary <- summary(ML_lm7) # Summary of fit
+tidy(ML_lm7, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+model_stats7 <- glance(ML_lm7) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values3$model7 <- model_stats7$r.squared
 
 # Create a gtsummary table directly from the model
-table_fit5 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
+table_fit7 <- tbl_regression(ML_lm7, exponentiate = FALSE, intercept = TRUE) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
-table_fit5
+table_fit7
 
 # Save regression output into table
-saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))
+saveRDS(table_fit7, file = here::here("results", "tables", "table_fit7.rds"))
 
@@
 # Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm6 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
+ML_lm8 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
 
 # Summarize model fit with tidiers
-ML_lm6_summary <- summary(ML_lm6) # Summary of fit
-tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
-glance(ML_lm6) # Model-level statistics
+ML_lm8_summary <- summary(ML_lm8) # Summary of fit
+tidy(ML_lm8, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+model_stats8 <- glance(ML_lm8) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values3$model8 <- model_stats8$r.squared
 
 # Create a gtsummary table directly from the model
-table_fit6 <- tbl_regression(ML_lm6, exponentiate = FALSE, intercept = TRUE) %>% 
+table_fit8 <- tbl_regression(ML_lm8, exponentiate = FALSE, intercept = TRUE) %>% 
   bold_p() %>% 
@@
   modify_column_unhide(columns = c(statistic, std.error))
-table_fit6
+table_fit8
 
 # Save regression output into table
-saveRDS(table_fit6, file = here::here("results", "tables", "table_fit6.rds"))
+saveRDS(table_fit8, file = here::here("results", "tables", "table_fit8.rds"))
 ```
+
+The $R^2$ for these models were `r round(r_squared_values3$model7, 3)` and `r round(r_squared_values3$model8, 3)`.
 
@@
 
-Now we will still consider `Birthweight` as the outcome of interest, but we will transform it into a categorical outcome of <=2500 (low birthweight, LBW) and >2500 (normal birthweight, NBW) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
+Now I will still consider `Birthweight` as the outcome of interest, but I will use the dummy variable created in the `eda-v1.qmd` file that categorizes infant birth weight into normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
 
-First, we will split data into training and testing sets to later calculate ROC_AUC.
+First, I will split data into training and testing sets to later calculate ROC_AUC.
 
@@
 
-Now we will run the first logistic model with `ModifiedColor` as the predictor.
+Now I will run the first logistic model with `ModifiedColor` as the predictor.
 
@@
 # Create recipe
-# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
+# Logistic model 1: Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
 log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
@@
 print(roc_plot1)
+
+# Store metrics in a list for dynamic inline text
+logistic_model_metrics1 <- list(
+  pseudo_r_squared = mcfadden_r2_1,
+  auc = roc_auc_result1$.estimate
+)
 ```
-With an ROC_AUC of 0.463, our model is performing worse than random guessing. Our pseudo R-squared is 0.004.
 
-Now let's run another logistic model with `Color` as the predictor.
+Then, I will create a list of lists to store my performance metrics for dynamic inline text later.
 
 ```{r}
-# Simple logistic regression model with Color as the predictor
+# Create a list to store performance metrics
+all_logistic_models <- list()
+all_logistic_models$model1 <- logistic_model_metrics1
+```
+
+With an ROC_AUC of `r round(all_logistic_models$model1$auc, 3)`, the model is performing worse than random guessing. The pseudo-$R^2$ is `r round(all_logistic_models$model1$pseudo_r_squared, 3)`.
+
+Now I will run another logistic model with `Color` as the predictor.
+
+```{r}
+# Logistic model 2: simple logistic regression model with Color as predictor
 # Check factor levels
@@
 
-# Calculate metrics using pseudo-R2
 # Extract model metrics using broom
@@
 print(roc_plot2)
+
+# Store metrics in a list for dynamic inline text
+logistic_model_metrics2 <- list(
+  pseudo_r_squared = mcfadden_r2_1,
+  auc = roc_auc_result1$.estimate
+)
+all_logistic_models$model2 <- logistic_model_metrics2 # add to list of lists
 ```
-The ROC_AUC for our second simple logistic regression model is 0.47, or still worse than random guessing. Our pseudo R-squared is 0.006.
 
-Now, we will run two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.
+The ROC_AUC for the second simple logistic regression model is `r round(all_logistic_models$model2$auc, 3)`, or still worse than random guessing.
+The pseudo- $R^2$ is `r round(all_logistic_models$model2$auc, 3)`.
+
+Now, I will run two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.
 
 ```{r}
-# Multiple logistic regression model with ModifiedColor as the predictor
+# Logistic model 3: Multiple logistic regression model with ModifiedColor as the predictor
 # Check factor levels
@@
 print(roc_plot3)
+
+# Store metrics in a list for dynamic inline text
+logistic_model_metrics3 <- list(
+  pseudo_r_squared = mcfadden_r2_1,
+  auc = roc_auc_result1$.estimate
+)
+all_logistic_models$model3 <- logistic_model_metrics3 # add to list of lists
 ```
-Our ROC_AUC here is worse, at 0.46, or worse than random guessing. Our pseudo-R2 is 0.0055.
+
+The ROC_AUC here is `r round(all_logistic_metrics$model3$auc, 3)`, still worse than random guessing. The pseudo-$R^2$ is `r round (all_logistic_metrics$model3$pseudo_r_squared, 3)`.
 
 ```{r}
-# Multiple logistic regression model with Color as the predictor
+# Logistic model 4: Multiple logistic regression model with Color as the predictor
 # Check factor levels
@@
 print(roc_plot4)
+
+# Store metrics in a list for dynamic inline text
+logistic_model_metrics4 <- list(
+  pseudo_r_squared = mcfadden_r2_1,
+  auc = roc_auc_result1$.estimate
+)
+all_logistic_models$model4 <- logistic_model_metrics4 # add to list of lists
 ```
-In our final model, our ROC_AUC is 0.477, still worse than random guessing. Our pseudo-R2 is 0.007.
+
+In the final model, the ROC_AUC is `r round(all_logistic_metrics$model4$auc, 3)` still worse than random guessing. The pseudo-$R^2$ is `r round(all_logistic_metrics$model4$pseudo_r_squared, 3)`.
 
@@
 ```
-```

### 2024-08-16 code/eda-code/eda-v1.qmd
Revisions for submission 3
Clean up of code

@@
 
-Now let's make some tables about the data to get a better idea of the overall numbers.
-
 Before we create a summary table, lets clean up our variables for easier analysis.
@@
 I maintained this categorization.
-However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category.
+However, I also created a new dummy variable for skin color, combining Black and mixed-Race patients into one category, `Afro-Descent`, and all White patients into a `Euro-Descent` category.
 
-For the `Status` variable, Table 1 will include all of the gravidity or parity categories from the original data.
-But I will also create a dummy variable for the linear analysis: one category for first pregnancy/birth and one for second or more pregnancies/births.
-With this, I am removing the distinction between parity and gravidity since it is unclear from the original data published in the clinical notes if the physicians were categorizing the women before the delivery or after.
+For the `Status` variable, the Supplementary Table 1 will include all of the gravidity or parity categories from the original data.
+But I will also create a dummy variable for the summary table for both the final paper and the linear analysis: one category for first pregnancy/birth and one for second or more pregnancies/births.
+With this, I am removing the distinction between parity and gravidity since it is unclear from the original data if the physicians were categorizing the women before the delivery or after.
 With this, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births.
@@
 For second or more pregnancies/births, I combine all multiple statuses (gravida or parous) into `multiparous`.
-This combines "multigesta", "secundigesta", and "trigesta" (-gesta meaning -gravida in English) into `multiparous`, since they mean the same thing (two or more pregnancies).
+This combines "multigesta", "secundigesta", and "trigesta" (-gesta in Portuguese meaning -gravida in English) into `multiparous`, since they mean the same thing (two or more pregnancies).
 I also combine "secundipara" and "multipara" into the category `multiparous`, since this means two or more births.
@@
 
-For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern".
+For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "Latin American", and all Middle Eastern nationalities into "Middle Eastern".
 I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations.
@@
 Finally, for the summary table, in which I want to include infant birth weight stratified by maternal skin color and by fetal sex, I will create new dummy variables here for that purpose.
+For maternal skin color, these variables are `Weightblack`, `Weightmixed`, and `Weightwhite`.
+These are categorical variables (yes/no).
+From these, I create new continuous dummy variables for birth weight according to maternal skin color: `WeightgramsBlack`, `WeightgramsMixed`, and `WeightgramsWhite`.
+I do the same process for fetal sex: `Weightfemale`, `Weightmale`, `WeightgramsFemale`, and `WeightgramsMale`.
+
+Using these new dummy variables, I create new dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color.
+These are `Afro_NBW` and `Afro_LBW`; and `Euro_NBW` and `Euro_LBW`.
+I will use these to see how many infants were low birth weight or normal birth weight for each maternal skin category in the `introanalysis-v1.qmd` file.
 
@@
 # Verify new variables
-str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale")])
+str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale", "Afro_LBW", "Afro_NBW", "Euro_LBW", "Euro_NBW")])
 
@@
 print(head(ML_summary))
-
-# See LBW and NBW with true denominator
-LBW_percentage <- ML_summary %>%
-# Remove NAs from both variables
-filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
-# Group by ModifiedColor
-group_by(ModifiedColor) %>%
-# Calculate the percentage for each group
-summarise(
-  total_count = n(),
-  lbw_count = sum(BirthweightCategory == "LBW"),
-  lbw_percentage = (lbw_count / total_count) * 100
-  )
-
-# Print the result
-print(LBW_percentage)
-
-# Perform difference in means test (t-test) for birth weight category
-# Modify code to create dataset for t-test
-LBW_data <- ML_summary %>%
-  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
-  mutate(is_LBW = as.numeric(BirthweightCategory == "LBW"))  # 1 if LBW, 0 if not
-
-# Print the first few rows to check the structure
-print(head(LBW_data))
-
-# Perform t-test
-t_test_result <- t.test(is_LBW ~ ModifiedColor, data = LBW_data, subset = ModifiedColor %in% c("Euro-Descent", "Afro-Descent"))
-
-# Print the result
-print(t_test_result)
 ```
-
 Let's also change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.
@@
 
-# Save new summary data with recoded factor variables
-saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
-
 #Summarize missing data using dplyr
@@
 missing_data_summary
+
+# Save new summary data with recoded factor variables
+saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
 ```
@@
 
-Now we will create the summary table for mean birth weights across time and place.
-
-```{r}
-#Path to summary data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")
-
-#load data
-Mean_Birthweights <- readRDS(data_location)
-
-# Replace NA with a dash for each column
-Mean_Birthweights <- Mean_Birthweights %>%
-  mutate(across(where(is.numeric), ~ ifelse(is.na(.), "-", as.character(.)))) %>%
-  mutate(across(where(is.character), ~ replace_na(., "-")))
-
-# Display the table using flextable
-table2 <- flextable(Mean_Birthweights) %>%
-  set_caption("Mean Birth Weight Comparison") %>%
-  align(j = 2, align = "left", part = "all") %>%
-  align(j = 4, align = "right", part = "all") %>%
-  set_header_labels(
-    Sample = "Sample",
-    Year = "Year", 
-    Mean = "Mean (grams)",
-    Source = "Source"
-  ) %>%
-  bold(part = "header") %>%
-  #autofit() %>%
-  set_table_properties(layout = "autofit", width = 1, align = "center")
-table2
-
-#Save table
-saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
-```
-
 For maternal race, we have a sample size of n = 2695.
@@
 
+Now we will create the summary table for mean birth weights across time and place.
+
+```{r}
+#Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")
+
+#load data
+Mean_Birthweights <- readRDS(data_location)
+
+# Replace NA with a dash for each column
+Mean_Birthweights <- Mean_Birthweights %>%
+  mutate(across(where(is.numeric), ~ ifelse(is.na(.), "-", as.character(.)))) %>%
+  mutate(across(where(is.character), ~ replace_na(., "-")))
+
+# Display the table using flextable
+table2 <- flextable(Mean_Birthweights) %>%
+  set_caption("Mean Birth Weight Comparison") %>%
+  align(j = 2, align = "left", part = "all") %>%
+  align(j = 4, align = "right", part = "all") %>%
+  set_header_labels(
+    Sample = "Sample",
+    Year = "Year", 
+    Mean = "Mean (grams)",
+    Source = "Source"
+  ) %>%
+  bold(part = "header") %>%
+  #autofit() %>%
+  set_table_properties(layout = "autofit", width = 1, align = "center")
+table2
+
+#Save table
+saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
+```
+
 Because I am working on a different subset of the data, `ML_linear` for the statistical analysis, I will also need to recategorize variables here.
@@
 
-# Save new summary data with recoded factor variables
+# Save new linear data with recoded factor variables
 saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear.rds"))

### 2024-08-16 code/processing-code/processingfile-v1.qmd
Revisions for submission 3
Clean up of code

@@
 ```
-Now, I will create a new categorical variable of Normal Birthweight (NBW) for all birth weights greater than 2500 grams and Low Birthweight (LBW) for all birth weights equal to or less than 2500 grams for a future logistic regression.
+Now, I will create a new categorical variable of normal birth weight (NBW) for all birth weights $>2500$ g and low birth weight (LBW) for all birth weights $\leq2500$ g for a future logistic regression.
 This will come from the `Weightgrams` variable.

### 2024-08-16 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission 3
Clean up of code

@@
 # Introduction
+
 Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
-In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research, as LBW is associated with higher rates of neonatal and infant mortality and possible health problems over the life course [@cardoso2005; @paixao2021; @vilanova2019].
-BILL COMMENT: ARE THERE LONGITUDINAL STUDIES OF LATER-LIFE MORTALITY OR LIFE EXPECTANCY WITH BIRTH WEIGHT?
+In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research [@hughes2017; @silva1998; @wilcox2001].
+LBW can be a function of PTB, of intra-uterine growth retardation (IUGR) (otherwise known as small-for-gestational-age or SGA) or of both [@hughes2017].
 
-Contemporary studies in Brazil have shown that non-White mothers have higher rates of LBW [@nyarko2013; @wehby2015a; @vilanova2019].
-These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns have higher rates of LBW than White newborns, with trends worsening over time [@pollock2021].
-Racism is a major factor in these outcomes, as studies demonstrate that U.S.-born Black women have higher rates of LBW and PTB than Black immigrant women [@braveman2021a; @clay2023a; @david1997].
-BILL COMMENT: RELATED TO RACISM, WHAT DOES LITERATURE ON BIRTH WEIGHT SAY ABOUT ITS CAUSES?
-THESE ARE CAUSAL STATEMENTS.
+LBW is associated with higher rates of neonatal and infant mortality [@paixao2021; @vilanova2019].
+Longitudinal studies also have demonstrated an association between at-term LBW and increased risk for cardiovascular diseases and other chronic conditions over the life course [@barker2004].
+But birth weight in general and LBW in particular are often proxies for unmeasurable variables.
+Thus, these categories should not be understood as a causal mechanism of outcomes such as infant mortality or later-life chronic disease; rather, they provide important population-level information [@wilcox2001].
+
+Contemporary studies in Brazil have shown that non-White mothers have higher odds of giving birth to LBW infants [@nyarko2013; @wehby2015a; @vilanova2019].
+Non-White mothers also had infants with lower mean birthweights than their White counterparts [@leal2006].
+These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns are at higher risk for LBW than White newborns, with trends worsening over time [@pollock2021].
+Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997].
 
@@
 In the nineteenth century, population-level health was poor for all Brazilians, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from nutritional deficiencies [@karasch1987; @kiple1989].
+EXPAND
 
 In the immediate aftermath of the abolition of slavery, which coincided with the implementation of the country's first republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into economic, civic, or political life occurred [@carvalho2004].
-Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016; @roth2020].
+Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016].
 
@@
 
-By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @martins2004; @mott2002; @roth2020].
+By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @martins2004; @roth2020].
 Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
@@
 At the same time, longstanding racial hierarchies forged under racial slavery morphed into "scientific" ways to categorize humans based on skin color in post-abolition contexts.
-In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than restricting reproduction through sterilization [@schwarcz1993; @stepan1991].
-Although some eugenicists called for sterilization of citizens they deemed unfit, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013].
+In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than restricting reproduction through sterilization [@stepan1991].
+Although some eugenicists called for sterilization of citizens they deemed unfit, they never couched these discussions in explicitly racial terms [@wegner2013].
 Whiteness remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.
@@
 Laranjeiras provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
-Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants.
+Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe.
 Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
@@
 Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022].
-The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not "pathologically" different or part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].
+The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].
 
@@
 The project's Github repository, particularly the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting.
-All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online [@roth2024].
+All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online.
 
@@
 The outcome was birth weight measured as a continuous variable in grams for live births only.
-In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW (<p>&ge;</p>2500 g) or LBW (<2500 g), according to current WHO classifications) [@worldhealthorganization2022].
-BILL COMMENT: THESE COME FROM WHERE AND WHEN?
+In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
+The use of this cutoff in historical studies can be questionable as it was not even an established category in the 1920s.
+A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification.
+This was a period in which it was difficult for physicians to accurately calculate gestational age, and many physicians used birth weight as a proxy for prematurity.
+The WHO adopted this cutoff in 1948, and it has remained as the standard ever since [@hughes2017].
 
@@
 
-Except at the extremes, racial classification in Brazil was, and continues to be, imprecise [@travassos2004].
-Racial categories and skin color exist on a spectrum rather than a Black-White binary [@telles2006].
+Except at the extremes, racial classification in Brazil was, and continues to be, imprecise as racial categories and skin color classifications in Brazil exist on a spectrum rather than as a Black-White binary as in the United States [@travassos2004].
 Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category.
@@
 
-I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
-Because most results were not statistically significant in these models, I do not include adjusted models here.
-Unadjusted and adjusted logistic models and results are included in the supplemental analysis.
-BILL COMMENT: WHY WEREN'T MODELS SIGNIFICANT?
-
 I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race to see if there were differential outcomes for Black and mixed-race women.
@@
 The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers.
-A difference in means test shows borderline significance between the means (t-statistic = 1.9119, df.
+A difference in means test showed borderline significance (t-statistic = 1.9119, df.
 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).
@@
 Age and nationality were not significant.
-
-Unadjusted ORs (Table 4) demonstrate that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019).
-When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were not significant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
-MAKING CONTINUOUS VARIBLE CATEGORICAL.
-LBW NOT A CATEGORY YET PER WHO.
 
@@
 *Strengths and Limitations*
+
 This study provides heretofore unpublished hospital records on one of Brazil's most important public maternity hospitals in the early twentieth century, a hospital that remains in operation.
@@
 I assume that infants at risk for PTB were also at-risk for LBW.
+WE CANNOT MEASURE THE VARIABLE THAT IS REALLY DRIVING THINGS.
 
@@
 *Conclusion*
-This study has shown that birth weight differed by maternal skin color.
-This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms.
-These findings, although not generalizable beyond this specific hospital, are consistent with present-day studies, demonstrating how racial inequality in health outcomes has persisted over time.
+
+This study has shown that lower birth weights were associated with Afro-descended women giving birth in the Maternidade Laranjeiras in Rio de Janeiro, Brazil between 1922 and 1926.
+This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which I do not have direct evidence.
+These findings, although not generalizable beyond this specific hospital, are consistent with present-day studies on racialized disparities in birth weight in Rio de Janeiro, demonstrating how racial inequality in health outcomes has persisted over time.
 
@@
 ```
+
 {{< pagebreak >}}
@@
 #| label: tbl-table4
-#| tbl-cap: "Unadjusted Odds Ratios"
-#| echo: FALSE
-table4 <- readRDS(here::here("results", "tables", "table5_final.rds"))
-table4
-```
-{{< pagebreak >}}
-
-```{r}
-#| label: tbl-table5
 #| tbl-cap: "Mean Birth Weights"

### 2024-08-16 products/manuscript/supplement/Supplemental_Materials.qmd
Revisions for submission 3
Clean up of code

@@
     highlight-style: github
-    bibliography: ../../../assets/Mada_Bib.bib
-    csl: ../../../assets/chicago-fullnote-bibliography.csl
-    suppress-bibliography: true
+    bibliography: ../../assets/Mada_Bib.bib
+    csl: ../../assets/vancouver.csl
+    suppress-bibliography: false
 ---
@@
 
-If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder; 2) `eda-v1.qmd` or `edacode.R` in the `eda-code` folder; 3) `introanalyis-v1.qmd` or `introanalysis.R` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` or `fullanalysiscode.R` in the `analysis-code` folder.
-Please defer to all `.qmd` files as the master over any `.R` files.
+If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` in the `processing-code` folder; 2) `eda-v1.qmd` in the `eda-code` folder; 3) `introanalyis-v1.qmd` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` in the `analysis-code` folder.
+Older versions of code in `.r` are available in the `old-code` folder.
+This is not updated since 15 August 2024.
 
@@
 The GitHub repository for this project includes all relevant materials.
-All discussions of data import are detailed in the files `processingfile-v1.qmd` or `processingcode.R` in the `processing-code` folder.
+All discussions of data import are detailed in `processingfile-v1.qmd` in the `processing-code` folder.
 The original raw data, the codebook, and the processed data are available in the `data` folder.
@@
 
-For all three models, the r-squared is very small (0.007, 0.006, and 0.008), indicating that the model does not explain much of the variance in birth weight.
-This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.
-
 The fourth exploratory model again runs a multivariate linear regression, this time using skin color as it was differentiated in the original data, or with three categories (White, Black, and Mixed Race).
@@
 Yet, as Claudia Goldin and Robert A.
-Margo have shown for nineteenth-century Philadelphia, the inclusion of gestational age greatly improved the regression fit in their studies, demonstrating that gestational age could explain much of the variance of birth weight in this study [@goldin1989].
-Nonetheless, other scholars have shown that even though differential rates of PTB explain some of the variance in birth weight according to maternal skin color in the United States, up to 91 percent of influenced birth weight, as the lack of historical data precludes this line of inquiry.
-Rather, it is exploring if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro.
+Margo have shown for nineteenth-century Philadelphia, the inclusion of gestational age greatly improved the regression fit in their studies, demonstrating that gestational age could explain much of the variance of birth weight [@goldin1989].
+Nonetheless, other scholars have shown that even though differential rates of PTB explain some of the variance in birth weight according to maternal skin color in the United States, up to 91 percent of the Black-White gap in full-term birth weight remained unexplained [@costa2004].
+
+Here, I am exploring if maternal skin color is associated with lower infant birth weight during a historical time period in Rio de Janeiro, Brazil.
 Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color is also probably associated with premature birth, which is correlated to infant birth weight.
@@
 
-Now we will still consider `Birthweight` as the outcome of interest, but we will transform it into a categorical outcome of <=2500 (low birth weight, LBW) and >2500 (normal birth weight, NBW) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
+In my introductory analysis, I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
+
+I consider `Birthweight` as the outcome of interest, but I transformed it into a categorical outcome of NBW $>2500$ g or LBW $\leq2500$ g) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
+These cutoffs are according to the WHO [@worldhealthorganization2022].
 
@@
 
-The second model is also a simple logistic regression, this time using `Color` as the predictor variable (so three categories, Black, mixed race, and White).
+Model 5 demonstrates that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019).
+
+The second model is also a simple logistic regression, this time using `Color` as the predictor variable (Black, mixed race, and White).
 
@@
 ```
+
+When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were not significant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
 
@@
 ```
-In these two models, the only significant finding is the final logistic model, in which infants born to mixed-race mothers had 21% higher odds of being born with low birth weight compared to infants born to white mothers.
+
+In these two models, the only significant finding is from the final logistic model, in which infants born to mixed-race mothers had 21% higher odds of being born with low birth weight compared to infants born to white mothers.
 The other models did not show significant results.
@@
 
-Given that this study is an inferential, hypothesis-supporting approach that is exploring how specific maternal predictors affect the outcome of interest, birth weight, please see the `fullanalysis-v1.qmd` for all predictive metrics, where test/train and cross-validation were applied to test performance.
+Given that this study is an inferential, hypothesis-supporting approach that is exploring how specific maternal predictors affect the outcome of interest, birth weight, please see the `fullanalysis-v1.qmd` for all predictive metrics, where I applied test/train (75-25 split) and cross-validation to test performance.
 Model performance did not significantly improve with these additional tests.
-When run on testing data, most results were insignificant.
-
-Test/train (75-25 split) and cross-validation were performed on all models.
-Model prediction performance did not improve greatly using these tests.
-Model performance resulted in insignificant results when run on the testing data given the smaller sample size.
+When run on testing data, most results were insignificant, given the smaller sample size.
 
@@
 The performance metrics for the logistic models are available in the `introanalysis-v1.qmd` file.
+
+# Appendix
+Immigrant low birth weight [@clay2023]

### 2024-08-21 code/analysis-code/fullanalysis-v1.qmd
Revisions for submission 4
Incorporated collinearity models

@@
 ML_linear <- readRDS(data_location)
+str(ML_linear)
 ```
@@
 Performing cross-validation results in an RMSE that is the same as our training model (488), lower than our full model.
-The r-squared (0.015) is also higher than our training and full models (0.01), demonstrating better model performance.
+The r-squared (0.017) is also higher than our training and full models (0.01), demonstrating better model performance.
 
@@
 
-# Merging Model 1 and 2 into Table
+## Models 1 and 2 Sub-Samples
 
 ```{r}
-# Combine the two tables using tbl_merge
+# Model 1: Sub-sample simple model (ML_lm1_sub)
+# Create a list to store R-squared values
+r_squared_values_sub <- list()
+
+# Filter data based on desired conditions
+filtered_data <- ML_linear %>%
+  filter(Brazilian == "Brazilian")
+
+# Set reference level for "Euro-descent"
+ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm1_sub <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = filtered_data)
+
+# Summarize model fit with tidiers
+ML_lm1_sub_summary <- summary(ML_lm1_sub) # Summary of fit
+tidy(ML_lm1_sub, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+ML_lm1_sub_stat <- glance(ML_lm1_sub) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values_sub$ML_lm1_sub <- ML_lm1_sub_stat$r.squared
+
+# Create a gtsummary table directly from the model
+ML_lm1_sub_table <- 
+  tbl_regression(ML_lm1_sub, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
+ML_lm1_sub_table
+
+# Save regression output into table
+saveRDS(ML_lm1_sub_table, file = here::here("results", "tables", "ML_lm1_sub_table.rds"))
+```
+```{r}
+# Model 2: Sub-sample multivariate model (ML_lm2_sub)
+# Filter data based on desired conditions
+filtered_data <- ML_linear %>%
+  filter(Brazilian == "Brazilian")
+
+# Set reference level for "Euro-descent"
+ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm2_sub <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + + ModifiedStatus, data = filtered_data)
+
+# Summarize model fit with tidiers
+ML_lm2_sub_summary <- summary(ML_lm2_sub) # Summary of fit
+tidy(ML_lm2_sub, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+ML_lm2_sub_stat <- glance(ML_lm2_sub) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values_sub$ML_lm2_sub <- ML_lm2_sub_stat$r.squared
+
+# Create a gtsummary table directly from the model
+ML_lm2_sub_table <- 
+  tbl_regression(ML_lm2_sub, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
+ML_lm2_sub_table
+
+# Save regression output into table
+saveRDS(ML_lm2_sub_table, file = here::here("results", "tables", "ML_lm2_sub_table.rds"))
+```
+
+# Merging Model 1 and 2 and sub-sample 1 and 2 into Table
+
+```{r}
+# Combine the four tables using tbl_merge
 table3 <- tbl_merge(
-  list(table_model1, table_model2),
-  tab_spanner = c("**Model 1**", "**Model 2**")) %>%
+  list(table_model1, ML_lm1_sub_table, table_model2, ML_lm2_sub_table),
+  tab_spanner = c("**Model 1**", "**Sub-Model 1**", "**Model 2**", "**Sub-Model 2**")) %>%
   modify_table_body(fun = ~.x %>% arrange(variable))

### 2024-08-21 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 4
Incorporated collinearity models

@@
 So I will perform some subsample modeling to understand this issue.
-We will create a `Brazilian` (yes, no) variable and then run a regression with this variable.
-Then, I will drop all non-Brazilians from the study and run the original regression with the covariates.
+We will create a `Brazilian` (yes, no) variable, drop all non-Brazilians from the study, and run the original regression with the covariates.
 
@@
 # Create dummy variables for status and change to English
-ML_linear$Brazilian <- recode_factor(ML_linear$ModifiedNationality, "European" = "Non-Brazilian", "Middle Eastern" = "Non-Brazilian", "Latin American" = "Non-Brazilian", "Brazilian" = "Brazilian")
+ML_linear <- mutate(ML_linear, 
+                    Brazilian = recode_factor(ModifiedNationality, 
+                                              "European" = "Non-Brazilian",
+                                              "Middle Eastern" = "Non-Brazilian",
+                                              "Latin American" = "Non-Brazilian",
+                                              "Brazilian" = "Brazilian"))
 
+# Check the structure of the updated data frame
 str(ML_linear)
@@
 # Model 6 sub-sample simple linear model
-# Set reference level for Brazilian
+# Set reference level for "Euro-descent"
 ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

### 2024-08-21 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission 4
Incorporated collinearity models

@@
 ```
+
 # Abstract
+
 ## Background
+
 In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence.
@@
 ## Methods
-This cross-sectional analysis uses hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital.
+
+This cross-sectional analysis used hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital.
 I used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.
@@
 ## Findings
+
 African ancestry was correlated with lower birth weights.
@@
 ## Interpretations
+
 The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for women of color -- and possible epigenetic effects from the lived experience of racism.
@@
 ## Funding
+
 This archival research for this project was funded by the National Science Foundation (NSF) and the Fulbright IIE Student Program.
@@
 ## Variables
+
 ### Outcome
@@
 In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
-The use of this cutoff in historical studies can be questionable as it was not even an established category in the 1920s.
+We can question the use of this cutoff in historical studies, as it was not an established category until the mid-twentieth century.
 A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification.
-This was a period in which it was difficult for physicians to accurately calculate gestational age, and many physicians used birth weight as a proxy for prematurity.
-The WHO adopted this cutoff in 1948, and it has remained as the standard ever since [@hughes2017].
+This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity [@wilcox CITE CORRECTLY].
+The WHO adopted this cutoff in 1948, and it has remained the standard since [@hughes2017].
 
@@
 
-Except at the extremes, racial classification in Brazil was, and continues to be, imprecise as racial categories and skin color classifications in Brazil exist on a spectrum rather than as a Black-White binary as in the United States [@travassos2004].
+Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum rather than as a Black-White binary as in the United States [@travassos2004].
 Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category.
@@
 
-I calculated summary statistics for the sample, including sample proportions for maternal skin color, maternal ancestry, parity, maternal nationality, birth outcome, maternal outcome, fetal outcome, and fetal sex.
-I also calculated mean maternal age, infant birth length, and infant birth weight, as a whole and stratified by fetal sex and maternal skin color.
+I calculated summary statistics, including sample proportions for maternal skin color, maternal ancestry, parity, maternal nationality, birth outcome, maternal outcome, fetal outcome, and fetal sex.
+I also calculated mean maternal age, infant birth length, and infant birth weight, unadjusted and then stratified by first fetal sex and then maternal skin color.
 
@@
 I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
-Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age, as is best practice for disentangling PTB from LBW [@cutland2017].
+Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age.
 I discuss this constraint in the discussion below.
@@
 
+Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts, both of color and White [@franken2019; @ball2020].
+Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 in two steps.
+First, I created a new variable `Brazilian` (yes/no) and ran a simple linear analysis with birth weight as the dependent variable and maternal nationality as the independent variable.
+Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value <0.001).
+Given that all women of color in the data set were Brazilian, there is collinearity occurring.
+Thus, second, I dropped all non-Brazilians from the study, and ran the original regressions from Models 1 and 2.
+
 # Results
@@
 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).
+Regardless of a difference between infants born to mothers of different skin colors, all LBW infants in the past, the probability of surviving the neonatal period was lower than today given higher neonatal and infant mortality rates [@costa1998; @roth2020].
 
@@
 
+UPDATE FOR CORRELATION STUFF
+
 In models 1 and 2 (Table 2), women who were nulliparous (delivering their first child) had infants weighing on average 113 grams less than multiparous women.
@@
 Age and nationality were not significant.
+UPDATE
 
@@
 
+WHO: birth weight "is strongly conditioned by the health and nutritional status of the mother"
+
+WARD: "In addition, because birth weight is sensitive to environmental factors during the nine months of gestation (especially the final three months) as well as the long-range effects of changes in living conditions, it reflects both short- and long-term trends." (5)
+
+WARD: "By and large, the most privileged social groups bear the heaviest babies while the most disadvantaged have the lightest.
+These differences, in turn, reflect the fact that fetal growth and weight at birth are influenced by a wide range of biological, social, and economic factors, not all of which are fully understood." (10)
+
+WARD: From the standpoint of neonatal mortality the optimum weight of a newborn child lies between 3500 and 3900 grams.
+Infants bom outside this range have higher mortality levels than do those within it.
+The riSks are especially great among those weighing considerably less than the optimum.
+Estimates suggest that the chance of death during the first week of life increases by 40 percent, and the risk of fetal death by 25 percent, for every 100-gram decrease below the optimum level of weight.
+(10)
+
+Optimal birth weight is the weight at which birth weight-specific mortality is lowest
+
 Because the hospital records excluded information on maternal health factors, we don't know which contributed to racial disparities in infant birth weight.
@@
 
-All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014].
+All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014; @ward1993].
 In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
@@
 I assume that infants at risk for PTB were also at-risk for LBW.
-WE CANNOT MEASURE THE VARIABLE THAT IS REALLY DRIVING THINGS.
 
-Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study.
+Lack of data on maternal nutrition and health status or maternal work routine and income, all of which cannot be corrected for by using maternal skin color as a proxy, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study.
+Maternal skin color may serve as a partial index of the unobservable variables that are acting upon infant birth weight [@costa2004; @ward1993].
 Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.
+
+HOW INFANTS WERE WEIGHED AND ACCURACY OF SCALE
 

### 2024-09-03 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 5
clean up of code and text

@@
 library(tidymodels) #for modeling
+library(gtExtras) # for easy summary visualization
 rm(list=ls()) #clear the environment
@@
 
-After performing a Welch Two Sample t-test to compare the rates of LBW between Euro- and Afro-descent mothers, the results were as follows: t = `r round(t_statistic, 2)`, df = `r round(df, 1)`, p-value = `r round(p_value, digits = 3)`, mean difference = `r round(mean_diff, 4)`, 95% CI = [`r round(ci_lower, 4)`, `r round(ci_upper, 4)`].
+After performing a Welch Two Sample t-test to compare the rates of LBW between Euro- and Afro-descent mothers, the results were as follows: t = `r round(t_statistic, 2)`, df = `r round(df, 1)`, p-value = `r round(p_value, digits = 3)`, mean difference = `r round(mean_diff, 4)`, 95% CI = \[`r round(ci_lower, 4)`, `r round(ci_upper, 4)`\].
 
@@
 str(ML_linear)
+
+# Save ML_Linear as RDS
+save_data_location <- here::here("data","processed-data","ML_linear.rds")
+saveRDS(ML_linear, file = save_data_location)
 ```
@@
 ```
+
+# Predominant and residual distributions
+
+```{r}
+# Calculate the mean and standard deviation of the predominant distribution
+predominant_stats <- ML_summary %>% 
+   filter(!is.na(Weightgrams)) %>% #filter out NAs
+  summarize(
+    mean_weight = mean(Weightgrams),
+    sd_weight = sd(Weightgrams)
+  )
+
+# Determine the range of the predominant distribution
+predominant_lower <- predominant_stats$mean_weight - 2 * predominant_stats$sd_weight
+predominant_upper <- predominant_stats$mean_weight + 2 * predominant_stats$sd_weight
+
+# Calculate the proportion of observations in the predominant distribution
+predominant_proportion <- ML_summary %>%
+  filter(Weightgrams >= predominant_lower & Weightgrams <= predominant_upper) %>%
+  nrow() / nrow(ML_summary)
+
+# Calculate the proportion of observations in the residual distribution
+residual_proportion <- 1 - predominant_proportion
+
+# Print the results
+cat("Predominant distribution:\n")
+cat("Mean:", predominant_stats$mean_weight, "grams\n")
+cat("Standard deviation:", predominant_stats$sd_weight, "grams\n")
+cat("Proportion:", predominant_proportion, "\n\n")
+
+cat("Residual distribution:\n")
+cat("Proportion:", residual_proportion, "\n")
+
+# Plot the distribution
+residual_plot_all <- ML_summary %>% filter(!is.na(Weightgrams)) %>% #filter out NAs
+  ggplot(aes(x = Weightgrams)) +
+  geom_density(fill="#69b3a2", color="#e9ecef") +
+  stat_function(fun = dnorm, args = list(mean = predominant_stats$mean_weight, sd = predominant_stats$sd_weight), color = "black", linewidth = 0.5) +
+  labs(
+    title = "Birth Weight Distribution for All Births, Maternidade Laranjeiras, 1922-1926",
+    x = "Birth Weight (g)"
+  ) +
+  theme_ipsum() +
+  theme(plot.title = element_text(size=12))
+plot(residual_plot_all)
+
+# Save plot to file
+figure_file = here("results","figures","residual_plot_all.png")
+ggsave(filename = figure_file, plot=residual_plot_all, bg="white") #background is white
+
+# Calculate two distributions: Afro and Euro
+# Calculate the mean and standard deviation of the predominant distribution Afro
+predominant_stats_afro <- ML_summary %>% 
+   filter(!is.na(WeightgramsAfro)) %>% #filter out NAs
+  summarize(
+    mean_weight_afro = mean(WeightgramsAfro),
+    sd_weight_afro = sd(WeightgramsAfro)
+  )
+
+# Determine the range of the predominant distribution
+predominant_lower_afro <- predominant_stats_afro$mean_weight_afro - 2 * predominant_stats_afro$sd_weight_afro
+predominant_upper_afro <- predominant_stats_afro$mean_weight_afro + 2 * predominant_stats_afro$sd_weight_afro
+
+# Calculate the proportion of observations in the predominant distribution
+predominant_proportion_afro <- ML_summary %>%
+  filter(WeightgramsAfro >= predominant_lower_afro & WeightgramsAfro <= predominant_upper_afro) %>%
+  nrow() / nrow(ML_summary)
+
+# Calculate the proportion of observations in the residual distribution
+residual_proportion_afro <- 1 - predominant_proportion_afro
+
+# Print the results
+cat("Predominant distribution:\n")
+cat("Mean:", predominant_stats_afro$mean_weight_afro, "grams\n")
+cat("Standard deviation:", predominant_stats_afro$sd_weight_afro, "grams\n")
+cat("Proportion:", predominant_proportion_afro, "\n\n")
+
+cat("Residual distribution:\n")
+cat("Proportion:", residual_proportion_afro, "\n")
+
+# Calculate the mean and standard deviation of the predominant distribution Euro
+predominant_stats_euro <- ML_summary %>% 
+   filter(!is.na(WeightgramsEuro)) %>% #filter out NAs
+  summarize(
+    mean_weight_euro = mean(WeightgramsEuro),
+    sd_weight_euro = sd(WeightgramsEuro)
+  )
+
+# Determine the range of the predominant distribution
+predominant_lower_euro <- predominant_stats_euro$mean_weight_euro - 2 * predominant_stats_euro$sd_weight_euro
+predominant_upper_euro <- predominant_stats_euro$mean_weight_euro + 2 * predominant_stats_euro$sd_weight_euro
+
+# Calculate the proportion of observations in the predominant distribution
+predominant_proportion_euro <- ML_summary %>%
+  filter(WeightgramsEuro >= predominant_lower_euro & WeightgramsEuro <= predominant_upper_euro) %>%
+  nrow() / nrow(ML_summary)
+
+# Calculate the proportion of observations in the residual distribution
+residual_proportion_euro <- 1 - predominant_proportion_euro
+
+# Print the results
+cat("Predominant distribution:\n")
+cat("Mean:", predominant_stats_euro$mean_weight_euro, "grams\n")
+cat("Standard deviation:", predominant_stats_euro$sd_weight_euro, "grams\n")
+cat("Proportion:", predominant_proportion_euro, "\n\n")
+
+cat("Residual distribution:\n")
+cat("Proportion:", residual_proportion_euro, "\n")
+
+# Plot both distributions
+residual_plot_divided <- ML_summary %>% 
+  filter(!is.na(WeightgramsAfro) | !is.na(WeightgramsEuro)) %>% #filter out NAs
+  ggplot() +
+  geom_density(aes(x = WeightgramsAfro), fill = "#69b3a2", alpha = 0.5, color = NA) +
+  geom_density(aes(x = WeightgramsEuro), fill = "#404080", alpha = 0.5, color = NA) +
+  stat_function(fun = dnorm, 
+        args = list(mean = predominant_stats_afro$mean_weight_afro,
+                sd = predominant_stats_afro$sd_weight_afro),
+                color = "black", linewidth = 0.5) +
+  stat_function(fun = dnorm, 
+                args = list(mean = predominant_stats_euro$mean_weight_euro, 
+                            sd = predominant_stats_euro$sd_weight_euro), 
+                color = "black", linewidth = 0.5) +
+  labs(
+    title = "Birth Weight Distribution for Afro- and Euro-Descended Infants",
+    subtitle = "Maternidade Laranjeiras, 1922-1926",
+    x = "Birth Weight (g)",
+    fill = "Infant Descent"
+  ) +
+  scale_x_continuous(limits = c(1000, 5000), breaks = seq(1000, 5000, by = 1000)) +
+  theme_ipsum() +
+  theme(plot.title = element_text(size=12),
+        plot.subtitle = element_text(size=10),
+        legend.position = "bottom")
+plot(residual_plot_divided)
+
+# Save plot to file
+figure_file = here("results","figures","residual_plot_divided.png")
+ggsave(filename = figure_file, plot=residual_plot_divided, bg="white") #background is white
+```

### 2024-09-03 code/eda-code/eda-v1.qmd
Revisions for submission 5
clean up of code and text

@@
 #Path to summary data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","ML_summary_processed.rds")
+data_location <- here::here("data","processed-data","ML_summary.rds")
 
@@
 
-Using these new dummy variables, I create new dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color.
+Using these new dummy variables, I created new dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color.
 These are `Afro_NBW` and `Afro_LBW`; and `Euro_NBW` and `Euro_LBW`.
@@
     WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
-    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
+    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_),
+    WeightgramsAfro = if_else(ModifiedColor == "Afro-Descent", Weightgrams, NA_real_),
+    WeightgramsEuro = if_else(ModifiedColor == "Euro-Descent", Weightgrams, NA_real_)
   )
@@
 ```
-Let's also change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.
+I will now change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.
 
@@
 
-I have now learned that `gtsummary` is a very easy way to create summary statistics (Table 1).
+`Gtsummary` is a very easy way to create summary statistics (Table 1).
 In the note section at the end of this document, I also include the code I worked on to create the tables manually to show my learning process.
@@
     WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
+    WeightgramsAfro = if_else(ModifiedColor == "Afro-Descent", Weightgrams, NA_real_),
+    WeightgramsEuro = if_else(ModifiedColor == "Euro-Descent", Weightgrams, NA_real_),
     WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
@@
   tbl_summary(
-    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale, BirthweightCategory),
+    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsAfro, WeightgramsEuro, WeightgramsFemale, WeightgramsMale, BirthweightCategory),
     missing = "no",
     label = list(
-      Color ~ "Color (n = 2,695)",
-      ModifiedColor ~ "Ancestry (n = 2,695)", 
-      ModifiedStatus ~ "Parity (n = 2,836)", 
-      Age ~ "Maternal Age (n = 2,783)", 
-      ModifiedNationality ~ "Combined Nationality (n = 2,773)", 
-      Birth ~ "Birth Outcome (n = 2,761)", 
-      MaternalOutcome ~ "Maternal Outcome (n = 2,829)", 
-      FetalOutcome ~ "Fetal Outcome (n = 2,666)", 
-      Sex ~ "Sex (n = 2,534)",
-      Lengthcentimeters ~ "Birth Length (cm) (n = 2,405)", 
-      Weightgrams ~ "Birth Weight (g) (n = 2,384)",
-      WeightgramsBlack ~ "Black (g) (n = 643)",
-      WeightgramsMixed ~ "Mixed-Race (g) (n = 675)",
-      WeightgramsWhite ~ "White (g) (n = 950)",
-      WeightgramsFemale ~ "Female (g) (n = 1,074)",
-      WeightgramsMale ~ "Male (g) (n = 1,270)",
-      BirthweightCategory ~ "Birth Weight Category (n = 2,231)"
+      Color ~ "Color",
+      ModifiedColor ~ "Ancestry", 
+      ModifiedStatus ~ "Parity", 
+      Age ~ "Maternal Age", 
+      ModifiedNationality ~ "Combined Nationality", 
+      Birth ~ "Birth Outcome", 
+      MaternalOutcome ~ "Maternal Outcome", 
+      FetalOutcome ~ "Fetal Outcome", 
+      Sex ~ "Sex",
+      Lengthcentimeters ~ "Birth Length (cm)", 
+      Weightgrams ~ "Birth Weight (g)",
+      WeightgramsBlack ~ "Black",
+      WeightgramsMixed ~ "Mixed-Race",
+      WeightgramsWhite ~ "White",
+      WeightgramsAfro ~ "Afro-Descent",
+      WeightgramsEuro ~ "Euro-Descent",
+      WeightgramsFemale ~ "Female",
+      WeightgramsMale ~ "Male",
+      BirthweightCategory ~ "Birth Weight Category"
     ),
@@
   ) %>%
-  #add_n() #%>%
+  add_n() %>%
   modify_table_body( #indent these three rows
@@
         variable == "WeightgramsMixed" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsAfro" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsEuro" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
         variable == "WeightgramsFemale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
@@
 #Path to summary data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")
+data_location <- here::here("data","processed-data","Mean_Birthweights.rds")
 
@@
 
+# ML Linear Data
+
 Because I am working on a different subset of the data, `ML_linear` for the statistical analysis, I will also need to recategorize variables here.
@@
 # Create dummy variables for color and change to English using recode factor from dplyr
-ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Preta" = "Afro-Descent", "Parda" = "Afro-Descent", "Branca" = "Euro-Descent")
+ML_linear$ModifiedColor <- factor(ifelse(ML_linear$Color %in% c("Black", "Mixed Race"), "Afro-Descent", "Euro-Descent"))
 
@@
 # Create dummy variables for nationality and change to English
-ML_linear$ModifiedNationality <- recode_factor(ML_linear$Nationality, "Alema" = "European", "Argentina" = "Latin American", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "Latin American", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "Middle Eastern", "Uruguaya" = "Latin American")
+ML_linear$ModifiedNationality <- factor(case_when(
+  ML_linear$Nationality %in% c("German", "Austrian", "Spanish", "French", "Italian", "Polish", "Portuguese", "Romanian", "Russian", "Swiss") ~ "European",
+  ML_linear$Nationality %in% c("Argentine", "Paraguayan", "Uruguayan") ~ "Latin American",
+  ML_linear$Nationality == "Brazilian" ~ "Brazilian",
+  ML_linear$Nationality == "Syrian" ~ "Middle Eastern"
+))
 

### 2024-09-03 code/processing-code/processingfile-v1.qmd
Revisions for submission 5
clean up of code and text

@@
 # Save ML_Linear as RDS
-save_data_location <- here::here("data","processed-data","ML_linear_processed.rds")
+save_data_location <- here::here("data","processed-data","ML_linear.rds")
 saveRDS(ML_linear, file = save_data_location)
@@
 # Save ML_summary as RDS
-save_data_location <- here::here("data","processed-data","ML_summary_processed.rds")
+save_data_location <- here::here("data","processed-data","ML_summary.rds")
 saveRDS(ML_summary, file = save_data_location)

### 2024-09-03 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission 5
clean up of code and text

@@
 In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence.
-After abolition, the state did not create programs to incorporate people formerly enslaved and their descendants into economic, civil, or political life.
+After abolition, the state did not create programs to incorporate formerly enslaved people and their descendants into economic, civil, or political life.
 Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century.
@@
 This cross-sectional analysis used hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital.
-I used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.
+It used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.
 
@@
 
+Birth weight and its relationship to infant health have been a focus of public health research for over a century [@hughes2017].
+In epidemiology, both low birth weight (LBW) and preterm birth (PTB) within and across racialized communities is a focal point of research [@braveman2021a; @david1997; @hughes2017; @silva1998; @wilcox2001].
+LBW can be a function of PTB, of fetal growth restriction (FGR, previously known as intra-uterine growth restriction or IUGR, which can include but is not limited to small-for-gestational-age or SGA infants -- those born at less than 10 percentile), or of both [@wilcox2001; @wilcox2024].
+Nonetheless, PTB is the primary determinant of both LBW and lower birth weight in general [@maruyama2020].
+
+Multiple social, economic, biological, and genetic factors influence birth weight, which is also partially conditioned by maternal nutrition, health, and health care utilization during pregnancy [@braveman2021a; @david1997; @ward1993].
+Recent genetic studies have detailed the specific alleles through which both the maternal and fetal genomes contribute to birth weight and gestational length, while acknowledging the continued importance of non-genetic factors [@juliusdottir2021; @sole-navais2023; @warrington2019].
+Given that these studies are based on well-off populations in countries with low neonatal and infant mortality rates, they might overestimate genetic impact on birth weight for historical studies of impoverished populations.
+
+Although not a causal mechanism, LBW is associated with higher rates of neonatal and infant mortality [@paixao2021; @vilanova2019].
+Originally, epidemiologists hypothesized an association between at-term LBW and increased risk for cardiovascular diseases and other chronic physical health conditions over the life course [@barker2004].
+Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest this.
+For example, Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure.
+Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.
+
+Historians have shown how birth weight can be an indicator for the often unmeasurable nutritional status of mothers of different social classes and racial classifications in the past [@goldin1989; @costa1998; @costa2004; @ward1993].
+Historically, women from more privileged groups had heavier infants than women from less advantaged groups [@stein1984; @ward1993].
+
+The differences between groups both today and in the past have given rise to much research stratified by racial group.
 Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
-In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research [@hughes2017; @silva1998; @wilcox2001].
-LBW can be a function of PTB, of intra-uterine growth retardation (IUGR) (otherwise known as small-for-gestational-age or SGA) or of both [@hughes2017].
-
-LBW is associated with higher rates of neonatal and infant mortality [@paixao2021; @vilanova2019].
-Longitudinal studies also have demonstrated an association between at-term LBW and increased risk for cardiovascular diseases and other chronic conditions over the life course [@barker2004].
-But birth weight in general and LBW in particular are often proxies for unmeasurable variables.
-Thus, these categories should not be understood as a causal mechanism of outcomes such as infant mortality or later-life chronic disease; rather, they provide important population-level information [@wilcox2001].
-
+Thus, racism and racist policies also can influence birth outcomes.
 Contemporary studies in Brazil have shown that non-White mothers have higher odds of giving birth to LBW infants [@nyarko2013; @wehby2015a; @vilanova2019].
-Non-White mothers also had infants with lower mean birthweights than their White counterparts [@leal2006].
+Non-White mothers also had infants with lower mean birth weights than their White counterparts [@leal2006].
 These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns are at higher risk for LBW than White newborns, with trends worsening over time [@pollock2021].
 Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997].
+Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].
 
@@
 Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere.
-In the nineteenth century, population-level health was poor for all Brazilians, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from nutritional deficiencies [@karasch1987; @kiple1989].
-EXPAND
+In the nineteenth century, population-level health was poor for all Brazilians, but particularly so for enslaved individuals, who experienced the epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever that affected the entire population [@karasch1987].
+Although caloric intake for the enslaved population was probably sufficient in terms of quantity, excluding pregnant and nursing women, scholars have shown how it lacked key nutrients, and thus the enslaved suffered from nutritional deficiencies, including thiamine deficiency, which was associated with high infant mortality rates [@karasch1987; @kiple1989].
+Today, thiamine deficiency is correlated with lower birth weight and adverse perinatal outcomes including neonatal death [@gr2024; @kareem2023].
 
@@
 
-By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @martins2004; @roth2020].
-Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
-There, obstetricians and gynecologists began implementing surgical advances in labor and delivery [@martins2004; @roth2021].
-Underlying these structural and technological changes was an ideology scholars have described as "scientific motherhood," supported by both physicians and elite women, which relied on a technocratic model towards pregnancy and motherhood and, in doing so, essentialized notions of gender and foregrounded physicians’ scientific authority [@freire2009; @martins2004; @otovo2016].
-
-At the same time, longstanding racial hierarchies forged under racial slavery morphed into "scientific" ways to categorize humans based on skin color in post-abolition contexts.
-In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than restricting reproduction through sterilization [@stepan1991].
-Although some eugenicists called for sterilization of citizens they deemed unfit, they never couched these discussions in explicitly racial terms [@wegner2013].
-Whiteness remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.
-
-Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces.
-Laranjeiras provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
+By the 1920s, in Rio de Janeiro public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @roth2020].
+Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
 Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe.
@@
 
-BILL COMMENT: I AM CURIOUS, DO WHITE SOUTH AFRICANS HAVE HIGHER BIRTH WEIGHT, CONTROLLING FOR INCOME? ANY STUDY OF THIS?
-
-This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993].
+This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights in major U.S.
+and Western European cities [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993].
 It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].
@@
 It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
-BILL COMMENT: PUT THIS PARAGRAPH MUCH EARLIER.
 
@@
 When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a].
-BILL COMMENT: FOR OBSTETRICIANS WHO APPEAR MORE THAN ONCE, IS THERE A RELATIONSHIP WITH BIRTH WEIGHT?
 
@@
 In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
-We can question the use of this cutoff in historical studies, as it was not an established category until the mid-twentieth century.
-A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification.
-This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity [@wilcox CITE CORRECTLY].
-The WHO adopted this cutoff in 1948, and it has remained the standard since [@hughes2017].
+Wilcox (@wilcox2001) has questioned the utility of the LBW category in epidemiological studies given its arbitrary nature.
+We can question the use of this cutoff in historical studies as well, as it was not an established category until the mid-twentieth century.
+A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification [@hughes2017].
+This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity.
+The WHO adopted this cutoff in 1948, and it has remained the standard ever since even as epidemiologists began to differentiate between small-for-gestational age (SGA, now FGR) and PTB in the 1970s [@hughes2017; @wilcox2001].
+Physicians in early twentieth-century Rio de Janeiro defined low birth weight as BLANK [@conrado1924a].
 
@@
 I recategorized the latter as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian).
+For a sub-sample analysis to test collinearity between nationality and skin color, I further created a dummy category between `Brazilian` and `non-Brazilian`.
 
@@
 A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
-The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009].
+The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009; @ward1993].
 But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
@@
 
-Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts, both of color and White [@franken2019; @ball2020].
-Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 in two steps.
-First, I created a new variable `Brazilian` (yes/no) and ran a simple linear analysis with birth weight as the dependent variable and maternal nationality as the independent variable.
+Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts of all skin colors [@franken2019; @ball2020].
+Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 (modified ancestry) in two steps.
+First, I created a new dichotomous variable, `Brazilian` (yes/no) and ran a simple linear analysis with birth weight as the outcome and maternal nationality as the independent variable.
 Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value <0.001).
-Given that all women of color in the data set were Brazilian, there is collinearity occurring.
-Thus, second, I dropped all non-Brazilians from the study, and ran the original regressions from Models 1 and 2.
+Given that all women of color in the data set were Brazilian, collinearity exists.
+Second, I dropped all non-Brazilians from the study, and ran the original simple and multiple regressions from Models 1 and 2.
 
@@
 The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers.
-A difference in means test showed borderline significance (t-statistic = 1.9119, df.
+A difference in means test showed borderline significance at the alpha 0.05 level (t-statistic = 1.9119, df.
 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).
-Regardless of a difference between infants born to mothers of different skin colors, all LBW infants in the past, the probability of surviving the neonatal period was lower than today given higher neonatal and infant mortality rates [@costa1998; @roth2020].
 
 Results for all linear models are displayed in Tables 2 and 3.
-All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race).
+All models demonstrated an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race).
 When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter than their Euro-descended counterparts (Table 2).
@@
 Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent.
-AT THIS POINT I NEED TO SEE A LITERATURE THAT TELLS ME 100 GRAMS MATTERS FOR SOMETHING, HEALTH-WISE, IS 100 GRAMS MORE THAN 2 STANDARD DEVIATIONS?
+Research has shown that below 3000 g, risk of death in the first week of life increases by 40% for every 100 g decrease in birth weight [@stein1984; @wilcox1983].
+More recent studies relying on IV have found that a 10% increase in birth weight decreases infant mortality by 13.3 deaths per 1000 live births [@maruyama2020].
+Applied to these data, 85 g difference in birth weight between Afro- and Euro-descended infants (sub-model 2) would have increased infant mortality among Afro-descended children by 3.7 deaths per 1000 births.
 
-UPDATE FOR CORRELATION STUFF
-
-In models 1 and 2 (Table 2), women who were nulliparous (delivering their first child) had infants weighing on average 113 grams less than multiparous women.
-In models 3 and 4 (Table 3), the average was 114 grams lighter.
-This is an association found both in other historical studies and today [@costa1998; @souza2020].
+In all models (Tables 2 and 3), women who were nulliparous (delivering their first child) had lighter infants weighing than multiparous women, an association recognized by physicians at the time and found in both other historical studies and today [@conrado1924a; @costa1998; @souza2020].
 Age and nationality were not significant.
-UPDATE
 
@@
 
-This study found that women of African descent in early twentieth-century Brazil both had infants with lower birth weights and higher rates of LBW than their White counterparts.
-When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today.
-Although the mean birth weight was still within today's normal range, the differences between racialized groups demonstrate differential health outcomes that have persisted until today.
+This study found that women of African descent in early twentieth-century Brazil both had infants with lower birth weights and higher rates of LBW than their White counterparts, perhaps indicative of higher rates of PTB.
+Regardless of a difference between infants born to mothers of different skin colors, for all LBW infants in the past, the probability of surviving the neonatal period was lower than today given higher neonatal and infant mortality rates and less effective medical interventions [@costa1998; @roth2020].
+This fact is born out by more recent time-series analysis of neonatal survival [@maruyama2020].
 
 Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight across the Americas and in New Zealand.
-
-Black infants born at Laranjeiras had the lowest mean birth weight (3037 grams) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004].
-Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period.
-They were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
-White infants born at Laranjeiras had a mean birth weight of 3133 grams, slightly more than 50 grams lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
-They were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 grams).
+Black infants born at Laranjeiras had the lowest mean birth weight (3037 g) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004].
+Black infants born at Laranjeiras were almost 150 g lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period.
+They were less than 100 g lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
+White infants born at Laranjeiras had a mean birth weight of 3133 g, slightly more than 50 g lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
+They were nearly 300 g lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 g).
 As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same historical period [@costa1998; @costa2004; @goldin1989].
 
-What can explain the differences in birth weight both historically and today?
+When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today.
+Although the mean birth weight at Laranjeiras was within today's normal range, the differences between racialized groups demonstrate differential health outcomes that continue to persist.
+Across populations, relatively high mean birth weights in the nineteenth and early twentieth centuries are probably an artifact of high stillbirth and neonatality rates [@costa1998; @costa2004].
+Today, PTB and lower birth weight infants that probably would have died in utero 100 years ago are now surviving, bringing the mean birth weight down.
+As our estimates of PTB demonstrate, the difference in mean birth weight is probably a reflection of higher rates of PTB among African-descended women.
+
+What can explain the differences in birth weight historically both between all women at Laranjeiras and those of other historical populations and among different skin-color groups within Laranjeiras?
 The intrauterine environment is a critical determinant of infant birth weight.
-Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care can all influence birth weight.
-Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986].
-Other factors include maternal age, parity, birth spacing, and gestational age.
-We can imagine that poorer women in early twentieth-century Rio de Janeiro performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
+Maternal age, parity, birth spacing, and gestational age all condition birth weight.
+In this study, maternal age was not a significant factor, although multiparous mothers had infants with higher birth weights, as is common today.
+The data did not provide information on birth spacing or gestational age, the latter which I discuss below in the limitations.
+
+What about external factors related to poverty?
+Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care also influence birth weight [@stein1984].
+Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986; @stein1984].
+Yet nutritional deficiencies that were found among African-descended populations in nineteenth-century Rio de Janeiro due to slavery are associated with lower birth weights and infant mortality [@karasch1987; @kiple1989; @gr2024; @kareem2023].
+Although working women of all skin colors had more control over their diets than did the enslaved, poor and working-class Cariocas (as residents of Rio de Janeiro are called) continued to rely heavily on manioc and dried meat, both deficient in thiamine.
+[CITE] We know that poorer women in early twentieth-century Rio de Janeiro also performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
+And working-class women of all skin colors often worked through the third trimester as domestic servants [@roth2020].
 These same women might have lived in crowded housing arrangements with lack of access to running water [@graham1992; @fischer2008; @roth2020].
-Generalized poverty might have influenced both maternal nutrition and the ability to access prenatal care.
-
-WHO: birth weight "is strongly conditioned by the health and nutritional status of the mother"
-
-WARD: "In addition, because birth weight is sensitive to environmental factors during the nine months of gestation (especially the final three months) as well as the long-range effects of changes in living conditions, it reflects both short- and long-term trends." (5)
-
-WARD: "By and large, the most privileged social groups bear the heaviest babies while the most disadvantaged have the lightest.
-These differences, in turn, reflect the fact that fetal growth and weight at birth are influenced by a wide range of biological, social, and economic factors, not all of which are fully understood." (10)
-
-WARD: From the standpoint of neonatal mortality the optimum weight of a newborn child lies between 3500 and 3900 grams.
-Infants bom outside this range have higher mortality levels than do those within it.
-The riSks are especially great among those weighing considerably less than the optimum.
-Estimates suggest that the chance of death during the first week of life increases by 40 percent, and the risk of fetal death by 25 percent, for every 100-gram decrease below the optimum level of weight.
-(10)
-
-Optimal birth weight is the weight at which birth weight-specific mortality is lowest
+Thus, generalized poverty might have influenced women of African descent's nutrition and ability to access prenatal care.
 
@@
 All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014; @ward1993].
+Physicians could have introduced systematic measurement error when weighing newborns in Laranjeiras, either through inconsistent practices or inaccurate equipment.
+The hospital records do not provide information on weighing practices, but published observations from medical students who trained at Laranjeiras show that the norm was to weigh all newborns without clothes on a calibrated scale [@conrado1924a].
+Thus, an upward bias to the mean weight through the inclusion of clothing or blankets for all infants seems unlikely as does any systematic bias according to infant skin color in any direction.
+It seems unlikely that physicians weighed infants of one skin color in a different method than those of another given that this was the premier teaching hospital associated with one of Brazil's only medical schools.
+Comparing the mean weights to those of other historical data sets (Table 4) shows that mean weights at Laranjeiras were lower than all others except for the estimated weights of slave infants in the antebellum southern United States, showing that if any upward bias existed, weights were even lower than presented here.
+Additionally, birth weights at Laranjeiras were rounded to the nearest 50 grams.
+If there was a systematic bias in any direction for any group of infants, it would have had to be at the hospital level since multiple physicians and medical students weighed babies over the five-year period.
+The theoretical upper bound on bias is 50 grams, and my results show a difference greater than 50 g.
+Claudia Goldin [@goldin1989] has shown for Philadelphia that heaping did not systematically bias the mean, and I assume the same here.
+
 In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
-Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age, SGA) and pre-term infants, or those born before 37 weeks' gestation.
+Thus, the birth weight variable captured both smaller babies born at term, whether with or without FGR, and pre-term infants (PTB), or those born before 37 weeks' gestation.
 We are thus unable to know how many of these infants were low birth weight at term [@falcao2020].
-This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@cutland2017].
-Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989].
-I assume that infants at risk for PTB were also at-risk for LBW.
+This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@wilcox2001; @wilcox2024].
+Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have data sets that include gestational age [@costa1998; @costa2004; @goldin1989].
+Here, I assume that infants at risk for PTB were also at-risk for lower birth weight.
+
+Today, lower mean birth weights between populations are not necessarily associated with worse health outcomes.
+In fact, Wilcox [@wilcox2001, p.
+1236] has shown that "populations with lighter babies do not necessarily have worse mortality." Yet continued differences between racialized groups hints at the possible role of socio-economic inequity and racism in determining health outcomes.
+In addition, studies relying on instrumental variable (IV) analysis have shown that even in the recent past in developed countries, the impact of birth weight on adverse perinatal and neonatal outcomes has declined over the past 50 years given advances in medical care [@maruyama2020].
+We can imagine that 100 years ago, the association of birth weight, most likely as a proxy for PTB, with adverse neonatal outcomes was even greater.
+Finally, birth weights studied historically are as important for understanding their mother's past lives as for predicting perinatal or neonatal outcomes [@ward1993].
 
@@
 Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.
-
-HOW INFANTS WERE WEIGHED AND ACCURACY OF SCALE
 

### 2024-09-06 code/analysis-code/fullanalysis-v1.qmd
Final Revisions for Submission
Cleaned up all files

@@
 ```{r, message=FALSE,  warning=FALSE}
-#Path to summary data. Note the use of the here() package and not absolute paths
+# Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear.rds")
-#load data
+
+# Load data
 ML_linear <- readRDS(data_location)
@@
 
-First, I will create train and test data (75/25) to be able to evaluate the performance of our models.
+First, I will create train and test data (75/25) to be able to evaluate the performance of the models.
 I will also set the seed to ensure reproducibility.
@@
 
-whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
+whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
 
@@
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Model 1: Final model for paper using all data
+# Create a list to store R-squared values
+r2_1 <- list()
+rmse_1 <- list()
+
 # Model using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit1_pred <- bind_cols(ML_fit1_pred, ML_linear %>% select(Weightgrams))
@@
 
-#Print table
+# Print table
 table_model1
 
+# Extract RMSE and R2 for in-line text
+r2_1$model1 <- metrics_fit1[metrics_fit1$.metric == "rsq", ".estimate"]
+rmse_1$model1 <- metrics_fit1[metrics_fit1$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
 saveRDS(table_model1, file = here::here("results", "tables", "table_model1_final.rds"))
-```
+
+# Create a gtsummary table directly from the model with CIs
+table_model1_CI <-
+  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
+table_model1_CI
+```
+
+The adjusted R-squared is `r round(r2_1$model1, 3)`, and the RMSE is `r round(rmse_1$model1)`.
+
 ## Model 1 Performance
 
-First we will evaluate the model on our training data.
-
-```{r}
+First, I will evaluate the model on the training data.
+
+```{r, message=FALSE,  warning=FALSE}
 # Training model 1
+# Create a list to store R-squared and RMSE values
+r2_1_train <- list()
+rmse_1_train <- list()
+
 # Model using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit1_pred_train <- bind_cols(ML_fit1_pred_train, train_data %>% select(Weightgrams))
@@
 
+# Extract R-squared and RMSE values for in-line text (because data frame with columns not list)
+r2_1_train$model1_train <- metrics_fit1_train[metrics_fit1_train$.metric == "rsq", ".estimate"]
+rmse_1_train$model1_train <- metrics_fit1_train[metrics_fit1_train$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 ```
-Our results here are not that different from the model run on all the data.
-Our adjusted R-squared is 0.01, the same as the full model.
-The RMSE is 488, which is slightly lower than the full model (494), which shows better performance.
-
-Now let's perform cross-validation on the training data.
+
+The results are similar to those from the model run on all the data.
+The adjusted R-squared is `r round(r2_1_train$model1_train, 3)`, the same as the full model.
+The RMSE is `r round(rmse_1_train$model1_train)`, which is slightly lower than the full model, `r round(rmse_1$model1)`, which shows better performance.
+
+Now, I will perform cross-validation on the training data.
 
@@
 folds
+
+# Create a list to store R-squared and RMSE values
+r2_1_cv <- list()
+rmse_1_cv <- list()
 
@@
 
+# Extract R-squared and RMSE values for in-line text
+# For R-squared
+r2_1_cv$model1_cv <- ML_cv1_metrics %>%
+  filter(.metric == "rsq") %>%
+  pull(mean)
+
+# For RMSE
+rmse_1_cv$model1_cv <- ML_cv1_metrics %>%
+  filter(.metric == "rmse") %>%
+  pull(mean)
+
 # Save regression output into table
@@
 ```
-Performing cross-validation results in an RMSE that is the same as our training model (488), lower than our full model.
-The r-squared (0.017) is also higher than our training and full models (0.01), demonstrating better model performance.
-
-Finally, let's run the model on our test data.
+Performing cross-validation results in an RMSE that is the same as the training model `r round(rmse_1_cv$model1_cv)`, lower than the full model.
+The r-squared `r round(r2_1_cv$model1_cv, 3)` is also higher than the training `r round(r2_1_train$model1_train, 3)` and full models `r round(r2_1$model1, 3)`, demonstrating better model performance.
+
+Finally, I'll run the model on the test data.
 
@@
 # Testing model 1
+# Create a list to store R-squared and RMSE values
+r2_1_test <- list()
+rmse_1_test <- list()
+
 # Model using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit1_pred_test <- bind_cols(ML_fit1_pred_test, test_data %>% select(Weightgrams))
@@
 
+# Extract R-squared and RMSE values for in-line text (because data frame with columns not list)
+r2_1_test$model1_test<- metrics_fit1_test[metrics_fit1_test$.metric == "rsq", ".estimate"]
+rmse_1_test$model1_test <- metrics_fit1_test[metrics_fit1_test$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 
-When running the model on our testing dataset, our results remain significant at the 0.05 level. The RMSE is 508 and the r-squared is 0.008. This shows a slightly worse fit than our training results.
+When running the model on the testing dataset, the results remain significant at the 0.05 level.
+The RMSE is `r round(rmse_1_test$model1_test)` and the r-squared is `r round(r2_1_test$model1_test, 3)`.
+This shows a slightly worse fit than the training results.
 
@@
 
-whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
+whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
 Results are displayed in Table 3.
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Model 2: Final model for paper using all data
+# Create a list to store R-squared values
+r2_2 <- list()
+rmse_2 <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit2_pred <- bind_cols(ML_fit2_pred, ML_linear %>% select(Weightgrams))
@@
 
+# Extract RMSE and R2 for in-line text
+r2_2$model2 <- metrics_fit2[metrics_fit2$.metric == "rsq", ".estimate"]
+rmse_2$model2 <- metrics_fit2[metrics_fit2$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
 saveRDS(table_model2, file = here::here("results", "tables", "table_model2_final.rds"))
-```
-Our RMSE is 490 and our r-squared is 0.026.
+
+# Create a gtsummary table directly from the model with CIs
+table_model2_CI <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
+table_model2_CI
+```
+
+The RMSE is `r round(rmse_2$model2)` and the r-squared is `r round(r2_2$model2, 3)`.
 
@@
 
-Now let's run the model on our training data.
-
-```{r}
+Now, I'll run the model on the training data.
+
+```{r, message=FALSE,  warning=FALSE}
 # Training model 2
+# Create a list to store R-squared and RMSE values
+r2_2_train <- list()
+rmse_2_train <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit2_pred_train <- bind_cols(ML_fit2_pred_train, train_data %>% select(Weightgrams))
@@
 
+# Extract RMSE and R2 for in-line text
+r2_2_train$model2_train <- metrics_fit2_train[metrics_fit2_train$.metric == "rsq", ".estimate"]
+rmse_2_train$model2_train <- metrics_fit2_train[metrics_fit2_train$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 ```
-Our RMSE is 483 and our r-squared is 0.03, a slightly better performance than the full model.
-
-Now let's run cross-validation.
+The RMSE is `r round(rmse_2_train$model2_train)` and the r-squared is `r round(r2_2_train$model2_train, 3)` a slightly better performance than the full model.
+
+Now I'll run cross-validation.
 
@@
 folds
+
+# Create a list to store R-squared and RMSE values
+r2_2_cv <- list()
+rmse_2_cv <- list()
 
@@
 
+# Extract R-squared and RMSE values for in-line text
+# For R-squared
+r2_2_cv$model2_cv <- ML_cv2_metrics %>%
+  filter(.metric == "rsq") %>%
+  pull(mean)
+
+# For RMSE
+rmse_2_cv$model2_cv <- ML_cv2_metrics %>%
+  filter(.metric == "rmse") %>%
+  pull(mean)
+
 # Save regression output into table
@@
 ```
-Here, our RMSE is 483 and our r-squared is 0.026, the same as the training model.
-
-Finally, let's run our model on the test data.
-
-```{r}
+Here, the RMSE is `r round(rmse_2_cv$model2_cv)` and the r-squared is `r round(r2_2_cv$model2_cv, 3)`, the same as the training model `r round(rmse_2_train$model2_train)`, `r round(r2_2_train$model2_train, 3)`.
+
+Finally, I'll run the model on the test data.
+
+```{r, message=FALSE,  warning=FALSE}
 # Testing model 2
+# Create a list to store R-squared and RMSE values
+r2_2_test <- list()
+rmse_2_test <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit2_pred_test <- bind_cols(ML_fit2_pred_test, test_data %>% select(Weightgrams))
@@
 
+# Extract R-squared and RMSE values for in-line text (because data frame with columns not list)
+r2_2_test$model2_test<- metrics_fit2_test[metrics_fit2_test$.metric == "rsq", ".estimate"]
+rmse_2_test$model2_test <- metrics_fit2_test[metrics_fit2_test$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 ```
-Running the model on our test data, we see no significant results at any alpha level.
+
+Running the model on the test data, there are no significant results at any alpha level.
 
@@
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Model 1: Sub-sample simple model (ML_lm1_sub)
@@
 ```
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Model 2: Sub-sample multivariate model (ML_lm2_sub)
@@
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Combine the four tables using tbl_merge
@@
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Model 3: Final model for paper using all data
+# Create a list to store R-squared and RMSE values
+r2_3 <- list()
+rmse_3 <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit3_pred <- bind_cols(ML_fit3_pred, ML_linear %>% select(Weightgrams))
@@
 
+# Extract RMSE and R2 for in-line text
+r2_3$model3 <- metrics_fit3[metrics_fit3$.metric == "rsq", ".estimate"]
+rmse_3$model3 <- metrics_fit3[metrics_fit3$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
 saveRDS(table_model3, file = here("results", "tables", "table_model3_final.rds"))
-```
-
-Our RMSE is 494 and our r-squared is 0.01.
+
+# Create a gtsummary table from model with CIs
+table_model3_CI <- tbl_regression(ML_fit3, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>%
+  add_glance_table(include = c(r.squared, nobs)) %>%
+  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
+table_model3_CI
+```
+
+The RMSE is `r round(rmse_3$model3)` and the r-squared is `r round(r2_3$model3, 3)`.
 
@@
 
-Let's now evaluate the model on our training data.
+Now I'll evaluate the model on our training data.
 
@@
 # Training model 3
+# Create a list to store R-squared and RMSE values
+r2_3_train <- list()
+rmse_3_train <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit3_pred_train <- bind_cols(ML_fit3_pred_train, train_data %>% select(Weightgrams))
@@
 
+# Extract RMSE and R2 for in-line text
+r2_3_train$model3_train <- metrics_fit3_train[metrics_fit3_train$.metric == "rsq", ".estimate"]
+rmse_3_train$model3_train <- metrics_fit3_train[metrics_fit3_train$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 
-Our RMSE is 489 and our r-squared is 0.01, slightly better performance than the full model. Our results are not significantly different from the full model.
-
-Now let's run cross-validation.
+The RMSE is `r round(rmse_3_train$model3_train)` and the r-squared is `r round(r2_3_train$model3_train, 3)`, slightly better performance than the full model `r round(r2_3$model3, 3)`.
+The results are not significantly different from the full model.
+
+Now I'll run cross-validation.
 
@@
 folds
+
+# Create list to store R-squared and RMSE values
+r2_3_cv <- list()
+rmse_3_cv <- list()
 
@@
 
+# Extract R-squared and RMSE values for in-line text
+# For R-squared
+r2_3_cv$model3_cv <- ML_cv3_metrics %>%
+  filter(.metric == "rsq") %>%
+  pull(mean)
+
+# For RMSE
+rmse_3_cv$model3_cv <- ML_cv3_metrics %>%
+  filter(.metric == "rmse") %>%
+  pull(mean)
+
 # Save regression output into table
@@
 ```
-Using cross-validation, our RMSE is 487 and our r-squared is 0.01, a similar performance to our training model.
-
-Finally, let's run our model on the test data.
+Using cross-validation, the RMSE is `r round(rmse_3_cv$model3_cv)` and the r-squared is `r round(r2_3_cv$model3_cv, 3)`, similar performances to the training model `r round(rmse_3_train$model3_train)`, `r round(r2_3_train$model3_train, 3)`.
+
+Finally, I'll run the model on the test data.
 
@@
 # Testing model 3
+# Create a list to store R-squared and RMSE values
+r2_3_test <- list()
+rmse_3_test <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit3_pred_test <- bind_cols(ML_fit3_pred_test, test_data %>% select(Weightgrams))
@@
 
+# Extract R-squared and RMSE values for in-line text (because data frame with columns not list)
+r2_3_test$model3_test<- metrics_fit3_test[metrics_fit3_test$.metric == "rsq", ".estimate"]
+rmse_3_test$model3_test <- metrics_fit3_test[metrics_fit3_test$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 
-Much like in the other testing models, our results are not significant (except the intercept). Our performance is also worse (RMSE of 507 and r-squared of 0.008).
+Much like in the other testing models, the results are not significant (except the intercept).
+The performance is also worse (RMSE of `r round(rmse_3_test$model3_test)` and r-squared of `r round(r2_3_test$model3_test, 3)`).
 
@@
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Model 4: Final model for paper using all data
+# Create a list to store R-squared values
+r2_4 <- list()
+rmse_4 <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit4_pred <- bind_cols(ML_fit4_pred, ML_linear %>% select(Weightgrams))
@@
 
+# Extract RMSE and R2 for in-line text
+r2_4$model4 <- metrics_fit4[metrics_fit4$.metric == "rsq", ".estimate"]
+rmse_4$model4 <- metrics_fit4[metrics_fit4$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
 saveRDS(table_model4, file = here("results", "tables", "table_model4_final.rds"))
-```
-Our RMSE is 490 and our r-squared is 0.026.
+
+# Create a gtsummary table from model with CIs
+table_model4_CI <- tbl_regression(ML_fit4, exponentiate = FALSE, intercept = TRUE) %>%
+  modify_column_unhide(columns = c(statistic, std.error)) %>%
+  add_significance_stars(hide_ci = FALSE, hide_p = FALSE) %>% # Hide confidence intervals and p-values
+  add_glance_table(include = c(r.squared, nobs)) %>% # Add R-squared and number of observations
+  modify_header(statistic = "**T-Statistic**")
+table_model4_CI
+```
+The RMSE is `r round(rmse_4$model4)` and the r-squared is `r round(r2_4$model4, 3)`.
 
@@
 
-Let's now evaluate the model on our training data.
+I will now evaluate the model on the training data.
 
@@
 # Model 4 training
+# Create list to store R-squared and RMSE values
+r2_4_train <- list()
+rmse_4_train <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit4_pred_train <- bind_cols(ML_fit4_pred_train, train_data %>% select(Weightgrams))
@@
 
+# Extract RMSE and R2 for in-line text
+r2_4_train$model4_train <- metrics_fit4_train[metrics_fit4_train$.metric == "rsq", ".estimate"]
+rmse_4_train$model4_train <- metrics_fit4_train[metrics_fit4_train$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 ```
-Our RMSE is 483 and our r-squared is 0.033, a slightly better performance than the full model.
-
-Now let's run cross-validation.
+
+The RMSE is `r round(rmse_4_train$model4_train)` and the r-squared is `r round(r2_4_train$model4_train, 3)` a slightly better performance than the full model `r round(r2_4$model4, 3)`.
+
+Now I'll run cross-validation.
 
@@
 folds
+
+# Create list to store R-squared and RMSE values
+r2_4_cv <- list()
+rmse_4_cv <- list()
 
@@
 
+# Extract R-squared and RMSE values for in-line text
+# For R-squared
+r2_4_cv$model4_cv <- ML_cv4_metrics %>%
+  filter(.metric == "rsq") %>%
+  pull(mean)
+
+# For RMSE
+rmse_4_cv$model4_cv <- ML_cv4_metrics %>%
+  filter(.metric == "rmse") %>%
+  pull(mean)
+
 # Save regression output into table
@@
 ``` 
-Here, our RMSE is 484 and our r-squared is 0.025, a better performance than both the full model and training model.
-
-Finally, let's run our model on the test data.
+Here, the RMSE is `r round(rmse_4_cv$model4_cv)` and the r-squared is `r round(r2_4_cv$model4_cv, 3)`, a better performance than both the full model `r round(rmse_4$model4)`, `r round(r2_4$model4, 3)` and the training model `r round(r2_4_train$model4_train, 3)`, `r round(rmse_4_train$model4_train)`.
+
+Finally, I will run model on the test data.
 
@@
 # Model 4 testing
+# Create list to store R-squared and RMSE values
+r2_4_test <- list()
+rmse_4_test <- list()
+
 # Using recipes and workflows
@@
 
-# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
+# Predicted numeric outcome named .pred. Match predicted values with corresponding observed outcome values
 ML_fit4_pred_test <- bind_cols(ML_fit4_pred_test, test_data %>% select(Weightgrams))
@@
 
+# Extract R-squared and RMSE values for in-line text (because data frame with columns not list)
+r2_4_test$model4_test<- metrics_fit4_test[metrics_fit4_test$.metric == "rsq", ".estimate"]
+rmse_4_test$model4_test <- metrics_fit4_test[metrics_fit4_test$.metric == "rmse", ".estimate"]
+
 # Save regression output into table
@@
 ```
-Again, most of our results are not significant except the intercept given the small sample size.
+
+Again, most results are not significant except the intercept given the small sample size.
 
@@
 
-```{r}
+```{r, message=FALSE,  warning=FALSE}
 # Combine the two tables using tbl_merge

### 2024-09-06 code/analysis-code/introanalysis-v1.qmd
Final Revisions for Submission
Cleaned up all files

@@
 ```{r}
-#Path to linear data. Note the use of the here() package and not absolute paths
+# Path to linear data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear.rds")
-#load data
+
+# Load data
 ML_linear <- readRDS(data_location)
 
-#Path to summary data. Note the use of the here() package and not absolute paths
+# Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_summary.rds")
-#load data
+
+# Load data
 ML_summary <- readRDS(data_location)
@@
   filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>% # Remove NAs from both variables
-
 group_by(ModifiedColor) %>% # Group by ModifiedColor
+
   # Calculate the percentage for each group
@@
 
-# Print the result
+# Print result
 print(LBW_percentage)
@@
 
-#Print table
+# Print table
 table_fit2
@@
 
-I am not trying to explain which general factors are trying to explain birth weight.
+I am not trying to pinpoint which general factors are trying to explain birth weight.
 Rather, I'm trying to understand if maternal skin color is associated with lower infant birth weight during this specific time period in Rio de Janeiro, Brazil.
@@
 Since all women of color in this sample were Brazilian, there is a problem with multicollinearity.
-So I will perform some subsample modeling to understand this issue.
+So I will perform some sub-sample modeling to understand this issue.
 We will create a `Brazilian` (yes, no) variable, drop all non-Brazilians from the study, and run the original regression with the covariates.
@@
 
-Here, non-Brazilians had babies on average weighing 100 grams more than Brazilians (p-value 0.001).
+Here, non-Brazilians had babies on average weighing 100 grams more than Brazilians (p-value = 0.001).
 So I will now run a regression of only Brazilian women using `Afro-Descent` and `Euro-descent` categories.
@@
 # Predominant and residual distributions
+
+Here, I have created a plot of the distribution of birth weights at Laranjeiras, following Wilcox's discussion of the residual and predominant birth weight distributions [@wilcox2001].
 

### 2024-09-06 code/eda-code/eda-v1.qmd
Final Revisions for Submission
Cleaned up all files

@@
 
-Load the data. First we will load the ML_summary set for summary statistics.
+Load the data. First, I will load the ML_summary set for summary statistics.
 
 ```{r}
-#Path to summary data. Note the use of the here() package and not absolute paths
+# Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_summary.rds")
 
-#load data
+# Load data
 ML_summary <- readRDS(data_location)
@@
 
-# save to file
+# Save to file
 summarytable_file = here("results", "tables", "summarytable.rds")
@@
 
-We are saving the results to the `results` folder.
+I am the results to the `results` folder.
 All final table/figures will have `_final` in the name.
@@
 
-Histogram plots for the continuous outcomes.
+Histogram plots for continuous outcomes.
 
@@
 Now infant birth length.
+
 ```{r}
@@
 
-For now, let's look at height as function of weight. We can see that length is positively correlated with weight, which is unsurprising.
+For now, let's look at height as function of weight. Length is positively correlated with weight, which is unsurprising.
 
@@
 
-Before we create a summary table, lets clean up our variables for easier analysis.
-First, we need to create dummy variables for categories.
-We will start with the `Color` variable, which is maternal skin color.
-We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation.
+Before we create a summary table, I will engage in another round of data cleaning for easier analysis.
+First, I need to create dummy variables for categories.
+I will start with the `Color` variable, which is maternal skin color.
+I will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation.
 The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*).
 I maintained this categorization.
-However, I also created a new dummy variable for skin color, combining Black and mixed-Race patients into one category, `Afro-Descent`, and all White patients into a `Euro-Descent` category.
+However, I also created a new dummy variable for skin color, combining Black and mixed-race patients into one category, `Afro-Descent`, and all White patients into a `Euro-Descent` category.
 
@@
 With this, I am removing the distinction between parity and gravidity since it is unclear from the original data if the physicians were categorizing the women before the delivery or after.
-With this, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births.
+Thus, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births.
 However, because these data are about women entering a hospital to give birth, rather than to receive prenatal care for a pregnancy, and this assumption can provide more statistical power for a linear regression, I assume this risk.
@@
 
-For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "Latin American", and all Middle Eastern nationalities into "Middle Eastern".
-I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations.
+For the `Nationality` variable, I combine all European nationalities into one category, "European", all Latin American nationalities into "Latin American", and all Middle Eastern nationalities into "Middle Eastern".
+I am combining categories because for many of the `European` and `Latin American` categories, there were very few observations.
 Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil.
@@
 I will now change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.
+In the future, I should perform this step in the processing script.
 
@@
 
-#Summarize missing data using dplyr
+# Summarize missing data using dplyr
 missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
@@
 
-`Gtsummary` is a very easy way to create summary statistics (Table 1).
-In the note section at the end of this document, I also include the code I worked on to create the tables manually to show my learning process.
+`Gtsummary` is a very easy way to create summary statistics (Table 1). For previous code I worked on in relation to summary stats, see the `edacode-extra.R` file.
 
@@
 
-For maternal race, we have a sample size of n = 2695.
-The majority of the sample is White (branca), with 42.45% of the sample.
-Mixed-race (parda) is 29.2% of the sample, and Black (preta) is 28.32% of the sample.
-Let's now compare White patients to all patients of color by combining Black and mixed-race categories.
-I will create a dummy variable for color, called modified_color with two categories: Afro-Descent (comprising parda and preta) and Euro-Descent (comprising branca).
+For maternal skin color, we have a sample size of n = 2695.
+The majority of the sample is White (*branca*), with 42.45% of the sample.
+Mixed-race (*parda*) is 29.2% of the sample, and Black (*preta*) is 28.32% of the sample.
+I'll now compare White patients to all patients of color by combining Black and mixed-race categories.
+I will create a dummy variable for color, called `ModifiedColor` with two categories: Afro-Descent (comprising *parda* and *preta*) and Euro-Descent (comprising *branca*).
 
-We can see that of the clinic patients, 1551 (57.6%) were of African descent (defined as preta or parda) and 1144 (42.4%) were of European descent (branca).
+Of the clinic patients, 1551 (57.6%) were of African descent (defined as *preta* or *parda*) and 1144 (42.4%) were of European descent (branca).
 
-We can see that of a sample size of n = 2761, 88% of outcomes were natural deliveries.
+Of a sample size of n = 2761, 88% of outcomes were natural deliveries.
 
-Now let's take abortion out of this sample size to just look at labor and delivery outcomes.
+I'll now remove abortion from the dataset to look solely at labor and delivery outcomes.
 
@@
 ```
-We can see that of a new sample size of n = 2672, 91% of outcomes were natural deliveries.
 
-Now let's look at maternal outcomes.
+Of the new sample of n = 2672, 91% of outcomes were natural deliveries.
 
-Now let's look at some common reproductive and perinatal health statistics: maternal mortality ratio (MMR), stillbirth rate (SBR), and sex ratio at birth (SRaB).
-The equation is MMR = MD/LB * 10,000.
-We can also look at stillbirth rate.
-The equation is SBR = SB/LB * 1000.
+Maternal outcomes are also important to understand, including the following reproductive and perinatal health statistics: maternal mortality ratio (MMR), stillbirth rate (SBR), and sex ratio at birth (SRaB).
+For MMR, the equation is MMR = MD/LB * 10,000.
+For SBR, the equation is SBR = SB/LB * 1000.
+For SRaB, the equation is SRaB = (M/F) * 100.
+
 ```{r}
@@
 ```
-The MMR is 94.1% in this specific clinic.
-The SBR is 82.9% in this specific clinic.
-The sex ratio is 1.2 or there were 120 male live births per 100 female live births at this specific clinic.
+
+The MMR is 94.1% in this specific hospital.
+The SBR is 82.9% in this specific hospital.
+The sex ratio is 1.2 or there were 120 male live births per 100 female live births at this hospital.
 This is much higher than the current range of between 103 and 107 male births per 100 female births.
 
-Now we will create the summary table for mean birth weights across time and place.
+Now I will create the summary table for mean birth weights across time and place.
 
@@
 ```{r}
-#Path to summary data. Note the use of the here() package and not absolute paths
+# Path to summary data. Note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","ML_linear.rds")
-#load data
+
+# Load data
 ML_linear <- readRDS(data_location)
@@
 # Create dummy variables
-
-#Check pre-structure
+# Check pre-structure
 str(ML_linear)
@@
 # Notes
-
-For the previous code I worked on in relation to summary stats, see the `edacode-extra.R` file.

### 2024-09-06 code/processing-code/processingfile-v1.qmd
Final Revisions for Submission
Cleaned up all files

@@
 
-Please note I used ChatGPT and CoPilot throughout to help troubleshoot code.
-
-# Processing script
-
-This contains the same code and comments/information as `processingcode.R`. This quarto file does not pull code from the R script but rather includes it directly.
+Please note I used ChatGPT, CoPilot, and Claude Anthropic throughout to help troubleshoot code. LLM were *not* used in analysis or writing.
 
@@
 ```{r}
-# path to data
-# note the use of the here() package and not absolute paths
+# Path to data
+# Note the use of the here() package and not absolute paths
 data_location <- here::here("data","raw-data","MaternidadeLaranjeiras.csv")
@@
 
-From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.
+This initial read shows that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.
 
@@
 
-First we can look at the Codebook and practice table making.
+First, I will look at the Codebook and practice table making.
+
 ```{r}
@@
 
-#Print table
+# Print table
 print(codebook_table)
@@
 These variables include qualitative information that can be excluded from our processed data, since we are not looking at qualitative notes on cases.
+Thus, I will remove the following variables as I process the data: `Nationality_Notes`, `Maternal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.
 Nonetheless, for `Fetal_Notes`, we need to keep the Fetal_Notes variable at the moment as it qualifies the `FetalOutcome` variable.
+Further, I will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).
 
-We also don't need the citation information to run regressions.
-So we will be removing the following variables as we process our data: `VolN`, `Page`, `Number`, `Nationality_Notes`, `Maternal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.
-Further, we will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).
+I also don't need the citation information to run regressions. So I will remove the following variables as I process our data: `VolN`, `Page`, `Number`.
 
@@
 
-Despite removing variables we won't use because of their qualitative nature, we still have many missing observations for some of the variables. Let's check for missing data.
+Despite removing variables I won't use because of their qualitative nature, I still have many missing observations for some of the variables.
 
 ```{r}
-#Summarize missing data using dplyr
+# Check for missing data
+# Summarize missing data using dplyr
 missing_data_summary <- ML_1 %>% dplyr::summarise_all(list(~sum(is.na(.))))
@@
 
-We can see that `CivilStatus` has 2834 missing observations, `GestationalAge_Months` has 2789 missing observations, `Fetal_Notes` has 2823 missing observations, and `Birth_Notes` also has 2330 missing values.
-This is a lot.
+The `CivilStatus` variable has 2834 missing observations, `GestationalAge_Months` has 2789 missing observations, `Fetal_Notes` has 2823 missing observations, and `Birth_Notes` also has 2330 missing values.
 Although civil status can serve as a proxy for income and class, I have too many missing values here for the variable to be meaningful, so I will exclude `CivilStatus`.
-Gestational age in months is also a valuable insight in relation to LBW and PTB, but my data set has too many missing values to be of use here, we can also exclude `GestationalAge_Months`.
+Gestational age in months is also a valuable insight in relation to LBW and PTB, but my data set also has too many missing values, and I will also exclude `GestationalAge_Months`.
 Some gestational age entries were also in relation to spontaneous abortions.
@@
 
-Before I exclude any variables, I need to make sure that the original hospital records did not misrecord stillbirths as neonatal deaths and vice versa before I exclude them.
-First, we will look at the values for `FetalOutcome` and `Fetal_Notes` that are not missing any values for both variables.
+Before I exclude any variables, I need to make sure that the original hospital records did not misrecord stillbirths as neonatal deaths and vice versa.
+First, I will look at the values for `FetalOutcome` and `Fetal_Notes` that are not missing any values for both variables.
 This gives us 22 entries.
-Most values in the `Fetal_Notes` variable refer to either a macerated (*macerado*) stillbirth, perhaps implying maternal syphilis or a neonatal death (*faleceu*) with a given time frame.
-We can see that the `FetalOutcome` variable is not always consistent with the `Fetal_Notes` variable.
+Most values in the `Fetal_Notes` variable refer to either a macerated (*macerado*) stillbirth, perhaps implying maternal syphilis, or a neonatal death (*faleceu*) with a specific time frame.
+Nonetheless, the `FetalOutcome` variable is not always consistent with the `Fetal_Notes` variable.
 For example, a live birth and subsequent neonatal death is marked as a stillbirth in the `FetalOutcome` variable.
@@
 
-I also need to extract information on gestation (twin versus singleton birth) from this `GestationalAge_Months` before I exclude it.
+I also need to extract information on gestation (twin versus singleton birth) from `GestationalAge_Months` before I exclude it.
 My data set will only include singleton births for the statistical analysis.
@@
 
-Now we can check for missing data again.
-
 ```{r}
-#Summarize missing data using dplyr
+# Re-check for missing data
+# Summarize missing data using dplyr
 missing_data_summary <- ML_4 %>% dplyr::summarise_all(list(~sum(is.na(.))))
@@
 Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study.
-Here, I will create two datasets: one for summary statistics and one for the linear regression model.
+Thus, I will create two datasets: one for summary statistics and one for the linear regression model.
 In addition, most contemporary studies exclude all birthweights less than 500 grams because they have a high chance of being stillborn.
@@
 
-Now, I can see that my `ML_linear` dataset has no missing data.
-It has 1919 observations and 12 variables.
-This might change as I begin analysis.
-For example, I might want to exclude infant birth length, which is included at the moment.
+Now, I can see that my `ML_linear` dataset has no missing data. It has 1919 observations and 12 variables. This might change as I begin analysis.
 
@@
 # Data loading for mean birth weights table
+
+I am also including a descriptive table of mean birth weights culled from other published papers. Here, I will create a table for these data.
+
 ```{r}
-# path to data
-# note the use of the here() package and not absolute paths
+# Path to data
+# Note the use of the here() package and not absolute paths
 data_location <- here::here("data","raw-data","Mean_Birthweights.csv")
@@
 # Save Mean_Birthweight as RDS
-save_data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")
+save_data_location <- here::here("data","processed-data","Mean_Birthweights.rds")
 saveRDS(Mean_Birthweights, file = save_data_location)

### 2024-09-06 products/manuscript/Manuscript_Final_Lancet.qmd
Final Revisions for Submission
Cleaned up all files

@@
 
-In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence.
-After abolition, the state did not create programs to incorporate formerly enslaved people and their descendants into economic, civil, or political life.
+In 1888, after nearly four centuries, Brazil became the last country in the Western Hemisphere to abolish slavery.
+In the aftermath, the state did not create programs to incorporate formerly enslaved people and their descendants into economic, civil, or political life.
 Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century.
-Public health studies have shown that health disparities continue to mirror patterns from over a century ago.
-The goal of this study is to quantify health disparities in infant birth weights by maternal characteristics in the post-abolition period in one of Brazil's first public maternity hospitals.
+Epidemiologic studies have shown that health disparities continue to mirror patterns from over a century ago.
+The goal of this study is to quantify health disparities in the post-abolition period using infant birth weight in one of Brazil's first public maternity hospitals.
 
@@
 This cross-sectional analysis used hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital.
-It used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.
+It used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality.
 
@@
 African ancestry was correlated with lower birth weights.
-Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant.
-Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.
+Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value = 0.002 [95% CI -138, -32]) than infants born to Euro-descendant women, whether Brazilian or immigrant.
+Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value = 0.019 [95% CI -131, -12]) than infants born to White women.
 
@@
 
-The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for women of color -- and possible epigenetic effects from the lived experience of racism.
+The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for women of African descent -- and possible epigenetic effects from the lived experience of racism.
 The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.
@@
 
+This study analyzes infant birth weight as a proxy for general maternal and fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras, a public maternity hospital in Rio de Janeiro, Brazil between 1922 and 1926.
+It aims to provide basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil, substantiating the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
+
 Birth weight and its relationship to infant health have been a focus of public health research for over a century [@hughes2017].
-In epidemiology, both low birth weight (LBW) and preterm birth (PTB) within and across racialized communities is a focal point of research [@braveman2021a; @david1997; @hughes2017; @silva1998; @wilcox2001].
-LBW can be a function of PTB, of fetal growth restriction (FGR, previously known as intra-uterine growth restriction or IUGR, which can include but is not limited to small-for-gestational-age or SGA infants -- those born at less than 10 percentile), or of both [@wilcox2001; @wilcox2024].
-Nonetheless, PTB is the primary determinant of both LBW and lower birth weight in general [@maruyama2020].
+In epidemiology, both low birth weight (LBW) and preterm birth (PTB), or infants born <37 weeks' gestation, within and across racialized communities have garnered attention [@braveman2021a; @david1997; @hughes2017; @silva1998; @wilcox2001].
+LBW can be a function of PTB, of fetal growth restriction (FGR, previously known as intra-uterine growth restriction or IUGR, which can include but is not limited to small-for-gestational-age or SGA infants -- those born at less than the 10th percentile), or of both [@wilcox2001; @wilcox2024].
+Nonetheless, PTB is an important determinant of both LBW and lower birth weight in general [@maruyama2020].
 
@@
 Originally, epidemiologists hypothesized an association between at-term LBW and increased risk for cardiovascular diseases and other chronic physical health conditions over the life course [@barker2004].
-Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest this.
-For example, Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure.
-Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.
+Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest long-term trends associated with "fetal programming" [@maruyama2020; @warrington2019].
 
-Historians have shown how birth weight can be an indicator for the often unmeasurable nutritional status of mothers of different social classes and racial classifications in the past [@goldin1989; @costa1998; @costa2004; @ward1993].
+Historians, using hospital records, have shown how birth weight can be an indicator for the often unmeasurable nutritional status of mothers of different social classes in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @ward1993].
 Historically, women from more privileged groups had heavier infants than women from less advantaged groups [@stein1984; @ward1993].
 
-The differences between groups both today and in the past have given rise to much research stratified by racial group.
+The differences between and within groups both today and in the past have given rise to much research stratified by racial classification.
 Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
@@
 Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997].
-Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].
 
 Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@jasienska2009; @owens2019; @martins2024; @steckel1986].
-In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2020; @telles2022].
+Studies in the human sciences have hypothesized an intergenerational effect of poor health conditions, including undernutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].
+In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@adamo1998; @otovo2016; @roth2020; @telles2022].
 
-Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country, compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
-Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission or the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
-
-Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere.
-In the nineteenth century, population-level health was poor for all Brazilians, but particularly so for enslaved individuals, who experienced the epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever that affected the entire population [@karasch1987].
-Although caloric intake for the enslaved population was probably sufficient in terms of quantity, excluding pregnant and nursing women, scholars have shown how it lacked key nutrients, and thus the enslaved suffered from nutritional deficiencies, including thiamine deficiency, which was associated with high infant mortality rates [@karasch1987; @kiple1989].
+Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021].
+It was also the last country to abolish slavery (1888) in the Western Hemisphere.
+In the nineteenth century, population-level health was poor for all Brazilians, but especially for enslaved individuals, who experienced epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever as well as undernutrition and violent corporal punishment [@karasch1987].
+Excluding pregnant and nursing women, historians have argued that caloric intake for the enslaved population was probably sufficient in terms of quantity, yet it lacked key nutrients, and thus the enslaved suffered from nutritional deficiencies, including insufficient thiamine, which was associated with infant mortality [@karasch1987; @kiple1989].
 Today, thiamine deficiency is correlated with lower birth weight and adverse perinatal outcomes including neonatal death [@gr2024; @kareem2023].
 
-In the immediate aftermath of the abolition of slavery, which coincided with the implementation of the country's first republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into economic, civic, or political life occurred [@carvalho2004].
-Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016].
+In the aftermath of abolition, no state-run efforts to incorporate formerly enslaved, African-descended peoples into economic, civic, or political life occurred [@fischer2008].
+Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower literacy levels, and had worse health outcomes than their White and White immigrant counterparts [@adamo1998; @ball2020; @fischer2008; @otovo2016].
 
@@
 The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe.
-Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@benchimol1999; @hochman2013].
-Although infant mortality rates declined from the nineteenth century, the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@roth2020].
-Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].
+Improved sanitation measures helped stem infectious disease outbreaks [@benchimol1999; @hochman2013].
+Infant mortality rates declined from the nineteenth century, although the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@adamo1998; @roth2020; @loudon1992; @woods2009].
 
-By the 1920s, in Rio de Janeiro public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @roth2020].
-Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
+Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, provided healthcare free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
 Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe.
-Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
-Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
-
-This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights in major U.S.
-and Western European cities [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993].
-It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].
 
 In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon.
-In our current postgenomic era, in which race as a biological category has been rejected, the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking remains very real [@meloni2022].
-Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022].
+In our current postgenomic era, race as a biological category has been rejected, yet the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking in epigenetics remains very real [@meloni2022].
+Although differences between racialized groups existed, they were not immutable or due to genetic differences based on race.
 The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].
-
-This study analyzes infant birth weight as a proxy for general fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras.
-It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
 
@@
 
-This analysis draws on a unique data set of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926.
-I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO).
-The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings.
-Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras.
+This analysis draws on a unique dataset of 2,845 recorded clinical visits, most of them deliveries, to Maternidade Laranjeiras between 1922 and 1926.
+I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO), which, between June 1922 and May 1926, published the monthly clinical reports of all women treated at Laranjeiras.
 I was unable to locate vol.
@@
 4 (1926).
-In the initial months of publication, the reports included more complete information, including labor and delivery time or detailed descriptions of surgical procedures.
-Over time, clinical notes became streamlined.
-When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a].
 
-From the available issues, I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.
+I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.
 
@@
 *Birth weight*.
-The outcome was birth weight measured as a continuous variable in grams for live births only.
-In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
-Wilcox (@wilcox2001) has questioned the utility of the LBW category in epidemiological studies given its arbitrary nature.
-We can question the use of this cutoff in historical studies as well, as it was not an established category until the mid-twentieth century.
-A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification [@hughes2017].
-This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity.
-The WHO adopted this cutoff in 1948, and it has remained the standard ever since even as epidemiologists began to differentiate between small-for-gestational age (SGA, now FGR) and PTB in the 1970s [@hughes2017; @wilcox2001].
-Physicians in early twentieth-century Rio de Janeiro defined low birth weight as BLANK [@conrado1924a].
+The outcome was birth weight measured as a continuous variable in grams for live births only, in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
+In a secondary analysis (see Supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
 
@@
 *Maternal skin color*.
-The original records categorized maternal racial categories as White (*branca*), the reference group; mixed-race (*parda*); and Black (*preta*).
+The original records categorized maternal skin color as White (*branca*), the reference group; mixed-race (*parda*); and Black (*preta*).
 I maintained this categorization.
@@
 
-Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum rather than as a Black-White binary as in the United States [@travassos2004].
+Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum rather than as a Black-White binary [@travassos2004].
 Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category.
-This method is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].
+This method also is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].
 
@@
 *Gestational status*.
-The original hospital records divided gestational status into eight categories of parity and gravidity: nulliparous, primiparous, secundiparous, multiparous; and primigravida, secundigravida, trigravida, and multigravida.
-Due to issues of statistical power, I combined these into two categories: 1) nulliparous, primigravida, and primiparous into `nulliparous` (or any woman who is giving birth for the first time); and 2) secundiparous, multiparous, secundigravida, trigravida, and multigravida into `multiparous` (or any woman who has given birth more than once).
+The original hospital records divided gestational status into eight categories of parity and gravidity: nulliparous, primiparous, secundiparous, and multiparous; and primigravida, secundigravida, trigravida, and multigravida.
+Due to issues of statistical power, I combined these into two categories: 1) `nulliparous`, which included nulliparous, primigravida, and primiparous (any woman giving birth for the first time); and 2) `multiparous`, which included secundiparous, multiparous, secundigravida, trigravida, and multigravida (any woman who has given birth more than once).
 
 *Maternal nationality*.
-The original hospital records recorded maternal nationality by country, which are available in the supplemental file.
-For analysis here, I created categories based on individual country (`nationality`) and a dummy variable based on region (`modifiednationality`).
+The original hospital records recorded maternal nationality by country.
+For analysis, I created categories based on individual country (`Nationality`) and a dummy variable based on region (`ModifiedNationality`).
 I recategorized the latter as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian).
-For a sub-sample analysis to test collinearity between nationality and skin color, I further created a dummy category between `Brazilian` and `non-Brazilian`.
+For a sub-sample analysis to test collinearity between nationality and skin color, I further created a dummy category between `Brazilian` and `Non-Brazilian`.
 
@@
 However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries), coded as operatory or natural by hospital records, as interventionist.
-I classified spontaneous abortion as a separate category within this variable.
+I classified spontaneous abortion as a separate category.
 
-*Infant length*.
-Infant length was recorded in centimeters in the original records.
-A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
-The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009; @ward1993].
-But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
-Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
+*Infant length*. Infant length was recorded in centimeters in the original records.
 
 *Maternal outcome*.
-Hospital records recorded maternal outcome as death, discharge, or hospital transferal, which I maintain here.
+Hospital records recorded maternal outcome as death, discharge, or hospital transferal, which I maintained.
 I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality below.
@@
 For the statistical models, I excluded both spontaneous abortions (recorded in the birth variable) and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data, which is standard methodology for looking at infant birth weight today [@nyarko2013; @wehby2015a; @wehby2017a].
-The clinic records recorded 20 neonatal deaths, which I included in the live birth category.
-These numbers could be higher, as the clinic only recorded deaths within the first week of life, probably because those infants died in the hospital rather than at home.
+The hospital records recorded 20 neonatal deaths, which I included in the live birth category.
+These numbers could be higher, as the hospital only recorded deaths within the first week of life, when the infant was still hospitalized.
 
@@
 
-I calculated summary statistics, including sample proportions for maternal skin color, maternal ancestry, parity, maternal nationality, birth outcome, maternal outcome, fetal outcome, and fetal sex.
+I calculated sample proportions for maternal skin color, maternal ancestry, parity, maternal nationality, birth outcome, maternal outcome, fetal outcome, and fetal sex.
 I also calculated mean maternal age, infant birth length, and infant birth weight, unadjusted and then stratified by first fetal sex and then maternal skin color.
 
-I also calculated the maternal mortality ratio (MMR), the mean stillbirth rate (SBR), and the sex ratio at birth (SRaB).
+Finally, I calculated the maternal mortality ratio (MMR) and the mean stillbirth rate (SBR). In the Supplement, I calculated the sex ratio at birth (SRaB).
 
@@
 where maternal deaths (MD) are divided by 10,000 live births (LB).
-In current studies, MD are divided by 100,000 live births, but I use 10,000 in the denominator, which is in line with historical studies given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].
+In current studies, MD are divided by 100,000 LB, but I use 10,000 in the denominator, which is in line with historical studies given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].
 
@@
 
-The SRaB is:
-
-$$SRaB = \frac{M}{F} \times 100$$
-
-where the total number of live male births (M) is divided by 100 live female births (F) in a given period.
-
 ## Statistical analysis
@@
 I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight.
-I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
-Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age.
+Due to the lack of comprehensive information on gestational age, I did not consider birth weight in reference to gestational age.
 I discuss this constraint in the discussion below.
@@
 
-The simple linear regression model for modified ancestry is as follows:
+The simple linear regression model for modified ancestry is:
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.
 
-The multiple linear regression model for modified ancestry is as follows:
+The multiple linear regression model for modified ancestry is:
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
 Results for both models are displayed in Table 2.
 
-The simple linear regression model for original skin color is as follows:
+The simple linear regression model for original skin color is:
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.
 
-The multiple linear regression model for original color is as follows:
+The multiple linear regression model for original color is:
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
 Results for both models are displayed in Table 3.
@@
 Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 (modified ancestry) in two steps.
-First, I created a new dichotomous variable, `Brazilian` (yes/no) and ran a simple linear analysis with birth weight as the outcome and maternal nationality as the independent variable.
-Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value <0.001).
+First, I created a new dichotomous variable, `Brazilian`/`Non-Brazilian` and ran a simple linear analysis with birth weight as the outcome and maternal nationality as the independent variable.
+Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value = 0.001 [95% CI 40, 161]).
 Given that all women of color in the data set were Brazilian, collinearity exists.
-Second, I dropped all non-Brazilians from the study, and ran the original simple and multiple regressions from Models 1 and 2.
+Second, I dropped all non-Brazilians and ran the original simple and multiple regressions from Models 1 and 2.
+Results for the sub-sample analysis are included in Table 2.
 
@@
 
-Table 1 provides summary statistics for the sample.
+Table 1 provides summary statistics for the dataset.
 In the hospital records, 58% were of African descent (defined as *preta* or *parda*) and 42% were of European descent (*branca*).
-First-time mothers ("nulliparous") were 41% of the sample.
-Of the total sample, 84% were Brazilian.
+First-time mothers (nulliparous) comprised 41%.
+Of all women, 84% were Brazilian.
 Of the nearly 16% immigrant patients, 15% were from Europe.
@@
 The SBR was 82.9%.
-The sex ratio was 1.2: there were 120 male live births per 100 female live births.
-This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
 
@@
 A difference in means test showed borderline significance at the alpha 0.05 level (t-statistic = 1.9119, df.
-2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).
+2008.2, p-value = 0.056, [95% CI -0.0007, 0.0532]).
 
@@
 All models demonstrated an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race).
-When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter than their Euro-descended counterparts (Table 2).
-In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers.
-Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent.
+When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter (p-value = 0.002 [95% CI -138, -32]) than their Euro-descended counterparts (Table 2).
+In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter (p-value = 0.019 [95% CI -131, -12]) than White mothers.
+Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of European descent.
+
 Research has shown that below 3000 g, risk of death in the first week of life increases by 40% for every 100 g decrease in birth weight [@stein1984; @wilcox1983].
-More recent studies relying on IV have found that a 10% increase in birth weight decreases infant mortality by 13.3 deaths per 1000 live births [@maruyama2020].
-Applied to these data, 85 g difference in birth weight between Afro- and Euro-descended infants (sub-model 2) would have increased infant mortality among Afro-descended children by 3.7 deaths per 1000 births.
+More recent studies relying on IV have found that a 10% increase in any birth weight decreases infant mortality by 13.3 deaths per 1000 live births [@maruyama2020].
+Applied to these data, an 85 g difference in birth weight between Afro- and Euro-descended infants (sub-model 2) would have increased infant mortality among Afro-descended children by 3.7 deaths per 1000 births.
 
@@
 
-In general, maternal and infant outcomes were not better in the Maternidade Laranjeiras.
+In general, maternal and infant outcomes were not better in the Maternidade Laranjeiras than among the wider population.
 Between 1922 and 1926, the clinic's MMR was 94.26%.
@@
 Thus, the city’s rates were probably higher.
-The same pattern holds true for the SBR, which was 84.8% for Laranjeiras and 73.68% for the city of Rio de Janeiro for the same period [@roth2020].
-Again, registration effects could explain the higher hospital rates.
-However, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures before the rise of antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
-
-Laranjeiras also had a high sex ratio (1.2).
-Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was a female.
-Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
-However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
-The skewed sex ratio deserves further study.
+The same pattern holds for the SBR, which was 84.8% for Laranjeiras and 73.68% for the city of Rio de Janeiro for the same period [@roth2020].
+However, these differences demonstrate that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and infant and supports other historical studies that interventionist obstetric procedures before antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
 
 This study found that women of African descent in early twentieth-century Brazil both had infants with lower birth weights and higher rates of LBW than their White counterparts, perhaps indicative of higher rates of PTB.
-Regardless of a difference between infants born to mothers of different skin colors, for all LBW infants in the past, the probability of surviving the neonatal period was lower than today given higher neonatal and infant mortality rates and less effective medical interventions [@costa1998; @roth2020].
-This fact is born out by more recent time-series analysis of neonatal survival [@maruyama2020].
+Wilcox [@wilcox2001] has questioned the utility of the LBW category in epidemiological studies given its arbitrary nature.
+We can question the use of this cutoff in historical studies as well, as it was not an established category until the mid-twentieth century.
+A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification [@hughes2017].
+This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity.
+The WHO adopted this cutoff in 1948, and it has remained the standard ever since even as epidemiologists began to differentiate between small-for-gestational age (SGA, now FGR) and PTB in the 1970s [@hughes2017; @wilcox2001].
+Physicians in early twentieth-century Rio de Janeiro defined low birth weight as $>2500 g$ and normal birth weight as between 3000 and 3500 g [@conrado1924a].
 
-Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight across the Americas and in New Zealand.
+Regardless of its arbitrariness, LBW was dangerous in the past given that the probability of surviving the neonatal period was lower than today due to higher neonatal and infant mortality rates and less effective medical interventions [@costa1998; @roth2020; @maruyama2020].
+
+Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight globally.
 Black infants born at Laranjeiras had the lowest mean birth weight (3037 g) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004].
 Black infants born at Laranjeiras were almost 150 g lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period.
-They were less than 100 g lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
-White infants born at Laranjeiras had a mean birth weight of 3133 g, slightly more than 50 g lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro.
-They were nearly 300 g lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 g).
-As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same historical period [@costa1998; @costa2004; @goldin1989].
+They were less than 100 g lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first-century Rio de Janeiro.
+White infants born at Laranjeiras had a mean birth weight of 3133 g, nearly 300 g lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 g).
+They were slightly more than 50 g lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first-century Rio de Janeiro.
+Differences between racialized groups existed both historically and continue to exist today.
+Moreover, when compared both historically and contemporaneously, we can see that women of all colors at Laranjeiras did not fair as well as their counterparts in other major cities.
 
-When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today.
-Although the mean birth weight at Laranjeiras was within today's normal range, the differences between racialized groups demonstrate differential health outcomes that continue to persist.
+As scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in other regions during the same historical period [@costa1998; @costa2004; @goldin1989].
 Across populations, relatively high mean birth weights in the nineteenth and early twentieth centuries are probably an artifact of high stillbirth and neonatality rates [@costa1998; @costa2004].
 Today, PTB and lower birth weight infants that probably would have died in utero 100 years ago are now surviving, bringing the mean birth weight down.
-As our estimates of PTB demonstrate, the difference in mean birth weight is probably a reflection of higher rates of PTB among African-descended women.
 
-What can explain the differences in birth weight historically both between all women at Laranjeiras and those of other historical populations and among different skin-color groups within Laranjeiras?
-The intrauterine environment is a critical determinant of infant birth weight.
-Maternal age, parity, birth spacing, and gestational age all condition birth weight.
+What can explain the differences in birth weight historically both between women at Laranjeiras and those of other historical populations and among different skin-color groups within Laranjeiras?
+Although both maternal and fetal genetic factors [@juliusdottir2021; @sole-navais2023; @warrington2019] and maternal age, parity, birth spacing, and gestational age all condition birth weight [@stein1984], the intrauterine environment remains a critical determinant.
 In this study, maternal age was not a significant factor, although multiparous mothers had infants with higher birth weights, as is common today.
@@
 Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986; @stein1984].
-Yet nutritional deficiencies that were found among African-descended populations in nineteenth-century Rio de Janeiro due to slavery are associated with lower birth weights and infant mortality [@karasch1987; @kiple1989; @gr2024; @kareem2023].
-Although working women of all skin colors had more control over their diets than did the enslaved, poor and working-class Cariocas (as residents of Rio de Janeiro are called) continued to rely heavily on manioc and dried meat, both deficient in thiamine.
-[CITE] We know that poorer women in early twentieth-century Rio de Janeiro also performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
+Yet nutritional deficiencies that were found among African-descended populations in nineteenth-century Rio de Janeiro due to slavery were associated with lower birth weights and infant mortality [@karasch1987; @kiple1989; @gr2024; @kareem2023].
+Although working women of all skin colors in the early twentieth century had more control over their diets than did the enslaved, food was expensive and many of the nineteenth-century trends in nutritional deficiencies continued in the early twentieth century [@adamo1998].
+
+We also know that poorer women in early twentieth-century Rio de Janeiro performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
 And working-class women of all skin colors often worked through the third trimester as domestic servants [@roth2020].
 These same women might have lived in crowded housing arrangements with lack of access to running water [@graham1992; @fischer2008; @roth2020].
-Thus, generalized poverty might have influenced women of African descent's nutrition and ability to access prenatal care.
+Thus, generalized poverty might have influenced women of African descent's nutrition, physical routines, and ability to access prenatal care.
 
@@
 Studies demonstrate that syphilis increases adverse birth outcomes including increased risk of LBW and PTB [@dasilva2024].
-The clinical records that comprise this dataset did not consistently record syphilis rates.
-Scholars have shown how medical and public health professionals viewed it as an important cause of adverse health outcomes and even a possible "scourge" on the future of the Brazilian race during this historical period [@carrara1996a].
-Although I don't have accurate rates for Laranjeiras, city-wide rates from the 1930s can provide some context for the prevalence of infection among pregnant women.
+The clinical records that comprise this dataset did not consistently record syphilis rates, although sporadic notes showed it existed.
+Medical and public health professionals of the time viewed it as an important cause of adverse health outcomes [@carrara1996a].
+City-wide rates from the 1930s can provide some context for the prevalence of infection among pregnant women.
 Of all prenatal syphilis exams performed between 1935 and 1938 at public clinics or hospitals across the city, unadjusted for maternal skin color, 7.92% came back positive [@fontenelle, p.
@@
 
-Discussions of epigenetic markers associated with the lived experience of racism might also be applicable here.
-Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by a variety of biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms.
-Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@leimert2020].
+Discussions of epigenetic markers associated with the lived experience of racism might also be applicable.
+Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms [@leimert2020].
+Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@jasienska2009; @leimert2020].
 Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health [@meloni2022].
@@
 
-This study provides heretofore unpublished hospital records on one of Brazil's most important public maternity hospitals in the early twentieth century, a hospital that remains in operation.
+This study contributes heretofore unpublished hospital records from one of Brazil's most important public maternity hospitals in the early twentieth century, a hospital that remains in operation today.
 It provides quantitative evidence of racialized disparities in maternal-infant health in early twentieth-century Brazil.
@@
 The hospital records do not provide information on weighing practices, but published observations from medical students who trained at Laranjeiras show that the norm was to weigh all newborns without clothes on a calibrated scale [@conrado1924a].
-Thus, an upward bias to the mean weight through the inclusion of clothing or blankets for all infants seems unlikely as does any systematic bias according to infant skin color in any direction.
-It seems unlikely that physicians weighed infants of one skin color in a different method than those of another given that this was the premier teaching hospital associated with one of Brazil's only medical schools.
-Comparing the mean weights to those of other historical data sets (Table 4) shows that mean weights at Laranjeiras were lower than all others except for the estimated weights of slave infants in the antebellum southern United States, showing that if any upward bias existed, weights were even lower than presented here.
-Additionally, birth weights at Laranjeiras were rounded to the nearest 50 grams.
-If there was a systematic bias in any direction for any group of infants, it would have had to be at the hospital level since multiple physicians and medical students weighed babies over the five-year period.
-The theoretical upper bound on bias is 50 grams, and my results show a difference greater than 50 g.
-Claudia Goldin [@goldin1989] has shown for Philadelphia that heaping did not systematically bias the mean, and I assume the same here.
+Thus, an upward bias to the mean through the inclusion of clothing or blankets for all infants seems unlikely.
+It also seems unlikely that physicians weighed infants of one skin color in a different method than those of another given that Laranjeiras was the premier teaching hospital associated with a major medical school.
+If any upward bias existed, weights were even lower than presented here.
+
+Additionally, physicians could have introduced rounding error as all birth weights at Laranjeiras were rounded to the nearest 50 g.
+Yet, the theoretical upper bound on bias is 50 g, and my results show a difference greater than 50 g.
+Goldin [@goldin1989] has shown for Philadelphia that heaping did not systematically bias the mean, and I assume the same here.
 
 In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
-Thus, the birth weight variable captured both smaller babies born at term, whether with or without FGR, and pre-term infants (PTB), or those born before 37 weeks' gestation.
-We are thus unable to know how many of these infants were low birth weight at term [@falcao2020].
-This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@wilcox2001; @wilcox2024].
-Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have data sets that include gestational age [@costa1998; @costa2004; @goldin1989].
-Here, I assume that infants at risk for PTB were also at-risk for lower birth weight.
+Thus, the birth weight variable captured both smaller babies born at term, whether with or without FGR, and PTB.
+[@falcao2020].
+This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@falcao2020; @wilcox2001; @wilcox2024].
+Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to many historical inquiries into birth weights.
+Yet, the factors discussed here could have contributed PTB, to FGR, or to both.
 
-Today, lower mean birth weights between populations are not necessarily associated with worse health outcomes.
-In fact, Wilcox [@wilcox2001, p.
-1236] has shown that "populations with lighter babies do not necessarily have worse mortality." Yet continued differences between racialized groups hints at the possible role of socio-economic inequity and racism in determining health outcomes.
-In addition, studies relying on instrumental variable (IV) analysis have shown that even in the recent past in developed countries, the impact of birth weight on adverse perinatal and neonatal outcomes has declined over the past 50 years given advances in medical care [@maruyama2020].
+A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
+The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009; @ward1993].
+
+Today, lower mean birth weights between populations are not necessarily associated with worse health outcomes [@wilcox2001].
+Yet continued differences between racialized groups hints at the possible role of socio-economic inequity and racism in determining health outcomes.
+In addition, studies relying on IV analysis have shown that even in the recent past in developed countries, the impact of birth weight on adverse perinatal and neonatal outcomes has declined over the past 35 years given advances in medical care [@maruyama2020].
 We can imagine that 100 years ago, the association of birth weight, most likely as a proxy for PTB, with adverse neonatal outcomes was even greater.
@@
 Lack of data on maternal nutrition and health status or maternal work routine and income, all of which cannot be corrected for by using maternal skin color as a proxy, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study.
-Maternal skin color may serve as a partial index of the unobservable variables that are acting upon infant birth weight [@costa2004; @ward1993].
+Maternal skin color may serve as a partial index of the unobservable variables that acted upon infant birth weight [@costa2004; @ward1993].
 Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.
@@
 This study has shown that lower birth weights were associated with Afro-descended women giving birth in the Maternidade Laranjeiras in Rio de Janeiro, Brazil between 1922 and 1926.
-This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which I do not have direct evidence.
+This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which we do not have direct evidence.
 These findings, although not generalizable beyond this specific hospital, are consistent with present-day studies on racialized disparities in birth weight in Rio de Janeiro, demonstrating how racial inequality in health outcomes has persisted over time.
@@
 {{< pagebreak >}}
+
+During the preparation of this work the author(s) used the LLMs ChatGPT, CoPilot, and Claude in order to troubleshoot coding errors in R.
+After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the publication.

### 2024-09-06 products/manuscript/supplement/Supplemental_Materials.qmd
Final Revisions for Submission
Cleaned up all files

@@
     highlight-style: github
-    bibliography: ../../assets/Mada_Bib.bib
-    csl: ../../assets/vancouver.csl
+    bibliography: ../../../assets/Mada_Bib.bib
+    csl: ../../../assets/vancouver.csl
     suppress-bibliography: false
@@
 ```{r, echo=FALSE, message=FALSE}
-#load packages
+# Load packages
 library(tidyverse) #for data processing/cleaning
@@
 ```
+
 ## Schematic of workflow
@@
 When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician.
-For more information on this as a source, see my other publications [@roth2020; @roth2020a].
+Ohter publications have more detailed information on this source [@roth2020; @roth2020a].
+Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
+Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
 
-The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca de Biomedicina-A, Universidade Estadual do Rio de Janeiro (BBA-UERJ), all in Rio de Janeiro, Brazil.
+The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.
 Between January 2012 and July 2013, I manually digitized the journal by photographing each volume.
@@
 There are nine categorical variables: skin color; ancestry; parity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.
+
+## Introductory Information
+
+Please note the excluded information from the manuscript draft.
+
+Fetal programming: Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure.
+Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.
+
+Racist policy and birth weight: Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].
+
+Slave population size: Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021], compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
+Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission -- the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
+
+Growth rates during pregnancy: But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
+Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
+
+Rounding error: Any systematic bias in any direction for any group of infants would have been at the hospital level since multiple physicians and medical students weighed babies over the five-year period.
 
@@
 
-Finally, Figure 4 visualizes the relationship of birth weight and length, stratified by sex since male infants are usually slightly heavier and longer than females [@crawford1987, @marsal1996].
+Figure 4 visualizes the relationship of birth weight and length, stratified by sex since male infants are usually slightly heavier and longer than females [@crawford1987, @marsal1996].
 
@@
 ```
+
+Finally, I calculated the sex ratio at birth (SRaB).
+
+The SRaB is:
+
+$$SRaB = \frac{M}{F} \times 100$$
+
+where the total number of live male births (M) is divided by 100 live female births (F) in a given period.
+
+The sex ratio was 1.2: there were 120 male live births per 100 female live births.
+This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
+Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was female.
+Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
+However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
+The skewed sex ratio deserves further study.
 
@@
 
-Model 5 demonstrates that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019).
+Model 5 demonstrates that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.45 [95% CI 1.07, 1.96, p-value = 0.016).
 
@@
 
-When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were not significant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).
+When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64 [95% CI	1.17, 2.30], p-value = 0.004), but results for the infants of Black women were not significant at the alpha 0.05 level (OR 1.25 [95% CI 0.87, 1.79], p-value = 0.2).
 
@@
 # Appendix
-Immigrant low birth weight [@clay2023]

### 2024-09-10 products/manuscript/Manuscript_Final_SSM.qmd
For Submission to SSM
Updated format

--- 2024-09-06 products/manuscript/Manuscript_Final_Lancet.qmd
@@
 
-## Background
+In 1888, Brazil became the last country in the Americas to abolish slavery.
+Historians have outlined the racialized health disparities of people of African descent in the post-abolition period.
+Epidemiologists have shown that health disparities continue to mirror patterns from over a century ago.
+This cross-sectional analysis quantifies health disparities in the post-abolition period using infant birth weight in one of Brazil's first public maternity hospitals.
+It relies on hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras in Rio de Janeiro, Brazil to run linear models assessing differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality.
+African ancestry was correlated with lower birth weights.
+Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value = 0.002 [95% CI -138, -32]) than infants born to Euro-descendant women.
+Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value = 0.019 [95% CI -131, -12]) than infants born to White women.
+The findings were likely the consequence of slavery’s legacy, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for people of African descent -- and possible epigenetic effects from the lived experience of racism.
+The findings are consistent with current-day research on racialized health disparities in Brazil.
 
-In 1888, after nearly four centuries, Brazil became the last country in the Western Hemisphere to abolish slavery.
-In the aftermath, the state did not create programs to incorporate formerly enslaved people and their descendants into economic, civil, or political life.
-Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century.
-Epidemiologic studies have shown that health disparities continue to mirror patterns from over a century ago.
-The goal of this study is to quantify health disparities in the post-abolition period using infant birth weight in one of Brazil's first public maternity hospitals.
+## Keywords
+Brazil; birth weight; racial disparities; maternal-infant health; slavery; history
 
-## Methods
-
-This cross-sectional analysis used hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital.
-It used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality.
-
-## Findings
-
-African ancestry was correlated with lower birth weights.
-Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value = 0.002 [95% CI -138, -32]) than infants born to Euro-descendant women, whether Brazilian or immigrant.
-Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value = 0.019 [95% CI -131, -12]) than infants born to White women.
-
-## Interpretations
-
-The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for women of African descent -- and possible epigenetic effects from the lived experience of racism.
-The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.
-
-## Funding
-
-This archival research for this project was funded by the National Science Foundation (NSF) and the Fulbright IIE Student Program.
 {{< pagebreak >}}
@@
 Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest long-term trends associated with "fetal programming" [@maruyama2020; @warrington2019].
+Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure.
+Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.
 
@@
 Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997].
+Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].
 
@@
 
-Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021].
-It was also the last country to abolish slavery (1888) in the Western Hemisphere.
+Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021], compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
+Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission -- the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
+Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere.
 In the nineteenth century, population-level health was poor for all Brazilians, but especially for enslaved individuals, who experienced epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever as well as undernutrition and violent corporal punishment [@karasch1987].
@@
 
-# Methods
+# Materials and Methods
 
@@
 4 (1926).
+
+In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures.
+Over time, clinical notes became streamlined.
+When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician [@roth2020; @roth2020a].
+Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
+Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
 
@@
 
+Finally, I calculated the sex ratio at birth (SRaB).
+
+The SRaB is:
+
+$$SRaB = \frac{M}{F} \times 100$$
+
+where the total number of live male births (M) is divided by 100 live female births (F) in a given period.
+
 ## Statistical analysis
@@
 The SBR was 82.9%.
+The sex ratio was 1.2: there were 120 male live births per 100 female live births.
 
@@
 However, these differences demonstrate that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and infant and supports other historical studies that interventionist obstetric procedures before antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
+
+The sex ratio of 120 male live births per 100 female live births is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
+Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was female.
+Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
+However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
+The skewed sex ratio deserves further study.
 
@@
 Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986; @stein1984].
+But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
+Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
 Yet nutritional deficiencies that were found among African-descended populations in nineteenth-century Rio de Janeiro due to slavery were associated with lower birth weights and infant mortality [@karasch1987; @kiple1989; @gr2024; @kareem2023].
@@
 Additionally, physicians could have introduced rounding error as all birth weights at Laranjeiras were rounded to the nearest 50 g.
+Any systematic bias in any direction for any group of infants would have been at the hospital level since multiple physicians and medical students weighed babies over the five-year period.
 Yet, the theoretical upper bound on bias is 50 g, and my results show a difference greater than 50 g.
@@
 {{< pagebreak >}}
-
-During the preparation of this work the author(s) used the LLMs ChatGPT, CoPilot, and Claude in order to troubleshoot coding errors in R.
-After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the publication.

### 2024-09-10 products/manuscript/supplement/Supplemental_Materials.qmd
For Submission to SSM
Updated format

@@
 
-In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures.
-Over time, clinical notes became streamlined.
-When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician.
-Ohter publications have more detailed information on this source [@roth2020; @roth2020a].
-Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
-Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
-
 The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.
@@
 There are nine categorical variables: skin color; ancestry; parity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.
-
-## Introductory Information
-
-Please note the excluded information from the manuscript draft.
-
-Fetal programming: Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure.
-Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.
-
-Racist policy and birth weight: Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].
-
-Slave population size: Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021], compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
-Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission -- the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
-
-Growth rates during pregnancy: But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
-Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
-
-Rounding error: Any systematic bias in any direction for any group of infants would have been at the hospital level since multiple physicians and medical students weighed babies over the five-year period.
 
@@
 # Appendix
+
+## AI Use
+
+During the preparation of this work the author used the LLMs ChatGPT, CoPilot, and Claude in order to troubleshoot coding errors in R.
+After using this tool/service, the author reviewed and edited the content as needed and takes full responsibility for the content of the publication.

### 2024-11-01 code/analysis-code/fullanalysis-v1.qmd
RRSSM1
Updated code that stopped working. Included Brazilian/non-Brazilian subsample missing code.

@@
 filtered_data <- ML_linear %>%
-  filter(Brazilian == "Brazilian")
+  filter(Brazilian == "yes")
 

### 2024-11-01 code/eda-code/eda-v1.qmd
RRSSM1
Updated code that stopped working. Included Brazilian/non-Brazilian subsample missing code.

@@
 "Brazilian" will remain the same.
+I also create a dummy variable to check for collinearity between immigration status and skin color.
+I create a `Brazilian` variable (yes/no), and then I run a sub-sample analysis dropping all non-Brazilians.
 
@@
 ```{r}
-# Create dummy variables
 # Check pre-structure
 str(ML_linear)
-
-# Create dummy variables for color and change to English using recode factor from dplyr
-ML_linear$ModifiedColor <- factor(ifelse(ML_linear$Color %in% c("Black", "Mixed Race"), "Afro-Descent", "Euro-Descent"))
-
-# Create dummy variables for status and change to English
-ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multiparous", "Trigesta" = "Multiparous", "Multigesta" = "Multiparous", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nulliparous", "Nulipara" = "Nulliparous", "Primipara" = "Nulliparous")
-
-# Create dummy variables for nationality and change to English
-ML_linear$ModifiedNationality <- factor(case_when(
-  ML_linear$Nationality %in% c("German", "Austrian", "Spanish", "French", "Italian", "Polish", "Portuguese", "Romanian", "Russian", "Swiss") ~ "European",
-  ML_linear$Nationality %in% c("Argentine", "Paraguayan", "Uruguayan") ~ "Latin American",
-  ML_linear$Nationality == "Brazilian" ~ "Brazilian",
-  ML_linear$Nationality == "Syrian" ~ "Middle Eastern"
-))
 
@@
 
+# Create new variables
+ML_linear <- ML_linear %>%
+  mutate(
+    ModifiedColor = factor(ifelse(Color %in% c("Black", "Mixed Race"), # dummy variables for color and change to English
+                                 "Afro-Descent", "Euro-Descent"),
+                          levels = c("Euro-Descent", "Afro-Descent")),
+    ModifiedStatus = factor(case_when( # dummy variables for status and change to English
+      Status %in% c("Secundigesta", "Trigesta", "Multigesta", "Secundipara", "Multipara") ~ "Multiparous",
+      Status %in% c("Primigesta", "Nulipara", "Primipara") ~ "Nulliparous",
+      TRUE ~ NA_character_
+    ), levels = c("Nulliparous", "Multiparous")),
+    ModifiedNationality = factor(case_when( # dummy variables for nationality and change to English
+      Nationality %in% c("German", "Austrian", "Spanish", "French", "Italian", 
+                        "Polish", "Portuguese", "Romanian", "Russian", "Swiss") ~ "European",
+      Nationality %in% c("Argentine", "Paraguayan", "Uruguayan") ~ "Latin American",
+      Nationality == "Brazilian" ~ "Brazilian",
+      Nationality == "Syrian" ~ "Middle Eastern",
+      TRUE ~ NA_character_
+    ), levels = c("Brazilian", "European", "Latin American", "Middle Eastern")),
+    Brazilian = factor(ifelse(ModifiedNationality == "Brazilian", # create dummy variable for Brazilian/non-Brazilian to check for collinearity between skin color and immigration status
+                             "yes", "no"),
+                      levels = c("yes", "no"))
+  )
+
+# Verify the new structure
 str(ML_linear)

### 2024-11-01 products/manuscript/Manuscript_Final_SSM.qmd
RRSSM1
Updated code that stopped working. Included Brazilian/non-Brazilian subsample missing code.

@@
     bibliography: ../../assets/Mada_Bib.bib
-    csl: ../../assets/vancouver.csl
+    csl: ../../assets/apa.csl
     suppress-bibliography: false

### 2024-11-14 code/analysis-code/fullanalysis-v1.qmd
RRSSM2
Age2; immigrants; year fixed effects

@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$
-
-whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$
+
+whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Par` is parity (nulliparous or multiparous), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term.
 Results are displayed in Table 3.
@@
 # Define recipe for model
-ML_recipe2 <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>% 
+ML_recipe2 <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality + Year, data = ML_linear) %>% 
   step_relevel(ModifiedColor, ref_level = "Euro-Descent") #%>% # Set reference level for ModifiedColor
@@
 # Define recipe for model
-ML_recipe2_train <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = train_data) %>% 
-  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
-  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
-  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+ML_recipe2_train <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality + Year, data = train_data) %>% 
+  step_relevel(ModifiedColor, ref_level = "Euro-Descent") #%>% # Set reference level for ModifiedColor
+  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
 
@@
 # Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm2_sub <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + + ModifiedStatus, data = filtered_data)
+ML_lm2_sub <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + Year, data = filtered_data)
 
@@
 
-The second two models are also simple and multiple linear regressions with outcome variable as birthweight in grams.
+The second two models are also simple and multiple linear regressions with outcome variable as birth weight in grams.
 However, in these two models, I maintained maternal skin color as three categories, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women.
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$
-
-whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$
+
+whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Par` is parity (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term.
 
@@
 # Specify recipe
-ML_recipe4 <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
+ML_recipe4 <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality + Year, data = ML_linear) %>%
   step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
@@
 # Specify recipe
-ML_recipe4_train <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = train_data) %>%
+ML_recipe4_train <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality + Year, data = train_data) %>%
   step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
-  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
-  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
-  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") #%>% # Set reference level for ModifiedStatus
+  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
 
@@
 # Define recipe for model
-ML_recipe4_cv <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
+ML_recipe4_cv <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality + Year, data = ML_linear) %>%
   step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
-  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
-  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
-  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") #%>% # Set reference level for ModifiedStatus
+  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
 
@@
 # Specify recipe
-ML_recipe4_test <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = test_data) %>%
+ML_recipe4_test <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality + Year, data = test_data) %>%
   step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
-  step_relevel(ModifiedStatus, ref_level = "Multiparous") %>% # Set reference level for ModifiedStatus
-  step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
-  step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
+  step_relevel(ModifiedStatus, ref_level = "Multiparous") #%>% # Set reference level for ModifiedStatus
+  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
+  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors
 

### 2024-11-14 code/analysis-code/introanalysis-v1.qmd
RRSSM2
Age2; immigrants; year fixed effects

@@
   group_by(ModifiedColor) %>% # Group by ModifiedColor
-
-  # Calculate the percentage for each group
-summarise(
+  summarise( # Calculate the percentage for each group
     total_count = n(),
@@
 
-# Linear Models
+# Linear Models ModifiedColor
 
@@
 
+## Model 1
+
 I will now run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry. I will use the recipe and workflow features of `tidymodels`.
@@
 # Define recipe for model
-ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
-  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level
-  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
+ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) #%>% 
+  #step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level
+  #step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable
 
@@
 ```
+## Model Age
 
@@
 # Create recipe
-ML_recipe_age <- recipe(Weightgrams ~ Age, data = ML_linear)
+ML_recipe_age <- recipe(Weightgrams ~ Age + AgeSquared, data = ML_linear)
 
@@
 ```
-
-Now I will run a slightly more complicated model that includes maternal skin color, age, nationality, and gestational status.
-
-```{r}
+## Model 3
+
+Now I will run a slightly more complicated model that includes maternal skin color, age, age squared, nationality, and parity (`ModifiedStatus`). This is the basis for my main model.
+
+```{r}
+# Model 3 using recipes and workflows
 # Set reference level for ModifiedColor
@@
 # Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
-ML_lm3 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)
+ML_lm3 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + AgeSquared + ModifiedStatus + ModifiedNationality, data = ML_linear)
 
@@
 In the second model, we can see that older mothers are associated with giving birth to infants with higher birth weights.
+In the third model, neither age nor age-squared are significant.
+Nonetheless, birthweight appears to increases with age, but above the maternal age of 40, infant weight decreases.
+Thus, there is a reverse U-shaped relationship.
+The coefficients suggests that while birthweight increases with maternal age, it does so at a decreasing rate until a certain maternal age after which it begins decreasing.
 
@@
 ```
+
+### Model 4
 
@@
 So I will perform some sub-sample modeling to understand this issue.
-We will create a `Brazilian` (yes, no) variable, drop all non-Brazilians from the study, and run the original regression with the covariates.
-
-```{r}
-# Create dummy variables for new category of Brazilian or not
-# Create dummy variables for status and change to English
-ML_linear <- mutate(ML_linear, 
-                    Brazilian = recode_factor(ModifiedNationality, 
-                                              "European" = "Non-Brazilian",
-                                              "Middle Eastern" = "Non-Brazilian",
-                                              "Latin American" = "Non-Brazilian",
-                                              "Brazilian" = "Brazilian"))
-
-# Check the structure of the updated data frame
-str(ML_linear)
-
-# Save ML_Linear as RDS
-save_data_location <- here::here("data","processed-data","ML_linear.rds")
-saveRDS(ML_linear, file = save_data_location)
-```
+In `eda-v1.qmd` I created a `Brazilian` (yes, no) variable, so here, I will drop all non-Brazilians from the study, and run the original regression with the covariates.
+
+### Model 5
 
@@
 # Set reference level for Brazilian
-ML_linear$Brazilian <- relevel(ML_linear$Brazilian, ref = "Brazilian")
+ML_linear$Brazilian <- relevel(ML_linear$Brazilian, ref = "yes")
 
@@
 
+### Model 6
+
 Here, non-Brazilians had babies on average weighing 100 grams more than Brazilians (p-value = 0.001).
@@
 
-## Multivariate analysis
+# Linear Models Color
 
@@
 ```
+
+## Model 7
 
@@
 saveRDS(table_fit7, file = here::here("results", "tables", "table_fit7.rds"))
-
-# Complex linear model
+```
+
+## Model 8
+
+```{r}
+# Model 8 multivariate linear model
 # Set reference level for Color
@@
 
-## Logistic models
+# Year Fixed-Effect
+
+## Model Years (9, 10, 11)
+
+In the first round of peer review, one reviewer suggested that I should justify why I provide this as a cross-sectional approach rather than test for fixed effects by year given periodic epidemics in infectious disease such as malaria and yellow fever and changes in real wages.
+So I will run a model that includes year as a fixed effect here with the `ModifiedColor` variable as the main predictor variable.
+
+```{r}
+# Model 9: multivariate linear model with year as fixed effect using ModifiedColor
+ML_linear <- ML_linear %>%
+  mutate(Year = as.factor( # Create a new column called "Year" and make it a factor
+    substr(Date, 1, 4) # Take the first 4 characters of the Date column
+  ))
+
+# Save new variable into dataset
+save_data_location <- here::here("data","processed-data","ML_linear.rds")
+saveRDS(ML_linear, file = save_data_location)
+
+# Set reference level for Color
+ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm9 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality + Year, data = ML_linear)
+
+# Summarize model fit with tidiers
+ML_lm9_summary <- summary(ML_lm9) # Summary of fit
+tidy(ML_lm9, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+model_stats9 <- glance(ML_lm9) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values3$model9 <- model_stats9$r.squared
+
+# Create a gtsummary table directly from the model
+table_fit9 <- tbl_regression(ML_lm9, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table_fit9
+
+# Save regression output into table
+saveRDS(table_fit9, file = here::here("results", "tables", "table_fit9.rds"))
+```
+
+I will also run the model five times, once per year.
+
+```{r}
+# Model 10: multivariate linear model for each year (1922-1924) with ModifiedColor
+# Set reference level for Color
+ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+
+# Ensure year is created as character vector
+years <- c("1922", "1923", "1924", "1925", "1926")
+
+# Create models for all years at once
+table_years <- ML_linear %>%
+  mutate(Year = as.character(substr(Date, 1, 4))) %>% # Create a new column called "Year" and make it a factor
+  filter(Year %in% years) %>%
+  group_by(Year) %>%
+  nest() %>%
+  mutate(
+    model = map(data, ~linear_reg() %>% 
+                  set_engine("lm") %>% 
+                  fit(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, 
+                      data = .x)),
+    tables = map(model, ~tbl_regression(.x, 
+                                      exponentiate = FALSE, 
+                                      intercept = TRUE) %>% 
+                   bold_p() %>% 
+                   add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+                   modify_column_unhide(columns = c(statistic, std.error)))
+  )
+
+# To view results for a specific year:
+table_years %>% 
+  filter(Year == "1922") %>% 
+  pull(tables) %>% 
+  .[[1]]
+
+# Create named list
+tables_list <- table_years %>%
+  arrange(Year) %>%
+  pull(tables) %>%
+  set_names(paste0("Year_", years))
+
+# Create gt summary table
+combined_table_years <- tbl_merge(
+ tbls = tables_list,
+ tab_spanner = c("**1922**", "**1923**", "**1924**", "**1925**", "**1926**")
+) %>%
+ modify_table_styling(
+   columns = everything(),
+   align = "center"
+ ) %>%
+ modify_table_body(fun = ~.x %>% arrange(variable))
+combined_table_years
+
+# Save regression output into table
+saveRDS(combined_table_years, file = here::here("results", "tables", "combined_table_years.rds"))
+```
+
+Given that 1923 demonstrates a coefficient for infant birth weights to Afro-descended mothers nearly 3 times the overall coefficient (-84 to -237), I will drop the observations from 1923 and re-run the regression to see if the results are still robust.
+The coefficient for infants born to Afro-descended mothers is not significant and is lower than in the full model (-40).
+This demonstrates that the births from 1923 seem to be driving the overall results.
+Nonetheless, the overall hypothesis still holds.
+Infants born to Afro-descended mothers had lower birth weights than those born to Euro-descended mothers, both in years where all birth weights were higher and years in which all birth weights were lower.
+Economic uncertainty affected women of African descent more than those of European descent.
+Additionally, these results could be due to smaller sample sizes and larger standard errors -- noise in the data.
+
+```{r}
+# Model 11: Multivariate model using ModifiedColor without observations from 1923
+# Filter out observations from 1923
+ML_linear_filtered <- ML_linear %>%
+  filter(Year != "1923")
+
+# Set reference level for Color
+ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")
+
+# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
+ML_lm11 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality + Year, data = ML_linear_filtered)
+
+# Summarize model fit with tidiers
+ML_lm11_summary <- summary(ML_lm11) # Summary of fit
+tidy(ML_lm11, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
+model_stats11 <- glance(ML_lm11) # Extract model-level statistics
+
+# Extract R-squared value
+r_squared_values3$model11 <- model_stats11$r.squared
+
+# Create a gtsummary table directly from the model
+table_fit11 <- tbl_regression(ML_lm11, exponentiate = FALSE, intercept = TRUE) %>% 
+  bold_p() %>% 
+  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
+  modify_column_unhide(columns = c(statistic, std.error))
+table_fit11
+
+# Save regression output into table
+saveRDS(table_fit11, file = here::here("results", "tables", "table_fit11.rds"))
+```
+
+# Logistic models
 
@@
 ```
+
+## Log Model 1
 
@@
 
+## Log Model 2
+
 Now I will run another logistic model with `Color` as the predictor.
@@
 logistic_model_metrics2 <- list(
-  pseudo_r_squared = mcfadden_r2_1,
+  pseudo_r_squared = mcfadden_r2_2,
   auc = roc_auc_result1$.estimate
@@
 The pseudo- $R^2$ is `r round(all_logistic_models$model2$auc, 3)`.
+
+## Log Model 3
 
@@
 logistic_model_metrics3 <- list(
-  pseudo_r_squared = mcfadden_r2_1,
+  pseudo_r_squared = mcfadden_r2_3,
   auc = roc_auc_result1$.estimate
@@
 The ROC_AUC here is `r round(all_logistic_metrics$model3$auc, 3)`, still worse than random guessing. The pseudo-$R^2$ is `r round (all_logistic_metrics$model3$pseudo_r_squared, 3)`.
+
+## Log Model 4
 
@@
 # Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
-log4_recipe <- recipe(BirthweightCategory ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
-  step_dummy(all_nominal(), -all_outcomes()) %>%
-  step_normalize(all_predictors())
+log4_recipe <- recipe(BirthweightCategory ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) #%>%
+ # step_dummy(all_nominal(), -all_outcomes()) %>%
+ # step_normalize(all_predictors())
 
@@
 logistic_model_metrics4 <- list(
-  pseudo_r_squared = mcfadden_r2_1,
+  pseudo_r_squared = mcfadden_r2_4,
   auc = roc_auc_result1$.estimate
@@
 
-# Merge unadjusted log models 1 and 2
+## Merge unadjusted log models 1 and 2
 
@@
 saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
+```
+
+## Log Model Stilbirth
+
+Reviewer 3 wrote: "A logit analysis using stillbirth as an outcome would bring robustness to the racial disparities that the author highlights.
+That there are 221 stillbirths creates this possibility and might even reveal greater racial disparities.
+While these could not be definitively linked to syphilis, this might provide further insight into that data limitation.
+Again, causation cannot be overstated, but given that the author cannot speak to prematurity with the data (another important indicator of maternal stress), this would be an interesting avenue to explore."
+
+```{r}
+# Check factor levels
+levels(ML_summary$ModifiedColor) # Reference is Euro-Descent
+levels(ML_summary$FetalOutcome) # Reference is Live Birth
+
+# Set the reference level explicitly to NBW
+ML_summary_log <- ML_summary %>%
+  filter(!is.na(FetalOutcome) & !is.na(ModifiedColor)) %>% #filter out NAs
+  mutate(FetalOutcome = relevel(FetalOutcome, ref = "Live Birth"),
+         ModifiedColor = relevel(ModifiedColor, ref = "Euro-Descent"))
+
+#set seed
+set.seed(456)
+ML_log_split2 <- initial_split(ML_summary_log, prop = 0.75, strata = FetalOutcome)
+
+# Create training and test datasets
+ML_log_train2 <- training(ML_log_split2)
+ML_log_test2 <- testing(ML_log_split2)
+
+# Logistic model 5: simple logistic regression model with stillbirth as outcome and ModifiedAncestry as the predictor
+# Create recipe
+# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
+log5_recipe <- recipe(FetalOutcome ~ ModifiedColor, data = ML_summary_log) # %>%
+  #step_dummy(all_nominal(), -all_outcomes()) %>%
+  #step_normalize(all_predictors())
+
+# Define model specification
+log5_spec <- logistic_reg() %>%
+  set_engine("glm") %>%
+  set_mode("classification")
+
+# Create workflow
+log5_wf <- workflow() %>%
+  add_recipe(log5_recipe) %>%
+  add_model(log5_spec)
+
+# Fit model
+log5_fit <- log5_wf %>%
+fit(data = ML_summary_log)
+
+# Extract fitted model from workflow
+log5_fitted <- extract_fit_parsnip(log5_fit)$fit
+
+# Create a gtsummary table directly from the model
+table_log5 <- tbl_regression(log5_fitted, exponentiate = TRUE) # exponentiate true for ORs
+
+#Bold significant p-values and add n
+table_log5 <- bold_p(table_log5, t = 0.05) %>%
+  add_n()
+table_log5
+
+# Save regression output into table
+saveRDS(table_log5, file = here::here("results", "tables", "table_log5.rds"))
+
+# Calculate metrics using pseudo-R2
+# Extract model metrics using broom
+model_metrics5 <- glance(log5_fitted)
+
+# Calculate McFadden's R-squared
+# McFadden's R-squared = 1 - (deviance(model) / deviance(null model))
+mcfadden_r2_5 <- 1 - (model_metrics5$deviance / model_metrics5$null.deviance)
+
+# Print McFadden's R-squared
+print(mcfadden_r2_5)
+
+# Calculate performance metrics using ROC_AUC
+# Fit the model on the training data
+log5_fit <- log5_wf %>% fit(data = ML_log_train2)
+
+# Predict probabilities on the test set
+prob_preds5 <- predict(log5_fit, new_data = ML_log_test2, type = "prob")
+
+# Combine predictions with true values
+results5 <- bind_cols(ML_log_test2, prob_preds5)
+
+# Calculate ROC-AUC
+roc_auc_result5 <- roc_auc(results5, truth = FetalOutcome, `.pred_Live Birth`)
+
+# Print ROC-AUC
+print(roc_auc_result5)
+
+# Create and plot the ROC curve
+roc_curve_data5 <- roc_curve(results5, truth = FetalOutcome, `.pred_Live Birth`)
+
+# Plotting the ROC curve
+roc_plot5 <- ggplot(roc_curve_data5, aes(x = 1 - specificity, y = sensitivity)) +
+  geom_path() +
+  geom_abline(lty = 2) +
+  labs(title = "ROC Curve",
+       x = "1 - Specificity",
+       y = "Sensitivity")
+
+# Print the plot
+print(roc_plot5)
+
+# Store metrics in a list for dynamic inline text
+logistic_model_metrics5 <- list(
+  pseudo_r_squared = mcfadden_r2_5,
+  auc = roc_auc_result1$.estimate
+)
+all_logistic_models$model5 <- logistic_model_metrics5 # add to list of lists1
 ```

### 2024-11-14 code/eda-code/eda-v1.qmd
RRSSM2
Age2; immigrants; year fixed effects

@@
 
+I will now change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.
+In the future, I should perform this step in the processing script.
+
 ```{r}
@@
 
+# Change observations from Portuguese to English
+ML_summary <- ML_summary %>%
+  mutate(
+    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
+    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
+    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
+    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
+    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth"),
+    Status = recode_factor(Status, "Multigesta" = "Multigravida", "Trigesta" = "Trigravida", "Secundigesta" = "Secundigravida", "Secundipara" = "Secundiparous", "Multipara" = "Multiparous", "Primigesta" = "Primigravida", "Nulipara" = "Nulliparous", "Primipara" = "Primiparous")
+    )
+
 # Recode factor using dplyr
-ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Parda" = "Afro-Descent", "Preta" = "Afro-Descent", "Branca" = "Euro-Descent")
+ML_summary$ModifiedColor <- recode_factor(ML_summary$Color, "Mixed Race" = "Afro-Descent", "Black" = "Afro-Descent", "White" = "Euro-Descent")
 
 # Create dummy variables for status and change to English
-ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multiparous", "Trigesta" = "Multiparous", "Multigesta" = "Multiparous", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nulliparous", "Nulipara" = "Nulliparous", "Primipara" = "Nulliparous")
+ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigravida" = "Multiparous", "Trigravida" = "Multiparous", "Multigravida" = "Multiparous", "Secundiparous" = "Multiparous", "Multiparous" = "Multiparous", "Primigravida" = "Nulliparous", "Nulliparous" = "Nulliparous", "Primiparous" = "Nulliparous")
 
-# Create dummy variables for nationality and change to English
-ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "Latin American", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "Latin American", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "Middle Eastern", "Uruguaya" = "Latin American")
+# Create dummy variables for nationality
+ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "German" = "European", "Argentine" = "Latin American", "Austrian" = "European", "Brazilian" = "Brazilian", "Spanish" = "European", "French" = "European", "Italian" = "European", "Paraguayan" = "Latin American", "Polish" = "European", "Portuguese" = "European", "Romanian" = "European", "Russian" = "European", "Swiss" = "European", "Syrian" = "Middle Eastern", "Uruguayan" = "Latin American")
 
@@
   mutate(
-    Weightblack = factor(if_else(Color == "Preta", "Yes", "No"), levels = c("No", "Yes")),
-    Weightmixed = factor(if_else(Color == "Parda", "Yes", "No"), levels = c("No", "Yes")),
-    Weightwhite = factor(if_else(Color == "Branca", "Yes", "No"), levels = c("No", "Yes")),
+    Weightblack = factor(if_else(Color == "Black", "Yes", "No"), levels = c("No", "Yes")),
+    Weightmixed = factor(if_else(Color == "Mixed Race", "Yes", "No"), levels = c("No", "Yes")),
+    Weightwhite = factor(if_else(Color == "White", "Yes", "No"), levels = c("No", "Yes")),
     Weightfemale = factor(if_else(Sex == "F", "Yes", "No"), levels = c("No", "Yes")),
@@
 
+# Create dummy variables for birth weight for infants born to mothers of different nationalities for summary table.
+ML_summary <- ML_summary %>%
+  mutate( # First create yes/not dummy variables for count
+    WeightPor = factor(if_else(Nationality == "Portuguese", "Yes", "No"), levels = c("No", "Yes")),
+    WeightGer = factor(if_else(Nationality == "German", "Yes", "No"), levels = c("No", "Yes")),
+    WeightIta = factor(if_else(Nationality == "Italian", "Yes", "No"), levels = c("No", "Yes")),
+    WeightAus = factor(if_else(Nationality == "Austrian", "Yes", "No"), levels = c("No", "Yes")),
+    WeightFr = factor(if_else(Nationality == "French", "Yes", "No"), levels = c("No", "Yes")),
+    WeightSp = factor(if_else(Nationality == "Spanish", "Yes", "No"), levels = c("No", "Yes")),
+    WeightPol = factor(if_else(Nationality == "Polish", "Yes", "No"), levels = c("No", "Yes")),
+    WeightSw = factor(if_else(Nationality == "Swiss", "Yes", "No"), levels = c("No", "Yes")),
+    WeightRom = factor(if_else(Nationality == "Romanian", "Yes", "No"), levels = c("No", "Yes")),
+    WeightRus = factor(if_else(Nationality == "Russian", "Yes", "No"), levels = c("No", "Yes")),
+    WeightArg = factor(if_else(Nationality == "Argentine", "Yes", "No"), levels = c("No", "Yes")),
+    WeightPar = factor(if_else(Nationality == "Paraguayan", "Yes", "No"), levels = c("No", "Yes")),
+    WeightUr = factor(if_else(Nationality == "Uruguayan", "Yes", "No"), levels = c("No", "Yes")),
+    WeightSy = factor(if_else(Nationality == "Syrian", "Yes", "No"), levels = c("No", "Yes")),
+    WeightEuro = factor(if_else(Nationality == "German" | 
+                                  Nationality == "Italian" |
+                                  Nationality == "Austrian" |
+                                  Nationality == "French" |
+                                  Nationality == "Spanish" |
+                                  Nationality == "Polish" |
+                                  Nationality == "Swiss" |
+                                  Nationality == "Romanian" |
+                                  Nationality == "Russian", "Yes", "No"), levels = c("No", "Yes")),
+    WeightBrWh = factor(if_else(Nationality == "Brazilian" & Color == "White", "Yes", "No"), levels = c("No", "Yes")),
+    WeightBrBl = factor(if_else(Nationality == "Brazilian" & Color == "Black", "Yes", "No"), levels = c("No", "Yes")),
+    WeightBrMx = factor(if_else(Nationality == "Brazilian" & Color == "Mixed Race", "Yes", "No"), levels = c("No", "Yes")),
+    WeightgramsPor = if_else(Nationality == "Portuguese", Weightgrams, NA_real_), # then create weight variables
+    WeightgramsGer = if_else(Nationality == "German", Weightgrams, NA_real_),
+    WeightgramsIta = if_else(Nationality == "Italian", Weightgrams, NA_real_),
+    WeightgramsAus = if_else(Nationality == "Austrian", Weightgrams, NA_real_),
+    WeightgramsFr = if_else(Nationality == "French", Weightgrams, NA_real_),
+    WeightgramsSp = if_else(Nationality == "Spanish", Weightgrams, NA_real_),
+    WeightgramsPol = if_else(Nationality == "Polish", Weightgrams, NA_real_),
+    WeightgramsSw = if_else(Nationality == "Swiss", Weightgrams, NA_real_),
+    WeightgramsRom = if_else(Nationality == "Romanian", Weightgrams, NA_real_),
+    WeightgramsRus = if_else(Nationality == "Russian", Weightgrams, NA_real_),
+    WeightgramsArg = if_else(Nationality == "Argentine", Weightgrams, NA_real_),
+    WeightgramsPar = if_else(Nationality == "Paraguayan", Weightgrams, NA_real_),
+    WeightgramsUr = if_else(Nationality == "Uruguayan", Weightgrams, NA_real_),
+    WeightgramsSy = if_else(Nationality == "Syrian", Weightgrams, NA_real_),
+    WeightgramsEuro = if_else(WeightEuro == "Yes", Weightgrams, NA_real_),
+    WeightgramsBrWh = if_else(Nationality == "Brazilian" & Color == "White", Weightgrams, NA_real_),
+    WeightgramsBrBl = if_else(Nationality == "Brazilian" & Color == "Black", Weightgrams, NA_real_),
+    WeightgramsBrMx = if_else(Nationality == "Brazilian" & Color == "Mixed Race", Weightgrams, NA_real_)
+  )
+
+# Verify new variables
+str(ML_summary[, c("WeightPor", "WeightgramsPor", 
+                   "WeightGer", "WeightgramsGer", 
+                   "WeightIta", "WeightgramsIta", 
+                   "WeightAus", "WeightgramsAus", 
+                   "WeightFr", "WeightgramsFr", 
+                   "WeightSp", "WeightgramsSp", 
+                   "WeightPol", "WeightgramsPol", 
+                   "WeightSw", "WeightgramsSw",
+                   "WeightRom", "WeightgramsRom",
+                   "WeightRus", "WeightgramsRus",
+                   "WeightArg", "WeightgramsArg",
+                   "WeightPar", "WeightgramsPar",
+                   "WeightUr", "WeightgramsUr",
+                   "WeightSy", "WeightgramsSy",
+                   "WeightEuro", "WeightgramsEuro",
+                   "WeightBrWh", "WeightgramsBrWh", 
+                   "WeightBrBl", "WeightgramsBrBl", 
+                   "WeightBrMx", "WeightgramsBrMx")])
+
 # View the updated dataframe
 print(head(ML_summary))
+
+# Calculate both mean and median weights by nationality
+Euro_stats <- ML_summary %>%
+  reframe(
+    Portuguese = list(
+      n = sum(!is.na(WeightgramsPor)),
+      mean = mean(WeightgramsPor, na.rm = TRUE),
+      median = median(WeightgramsPor, na.rm = TRUE)
+    ),
+    German = list(
+      n = sum(!is.na(WeightgramsGer)),
+      mean = mean(WeightgramsGer, na.rm = TRUE),
+      median = median(WeightgramsGer, na.rm = TRUE)
+    ),
+    Italian = list(
+      n = sum(!is.na(WeightgramsIta)),
+      mean = mean(WeightgramsIta, na.rm = TRUE),
+      median = median(WeightgramsIta, na.rm = TRUE)
+    ),
+    Austrian = list(
+      n = sum(!is.na(WeightgramsAus)),
+      mean = mean(WeightgramsAus, na.rm = TRUE),
+      median = median(WeightgramsAus, na.rm = TRUE)
+    ),
+    French = list(
+      n = sum(!is.na(WeightgramsFr)),
+      mean = mean(WeightgramsFr, na.rm = TRUE),
+      median = median(WeightgramsFr, na.rm = TRUE)
+    ),
+    Spanish = list(
+      n = sum(!is.na(WeightgramsSp)),
+      mean = mean(WeightgramsSp, na.rm = TRUE),
+      median = median(WeightgramsSp, na.rm = TRUE)
+    ),
+    Polish = list(
+      n = sum(!is.na(WeightgramsPol)),
+      mean = mean(WeightgramsPol, na.rm = TRUE),
+      median = median(WeightgramsPol, na.rm = TRUE)
+    ),
+    Swiss = list(
+      n = sum(!is.na(WeightgramsSw)),
+      mean = mean(WeightgramsSw, na.rm = TRUE),
+      median = median(WeightgramsSw, na.rm = TRUE)
+    ),
+    Romanian = list(
+      n = sum(!is.na(WeightgramsRom)),
+      mean = mean(WeightgramsRom, na.rm = TRUE),
+      median = median(WeightgramsRom, na.rm = TRUE)
+    ),
+    Russian = list(
+      n = sum(!is.na(WeightgramsRus)),
+      mean = mean(WeightgramsRus, na.rm = TRUE),
+      median = median(WeightgramsRus, na.rm = TRUE)
+    ),
+    Euro = list(
+      n = sum(!is.na(WeightgramsEuro)),
+      mean = mean(WeightgramsEuro, na.rm = TRUE),
+      median = median(WeightgramsEuro, na.rm = TRUE)
+    ),
+    Argentine = list(
+      n = sum(!is.na(WeightgramsArg)),
+      mean = mean(WeightgramsArg, na.rm = TRUE),
+      median = median(WeightgramsArg, na.rm = TRUE)
+    ),
+    Paraguayan = list(
+      n = sum(!is.na(WeightgramsPar)),
+      mean = mean(WeightgramsPar, na.rm = TRUE),
+      median = median(WeightgramsPar, na.rm = TRUE)
+    ),
+    Uruguayan = list(
+      n = sum(!is.na(WeightgramsUr)),
+      mean = mean(WeightgramsUr, na.rm = TRUE),
+      median = median(WeightgramsUr, na.rm = TRUE)
+    ),
+    Syrian = list(
+      n = sum(!is.na(WeightgramsSy)),
+      mean = mean(WeightgramsSy, na.rm = TRUE),
+      median = median(WeightgramsSy, na.rm = TRUE)
+    ),
+    Brazil_white = list(
+      n = sum(!is.na(WeightgramsBrWh)),
+      mean = mean(WeightgramsBrWh, na.rm = TRUE),
+      median = median(WeightgramsBrWh, na.rm = TRUE)
+    ),
+    Brazil_black = list(
+      n = sum(!is.na(WeightgramsBrBl)),
+      mean = mean(WeightgramsBrBl, na.rm = TRUE),
+      median = median(WeightgramsBrBl, na.rm = TRUE)
+    ),
+    Brazil_mixed = list(
+      n = sum(!is.na(WeightgramsBrMx)),
+      mean = mean(WeightgramsBrMx, na.rm = TRUE),
+      median = median(WeightgramsBrMx, na.rm = TRUE)
+    )
+  ) %>%
+  tidyr::unnest_longer(col = everything())
+Euro_stats
+
+# T-test between Portuguese and all other European nationalities combined
+t_test_euro <- t.test(ML_summary$WeightgramsPor, ML_summary$WeightgramsEuro)
+t_test_euro
+
+t_test_brazil <- t.test(ML_summary$WeightgramsPor, ML_summary$WeightgramsBrWh)
+t_test_brazil
+
+# Put t-test into gtsummary table for supplemental analysis
+# Create a table with both t-tests
+table_ttest <- tibble(
+  Comparison = c("Portuguese vs Other European", "Portuguese vs Brazilian White"),
+  `t-statistic` = c(t_test_euro$statistic, t_test_brazil$statistic),
+  `p-value` = c(t_test_euro$p.value, t_test_brazil$p.value),
+  `Mean Difference` = c(diff(t_test_euro$estimate), diff(t_test_brazil$estimate)),
+  `95% CI Lower` = c(t_test_euro$conf.int[1], t_test_brazil$conf.int[1]),
+  `95% CI Upper` = c(t_test_euro$conf.int[2], t_test_brazil$conf.int[2])
+    ) %>%
+  gt::gt() %>%
+  gt::fmt_number(
+    columns = c(`t-statistic`),
+    decimals = 2
+  ) %>%
+  gt::fmt_number(
+    columns = c(`p-value`),
+    decimals = 3
+  ) %>%
+  gt::fmt_number(
+    columns = c(`Mean Difference`, `95% CI Lower`, `95% CI Upper`),
+    decimals = 1
+  ) %>%
+  gt::tab_header(
+    title = "T-Test Results"
+  )
+table_ttest
+
+# Save the table
+saveRDS(table_ttest, file = here::here("results", "tables", "table_ttest_final.rds"))
+
+# Put means into gtsummary table for supplemental analysis
+table_nationality_stats <- ML_summary %>%
+  select(WeightgramsPor, WeightgramsGer, WeightgramsIta, WeightgramsAus,
+         WeightgramsFr, WeightgramsSp, WeightgramsPol, WeightgramsSw,
+         WeightgramsRom, WeightgramsRus, WeightgramsEuro, WeightgramsArg,
+         WeightgramsPar, WeightgramsUr, WeightgramsSy, WeightgramsBrWh,
+         WeightgramsBrBl, WeightgramsBrMx) %>%
+  tbl_summary(
+    missing = "no",
+    label = list(
+      WeightgramsPor ~ "Portuguese",
+      WeightgramsGer ~ "German",
+      WeightgramsIta ~ "Italian",
+      WeightgramsAus ~ "Austrian",
+      WeightgramsFr ~ "French",
+      WeightgramsSp ~ "Spanish",
+      WeightgramsPol ~ "Polish",
+      WeightgramsSw ~ "Swiss",
+      WeightgramsRom ~ "Romanian",
+      WeightgramsRus ~ "Russian",
+      WeightgramsEuro ~ "European (Exc. Portuguese)",
+      WeightgramsArg ~ "Argentine",
+      WeightgramsPar ~ "Paraguayan",
+      WeightgramsUr ~ "Uruguayan",
+      WeightgramsSy ~ "Syrian",
+      WeightgramsBrWh ~ "Brazilian White",
+      WeightgramsBrBl ~ "Brazilian Black",
+      WeightgramsBrMx ~ "Brazilian Mixed-Race"
+    ),
+    type = list(everything() ~ "continuous"),
+    statistic = list(all_continuous() ~ "{mean} ({p25}, {p75})"),
+    digits = everything() ~ 0 # all numbers to 0 decimal places
+  ) %>%
+  add_n() %>%
+  modify_header(label ~ "**Nationality**",
+                stat_0 ~ "**Mean (Q1, Q3)**",
+                n ~ "**N**")
+table_nationality_stats
+
+# Save table for supplemental analysis
+saveRDS(table_nationality_stats, file = here::here("results", "tables", "table_nationality_stats_final.rds"))
 ```
-I will now change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.
-In the future, I should perform this step in the processing script.
+The t-tests demonstrate no significant difference in the mean birthweight between Portuguese and all other European immigrants and between Portuguese and White Brazilian women.
+This suggests that differentiating European nationality further will not influence the regression results.
+
+Further, because there is only 1 observation for Austrian nationality; 3 observations for French nationality; and 2 observations for Polish and Swiss nationalities, I will drop these from my larger summary table.
 
 ```{r}
-# Change rest of observations from Portuguese to English
-ML_summary <- ML_summary %>%
-  mutate(
-    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
-    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
-    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
-    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
-    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth"),
-    Status = recode_factor(Status, "Multigesta" = "Multigravida", "Trigesta" = "Trigravida", "Secundigesta" = "Secundigravida", "Secundipara" = "Secundiparous", "Multipara" = "Multiparous", "Primigesta" = "Primigravida", "Nulipara" = "Nulliparous", "Primipara" = "Primiparous")
-    )
-
 # Summarize missing data using dplyr
@@
 
-`Gtsummary` is a very easy way to create summary statistics (Table 1). For previous code I worked on in relation to summary stats, see the `edacode-extra.R` file.
+`Gtsummary` is a very easy way to create summary statistics (Table 1).
+For previous code I worked on in relation to summary stats, see the `edacode-extra.R` file.
+I have excluded all nationalities with sample sizes smaller than 10 for infant birth weight (Argentine, Paraguayan, Syrian, Uruguayan, Romanian, French, Polish, Swiss, and Austrian).
 
@@
   tbl_summary(
-    include = c(Color, ModifiedColor, Status, ModifiedStatus, Age, Nationality, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams),
+    include = c(Color, Status, Age, Nationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams),
     missing = "no",
-    label = list(ModifiedColor ~ "Ancestry", Status ~ "Parity or Gravidity", ModifiedStatus ~ "Parity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Lengthcentimeters ~ "Birth Length (cms)", Weightgrams ~ "Birth Weight (grms)"),
+    label = list(
+      Color ~ "Skin color",
+      Status ~ "Parity or Gravidity",
+      Age ~ "Maternal Age",
+      Nationality ~ "Nationality",
+      Birth ~ "Birth Outcome", 
+      MaternalOutcome ~ "Maternal Outcome", 
+      FetalOutcome ~ "Fetal Outcome",
+      Sex ~ "Infant Sex",
+      Lengthcentimeters ~ "Birth Length (cms)", 
+      Weightgrams ~ "Birth Weight (grms)"
+      ),
     statistic = list(all_continuous() ~ "{mean} ({p25}, {p75})")
@@
   ML_summary %>%
-  mutate(
-    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
-    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
-    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
-    WeightgramsAfro = if_else(ModifiedColor == "Afro-Descent", Weightgrams, NA_real_),
-    WeightgramsEuro = if_else(ModifiedColor == "Euro-Descent", Weightgrams, NA_real_),
-    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
-    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
-  ) %>%
   tbl_summary(
-    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsAfro, WeightgramsEuro, WeightgramsFemale, WeightgramsMale, BirthweightCategory),
+    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsAfro, WeightgramsEuro, WeightgramsPor, WeightgramsGer, WeightgramsIta, WeightgramsSp, WeightgramsRus, WeightgramsBrWh, WeightgramsFemale, WeightgramsMale, BirthweightCategory),
     missing = "no",
+    type = list(WeightgramsRus ~ "continuous"),
     label = list(
@@
       WeightgramsEuro ~ "Euro-Descent",
+      WeightgramsPor ~ "Portuguese",
+      WeightgramsGer ~ "German",
+      WeightgramsIta ~ "Italian",
+      WeightgramsSp ~ "Spanish",
+      WeightgramsRus ~ "Russian",
+      WeightgramsBrWh ~ "White Brazilian",
       WeightgramsFemale ~ "Female",
@@
        all_continuous() ~ "{mean} ({p25}, {p75})",
-      WeightgramsBlack ~ "{mean} ({p25}, {p75})",
-      WeightgramsMixed ~ "{mean} ({p25}, {p75})",
-      WeightgramsWhite ~ "{mean} ({p25}, {p75})",
-      WeightgramsFemale ~ "{mean} ({p25}, {p75})",
-      WeightgramsMale ~ "{mean} ({p25}, {p75})"
-    ),
-    digits = list(WeightgramsBlack ~ 0, WeightgramsMixed ~ 0, WeightgramsWhite ~ 0, WeightgramsFemale ~ 0, WeightgramsMale ~ 0)
+       WeightgramsRus ~ "{mean} ({p25}, {p75})"
+    )
   ) %>%
@@
         variable == "WeightgramsEuro" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsPor" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsGer" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsIta" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsSp" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsRus" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
+        variable == "WeightgramsBrWh" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
         variable == "WeightgramsFemale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
@@
   filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
-  filter(Birth != "abortion") %>% # Exclude rows with abortion
+  filter(Birth != "Abortion") %>% # Exclude rows with abortion
   count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
@@
 
+Now I will compute total fertility rates for Rio de Janeiro, São Paulo, and Brazil from 1940 data.
+
+```{r}
+# Path to summary data. Note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","Fertility_Rates.rds")
+
+# Load data
+Fertility_Rates <- readRDS(data_location)
+
+# Check structure
+str(Fertility_Rates)
+
+tfr_region <- Fertility_Rates %>%
+  group_by(Region) %>%
+  summarize(TFR = round(sum(`Age-Specific Rate`)/100, 1))
+
+# To view the results
+print(tfr_region)
+```
+
 # ML Linear Data
@@
                              "yes", "no"),
-                      levels = c("yes", "no"))
-  )
+                      levels = c("yes", "no")),
+    AgeSquared = Age^2) # create squared term for age
 

### 2024-11-14 code/processing-code/processingfile-v1.qmd
RRSSM2
Age2; immigrants; year fixed effects

@@
 ```
-# Notes
+# Data loading for total fertility rates (TFR)
+
+I will also discuss total fertility rates (TFR) for Rio de Janeiro and Brazil.
+
+```{r}
+# Path to data
+# Note the use of the here() package and not absolute paths
+data_location <- here::here("data","raw-data","1940_FertilityRates.csv")
+Fertility_Rates <- read_csv(data_location)
+
+# Structure of Data
+glimpse(Fertility_Rates)
+
+# Save Mean_Birthweight as RDS
+save_data_location <- here::here("data","processed-data","Fertility_Rates.rds")
+saveRDS(Fertility_Rates, file = save_data_location)
+```

### 2024-11-14 products/manuscript/Manuscript_Final_SSM.qmd
RRSSM2
Age2; immigrants; year fixed effects

@@
 Epidemiologists have shown that health disparities continue to mirror patterns from over a century ago.
-This cross-sectional analysis quantifies health disparities in the post-abolition period using infant birth weight in one of Brazil's first public maternity hospitals.
-It relies on hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras in Rio de Janeiro, Brazil to run linear models assessing differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality.
+This cross-sectional analysis quantifies health disparities in a post-abolition urban center using infant birth weight in one of Brazil's first public maternity hospitals, the Maternidade Laranjeiras.
+It relies on hospital records on infants delivered between 1922-1926 (n=2,845) at the Maternidade Laranjeiras in the city of Rio de Janeiro, Brazil to run linear models assessing differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality.
 African ancestry was correlated with lower birth weights.
@@
 The findings were likely the consequence of slavery’s legacy, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for people of African descent -- and possible epigenetic effects from the lived experience of racism.
-The findings are consistent with current-day research on racialized health disparities in Brazil.
+The findings are consistent with current-day research on racialized health disparities in Brazil and demonstrate the importance of historical findings to public health research.
 
@@
 This study analyzes infant birth weight as a proxy for general maternal and fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras, a public maternity hospital in Rio de Janeiro, Brazil between 1922 and 1926.
-It aims to provide basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil, substantiating the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
+It aims to provide basic quantitative associations of the health status of the descendants of enslaved individuals in <an important urban city> in post-abolition Brazil, substantiating the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
 
@@
 Studies in the human sciences have hypothesized an intergenerational effect of poor health conditions, including undernutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].
-In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@adamo1998; @otovo2016; @roth2020; @telles2022].
+<Historians have emphasized how race and socioeconomic status (SES) were deeply intertwined in post-slave societies, including Brazil and thus disparities in literacy and access to education likely contributed to a health gap [@adamo1998; @ball2020; @franken2019; @franken2025].> In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@adamo1998; @otovo2016; @roth2020; @telles2022].
 
-Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021], compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
-Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission -- the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
+Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery, compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
+Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission -- the freeing of an enslaved person -- can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
 Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere.
@@
 In the aftermath of abolition, no state-run efforts to incorporate formerly enslaved, African-descended peoples into economic, civic, or political life occurred [@fischer2008].
-Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower literacy levels, and had worse health outcomes than their White and White immigrant counterparts [@adamo1998; @ball2020; @fischer2008; @otovo2016].
+Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower literacy levels, <fewer employment opportunities>, and had worse health outcomes than their White and White immigrant counterparts [@adamo1998; @ball2020; @fischer2008; @otovo2016].
 
-Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019].
+Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in urban centers in Brazil's southeastern and southern regions, where Rio de Janeiro is located [@franken2025; @monasterio2014].
 The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe.
-Improved sanitation measures helped stem infectious disease outbreaks [@benchimol1999; @hochman2013].
+In cities, improved sanitation measures helped stem infectious disease outbreaks [@benchimol1999; @hochman2013].
 Infant mortality rates declined from the nineteenth century, although the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@adamo1998; @roth2020; @loudon1992; @woods2009].
+Yet, these advancements were not distributed evenly across the country, as rural-urban and regional divides in health outcomes including maternal mortality, infant mortality (particularly neonatal mortality), and child stunting from malnutrition developed that only began to decrease in the new Millennium [@scheper-hughes1992; @victora2011].
 
-Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, provided healthcare free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
-Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe.
+<A rural-urban and regional divide was also noticeable in fertility patterns.
+Women in cities in the wealthier southeastern and southern regions, particularly Rio de Janeiro, had begun having fewer children at least as early as the 1920s.
+By the early 1930s, scholars demonstrate that Rio de Janeiro, although not Brazil as a whole, had begun its fertility transition from a period of high birth rates to lower birth rates.
+The total fertility rate (TFR), which summarizes fertility rates across all women of reproductive age in a particular time period, was less than 5 in the city of Rio de Janeiro compared to 6 for São Paulo state [@goncalves2019].
+By 1940, TFR for Rio de Janeiro had reached 3.8, compared to 4.7 for the city of São Paulo and 6.4 for Brazil [@ibge1952].>
+
+By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2015; @barreto2016; @mott2002; @roth2020].
+Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
+There, obstetricians and gynecologists began implementing surgical advances [@martins2004; @rohden2009; @2021].
+Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which relied on advances in medical knowledge and clinical infrastructure to support a technocratic model towards pregnancy and motherhood.
+This movement essentialized notions of gender, foregrounding physicians’ scientific authority [@besse1996; @freire2009; @martins2008].
+
+Maternidade Laranjeiras opened in 1904 and became the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school in 1918.
+In the mid-1920s, it was one of only three maternity hospitals for a total of 164 beds in a city of nearly 1.2 million inhabitants [@carvalho1924; @penteado1924; @roth2020].
+
+<Laranjeiras provided healthcare free of charge.
+The hospital was built in a neighborhood near weaving and textile factories that employed women and was adjacent to wealthier neighorhoods.
+Scholars have found that the majority of its patients came were neighborhood working-class women who were poor but not indigent.
+Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe [@barreto2016; @roth2020].
+Because it was the medical school's teaching hospital, it also attracted difficult labors and deliveries, with women of all classes traveling from urban and suburban areas to the city center after referral by a physician or midwife who attended the birth at home [@pintofilho1923].
+Other maternity hospitals in the 1920s had ambulatory systems, which incorporated midwives into home-based prenatal and labor and delivery care, and which the city's health department expanded in the 1930s.
+But during the years of this sample, Laranjeiras did not have such a service [@roth2020].
+In conclusion, most of the hospital's patients were working-class women from the surrounding neighborhood.
+Patients experiencing difficult labors came from across the city and from all social classes.
+However, in 1920s Rio de Janeiro, most mothers of all vary SES gave birth at home.
+For example, at the turn of the century, around 99% of births occurred at home.
+By the mid-1930s, around 80% of births occurred at home.
+For normal deliveries, wealthier women were least likely to give birth in a hospital [@roth2020].>
+
+<During the same time, old racial hierarchies based on legal status morphed into new allegedly scientific ways to categorize humans based on skin color.
+In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of whitening in which they believed that racial mixture would eventually lead to a White population [@schwarcz1993; @skidmore1993].
+Both government- and private-run initiatives to attract European immigrants had begun as early as the 1870s, in an effort to replace the slave labor force not with freed Blacks but with White immigrants [@ball2020; @skidmore1993] In the 1920s and 1930s, scientists and policymakers turned toward preventive eugenic theories that focused on improving environmental factors rather than coercively restricting reproduction through sterilization [@stepan1991].
+Although some eugenicists called for sterilization of unfit citizens, they never couched these discussions in explicitly racial terms [@wegner2013; @roth2020b].
+In the end, however, whiteness still remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.>
 
@@
 
-This analysis draws on a unique dataset of 2,845 recorded clinical visits, most of them deliveries, to Maternidade Laranjeiras between 1922 and 1926.
-I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO), which, between June 1922 and May 1926, published the monthly clinical reports of all women treated at Laranjeiras.
+This analysis draws on a unique dataset of 2,845 recorded clinical visits, most of them deliveries, to Maternidade Laranjeiras in <the city of Rio de Janeiro, Brazil> between 1922 and 1926.
+I extracted the sample from the country's premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO), which, between June 1922 and May 1926, published the monthly clinical reports of all women treated at Laranjeiras.
 I was unable to locate vol.
@@
 
-I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.
+I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), <twin status (singleton or twin)>, and the mother’s reproductive history.
+<Unfortunately, hospital clinicians erratically recorded civil status and gestational age, and thus I exclude them here.
+I discuss this further in the limitations section.>
 
@@
 The outcome was birth weight measured as a continuous variable in grams for live births only, in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
-In a secondary analysis (see Supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
+In a secondary analysis (Supplement), I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
 
@@
 
-*Gestational status*.
-The original hospital records divided gestational status into eight categories of parity and gravidity: nulliparous, primiparous, secundiparous, and multiparous; and primigravida, secundigravida, trigravida, and multigravida.
+*Maternal parity*.
+The original hospital records divided maternal parity and gravidity into eight categories of parity and gravidity: nulliparous, primiparous, secundiparous, and multiparous; and primigravida, secundigravida, trigravida, and multigravida.
 Due to issues of statistical power, I combined these into two categories: 1) `nulliparous`, which included nulliparous, primigravida, and primiparous (any woman giving birth for the first time); and 2) `multiparous`, which included secundiparous, multiparous, secundigravida, trigravida, and multigravida (any woman who has given birth more than once).
@@
 These numbers could be higher, as the hospital only recorded deaths within the first week of life, when the infant was still hospitalized.
+I also excluded all twin births, running the models on singleton births only, which is in line with current studies on birth weight [@david1997; @leal2006; @nyarko2013; @pollock2021; @silva1998; @silveira2019; @vilanova2019; @wehby2015a; @wehby2017a].
 
 *Fetal sex*. I maintained the original hospital categories of fetal sex as male and female.
+
+*Year*.
+I construct year fixed effects by including dummy variables for each year from 1923 to 1926.
+1922 is the base year.
+The coefficient on these dummy variables is the year fixed effect, which indicates how the average birth weight is different from the base year.
 
@@
 
-Finally, I calculated the maternal mortality ratio (MMR) and the mean stillbirth rate (SBR). In the Supplement, I calculated the sex ratio at birth (SRaB).
+Finally, I calculated the maternal mortality ratio (MMR) and the mean stillbirth rate (SBR).
 
@@
 where maternal deaths (MD) are divided by 10,000 live births (LB).
-In current studies, MD are divided by 100,000 LB, but I use 10,000 in the denominator, which is in line with historical studies given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].
+In current studies, MD are divided by 100,000 live births, but I use 10,000 live births in the denominator, which is in line with historical studies given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].
+The MMR measures obstetric risk: the risk of death once a person is pregnant [@worldhealthorganization2006].
 
@@
 I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race to see if there were differential outcomes for Black and mixed-race women.
-The independent variables include maternal age, maternal skin color, gestational status, and maternal nationality.
+The independent variables include maternal age, maternal skin color, parity, and maternal nationality.
 
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Par` is parity status (nulliparous or multiparous), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term.
 Results for both models are displayed in Table 2.
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), `Par` is parity (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term.
 Results for both models are displayed in Table 3.
@@
 Given that all women of color in the data set were Brazilian, collinearity exists.
-Second, I dropped all non-Brazilians and ran the original simple and multiple regressions from Models 1 and 2.
+Second, I dropped all non-Brazilians and ran the original regressions from Models 1 and 2.
 Results for the sub-sample analysis are included in Table 2.
@@
 In the hospital records, 58% were of African descent (defined as *preta* or *parda*) and 42% were of European descent (*branca*).
-First-time mothers (nulliparous) comprised 41%.
+<The city's female population in 1890 (Brazil did not conduct censuses in 1900 or 1910 and the 1920 census excluded skin color) comprised 56% White women, 29% mixed-race women, and 15% Black women.
+The city's female population in 1940 comprised 69% White women, 18% mixed-race women, and 12.5% Black women.
+The shift from mixed-race to white women was probably an artifice of reporting rather than any real racial change.
+Nonetheless, compared to either census, women of color were over-represented among hospital patients [@roth2020].> First-time mothers (nulliparous) comprised 41%.
 Of all women, 84% were Brazilian.
@@
 The remaining 9% were interventionist or operative deliveries.
-For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94.1%.
-The SBR was 82.9%.
-The sex ratio was 1.2: there were 120 male live births per 100 female live births.
+For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for <an MMR of 94 per 10,000 live births.
+The SBR was 83 stillbirths per 1000 live births.> The sex ratio was 1.2: there were 120 male live births per 100 female live births.
 
@@
 In general, maternal and infant outcomes were not better in the Maternidade Laranjeiras than among the wider population.
-Between 1922 and 1926, the clinic's MMR was 94.26%.
-For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2020].
+Between 1922 and 1926, the clinic's MMR was 94 maternal deaths per 10,000 live births.
+For those same years, the MMR for the city of Rio de Janeiro was 66 maternal deaths per 10,000 live births [@roth2020].
 Hospital rates probably reflected a registration effect, as medical personnel recorded all births and deaths, whereas reporting at the city level was less reliable [@loudon1992].
@@
 Thus, the city’s rates were probably higher.
-The same pattern holds for the SBR, which was 84.8% for Laranjeiras and 73.68% for the city of Rio de Janeiro for the same period [@roth2020].
+The same pattern holds for the SBR, which was 85 stillbirths per 1000 live births for Laranjeiras and 74 stillbirths per 1000 live births for the city of Rio de Janeiro for the same period [@roth2020].
 However, these differences demonstrate that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and infant and supports other historical studies that interventionist obstetric procedures before antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
@@
 However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
-The skewed sex ratio deserves further study.
+<Yet, the skewed sex ratio remains important to note even if we don't understand the underlying cause.>
 
@@
 They were slightly more than 50 g lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first-century Rio de Janeiro.
-Differences between racialized groups existed both historically and continue to exist today.
+<Historically, these babies were very light when compared to other urban centers across the globe.
+Brazil's late abolition of slavery, the continued presence of epidemic and endemic diseases such as malaria, yellow fever, and syphilis, nutritional deficiencies, and underresourced health services all contributed to this striking difference.> Further, differences between racialized groups existed both historically and continue to exist today.
 Moreover, when compared both historically and contemporaneously, we can see that women of all colors at Laranjeiras did not fair as well as their counterparts in other major cities.
@@
 
-This study contributes heretofore unpublished hospital records from one of Brazil's most important public maternity hospitals in the early twentieth century, a hospital that remains in operation today.
+<This study contributes heretofore unpublished hospital records from one of Brazil's most important public maternity hospitals in a major city in the early twentieth century, a hospital that remains in operation today.
 It provides quantitative evidence of racialized disparities in maternal-infant health in early twentieth-century Brazil.
 These findings are thus useful to compare to present-day research to assess the extent of progress in improved outcomes over time.
-They also complement descriptive historical studies and provide a base for qualitative research to analyze individual experience in the past.
+They also provide direct evidence of historical racial disparities in Brazil, which is rare for this time period, and thus complement descriptive historical studies and provide a base for qualitative research to analyze individual experience in the past.>
 
@@
 Yet, the theoretical upper bound on bias is 50 g, and my results show a difference greater than 50 g.
-Goldin [@goldin1989] has shown for Philadelphia that heaping did not systematically bias the mean, and I assume the same here.
+Scholars such as Goldin [@goldin1989] for birth weights in Philadelphia and <Steckel [@steckel1994] and Monasterio [@monasterio2014] for adult heights in the United States and Brazil,> respectively have shown that heaping did not systematically bias the mean, and I assume the same here.
 
-In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
+<Two sets of missing data are notable.
+First, hospital clinicians rarely recorded civil status of the mother (0.4% of all records included marital status).
+Other scholars have found that marital status was a good proxy for social support and economic status [@costa2004; @goldin1989].
+The hospital's exclusion of this variable precludes a full understanding of the mothers' socioeconomic status.>
+
+A second, major limitation was the original published clinical notes' exclusion of gestational age.
 Thus, the birth weight variable captured both smaller babies born at term, whether with or without FGR, and PTB.
@@
 This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which we do not have direct evidence.
-These findings, although not generalizable beyond this specific hospital, are consistent with present-day studies on racialized disparities in birth weight in Rio de Janeiro, demonstrating how racial inequality in health outcomes has persisted over time.
+These findings, although not generalizable beyond this specific hospital, <are consistent with both other historical studies and present-day studies on racialized disparities in birth weight in Rio de Janeiro,> demonstrating how racial inequality in health outcomes has persisted over time.
 

### 2024-11-14 products/manuscript/supplement/Supplemental_Materials.qmd
RRSSM2
Age2; immigrants; year fixed effects

@@
     bibliography: ../../../assets/Mada_Bib.bib
-    csl: ../../../assets/vancouver.csl
+    csl: ../../../assets/apa.csl
     suppress-bibliography: false
@@
 Older versions of code in `.r` are available in the `old-code` folder.
-This is not updated since 15 August 2024.
+These files have not been updated since 15 August 2024.
 
@@
 
-In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
+In this paper, I analyze a unique dataset of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
 I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO).
@@
 There are nine categorical variables: skin color; ancestry; parity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.
+
+In the first round of peer review, the reviewers suggested calculating median birth weight for different immigrant groups by nationality to better justify groupings.
+Ball [@ball2020] found that Portuguese immigrants had worse economic opportunities in São Paulo during the same time period.
+Although the Portuguese immigrant experience in Rio de Janeiro was likely better due to community, merchant connection, Portuguese business owners, and job availability, I needed to better justify my nationality categorization.
+In the `eda-v1.qmd` file, I calculated the mean birth weight for individual nationalities and performed a t-test between Portuguese infants and infants born to all other European nationalities and Portuguese infants and infants born to White Brazilian mothers.
+All results were not significant at the alpha 0.05 level.
+Thus, I will maintain the nationality categorizations of European, Brazilian, Latin American, and Middle Eastern.
+
+```{r}
+#| label: tbl-mean
+#| tbl-cap: "Mean Birthweights by Nationality"
+#| echo: FALSE
+table_mean <- readRDS(here::here("results", "tables", "table_nationality_stats_final.rds"))
+flex_table <- as_flex_table(table_mean)
+flex_table
+```
+```{r}
+#| label: tbl-ttest
+#| tbl-cap: "T-Test Results"
+#| echo: FALSE
+table_ttest <- readRDS(here::here("results", "tables", "table_ttest_final.rds"))
+table_ttest
+```
 
@@
 
-Finally, I calculated the sex ratio at birth (SRaB).
-
-The SRaB is:
-
-$$SRaB = \frac{M}{F} \times 100$$
-
-where the total number of live male births (M) is divided by 100 live female births (F) in a given period.
-
-The sex ratio was 1.2: there were 120 male live births per 100 female live births.
-This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
-Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was female.
-Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
-However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
-The skewed sex ratio deserves further study.
-
 ## Models
@@
 For the basic statistical analysis, I ran three linear models to understand the relationship between maternal factors and infant birth weight.
-The first model is a simple linear regression with outcome variable (birthweight in grams) and exposure variable maternal ancestry (`Euro-descent` or `Afro-descent`).
+The first model is a simple linear regression with outcome variable (birth weight in grams) and exposure variable maternal ancestry (`Euro-descent`, the reference, and `Afro-descent`).
+There is an association between maternal skin color and infant birth weight, with Euro-descended women having infants with higher birth weights than Afro-descended women.
+This initial finding serves as the impetus for the rest of the models.
 
@@
 
-The second simple bivariate linear analysis, looks at infant birth weight as a function of maternal age.
+The second simple bivariate linear analysis, looks at infant birth weight as a function of maternal age on both a linear scale and age-squared with `ModifiedColor` as the main predictor variable.
+Neither age nor age-squared were significant.
+Peer reviewers asked me to include $age^2$ in the model to improve the fit.
+Including $age^2$ in the model did not improve the fit, and the coefficient was not significant.
+Despite a lack of statistical significance, the coefficients suggests that while birth weight increases with maternal age, it does so at a decreasing rate.
+At certain maternal age, infant birth weight begins decreasing, presenting a reverse U-shaped curve.
 
@@
 
-The third exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, nationality, and gestational status on infant birth weight.
+The third exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, age squared, nationality, and gestational status on infant birth weight with the `ModifiedColor` variable as the main predictor variable.
 
 ```{r}
-#| label: tbl-table3
-#| tbl-cap: "Model 3"
+#| label: tbl-age2
+#| tbl-cap: "Age-Squared"
 #| echo: FALSE
-table3 <- readRDS(here::here("results", "tables", "table_fit3.rds"))
-flex_table <- as_flex_table(table3)
+table_age2 <- readRDS(here::here("results", "tables", "table_fit3.rds"))
+flex_table <- as_flex_table(table_age2)
 flex_table
@@
 
-In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women.
-In the second model, older mothers are associated with giving birth to infants with higher birth weights.
-
-The fourth exploratory model again runs a multivariate linear regression, this time using skin color as it was differentiated in the original data, or with three categories (White, Black, and Mixed Race).
+Peer review suggested that I control for year-specific effects given possible periodic epidemics in infectious disease such as malaria and yellow fever and changes in real wages.
+I do this by including year fixed effects.
+The fourth model looks at the relationship between maternal skin color, age, nationality, gestational status, and year fixed effects on infant birth weight with `ModifiedColor` as the main predictor variable.
+Generally, differences across years are common given changes in macroeconomic conditions such as real wages and epidemic disease.
+The coefficient for the main predictor variable `Afro-descent` is -84, compared to -86 in the full model without year fixed effects.
+Thus, controlling for year-specific effects, the model provides similar coefficients for birth weight, which were highest in 1922.
+This result is consistent with lower birth weights due to significant inflation and increased cost of living in Rio de Janeiro from 1923 to 1925 [@lobo2024].
 
@@
 #| echo: FALSE
-table4 <- readRDS(here::here("results", "tables", "table_fit4.rds"))
-flex_table <- as_flex_table(table4)
+table4 <- readRDS(here::here("results", "tables", "table_fit9.rds"))
+flex_table <-as_flex_table(table4)
+flex_table
+```
+
+I also tested the model five times, one for each year.
+
+```{r}
+#| label: tbl-table5
+#| tbl-cap: "Model 5"
+#| echo: FALSE
+table5 <- readRDS(here::here("results", "tables", "combined_table_years.rds"))
+flex_table <-as_flex_table(table5)
+flex_table
+```
+
+In every year, the coefficient for Afro-descended infants is consistently negative.
+However, the coefficient is statistically significant at conventional levels only for the 1923 sub-sample, in which the coefficient for infant birth weights to Afro-descended mothers was nearly 3 times the sample coefficient (-237 compared to -84).
+Thus, I then dropped the observations from 1923 and re-ran the model to see if the results remained robust.
+
+```{r}
+#| label: tbl-table6
+#| tbl-cap: "Model 6: 1922, 1924, 1925, 1926"
+#| echo: FALSE
+table6 <- readRDS(here::here("results", "tables", "table_fit11.rds"))
+flex_table <- as_flex_table(table6)
+flex_table
+```
+
+In this sub-sample model (Model 6), the coefficient for infants born to Afro-descended mothers was both not significant and lower than in the full model (-40).
+However, dropping the observations from 1923 has reduced my sample size by about 25%.
+It appears that births from 1923 are driving the overall results.
+In 1922, Brazil experienced an economic crisis with a drop in coffee prices, high inflation, and a resulting fiscal emergency, which most likely affected the port city and capital of Rio de Janeiro [@fritsch1993].
+It appears that economic uncertainty affected women of African descent more than those of European descent in this year.
+
+Additionally, these results could be due to smaller sample sizes and larger standard errors -- noise in the data.
+This reflects the limitations of historical data and historical research given small sample sizes.
+
+The last exploratory model again runs a multivariate linear regression, this time using skin color as it was differentiated in the original data, or with three categories (White, Black, and Mixed Race).
+
+```{r}
+#| label: tbl-table7
+#| tbl-cap: "Model 7"
+#| echo: FALSE
+table7 <- readRDS(here::here("results", "tables", "table_fit8.rds"))
+flex_table <- as_flex_table(table7)
 flex_table
@@
 Given my hypothesis, that the legacies of slavery affected maternal-fetal health, then maternal skin color is also probably associated with premature birth, which is correlated to infant birth weight.
+
+All code for these models is available in the `introanalysis-v1.qmd` file.
 
@@
 
+In the first round of peer review, the reviewers suggested that a logit analysis using stillbirth as an outcome would bring robustness to the racial disparities I highlight in the linear regression.
+After running a simple logistic regression with `FetalOutcome` as the predictor variable and `ModifiedColor` as the outcome, infants born to African-descended women had a 9% higher odds of having a stillbirth than Euro-descended women.
+However, all results were not significant [95% CI 0.82, 1.45; p-value 0.6], and thus I will not include them in the final analysis.
+
+```{r}
+#| label: tbl-table9
+#| tbl-cap: "Model 9"
+#| echo: FALSE
+table9 <- readRDS(here::here("results", "tables", "table_log5.rds"))
+flex_table <- as_flex_table(table9)
+flex_table
+```
+
 ## Performance Measures
@@
 ```{r}
-#| label: tbl-table9
+#| label: tbl-table10
 #| tbl-cap: "Cross-Validation Model 1"
@@
 ```{r}
-#| label: tbl-table10
+#| label: tbl-table11
 #| tbl-cap: "Cross-Validation Model 2"
@@
 ```{r}
-#| label: tbl-table11
+#| label: tbl-table12
 #| tbl-cap: "Cross-Validation Model 3"
@@
 ```{r}
-#| label: tbl-table12
+#| label: tbl-table13
 #| tbl-cap: "Cross-Validation Model 4"
@@
 
-# Appendix
-
-## AI Use
+# AI Use
 
@@
 After using this tool/service, the author reviewed and edited the content as needed and takes full responsibility for the content of the publication.
+
+# References

### 2024-11-22 code/eda-code/eda-v1.qmd
Final Changes for RR
Created DiB; updates for all changes

@@
 The words in Portuguese and English here are the same.
+Gravidity refers to the total number of pregnancies a woman has, regardless of duration.
+Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013].
 
@@
 
-Using these new dummy variables, I created new dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color.
+Using these dummy variables, I created further dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color.
 These are `Afro_NBW` and `Afro_LBW`; and `Euro_NBW` and `Euro_LBW`.
@@
 ```
-The t-tests demonstrate no significant difference in the mean birthweight between Portuguese and all other European immigrants and between Portuguese and White Brazilian women.
+The t-tests demonstrate no significant difference in the mean birth weight between Portuguese and all other European immigrants and between Portuguese and White Brazilian women.
 This suggests that differentiating European nationality further will not influence the regression results.
@@
   #autofit() %>%
-  set_table_properties(layout = "autofit", width = 1, align = "center")
+  set_table_properties(layout = "autofit", width = 1, align = "center") %>%
+  add_footer_lines("*All values are sample means except the Steckel estimate, which is a regression coefficient.")
 table2

### 2024-11-22 code/processing-code/processingfile-v1.qmd
Final Changes for RR
Created DiB; updates for all changes

@@
 Thus, I will remove the following variables as I process the data: `Nationality_Notes`, `Maternal_Notes`, `CauseofDeath`, `PreviousHistory`, and `Notes`.
-Nonetheless, for `Fetal_Notes`, we need to keep the Fetal_Notes variable at the moment as it qualifies the `FetalOutcome` variable.
+Nonetheless, for `Fetal_Notes`, we need to keep the Fetal_Notes variable as it qualifies the `FetalOutcome` variable.
 Further, I will exclude `Birth_Notes`, but first I will need to extract information on gestation (twin versus singleton birth).
 
-I also don't need the citation information to run regressions. So I will remove the following variables as I process our data: `VolN`, `Page`, `Number`.
+I also don't need the citation information to run regressions. I will remove the following variables as I process our data: `VolN`, `Page`, `Number`.
 
@@
 
-# Remove median and maternal age variables given missing data
-Mean_Birthweights <- Mean_Birthweights %>% dplyr::select(-Median, -`Maternal Age (mean)`, -"<2500gm (%)", -`Page #`)
+# Remove median,maternal age, low brith weight, page numbers, and types of measurement variables given missing data
+Mean_Birthweights <- Mean_Birthweights %>% dplyr::select(-Median, -`Maternal Age (mean)`, -"<2500gm (%)", -`Page #`, -"Measurement")
 

### 2024-11-22 products/manuscript/Manuscript_Final_SSM.qmd
Final Changes for RR
Created DiB; updates for all changes

--- 2024-11-01 products/manuscript/Manuscript_Final_SSM.qmd
@@
 Historians have outlined the racialized health disparities of people of African descent in the post-abolition period.
-Epidemiologists have shown that health disparities continue to mirror patterns from over a century ago.
-This cross-sectional analysis quantifies health disparities in the post-abolition period using infant birth weight in one of Brazil's first public maternity hospitals.
-It relies on hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras in Rio de Janeiro, Brazil to run linear models assessing differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality.
+Epidemiologists have shown that twenty-first-century health disparities continue to mirror patterns from over a century ago.
+This cross-sectional analysis quantifies health disparities in a post-abolition maternity hospital using infant birth weight.
+It relies on hospital records on infants delivered between 1922-1926 (n=2,845) at the Maternidade Laranjeiras in the city of Rio de Janeiro, Brazil to run linear models assessing differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality.
 African ancestry was correlated with lower birth weights.
-Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value = 0.002 [95% CI -138, -32]) than infants born to Euro-descendant women.
-Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value = 0.019 [95% CI -131, -12]) than infants born to White women.
-The findings were likely the consequence of slavery’s legacy, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for people of African descent -- and possible epigenetic effects from the lived experience of racism.
-The findings are consistent with current-day research on racialized health disparities in Brazil.
+Infants born to Afro-descendant women had birth weights estimated to be 84 grams lighter (p-value = 0.002 [95% CI -137, -32]) than infants born to Euro-descendant women.
+Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) and infants born to mixed-race (*parda*) women had birth weights estimated to be 70 grams lighter (p-value = 0.022 [95% CI -130, -10]) than infants born to White women.
+The findings were likely the consequence of slavery’s legacy, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for people of African descent.
+The findings are consistent with current-day research on racialized health disparities in Brazil and demonstrate the importance of historical findings to public health research.
 
@@
 
-This study analyzes infant birth weight as a proxy for general maternal and fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras, a public maternity hospital in Rio de Janeiro, Brazil between 1922 and 1926.
-It aims to provide basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil, substantiating the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
+This study analyzes infant birth weight as a proxy for general maternal, fetal, and infant health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras, a public maternity hospital in the city of Rio de Janeiro, Brazil between 1922 and 1926.
+It provides basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil, substantiating researcher's qualitative findings on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.
 
-Birth weight and its relationship to infant health have been a focus of public health research for over a century [@hughes2017].
+Birth weight and its relationship to fetal and infant health have been a focus of public health research for over a century [@hughes2017; @schneider2017].
 In epidemiology, both low birth weight (LBW) and preterm birth (PTB), or infants born <37 weeks' gestation, within and across racialized communities have garnered attention [@braveman2021a; @david1997; @hughes2017; @silva1998; @wilcox2001].
@@
 
-Multiple social, economic, biological, and genetic factors influence birth weight, which is also partially conditioned by maternal nutrition, health, and health care utilization during pregnancy [@braveman2021a; @david1997; @ward1993].
-Recent genetic studies have detailed the specific alleles through which both the maternal and fetal genomes contribute to birth weight and gestational length, while acknowledging the continued importance of non-genetic factors [@juliusdottir2021; @sole-navais2023; @warrington2019].
-Given that these studies are based on well-off populations in countries with low neonatal and infant mortality rates, they might overestimate genetic impact on birth weight for historical studies of impoverished populations.
+Multiple social, economic, biological, and genetic factors influence birth weight, which is also partially conditioned by maternal nutrition, health, and health care utilization during pregnancy [@braveman2021a; @david1997; @schneider2017; @ward1993].
+Recent genetic studies have detailed the specific alleles through which both the maternal and fetal genomes contribute to birth weight and gestational timing, while acknowledging the continued importance of non-genetic factors [@sole-navais2023; @warrington2019].
 
@@
 Originally, epidemiologists hypothesized an association between at-term LBW and increased risk for cardiovascular diseases and other chronic physical health conditions over the life course [@barker2004].
-Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest long-term trends associated with "fetal programming" [@maruyama2020; @warrington2019].
-Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure.
-Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.
+Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest long-term trends associated with "fetal programming." Warrington et al [-@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure.
+Maruyama and Heinesen [-@maruyama2020] have found that although lower birth weight is associated with both infant mortality and other health issues in the neonatal period, its association with long-term health issues declines over time.
 
-Historians, using hospital records, have shown how birth weight can be an indicator for the often unmeasurable nutritional status of mothers of different social classes in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @ward1993].
-Historically, women from more privileged groups had heavier infants than women from less advantaged groups [@stein1984; @ward1993].
+Historians, using hospital records, have shown how birth weight can be an indicator for the often-unmeasurable nutritional status of mothers in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @ward1993].
+Historically, women from more privileged groups had heavier infants than less-advantaged women [@stein1984; @ward1993].
 
-The differences between and within groups both today and in the past have given rise to much research stratified by racial classification.
-Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
-Thus, racism and racist policies also can influence birth outcomes.
+The differences between and within groups both today and in the past have prompted racially-stratified research.
+While race is a social construct and thus biological fiction, the social consequences of racism influence health and disease patterns, including birth outcomes, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
 Contemporary studies in Brazil have shown that non-White mothers have higher odds of giving birth to LBW infants [@nyarko2013; @wehby2015a; @vilanova2019].
-Non-White mothers also had infants with lower mean birth weights than their White counterparts [@leal2006].
+Non-White mothers also have infants with lower mean birth weights than their White counterparts [@leal2006].
 These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns are at higher risk for LBW than White newborns, with trends worsening over time [@pollock2021].
 Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997].
-Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].
 
-Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@jasienska2009; @owens2019; @martins2024; @steckel1986].
-Studies in the human sciences have hypothesized an intergenerational effect of poor health conditions, including undernutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].
-In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@adamo1998; @otovo2016; @roth2020; @telles2022].
+Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@jasienska2009; @owens2019; @steckel1986].
+Studies in the human sciences have hypothesized an intergenerational effect of poor health conditions, including undernutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020].
+Historians have emphasized how race and socioeconomic status (SES) were deeply intertwined in post-abolition Brazil and thus disparities in literacy and access to education likely contributed to health gaps [@adamo1998; @ball2020; @franken2025].
+In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal, fetal, and infant health outcomes [@adamo1998; @otovo2016; @roth2020; @telles2022].
 
-Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021], compared to around 300,000 for the United States [@slavevoyagesconsortium2021].
-Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission -- the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020].
+Brazil imported over five million enslaved Africans over nearly four centuries [@slavevoyagesconsortium2021].
+Mortality rates that surpassed fertility rates, in addition to higher rates of individual manumission -- the freeing of an enslaved person -- helps explain Brazil's reliance on the transatlantic slave trade.
 Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere.
-In the nineteenth century, population-level health was poor for all Brazilians, but especially for enslaved individuals, who experienced epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever as well as undernutrition and violent corporal punishment [@karasch1987].
-Excluding pregnant and nursing women, historians have argued that caloric intake for the enslaved population was probably sufficient in terms of quantity, yet it lacked key nutrients, and thus the enslaved suffered from nutritional deficiencies, including insufficient thiamine, which was associated with infant mortality [@karasch1987; @kiple1989].
-Today, thiamine deficiency is correlated with lower birth weight and adverse perinatal outcomes including neonatal death [@gr2024; @kareem2023].
+In the nineteenth century, population-level health was poor for all Brazilians, but especially for enslaved individuals, who experienced epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever as well as undernutrition and violent punishment [@karasch1987].
+Excluding pregnant and nursing women, historians have argued that caloric intake for the enslaved population was sufficient in terms of quantity, yet it lacked key nutrients.
+Thus, the enslaved suffered from nutritional deficiencies, including thiamine, which was associated with infant mortality [@karasch1987; @kiple1989].
+Today, thiamine deficiency is correlated with lower birth weight and adverse perinatal outcomes [@kareem2023].
 
-In the aftermath of abolition, no state-run efforts to incorporate formerly enslaved, African-descended peoples into economic, civic, or political life occurred [@fischer2008].
-Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower literacy levels, and had worse health outcomes than their White and White immigrant counterparts [@adamo1998; @ball2020; @fischer2008; @otovo2016].
+In the aftermath of abolition, no state-run efforts to incorporate formerly enslaved peoples into economic, civic, or political life occurred [@fischer2008].
+Further, the state supported a policy of whitening in which elites believed that both racial mixture and White immigration would whiten the population (Ball, 2020; Stepan, 1991).
+Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, and had lower literacy levels, fewer employment opportunities, and worse health outcomes than their White and White immigrant counterparts [@adamo1998; @ball2020; @fischer2008; @otovo2016].
 
-Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019].
-The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe.
-Improved sanitation measures helped stem infectious disease outbreaks [@benchimol1999; @hochman2013].
-Infant mortality rates declined from the nineteenth century, although the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@adamo1998; @roth2020; @loudon1992; @woods2009].
+Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in urban centers in Brazil's southeastern and southern regions, including Rio de Janeiro [@franken2025].
+The early twentieth century was a period of advancement in both clinical care and public health in Brazil and globally.
+In cities, improved sanitation measures stemmed infectious disease outbreaks [@benchimol1999; @hochman2013].
+Infant mortality rates declined from the nineteenth century, although the major advancements including blood transfusions and antibiotics that resulted in sustained improvements in maternal and infant mortality and stillbirth rates only began in the late 1930s [@adamo1998; @roth2020; @loudon1992; @woods2009].
+Further, these advancements were concentrated in cities, and rural-urban and regional divides in maternal-infant health outcomes developed across Brazil that only decreased in the new Millennium [@scheper-hughes1992; @victora2011].
 
-Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, provided healthcare free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020].
-Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe.
+A rural-urban and regional divide was also noticeable in fertility patterns.
+Women in the city of Rio de Janeiro started their fertility transition in the 1920s.
+In the early 1930s, the total fertility rate (TFR), which summarizes fertility rates across all women of reproductive age in a particular period, was less than 5 in Rio de Janeiro compared to 6 for São Paulo state [@goncalves2019].
+By 1940, TFR for Rio de Janeiro had reached 3.8, compared to 4.7 for the city of São Paulo and 6.4 for Brazil [@ibge1952].
 
-In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon.
-In our current postgenomic era, race as a biological category has been rejected, yet the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking in epigenetics remains very real [@meloni2022].
-Although differences between racialized groups existed, they were not immutable or due to genetic differences based on race.
-The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].
+Maternidade Laranjeiras opened in 1904 and became the obstetrical teaching hospital at Rio de Janeiro’s medical school in 1918.
+In the mid-1920s, it was one of Rio de Janeiro's three maternity hospitals for a total of 164 beds in a city of nearly 1.2 million inhabitants [@carvalho1924; @penteado1924; @roth2020].
+
+Laranjeiras provided healthcare free of charge.
+The hospital was built in a neighborhood near weaving and textile factories that employed women workers.
+Scholars have found that most of its patients were neighborhood working-class women who were poor but not indigent.
+Most patients were women of color defined as either Black or mixed race, and most White patients were European immigrants [@barreto2016; @roth2020].
+Because it was the medical school's teaching hospital, it also attracted difficult labors and deliveries, with women of all classes referred by a physician or midwife who initially attended the birth at home [@pintofilho1923].
+Although other maternity hospitals in the 1920s had ambulatory systems, which provided home-based prenatal and labor and delivery care, Laranjeiras did not [@roth2020].
+
+Because most of the hospital's patients were working-class women from the surrounding neighborhood, its clientele was relatively representative of the city's population, which was majority lower SES.
+But the hospital probably had a disproportionate number of difficult births given that medical practitioners referred obstructed labors to Laranjeiras.
+Further, in 1920s Rio de Janeiro, most mothers of all SES gave birth at home.
+At the turn of the century, around 99% of births were home-based.
+By the mid-1930s, around 80% of births occurred at home.
+For normal deliveries, wealthier women were least likely to give birth in a hospital [@roth2020].
+
+In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Rio de Janeiro, my aim is not to reify race as a genetic or epigenetic phenomenon [@meloni2022].
+Although differences between racialized groups existed, they were not immutable or due to race-based genetic differences.
+The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not part of permanently "damaged" communities (socially or biologically) [@tuck2009] but rather were living the consequences of slavery and socioeconomic and racial inequality.
 
@@
 
-This analysis draws on a unique dataset of 2,845 recorded clinical visits, most of them deliveries, to Maternidade Laranjeiras between 1922 and 1926.
-I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO), which, between June 1922 and May 1926, published the monthly clinical reports of all women treated at Laranjeiras.
+This analysis draws on a unique dataset of 2,845 recorded clinical visits, majority births, to Maternidade Laranjeiras in the city of Rio de Janeiro, Brazil between 1922 and 1926.
+I extracted data from the country's premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO), which, between June 1922 and May 1926, published the monthly clinical reports of all women treated at Laranjeiras.
 I was unable to locate vol.
@@
 
-In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures.
-Over time, clinical notes became streamlined.
-When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician [@roth2020; @roth2020a].
-Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021].
-Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
-
-I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.
+I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), twin status (singleton or twin), and maternal reproductive history.
+Hospital clinicians erratically recorded civil status and gestational age, and thus I excluded them.
 
 The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.
-Between January 2012 and July 2013, I manually digitized the journal by photographing each volume.
+Between January 2012 and July 2013, I photographed each volume.
 Then, between January and August 2017, I manually input the data into Excel from the digital reproductions.
@@
 The project's Github repository, particularly the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting.
-All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online.
+All materials, including the data set and code, are open access and reproducible.
 
@@
 *Birth weight*.
-The outcome was birth weight measured as a continuous variable in grams for live births only, in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
-In a secondary analysis (see Supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].
+The outcome was birth weight measured as a continuous variable in grams for live births only [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019].
+In `Supplemental_Materials`, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@who2022].
 
@@
 I maintained this categorization.
-However, I also created a new dummy variable for skin color, combining Black and mixed-race patients into one category, `Afro-Descent`, and White patients into a `Euro-Descent` category.
+I also created a new skin-color dummy variable, combining Black and mixed-race patients as `Afro-Descent` and White patients as `Euro-Descent`.
 
-Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum rather than as a Black-White binary [@travassos2004].
+Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum [@travassos2004].
 Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category.
-This method also is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].
+This method also aligns with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].
 
@@
 I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention and a vaginal delivery; interventionist, indicating medium medical intervention using forceps and a vaginal delivery; and operatory, indicating a cesarean section or embryotomy.
-However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries), coded as operatory or natural by hospital records, as interventionist.
+I recategorized external manipulations including version and Mauriceau maneuver, coded as operatory or natural by hospital records, as interventionist.
 I classified spontaneous abortion as a separate category.
 
-*Infant length*. Infant length was recorded in centimeters in the original records.
+*Infant length*. Hospital records included infant length in centimeters.
 
@@
 Hospital records recorded maternal outcome as death, discharge, or hospital transferal, which I maintained.
-I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality below.
+I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality.
 
@@
 I recorded fetal outcome as live birth or stillbirth.
-For the statistical models, I excluded both spontaneous abortions (recorded in the birth variable) and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data, which is standard methodology for looking at infant birth weight today [@nyarko2013; @wehby2015a; @wehby2017a].
-The hospital records recorded 20 neonatal deaths, which I included in the live birth category.
-These numbers could be higher, as the hospital only recorded deaths within the first week of life, when the infant was still hospitalized.
+For the statistical models, I excluded both spontaneous abortions (recorded in the birth variable) and any recorded live births weighing \<500 grams to avoid the inclusion of possible stillbirths that were misrecorded as live births in the original data [@nyarko2013; @wehby2015a].
+The hospital records listed 20 neonatal deaths, which I included as live births.
+These numbers could have been higher, as the hospital only recorded deaths within the first week of life, when the infant was still hospitalized.
+I ran all linear models on singleton births only, excluding twins [@david1997; @leal2006; @nyarko2013; @pollock2021; @silva1998; @silveira2019; @vilanova2019; @wehby2015a].
 
 *Fetal sex*. I maintained the original hospital categories of fetal sex as male and female.
+
+*Year*.
+I construct year fixed effects by including dummy variables for each year from 1923 to 1926.
+1922 is the base year.
+The coefficient on these dummy variables is the year fixed effect, which indicates how the average birth weight is different from the base year.
 
@@
 where maternal deaths (MD) are divided by 10,000 live births (LB).
-In current studies, MD are divided by 100,000 LB, but I use 10,000 in the denominator, which is in line with historical studies given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].
+Today, MD are divided by 100,000 live births, but I use 10,000 live births in the denominator given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].
+The MMR measures obstetric risk: the risk of death once a person is pregnant [@who2006].
 
@@
 
-where the total number of stillbirths (SB) is divided by 1000 total births (TB).
-
-Finally, I calculated the sex ratio at birth (SRaB).
-
-The SRaB is:
-
-$$SRaB = \frac{M}{F} \times 100$$
-
-where the total number of live male births (M) is divided by 100 live female births (F) in a given period.
+where the total number of stillbirths (SB) is divided by 1,000 total births (TB).
 
@@
 I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight.
-Due to the lack of comprehensive information on gestational age, I did not consider birth weight in reference to gestational age.
-I discuss this constraint in the discussion below.
-
 I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race to see if there were differential outcomes for Black and mixed-race women.
-The independent variables include maternal age, maternal skin color, gestational status, and maternal nationality.
+The independent variables include maternal age, skin color, parity, and nationality, and year fixed effects.
 
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Par` is parity (nulliparous or multiparous), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term.
 Results for both models are displayed in Table 2.
@@
 
-$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.
+$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), `Par` is parity (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term.
 Results for both models are displayed in Table 3.
 
-Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts of all skin colors [@franken2019; @ball2020].
+Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts of all skin colors [@ball2020; @franken2025].
 Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 (modified ancestry) in two steps.
@@
 Table 1 provides summary statistics for the dataset.
-In the hospital records, 58% were of African descent (defined as *preta* or *parda*) and 42% were of European descent (*branca*).
-First-time mothers (nulliparous) comprised 41%.
+Of hospital patients, 58% were of African descent (defined as *preta* or *parda*) and 42% were of European descent (*branca*).
+An average of census data for the city's female population's racial makeup between 1890 and 1940, years for which data exist, show that White women comprised 63%, mixed-race women 24%, and Black women 14% (with rounding errors) [@roth2020].
+Thus, at Laranjeiras, women of color were over-represented among hospital patients.
+First-time mothers (nulliparous), comprised 41%.
 Of all women, 84% were Brazilian.
-Of the nearly 16% immigrant patients, 15% were from Europe.
+Of the nearly 16% immigrant patients, 15% were European.
 
@@
 The remaining 9% were interventionist or operative deliveries.
-For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94.1%.
-The SBR was 82.9%.
-The sex ratio was 1.2: there were 120 male live births per 100 female live births.
+For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94 per 10,000 live births.
+The SBR was 83 stillbirths per 1,000 live births.
 
@@
 All models demonstrated an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race).
-When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter (p-value = 0.002 [95% CI -138, -32]) than their Euro-descended counterparts (Table 2).
-In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter (p-value = 0.019 [95% CI -131, -12]) than White mothers.
+When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 84 grams lighter (p-value = 0.002 [95% CI -137, -32]) than their Euro-descended counterparts (Table 2).
+In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) than White mothers, while mixed-race mothers gave birth to infants weighing on average 70 grams lighter (p-value = 0.022 [95% CI -130, -10]) than White mothers.
 Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of European descent.
@@
 More recent studies relying on IV have found that a 10% increase in any birth weight decreases infant mortality by 13.3 deaths per 1000 live births [@maruyama2020].
-Applied to these data, an 85 g difference in birth weight between Afro- and Euro-descended infants (sub-model 2) would have increased infant mortality among Afro-descended children by 3.7 deaths per 1000 births.
+Applied to these data, an 84 g difference in birth weight between Afro- and Euro-descended infants (sub-model 2) would have increased infant mortality among Afro-descended children by 3.7 deaths per 1,000 births.
 
@@
 Age and nationality were not significant.
+The year fixed effects variable demonstrates that birthweights were highest in 1922.
 
@@
 
-In general, maternal and infant outcomes were not better in the Maternidade Laranjeiras than among the wider population.
-Between 1922 and 1926, the clinic's MMR was 94.26%.
-For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2020].
-Hospital rates probably reflected a registration effect, as medical personnel recorded all births and deaths, whereas reporting at the city level was less reliable [@loudon1992].
-Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s [@roth2020].
-Thus, the city’s rates were probably higher.
-The same pattern holds for the SBR, which was 84.8% for Laranjeiras and 73.68% for the city of Rio de Janeiro for the same period [@roth2020].
-However, these differences demonstrate that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and infant and supports other historical studies that interventionist obstetric procedures before antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
+In general, maternal, fetal, and infant outcomes were not better in the Maternidade Laranjeiras than among the city's wider population.
+Between 1922 and 1926, the clinic's MMR was 94 maternal deaths per 10,000 live births compared to 66 maternal deaths per 10,000 live births for the city [@roth2020].
+Hospital rates probably reflected a registration effect [@loudon1992], as medical personnel recorded all births and deaths, whereas the city's vital statistics were still haphazardly collected in the 1920s [@roth2020].
+The same pattern holds for the SBR, which was 85 stillbirths per 1,000 live births for Laranjeiras and 74 stillbirths per 1,000 live births for the city [@roth2020].
 
-The sex ratio of 120 male live births per 100 female live births is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
-Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was female.
-Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
-However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
-The skewed sex ratio deserves further study.
-
-This study found that women of African descent in early twentieth-century Brazil both had infants with lower birth weights and higher rates of LBW than their White counterparts, perhaps indicative of higher rates of PTB.
-Wilcox [@wilcox2001] has questioned the utility of the LBW category in epidemiological studies given its arbitrary nature.
+This study found that women of African descent in early twentieth-century Rio de Janeiro had infants with lower birth weights and higher rates of LBW than their White counterparts, perhaps indicative of higher rates of PTB.
+@wilcox2001 has questioned the utility of the LBW category in epidemiological studies given its arbitrary nature.
 We can question the use of this cutoff in historical studies as well, as it was not an established category until the mid-twentieth century.
@@
 This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity.
-The WHO adopted this cutoff in 1948, and it has remained the standard ever since even as epidemiologists began to differentiate between small-for-gestational age (SGA, now FGR) and PTB in the 1970s [@hughes2017; @wilcox2001].
+The WHO adopted this cutoff in 1948, and it has remained the standard even as epidemiologists began to differentiate between small-for-gestational age (SGA, now FGR) and PTB in the 1970s [@hughes2017; @wilcox2001].
 Physicians in early twentieth-century Rio de Janeiro defined low birth weight as $>2500 g$ and normal birth weight as between 3000 and 3500 g [@conrado1924a].
 
-Regardless of its arbitrariness, LBW was dangerous in the past given that the probability of surviving the neonatal period was lower than today due to higher neonatal and infant mortality rates and less effective medical interventions [@costa1998; @roth2020; @maruyama2020].
+Regardless, LBW was more dangerous in the past given that the probability of surviving the neonatal period was lower than today due to higher mortality rates and less effective medical interventions [@costa1998; @roth2020; @maruyama2020].
 
-Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight globally.
-Black infants born at Laranjeiras had the lowest mean birth weight (3037 g) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004].
-Black infants born at Laranjeiras were almost 150 g lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period.
-They were less than 100 g lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first-century Rio de Janeiro.
-White infants born at Laranjeiras had a mean birth weight of 3133 g, nearly 300 g lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 g).
-They were slightly more than 50 g lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first-century Rio de Janeiro.
-Differences between racialized groups existed both historically and continue to exist today.
-Moreover, when compared both historically and contemporaneously, we can see that women of all colors at Laranjeiras did not fair as well as their counterparts in other major cities.
+Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day mean birth weights from various regions in Brazil and historical studies on birth weight globally.
+Black infants born at Laranjeiras had the lowest mean birth weight (3037 g) when compared to other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004].
+Historically, these babies were very light when compared to other urban centers across the globe.
+Black infants born at Laranjeiras were 146 g lighter than Black infants born at the Johns Hopkins hospital during an equivalent period.
+They were 85 g lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first-century Rio de Janeiro.
+White infants born at Laranjeiras had a mean birth weight of 3133 g, 290 g lighter than White infants born at the Johns Hopkins hospital during an equivalent period (3423 g).
+They were 53 g lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first-century Rio de Janeiro.
+Differences between racialized groups existed both historically and today.
 
-As scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in other regions during the same historical period [@costa1998; @costa2004; @goldin1989].
-Across populations, relatively high mean birth weights in the nineteenth and early twentieth centuries are probably an artifact of high stillbirth and neonatality rates [@costa1998; @costa2004].
-Today, PTB and lower birth weight infants that probably would have died in utero 100 years ago are now surviving, bringing the mean birth weight down.
+As scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in other regions during the same period [@costa1998; @costa2004; @goldin1989].
+Across populations, relatively high mean birth weights in the nineteenth and early twentieth centuries could be an artifact of high stillbirth rates, which selected upon fetuses that were already small and unhealthy [@costa1998; @costa2004; @schneider2017].
+Today, PTB and lower birth weight infants that might have died in utero 100 years ago now survive, bringing the mean birth weight down.
 
 What can explain the differences in birth weight historically both between women at Laranjeiras and those of other historical populations and among different skin-color groups within Laranjeiras?
-Although both maternal and fetal genetic factors [@juliusdottir2021; @sole-navais2023; @warrington2019] and maternal age, parity, birth spacing, and gestational age all condition birth weight [@stein1984], the intrauterine environment remains a critical determinant.
-In this study, maternal age was not a significant factor, although multiparous mothers had infants with higher birth weights, as is common today.
-The data did not provide information on birth spacing or gestational age, the latter which I discuss below in the limitations.
+Although both maternal and fetal genetic factors [@sole-navais2023; @warrington2019] and maternal age, parity, birth spacing, and gestational age all condition birth weight [@stein1984], the intrauterine environment remains a critical determinant.
+Here, maternal age was not significant, although multiparous mothers had infants with higher birth weights.
+The data did not provide information on birth spacing or gestational age.
 
 What about external factors related to poverty?
-Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care also influence birth weight [@stein1984].
-Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986; @stein1984].
-But other health problems related to maternal undernutrition during pregnancy depend on gestational period.
-Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].
-Yet nutritional deficiencies that were found among African-descended populations in nineteenth-century Rio de Janeiro due to slavery were associated with lower birth weights and infant mortality [@karasch1987; @kiple1989; @gr2024; @kareem2023].
-Although working women of all skin colors in the early twentieth century had more control over their diets than did the enslaved, food was expensive and many of the nineteenth-century trends in nutritional deficiencies continued in the early twentieth century [@adamo1998].
+Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care influence birth weight [@schneider2017; @stein1984].
+Maternal nutrition can too, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a malnutrition threshold must be met before adverse effects occur [@costa1998; @steckel1986; @stein1984].
+Other health problems related to maternal undernutrition during pregnancy depend on gestational period.
+Research on the Dutch famine during World War II shows that adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006].
+But overall nutritional deficiencies that were found among slave populations in nineteenth-century Rio de Janeiro were associated with lower birth weights and infant mortality [@karasch1987; @kiple1989; @kareem2023].
+Although working women of all skin colors in the early twentieth century had more control over their diets than did the enslaved, food was expensive and nutritional deficiencies continued [@adamo1998].
+The late abolition of slavery could have contributed to differences in the historical mean birth weights in Table 4.
 
-We also know that poorer women in early twentieth-century Rio de Janeiro performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992].
-And working-class women of all skin colors often worked through the third trimester as domestic servants [@roth2020].
-These same women might have lived in crowded housing arrangements with lack of access to running water [@graham1992; @fischer2008; @roth2020].
-Thus, generalized poverty might have influenced women of African descent's nutrition, physical routines, and ability to access prenatal care.
+We also know that poorer women in early twentieth-century Rio de Janeiro performed physically taxing jobs that required them to stand for long periods and work throughout the third trimester, as washerwomen, domestic help, and street vendors [@graham1992; @roth2020].
+These same women might have lived in crowded housing arrangements without running water [@graham1992; @fischer2008; @roth2020].
+Outbreaks of epidemic disease like malaria and yellow fever might have had differential effects on women based on their living and working conditions [@benchimol1999].
+Generalized poverty might have influenced women of African descent's nutrition, disease environments, physical routines, and ability to access prenatal care.
+Comparing these factors to those present in other locations might further explain the historical differences in mean birth weights.
 
-Because the hospital records excluded information on maternal health factors, we don't know which contributed to racial disparities in infant birth weight.
+Because the hospital records excluded information on maternal health factors, we don't know which contributed to racial disparities in birth weight.
 One possible cause that we do have some data for is syphilis.
 Like Costa found for early twentieth-century Baltimore, higher syphilis rates among African-descended women in Rio de Janeiro might have explained lower birth weights.
-Studies demonstrate that syphilis increases adverse birth outcomes including increased risk of LBW and PTB [@dasilva2024].
-The clinical records that comprise this dataset did not consistently record syphilis rates, although sporadic notes showed it existed.
-Medical and public health professionals of the time viewed it as an important cause of adverse health outcomes [@carrara1996a].
-City-wide rates from the 1930s can provide some context for the prevalence of infection among pregnant women.
-Of all prenatal syphilis exams performed between 1935 and 1938 at public clinics or hospitals across the city, unadjusted for maternal skin color, 7.92% came back positive [@fontenelle, p.
-275].
-
-Discussions of epigenetic markers associated with the lived experience of racism might also be applicable.
-Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms [@leimert2020].
-Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@jasienska2009; @leimert2020].
-Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health [@meloni2022].
+Syphilis results in adverse birth outcomes including increased risk of LBW and PTB [@dasilva2024].
+The clinical records only sporadically recorded syphilis rates in women’s reproductive histories or fetal outcomes (maceration).
+Physicians viewed it as an important cause of adverse health outcomes [@fontenelle; @penteado1924].
+City-wide rates from the 1930s can provide prevalence of infection among pregnant women.
+Of all prenatal syphilis exams performed between 1935 and 1938 at public clinics or hospitals, unadjusted for maternal skin color, around 8% were positive [@fontenelle].
 
@@
 
-This study contributes heretofore unpublished hospital records from one of Brazil's most important public maternity hospitals in the early twentieth century, a hospital that remains in operation today.
-It provides quantitative evidence of racialized disparities in maternal-infant health in early twentieth-century Brazil.
-These findings are thus useful to compare to present-day research to assess the extent of progress in improved outcomes over time.
-They also complement descriptive historical studies and provide a base for qualitative research to analyze individual experience in the past.
+This study contributes heretofore unpublished hospital records from one of Brazil's most important public maternity hospitals in a major early twentieth-century city, a hospital that remains in operation today.
+It provides quantitative evidence of racialized disparities in maternal, fetal, and infant health.
+These findings are thus useful to compare to present-day research to assess the extent of progress in outcomes.
+They also provide direct evidence of historical racial disparities in Brazil, rare for this period, and thus complement descriptive historical studies.
 
-All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014; @ward1993].
-Physicians could have introduced systematic measurement error when weighing newborns in Laranjeiras, either through inconsistent practices or inaccurate equipment.
+All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@ball2020; @franken2025; @steckel1998; @roberts2014; @ward1993].
+Physicians could have introduced systematic measurement error when weighing newborns, either through inconsistent practices or inaccurate equipment.
 The hospital records do not provide information on weighing practices, but published observations from medical students who trained at Laranjeiras show that the norm was to weigh all newborns without clothes on a calibrated scale [@conrado1924a].
 Thus, an upward bias to the mean through the inclusion of clothing or blankets for all infants seems unlikely.
-It also seems unlikely that physicians weighed infants of one skin color in a different method than those of another given that Laranjeiras was the premier teaching hospital associated with a major medical school.
-If any upward bias existed, weights were even lower than presented here.
+Physicians most likely did not weigh infants of one skin color differently than those of another given that Laranjeiras was a major medical school's teaching hospital.
+If any upward bias existed, weights were lower than presented here.
 
-Additionally, physicians could have introduced rounding error as all birth weights at Laranjeiras were rounded to the nearest 50 g.
-Any systematic bias in any direction for any group of infants would have been at the hospital level since multiple physicians and medical students weighed babies over the five-year period.
+Additionally, physicians could have introduced rounding error as all birth weights were rounded to the nearest 50 g.
+Any systematic bias in any direction for any group of infants would have been at the hospital level since multiple clinicians weighed babies over the five-year period.
 Yet, the theoretical upper bound on bias is 50 g, and my results show a difference greater than 50 g.
-Goldin [@goldin1989] has shown for Philadelphia that heaping did not systematically bias the mean, and I assume the same here.
+Scholars such as Goldin and Margo [-@goldin1989] for birth weights in Philadelphia and Steckel [-@steckel1994] for adult heights in the United States> have shown that heaping did not systematically bias the mean.
+I assume the same.
 
-In this study, a major limitation was the original published clinical notes' exclusion of gestational age.
-Thus, the birth weight variable captured both smaller babies born at term, whether with or without FGR, and PTB.
+Two sets of missing data are notable.
+First, hospital records excluded maternal marital status (included in 0.4% of records).
+Scholars have found that marital status was a good proxy for SES [@costa2004; @goldin1989].
+The hospital's exclusion of this variable precludes a full understanding of mothers' SES.
+
+A second, major limitation was the hospital records' exclusion of gestational age.
+Thus, the birth weight variable captures both smaller babies born at term, whether with or without FGR, and PTB.
 [@falcao2020].
-This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@falcao2020; @wilcox2001; @wilcox2024].
+This limits our ability to understand the relationship between maternal variables and infant birth weight, as pre-term infants are more likely to have lower birth weights [@falcao2020; @wilcox2001; @wilcox2024].
 Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to many historical inquiries into birth weights.
-Yet, the factors discussed here could have contributed PTB, to FGR, or to both.
-
-A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
-The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009; @ward1993].
+The factors discussed here could have contributed PTB, to FGR, or to both.
 
 Today, lower mean birth weights between populations are not necessarily associated with worse health outcomes [@wilcox2001].
-Yet continued differences between racialized groups hints at the possible role of socio-economic inequity and racism in determining health outcomes.
+Yet continued differences between racialized groups hints at the possible role of socio-economic inequity and racism in determining outcomes.
 In addition, studies relying on IV analysis have shown that even in the recent past in developed countries, the impact of birth weight on adverse perinatal and neonatal outcomes has declined over the past 35 years given advances in medical care [@maruyama2020].
 We can imagine that 100 years ago, the association of birth weight, most likely as a proxy for PTB, with adverse neonatal outcomes was even greater.
+Some scholars argue that reductions in perinatal mortality over the past 150 years have occurred due to a reduction of deaths among the most at-risk fetuses and neonates, not to an improvement in overall fetal health [@schneider2017].
 Finally, birth weights studied historically are as important for understanding their mother's past lives as for predicting perinatal or neonatal outcomes [@ward1993].
+Although studies have called into question the utility of birth weight as a marker of fetal health [@schneider2017; @wilcox2001], in many historical contexts, it remains one of the only available indicators.
 
-Lack of data on maternal nutrition and health status or maternal work routine and income, all of which cannot be corrected for by using maternal skin color as a proxy, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study.
+Lack of data on maternal nutrition, health, work routine, civil status, and income, all of which cannot be corrected for by using maternal skin color as a proxy, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study.
 Maternal skin color may serve as a partial index of the unobservable variables that acted upon infant birth weight [@costa2004; @ward1993].
-Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.
+Given the paucity of historical understandings of current-day health disparities in maternal, fetal, and infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past, particularly in non-North American and Western European countries.
 
@@
 
-This study has shown that lower birth weights were associated with Afro-descended women giving birth in the Maternidade Laranjeiras in Rio de Janeiro, Brazil between 1922 and 1926.
-This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which we do not have direct evidence.
-These findings, although not generalizable beyond this specific hospital, are consistent with present-day studies on racialized disparities in birth weight in Rio de Janeiro, demonstrating how racial inequality in health outcomes has persisted over time.
+This study has shown that lower birth weights were associated with Afro-descended women giving birth in the Maternidade Laranjeiras in Rio de Janeiro, Brazil between 1922 and 1926, suggesting that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which we do not have direct evidence.
+These findings, although not generalizable beyond this specific hospital, are consistent with both historical and present-day studies on racialized disparities in birth weight in Rio de Janeiro, demonstrating how racial inequality in health outcomes has persisted over time.
 
@@
 {{< pagebreak >}}
+
+# References

### 2024-11-22 products/manuscript/data-in-brief/Data_in_Brief.qmd
Final Changes for RR
Created DiB; updates for all changes

---
title: "Data_in_Brief"
author: Cassia Roth
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
    bibliography: ../../../assets/Mada_Bib.bib
    csl: ../../../assets/apa.csl
    suppress-bibliography: false
---
# Data-in-Brief

```{r}
# Load packages
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(gtsummary) #for summary tables
library(flextable) #for nice tables
```

## Abstract

This article presents clinical data collected from historical archival sources in Rio de Janeiro, Brazil between 1922 and 1926 with the aim of exploring racial and socio-economic disparities in birth weight, and thus providing insights into maternal, fetal, and infant health, in a major urban center in early twentieth-century Latin America.

The dataset comprises 2845 recorded clinical visits to a public maternity hospital, the Maternidade Laranjeiras, between June 1922 and May 1926 in Rio de Janeiro, Brazil. The records were extracted from Brazil’s major obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro. The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*. In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*. In 1922, it became the *Revista de Gynecologia e D’Obstetricia*. Physical copies of the journal are located in various archives in Rio de Janeiro, Brazil.

The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras. Vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926) were not locatable in the historical archives.

In the initial months of publication, the individual clinical reports included more complete information, including duration of labor or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician [@roth2020; @roth2020a]. From the available issues, the following information was recorded, when available, for all patients: patient number, gravidity or parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history. Gravidity refers to the total number of pregnancies a woman has, regardless of duration. Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013]. Most clinical notes were short; only those that described surgical births such as cesarean sections included details [@roth2021].

The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil. Between January 2012 and July 2013, all volumes were photographed. Then, between January and August 2017, data was manually put into Excel from the digital reproductions and then converted into a .csv file for upload into R.

The dataset presented in the article combines raw data of hospital records with additional analysis data derived from cleaning, processing, and engineering new variables from the raw data.

The data provided in this article offers a valuable historical resource for historians and epidemiologists interested in historical trends in maternal, fetal, and infant health. It enables comparative analyses with historical datasets available for North America and Western Europe.

## Specifications

### Subject

History

### Specific subject area

Historical hospital records that provide evidence of infant birth weight in Brazil.

### Type of data

Raw

Additional variables created by cleaning and processing data.

Table (.csv format)

Supporting materials (codebook and excel versions)

### Data collection

Clinical patient records for all patient visits to the Maternidade Laranjeiras between 1922 and 1926 were published in a historical medical journal. Between January 2012 and July 2013, relevant journal volumes and issues were photographed. Between January and August 2017, data was input into Excel from the digital reproductions and converted into a .csv file for upload into R).

### Data source location

The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.

### Data accessibility

### Related research article

## Value of the data

*	This dataset provides heretofore unpublished archival data into health outcomes in an understudied geographic region in the fields of history and epidemiology. It offers valuable insights into historical health outcomes related to racial and socioeconomic inequality in a major urban location in the Western Hemisphere. The data provide evidence to validate or challenge the existing body of research concerning birth weight, race, and socioeconomic inequality in other locations.

*	Researchers can utilize the data to make various comparisons in birth weight outcomes according to maternal variables across time and place. Although the accompanying article has utilized the data as a cross-sectional analysis, other researchers can combine this data set with other historical sets to detect trends, employing various methodological approaches.

*	Open-access historical data sets on health outcomes are rare for all locations and historical periods, but particularly for non-North American and Western European locals. This dataset serves as a fundamental resource for historians and historical demographers and epidemiologists studying maternal, fetal, and infant health in the past and today.

## Background

I initially collected this data during archival research for a PhD dissertation on the intertwined medical histories of abortion and childbirth in Rio de Janeiro, Brazil [@roth2020]. Historical datasets on quantitative outcomes of maternal, fetal, and infant health are sparse for all contexts, but particularly for non-North American and Western European historical contexts [@costa1998; @costa2004; @steckel1986; @ward1993]. Originally, the records were recorded for qualitative analysis of changes in operatory techniques including cesarean sections and embryotomies within a major public hospital in Brazil’s then-capital city of Rio de Janeiro, the Maternidade Laranjeiras. Thus, I did not extract all information from the hospital records given time constraints. Excluded information is as follows: pelvic measurements, length of various labor stages (first, second, and third), fetal presentation during delivery, placental weight, umbilical cord length, maternal pulse and temperature after delivery, timing of milk production, and timing of umbilical cord healing.

The extracted data includes demographic information on all women entering the hospital for gynecological and obstetric treatment, the majority to deliver an infant. It thus allows researchers to test associations between maternal variables including skin color, age, and parity and infant outcomes including birth weight, birth length, and stillbirth and neonatal mortality rates. It also provides information on maternal mortality ratios.

In addition to the published article, this dataset offers detailed information on the raw data, allowing further comparative investigations with other historical datasets on birth weight.

## Data description

The files associated with this data-in-brief article includes:
* Raw data: The original raw data, the codebook, and the processed data are available in the `data` folder.

- Hospital records data in .csv and .xlsx format (MaternidadeLaranjeiras.csv; MaternidadeLaranjeiras.xlsx): The hospital records data file contains 23 variables and 2845 entries of clinical visits to the Maternidade Laranjeiras hospital. Blank entries in the dataset indicate the information was missing from the original published archival records.

-	Codebook in .csv and .xlsx format (Codebook.csv; Codebook.xlsx): The codebook provides explanations and details regarding all 23 variables included in the original raw data hospital records.

- Mean birth weights in .csv and .xlsx format (Mean_Birthweights.csv; Mean_Birthweights.xlsx): The mean birth weight files provide data on published mean birth weights from different historical and epidemiological studies for comparison.

- Fertility rates in .csv and .xlsx format (1940_FertilityRates.csv; 1940_FertilityRates.xlsx): The fertility rate files provide archival and published data on total fertility rates (TFR) for Rio de Janeiro, São Paulo (city and state), and Brazil between 1930 and 1940.

- README.md file outlines content of the `raw-data` folder.

* Processed data: Processed data comprises all data sets cleaned and processed throughout the exploratory, intro analysis, and full analysis stages.

- ML_summary.rds: Processed data for all summary statistics. The processed summary data comprises 67 variables and 2845 observations.

- ML_linear.rds: Processed data for all linear models. The processed linear data comprises 19 variables and 1920 observations.

- Mean_Birthweights.rds: Processed data for mean birth weights table.

- Fertility_Rates_rds: Processed data for information on total fertility rates (TFR).

- README.md file outlines content of the `processed-data` folder.

## Experimental design, materials and methods

I collected the data between January 2012 and July 2013, when I photographed relevant journal volumes and issues. Between January and August 2017, I manually input data into Excel from the digital reproductions and converted into a .csv file for upload into R.

In R, I conducted multiple processing, cleaning, and engineering of recorded variables. If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` in the `processing-code` folder, which includes all discussions of data import; 2) `eda-v1.qmd` in the `eda-code` folder; 3) `introanalysis-v1.qmd` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` in the `analysis-code` folder. Older versions of code in `.r` are available in the `old-code` folder, which has not been updated since 15 August 2024.

All discussions of variable cleaning, processing, and feature engineering are available in the files above with accompanying code that is fully reproducible.

## Limitations

Historical data is limited by the people and administrative structures that initially created it. Historical datasets for Brazil are characterized by limited sample sizes, missing data, and imprecise or undefined measurements [@ball2020; @franken2025].

This dataset excludes multiple months of clinical records due to gaps in the archival sources. During data collection, I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926) in the historical archives, and thus this dataset has missing data. Further, the dataset is of limited size (n=2,845) and restricted to a short historical period (1922–1926).

Birth weight measurements were also relatively imprecise and the methods for weighing newborns was not included in the original records. All birth weights were rounded to the nearest 50 g.

Finally, the initial clinical records excluded important information related to birth weight as a health outcome, including gestational age, maternal marital or civil status, and post-natal feeding regimens.

## Ethics statement

The author has read and follow the ethical requirements for publication in Data in Brief and confirms that the current work does not involve human subjects, animal experiments, or any data collected from social media platforms.

## CRediT author statement

Cassia Roth was the sole author for this dataset.

## Acknowledgments

The author acknowledges Andreas Handel for help on creating reproducible code and William Summerhill for mentorship during the data collection and analysis process. The National Science Foundation (1226599) and the Fulbright IIE funded the archival research portion of this research.

## Declaration of competing interests

The author declares that she has no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# References

### 2024-11-22 products/manuscript/supplement/Supplemental_Materials.qmd
Final Changes for RR
Created DiB; updates for all changes

@@
 
-## Schematic of workflow
-
-If reproducing this analysis, please run code in the following order: 1) `processingfile-v1.qmd` in the `processing-code` folder; 2) `eda-v1.qmd` in the `eda-code` folder; 3) `introanalyis-v1.qmd` in the `analysis-code` folder; 4) `fullanalysis-v1.qmd` in the `analysis-code` folder.
-Older versions of code in `.r` are available in the `old-code` folder.
-These files have not been updated since 15 August 2024.
-
-## Data import and cleaning
-
-The GitHub repository for this project includes all relevant materials.
-All discussions of data import are detailed in `processingfile-v1.qmd` in the `processing-code` folder.
-The original raw data, the codebook, and the processed data are available in the `data` folder.
-
-## Data Source
+## Background
 
 In this paper, I analyze a unique dataset of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
-I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO).
-The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro.
-The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*.
-In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*.
-In 1922, it became the *Revista de Gynecologia e D’Obstetricia*.
+These hospital records were a result of a public-private expansion in what was then called maternal-infant healthcare services in Brazil's urban centers, specifically the capital city of Rio de Janeiro.
+By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2015; @barreto2016; @mott2002; @roth2020].
+Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
+There, obstetricians and gynecologists, all male and mostly White, began implementing surgical advances [@martins2004; @rohden2009; @roth2021].Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
+Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which relied on advances in medical knowledge and clinical infrastructure to support a technocratic model towards pregnancy and motherhood.
+This movement essentialized notions of gender, foregrounding physicians’ scientific authority [@besse1996; @freire2009; @martins2008].
+As the higher Maternal Mortality Ratio (MMR) and stillbirth rate (SBR) in the clinic compared to the city of Rio de Janeiro demonstrate, delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and infant and supports other historical studies that interventionist obstetric procedures before antibiotics and blood transfusions could be detrimental to maternal-infant health [@goldin1989].
 
-The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings.
-Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras.
-I was unable to locate vol.
-18, nos.
-4, 5, 6, 8 (1924) and vol.
-20, no.
-4 (1926).
-From the available issues, I recorded the following information, when available, for all patients: patient number, gravidity and parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.
-Gravidity refers to the total number of pregnancies a woman has, regardless of duration.
-Parity refers to a woman’s number of past pregnancies that reached viability and have been delivered, regardless of the number of children [@posner2013].
+During the same period, centuries-old racial hierarchies forged under racial slavery morphed into new allegedly scientific ways to categorize humans based on skin color in post-abolition contexts.
+In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of whitening in which they believed that racial mixture would eventually lead to a White population [@schwarcz1993; @skidmore1993].
+Both government- and private-run initiatives to attract European immigrants had begun as early as the 1870s, in an effort to replace the slave labor force not with freed Blacks but with White immigrants [@ball2020; @skidmore1993].
+In the 1920s and 1930s, scientists and policymakers turned toward preventive eugenic theories that focused on improving environmental factors rather than coercively restricting reproduction through sterilization [@stepan1991].
+Although some eugenicists called for sterilization of unfit citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @Wegner2013].
+In the end, however, whiteness still remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.
 
-The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.
-Between January 2012 and July 2013, I manually digitized the journal by photographing each volume.
-Then, between January and August 2017, I manually input the data into Excel from the digital reproductions.
-I then converted this into a .csv file for upload into R.
+Census data reflects these racial currents.
+Brazil did not conduct censuses in 1900 or 1910, and the 1920 census excluded skin color.
+In 1890, Rio de Janeiro's female population comprised 56% White, 29% mixed race, and 15% Black.
+In 1940, the next census to include skin color, the city’s female population in 1940 comprised 69% White women, 18% mixed-race women, and 12.5% Black women.
+Brazil did not record censuses in 1900 or 1910 and excluded skin color in 1920.
+The shift from mixed-race to white women was probably an reporting artifice rather than true change [@roth2020].
 
@@
 
-In the first round of peer review, the reviewers suggested calculating median birth weight for different immigrant groups by nationality to better justify groupings.
-Ball [@ball2020] found that Portuguese immigrants had worse economic opportunities in São Paulo during the same time period.
+To justify the nationality groupings, I calculated mean birth weight for different immigrant groups by nationality.
+Ball [-@ball2020] found that Portuguese immigrants had worse economic opportunities in São Paulo during the same time period.
 Although the Portuguese immigrant experience in Rio de Janeiro was likely better due to community, merchant connection, Portuguese business owners, and job availability, I needed to better justify my nationality categorization.
@@
 
-Here is the code for the full summary table of all variables.
+This summary table includes all original variables.
 
@@
 
-## Basic Statistical Analysis
+## Exploratory Analysis
 
@@
 ```
+Here, I also create a plot of the distribution of birth weights at Laranjeiras, following Wilcox's discussion of the residual and predominant birth weight distributions [@wilcox2001].
+
+```{r}
+#| label: fig-schematic-5
+#| fig-cap: Birth Weight Distribution for Afro- and Euro-Descended Infants, Maternidade Laranjeiras, 1922-1926
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","residual_plot_divided.png"))
+```
+
+I also calculate (SRaB) in the hospital:
+
+$$SRaB = \frac{M}{F} \times 100$$
+
+where the total number of live male births (M) is divided by 100 live female births (F) in a given period.
+
+The sex ratio in Maternidade Laranjeiras was 1.2: there were 120 male live births per 100 female live births.
+The sex ratio of 120 male live births per 100 female live births is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
+Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was female.
+Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016].
+However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants.
+Yet, the skewed sex ratio remains important to note even if we don't understand the underlying cause.
 
@@
 Neither age nor age-squared were significant.
-Peer reviewers asked me to include $age^2$ in the model to improve the fit.
 Including $age^2$ in the model did not improve the fit, and the coefficient was not significant.
 Despite a lack of statistical significance, the coefficients suggests that while birth weight increases with maternal age, it does so at a decreasing rate.
-At certain maternal age, infant birth weight begins decreasing, presenting a reverse U-shaped curve.
+At a certain maternal age, infant birth weight begins decreasing, presenting a reverse U-shaped curve.
 
@@
 ```{r}
-#| label: tbl-table5
-#| tbl-cap: "Model 5"
+#| label: tbl-table8
+#| tbl-cap: "Model 8"
 #| echo: FALSE
-table5 <- readRDS(here::here("results", "tables", "table_log1.rds"))
-flex_table <- as_flex_table(table5)
+table8 <- readRDS(here::here("results", "tables", "table_log1.rds"))
+flex_table <- as_flex_table(table8)
 flex_table
@@
 ```{r}
-#| label: tbl-table6
-#| tbl-cap: "Model 6"
+#| label: tbl-table9
+#| tbl-cap: "Model 9"
 #| echo: FALSE
-table6 <- readRDS(here::here("results", "tables", "table_log2.rds"))
-flex_table <- as_flex_table(table6)
+table9 <- readRDS(here::here("results", "tables", "table_log2.rds"))
+flex_table <- as_flex_table(table9)
 flex_table
@@
 ```{r}
-#| label: tbl-table7
-#| tbl-cap: "Model 7"
+#| label: tbl-table10
+#| tbl-cap: "Model 10"
 #| echo: FALSE
-table7 <- readRDS(here::here("results", "tables", "table_log3.rds"))
-flex_table <- as_flex_table(table7)
+table10 <- readRDS(here::here("results", "tables", "table_log3.rds"))
+flex_table <- as_flex_table(table10)
 flex_table
@@
 ```{r}
-#| label: tbl-table8
-#| tbl-cap: "Model 8"
+#| label: tbl-table11
+#| tbl-cap: "Model 11"
 #| echo: FALSE
-table8 <- readRDS(here::here("results", "tables", "table_log4.rds"))
-flex_table <- as_flex_table(table8)
+table11 <- readRDS(here::here("results", "tables", "table_log4.rds"))
+flex_table <- as_flex_table(table11)
 flex_table
@@
 
-In the first round of peer review, the reviewers suggested that a logit analysis using stillbirth as an outcome would bring robustness to the racial disparities I highlight in the linear regression.
+Performing a logistic regression with stillbirth as an outcome could bring robustness to the racial disparities I highlight in the linear regression.
 After running a simple logistic regression with `FetalOutcome` as the predictor variable and `ModifiedColor` as the outcome, infants born to African-descended women had a 9% higher odds of having a stillbirth than Euro-descended women.
-However, all results were not significant [95% CI 0.82, 1.45; p-value 0.6], and thus I will not include them in the final analysis.
+However, all results were not significant [95% CI 0.82, 1.45; p-value 0.6], and thus I exclude them from the final analysis.
 
 ```{r}
-#| label: tbl-table9
-#| tbl-cap: "Model 9"
+#| label: tbl-table12
+#| tbl-cap: "Model 12"
 #| echo: FALSE
-table9 <- readRDS(here::here("results", "tables", "table_log5.rds"))
-flex_table <- as_flex_table(table9)
+table12 <- readRDS(here::here("results", "tables", "table_log5.rds"))
+flex_table <- as_flex_table(table12)
 flex_table
@@
 
-Here I include the cross-validation results for the linear models.
+The performance metrics for the logistic models are available in the `introanalysis-v1.qmd` file.
 
-```{r}
-#| label: tbl-table10
-#| tbl-cap: "Cross-Validation Model 1"
-#| echo: FALSE
-ML_cv1_metrics <- readRDS(here::here("results", "tables", "ML_cv1_metrics.rds"))
-ML_cv1_metrics
-```
+## Discussion
 
-```{r}
-#| label: tbl-table11
-#| tbl-cap: "Cross-Validation Model 2"
-#| echo: FALSE
-ML_cv2_metrics <- readRDS(here::here("results", "tables", "ML_cv2_metrics.rds"))
-ML_cv2_metrics
-```
+Differences in birth weight between racialized groups is the paper's significant result and supports findings in other contexts.
+For example, differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and “colored” children that starts at birth and increases over time even as all children have gotten larger (@cameron2003).
 
-```{r}
-#| label: tbl-table12
-#| tbl-cap: "Cross-Validation Model 3"
-#| echo: FALSE
-ML_cv3_metrics <- readRDS(here::here("results", "tables", "ML_cv3_metrics.rds"))
-ML_cv3_metrics
-```
+Although not discussed in the main analysis, discussions of epigenetic markers associated with the lived experience of racism might also be applicable to racial differences in birth weight outcomes.
+Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms (@leimert2020).
+Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance (@jasienska2009; @leimert2020).
+Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health (@meloni2022).
 
-```{r}
-#| label: tbl-table13
-#| tbl-cap: "Cross-Validation Model 4"
-#| echo: FALSE
-ML_cv4_metrics <- readRDS(here::here("results", "tables", "ML_cv4_metrics.rds"))
-ML_cv4_metrics
-```
-
-The performance metrics for the logistic models are available in the `introanalysis-v1.qmd` file.
+A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy (@santana2021).
+The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight (@jasienska2009; @ward1993).
 

### 2024-11-23 code/analysis-code/introanalysis-v1.qmd
Final Changes for RR
Supplemental

@@
                                       exponentiate = FALSE, 
-                                      intercept = TRUE) %>% 
+                                      intercept = TRUE,
+                                      conf.int = FALSE) %>% 
                    bold_p() %>% 

### 2024-11-23 products/manuscript/supplement/Supplemental_Materials.qmd
Final Changes for RR
Supplemental

@@
 
-In this paper, I analyze a unique dataset of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926.
-These hospital records were a result of a public-private expansion in what was then called maternal-infant healthcare services in Brazil's urban centers, specifically the capital city of Rio de Janeiro.
+In this paper, I analyze a unique dataset of 2845 recorded clinical visits to the public maternity hospital, Maternidade Laranjeiras, between June 1922 and May 1926 in Rio de Janeiro, Brazil.
+These hospital records resulted from public-private expansions in the provision of what was then called maternal-infant healthcare services in Brazil's urban centers, specifically the capital city of Rio de Janeiro.
 By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2015; @barreto2016; @mott2002; @roth2020].
-Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process.
-There, obstetricians and gynecologists, all male and mostly White, began implementing surgical advances [@martins2004; @rohden2009; @roth2021].Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
+Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization parocess.
+There, obstetricians and gynecologists, all male and mostly White, began implementing surgical advances [@martins2004; @rohden2009; @roth2021].
+Although the records do not show direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.
+
 Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which relied on advances in medical knowledge and clinical infrastructure to support a technocratic model towards pregnancy and motherhood.
@@
 In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of whitening in which they believed that racial mixture would eventually lead to a White population [@schwarcz1993; @skidmore1993].
-Both government- and private-run initiatives to attract European immigrants had begun as early as the 1870s, in an effort to replace the slave labor force not with freed Blacks but with White immigrants [@ball2020; @skidmore1993].
+Both government- and private-run initiatives to attract European immigrants, in an effort to replace the slave labor force not with freed Blacks but with White immigrants, had begun as early as the 1870s [@ball2020; @skidmore1993].
 In the 1920s and 1930s, scientists and policymakers turned toward preventive eugenic theories that focused on improving environmental factors rather than coercively restricting reproduction through sterilization [@stepan1991].
-Although some eugenicists called for sterilization of unfit citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @Wegner2013].
-In the end, however, whiteness still remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.
+Although some eugenicists called for sterilization of unfit citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013].
+In the end, whiteness remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.
 
@@
 In 1940, the next census to include skin color, the city’s female population in 1940 comprised 69% White women, 18% mixed-race women, and 12.5% Black women.
-Brazil did not record censuses in 1900 or 1910 and excluded skin color in 1920.
-The shift from mixed-race to white women was probably an reporting artifice rather than true change [@roth2020].
+The shift from mixed-race to white women was probably a reporting artifice rather than true change [@roth2020].
 
@@
 To justify the nationality groupings, I calculated mean birth weight for different immigrant groups by nationality.
-Ball [-@ball2020] found that Portuguese immigrants had worse economic opportunities in São Paulo during the same time period.
-Although the Portuguese immigrant experience in Rio de Janeiro was likely better due to community, merchant connection, Portuguese business owners, and job availability, I needed to better justify my nationality categorization.
-In the `eda-v1.qmd` file, I calculated the mean birth weight for individual nationalities and performed a t-test between Portuguese infants and infants born to all other European nationalities and Portuguese infants and infants born to White Brazilian mothers.
-All results were not significant at the alpha 0.05 level.
-Thus, I will maintain the nationality categorizations of European, Brazilian, Latin American, and Middle Eastern.
+Ball [-@ball2020] found that Portuguese immigrants had worse economic opportunities in São Paulo during the same period.
+The Portuguese immigrant experience in Rio de Janeiro was likely better due to community, merchant connection, Portuguese business owners, and job availability [@ball2020].
+In the `eda-v1.qmd` file, I calculated the mean birth weight for individual nationalities (Table 1) and performed a t-test between (Table 2) Portuguese infants and infants born to all other European nationalities and Portuguese infants and infants born to White Brazilian mothers.
 
@@
 ```
+
 ```{r}
@@
 ```
-
-### Full Summary Table
-
-This summary table includes all original variables.
-
-```{r}
-#| label: tbl-tablesupplemental
-#| tbl-cap: "Supplemental Table"
-#| echo: FALSE
-table_supplemental <- readRDS(here::here("results", "tables", "table_supplemental_final.rds"))
-flex_table <- as_flex_table(table_supplemental)
-flex_table
-```
+All results were not significant at the alpha 0.05 level (Table 2). Thus, I will maintain the nationality categorizations of European, Brazilian, Latin American, and Middle Eastern.
 
@@
 To better understand our outcome of interest, birth weight, the exploratory analysis includes the distribution of the data.
-Figure 1 is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO now defines as normal birth weight [@worldhealthorganization2022].
+Figure 1 is a histogram of the birth weight data, with the two dotted red lines marking the upper and lower limits of what the WHO now defines as normal birth weight [@who2022].
 
@@
 
-Figure 4 visualizes the relationship of birth weight and length, stratified by sex since male infants are usually slightly heavier and longer than females [@crawford1987, @marsal1996].
+Figure 4 visualizes the relationship of birth weight and length, stratified by sex since male infants are usually slightly heavier and longer than females [@crawford1987; @marsal1996].
 
@@
 ```
-Here, I also create a plot of the distribution of birth weights at Laranjeiras, following Wilcox's discussion of the residual and predominant birth weight distributions [@wilcox2001].
+I also create a plot of the distribution of birth weights at Laranjeiras, following Wilcox's [-@wilcox2001] discussion of the residual and predominant birth weight distributions.
 
@@
 
-I also calculate (SRaB) in the hospital:
+I also calculate sex ratio at birth (SRaB) in the hospital:
 
@@
 The sex ratio in Maternidade Laranjeiras was 1.2: there were 120 male live births per 100 female live births.
-The sex ratio of 120 male live births per 100 female live births is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
+This is higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024].
 Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was female.
@@
 For the basic statistical analysis, I ran three linear models to understand the relationship between maternal factors and infant birth weight.
-The first model is a simple linear regression with outcome variable (birth weight in grams) and exposure variable maternal ancestry (`Euro-descent`, the reference, and `Afro-descent`).
+The first model is a simple linear regression with the outcome variable, birth weight in grams, and exposure variable maternal ancestry (`Euro-descent`, the reference, and `Afro-descent`).
 There is an association between maternal skin color and infant birth weight, with Euro-descended women having infants with higher birth weights than Afro-descended women.
@@
 
-The second simple bivariate linear analysis, looks at infant birth weight as a function of maternal age on both a linear scale and age-squared with `ModifiedColor` as the main predictor variable.
-Neither age nor age-squared were significant.
+The second simple bivariate linear analysis looks at infant birth weight as a function of maternal age on both a linear scale and $age^2$, with `ModifiedColor` as the main predictor variable.
+Neither age nor $age^2$ were significant.
 Including $age^2$ in the model did not improve the fit, and the coefficient was not significant.
@@
 #| label: tbl-age2
-#| tbl-cap: "Age-Squared"
+#| tbl-cap: "Model 3"
 #| echo: FALSE
@@
 
-Peer review suggested that I control for year-specific effects given possible periodic epidemics in infectious disease such as malaria and yellow fever and changes in real wages.
+I control for year-specific effects given possible periodic epidemics in infectious disease such as malaria and yellow fever and changes in real wages [@benchimol1999; @lobo2024].
 I do this by including year fixed effects.
@@
 In this sub-sample model (Model 6), the coefficient for infants born to Afro-descended mothers was both not significant and lower than in the full model (-40).
-However, dropping the observations from 1923 has reduced my sample size by about 25%.
 It appears that births from 1923 are driving the overall results.
@@
 
-Additionally, these results could be due to smaller sample sizes and larger standard errors -- noise in the data.
+However, dropping the observations from 1923 has reduced my sample size by about 25%.
+These results could be due to smaller sample sizes and larger standard errors -- noise in the data.
 This reflects the limitations of historical data and historical research given small sample sizes.
 
-The last exploratory model again runs a multivariate linear regression, this time using skin color as it was differentiated in the original data, or with three categories (White, Black, and Mixed Race).
+The last exploratory model again runs a multivariate linear regression, this time using skin color as it was differentiated in the original data, or with three categories (White, Black, and mixed Race).
 
@@
 
-In my introductory analysis, I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
+In my introductory analysis, I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a].
 
-I consider `Birthweight` as the outcome of interest, but I transformed it into a categorical outcome of NBW $>2500$ g or LBW $\leq2500$ g) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
-These cutoffs are according to the WHO [@worldhealthorganization2022].
+I consider `Birthweight` as the outcome of interest, but I transformed it into a categorical outcome of NBW $>2500$ g or LBW $\leq2500$ g) [@who2022] and fit a logistic model, using the main predictor of interest, `ModifiedColor`.
 
-The first model is a simple logistic regression, with modified maternal skin color as the predictor variable.
+The first model (Model 8) is a simple logistic regression, with modified maternal skin color as the predictor variable.
 
@@
 
-Model 5 demonstrates that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.45 [95% CI 1.07, 1.96, p-value = 0.016).
+Model 8 demonstrates that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.45 [95% CI 1.07, 1.96, p-value = 0.016).
 
-The second model is also a simple logistic regression, this time using `Color` as the predictor variable (Black, mixed race, and White).
+The second logistic model (Model 9) is also a simple logistic regression, this time using `Color` as the predictor variable (White, Black, and mixed race).
 
@@
 Then, I ran two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.
-The results are below.
 
@@
 
-In these two models, the only significant finding is from the final logistic model, in which infants born to mixed-race mothers had 21% higher odds of being born with low birth weight compared to infants born to white mothers.
+In these two models, the only significant finding is from the final logistic model, in which infants born to mixed-race mothers had 53% higher odds of being born with low birth weight compared to infants born to white mothers.
 The other models did not show significant results.
@@
 Best practice is to not categorize continuous variables, as this can include arbitrary cutoffs for variables that are best understood continuously [@senn2005].
-Thus, I will not include these logistic regressions in the final analysis.
+Further, scholars have called into question the use of LBW as a marker [@wilcox2001].
 
@@
 After running a simple logistic regression with `FetalOutcome` as the predictor variable and `ModifiedColor` as the outcome, infants born to African-descended women had a 9% higher odds of having a stillbirth than Euro-descended women.
-However, all results were not significant [95% CI 0.82, 1.45; p-value 0.6], and thus I exclude them from the final analysis.
+However, all results were not significant at the alpha 0.05 level [95% CI 0.82, 1.45; p-value 0.6].
 
@@
 When run on testing data, most results were insignificant, given the smaller sample size.
-
 The performance metrics for the logistic models are available in the `introanalysis-v1.qmd` file.
@@
 Differences in birth weight between racialized groups is the paper's significant result and supports findings in other contexts.
-For example, differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and “colored” children that starts at birth and increases over time even as all children have gotten larger (@cameron2003).
+For example, differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and “colored” children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].
 
 Although not discussed in the main analysis, discussions of epigenetic markers associated with the lived experience of racism might also be applicable to racial differences in birth weight outcomes.
-Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms (@leimert2020).
-Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance (@jasienska2009; @leimert2020).
-Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health (@meloni2022).
+Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms [@leimert2020].
+Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@jasienska2009; @leimert2020].
+Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health [@meloni2022].
 
-A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy (@santana2021).
-The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight (@jasienska2009; @ward1993).
+A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021].
+The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009; @ward1993].
 

### 2025-01-10 products/manuscript/Manuscript_Final_SSM.qmd
Final Commit
After copy edits

@@
 
-The differences between and within groups both today and in the past have prompted racially-stratified research.
+The differences between and within groups both today and in the past have prompted research stratified by race.
 While race is a social construct and thus biological fiction, the social consequences of racism influence health and disease patterns, including birth outcomes, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004].
@@
 
-The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.
+The journal is held at the Biblioteca Nacional (BN); the Maternidade Escola, Rio de Janeiro (ME-UFRJ); and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil.
 Between January 2012 and July 2013, I photographed each volume.
@@
 The original hospital records divided maternal parity and gravidity into eight categories of parity and gravidity: nulliparous, primiparous, secundiparous, and multiparous; and primigravida, secundigravida, trigravida, and multigravida.
-Due to issues of statistical power, I combined these into two categories: 1) `nulliparous`, which included nulliparous, primigravida, and primiparous (any woman giving birth for the first time); and 2) `multiparous`, which included secundiparous, multiparous, secundigravida, trigravida, and multigravida (any woman who has given birth more than once).
+Due to issues of statistical power, I combined these into two categories: 1) `nulliparous`, including nulliparous, primigravida, and primiparous (any woman giving birth for the first time); and 2) `multiparous`, including secundiparous, multiparous, secundigravida, trigravida, and multigravida (any woman who has given birth more than once).
 
@@
 *Year*.
-I construct year fixed effects by including dummy variables for each year from 1923 to 1926.
+I constructed year fixed effects by including dummy variables for each year from 1923 to 1926.
 1922 is the base year.
@@
 First-time mothers (nulliparous), comprised 41%.
-Of all women, 84% were Brazilian.
-Of the nearly 16% immigrant patients, 15% were European.
+Of all women, 84% were Brazilian, 15% were European immigrants, and the remaining were immigrants from other regions.
 
@@
 
-In all models (Tables 2 and 3), women who were nulliparous (delivering their first child) had lighter infants weighing than multiparous women, an association recognized by physicians at the time and found in both other historical studies and today [@conrado1924a; @costa1998; @souza2020].
+In all models (Tables 2 and 3), women who were nulliparous (delivering their first child) had lighter infants than multiparous women, an association recognized by physicians at the time and found in both other historical studies and today [@conrado1924a; @costa1998; @souza2020].
 Age and nationality were not significant.
@@
 Across populations, relatively high mean birth weights in the nineteenth and early twentieth centuries could be an artifact of high stillbirth rates, which selected upon fetuses that were already small and unhealthy [@costa1998; @costa2004; @schneider2017].
-Today, PTB and lower birth weight infants that might have died in utero 100 years ago now survive, bringing the mean birth weight down.
+Today, PTB and lower birth weight infants that might have died in utero 100 years ago survive, bringing the mean birth weight down.
 
@@
 Two sets of missing data are notable.
-First, hospital records excluded maternal marital status (included in 0.4% of records).
-Scholars have found that marital status was a good proxy for SES [@costa2004; @goldin1989].
-The hospital's exclusion of this variable precludes a full understanding of mothers' SES.
+First, hospital records excluded maternal marital status (included in only 0.4% of records).
+As scholars have found that marital status was a good proxy for SES [@costa2004; @goldin1989], the hospital's exclusion of this variable precludes a full understanding of mothers' SES.
 
@@
 
-This study has shown that lower birth weights were associated with Afro-descended women giving birth in the Maternidade Laranjeiras in Rio de Janeiro, Brazil between 1922 and 1926, suggesting that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which we do not have direct evidence.
+This study has shown that lower birth weights were associated with Afro-descended women giving birth in the Maternidade Laranjeiras in the city of Rio de Janeiro, Brazil between 1922 and 1926, suggesting that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms for which we do not have direct evidence.
 These findings, although not generalizable beyond this specific hospital, are consistent with both historical and present-day studies on racialized disparities in birth weight in Rio de Janeiro, demonstrating how racial inequality in health outcomes has persisted over time.

### 2025-01-10 products/manuscript/supplement/Supplemental_Materials.qmd
Final Commit
After copy edits

@@
 ```
-I also create a plot of the distribution of birth weights at Laranjeiras, following Wilcox's [-@wilcox2001] discussion of the residual and predominant birth weight distributions.
+I also created a plot of the distribution of birth weights at Laranjeiras, following Wilcox's [-@wilcox2001] discussion of the residual and predominant birth weight distributions.
 
@@
 
-I also calculate sex ratio at birth (SRaB) in the hospital:
+I also calculated sex ratio at birth (SRaB) in the hospital:
 
@@
 
-The third exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, age squared, nationality, and gestational status on infant birth weight with the `ModifiedColor` variable as the main predictor variable.
+The third exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, $age^2$, nationality, and gestational status on infant birth weight with the `ModifiedColor` variable as the main predictor variable.
 
